### 2013-05-01 article.tex
get rid of TMR tex file, replacing with own

% -*- LaTeX -*-
\documentclass{tmr}

\usepackage{mflogo}

%include polycode.fmt

\title{Parsers or something}
\author{Matt Fenwick\email{mfenwick100@gmail.com}}

\begin{document}

\begin{introduction}
\end{introduction}

\section{Parser combinators}

\section{Monad transformers}

\section{Example 1: recognizing}

\section{Example 2: error-reporting}

\section{Example 3: whitespace/comment-logging}

\section{Verbatim and code}

Um wow this is code:
\begin{spec}
class Foo f where
  foo :: f a -> f b -> f (a,b)
  -- use for wibbling

bar :: Int -> Bool
bar  23  =  False
bar  x   =  not (x `elem` [1,2,3])

\end{spec}
that was some pretty code.

\section{Transformer instance semantics}

Pass-through vs. deal-with, examples of each

\section{Limitations of the Monad Transformer Approach}

1. Exact vs. relative layer order; 
2. one instance per stack, even when more than one is applicable --
how to specify which one?
3. an inappropriate instance higher-up in the stack can mask a 
desired instance lower down in the stack -- that's why we had to
create a replacement for the Alternative type class and provide
some new instances for it -- to be able to use the right instances

\section{Conclusion}

Parser combinators and monad transformers FTW!

\bibliography{Author}

\end{document}

### 2013-05-01 article.tex
get article under way

@@
 
-\title{Parsers or something}
+\title{Combinator Parsers:  a Monad Transformer Approach}
 \author{Matt Fenwick\email{mfenwick100@gmail.com}}
@@
 \begin{introduction}
+Parsing, or "the process of analysing a 
+string of symbols according to the rules of a formal grammar" (Wikipedia),
+is one of the most fundamental and interesting problems in programming.
+The general goal of parsing is to build a tree representing the structure
+of the parsed input; further operations such as interpretation or
+code generations are then performed on the tree.
+
+However, in addition to recognizing a stream of tokens and building a parse
+tree, a practical parser has many additional duties.  If parsing fails,
+the location and cause of the failure must be accurately reported so that
+the user can address the problem; specially formatted comments may be need
+to be capture in order to be used
+to build hyperlinked code documentation, instead of simply thrown away; 
+failed parses may be accompanied by partial results to save the cost
+of reparsing the input before the error.
+
+Monad transformers offer an elegant solution, both to the problem of 
+implementing parser combinators themselves, as well as to the issue of
+building clean and modular parsers that support multiple effects.
+We'll explore what they have to offer using a simple language -- Woof --
+and build a number of separate parsers for it!
 \end{introduction}
 
-\section{Parser combinators}
+\section{Parser Combinators}
+Parser combinators have long been used for parsing in the 
+functional programming community, represent a powerful and simple solution.
+Their main advantages are ability to deal with complicated formats (see
+Chomsky levels), language integration, and the declarativeness and
+succinctness of the resulting parsers.
+Parser combinator libraries typically include a set of basic parsers and a
+set of combinators for combining little parsers into big ones.  Parsers are
+typically functions whose type is something like:
+\begin{verbatim}
+[t] -> m ([t], a)
+\end{verbatim}
+where t is the token type, a is the result type, and m represents some
+computational effects.
+A very simple parser succeeds, consuming a single token, when the token stream
+is not empty, and fails otherwise.  We can make the substitutions t -> Char,
+m -> Maybe, and a -> Char and implement the function:
+\begin{verbatim}
+    item :: [Char] -> Maybe ([Char], Char)
+    item (t:ts) = Just (ts, t)
+    item []     = Nothing
+
+    -- examples
+    ghci$ item "abcde"
+    Just ("bcde", 'a')
+
+    ghci$ item []
+    Nothing
+\end{verbatim}
+
+We can also come up with a simple combinator that takes a parser and a function
+as input, and succeeds if the parser succeeds and the function applied to the
+parser's result succeeds:
+\begin{verbatim}
+    check :: (a -> Bool) -> ([Char] -> Maybe ([Char], a)) -> [Char] -> Maybe ([Char], a)
+    check f p = p >>= \(ts, x) -> if (f x) then Just (ts, x) else Nothing
+
+    -- examples
+    ghci$ check (== 'a') item "abcde"
+    Just ("bcde", 'a')
+
+    ghci$ check (== 'b') item "abcde"
+    Nothing
+\end{verbatim}
+However, in these examples our computational effects are limited to Maybe -- we're
+unable to do error reporting, logging, and so on.  Coming up with more general 
+parsers and combinators will be the topic of the rest of the article.
 
@@
 
+\section{Introducing Woof:  a Simple Lisp}
+
+The following examples will all use Woof, a simple dialect of Lisp,
+progressively adding features to a simplistic initial implementation
+to create a usable parser.
+
+The language definition in pseudo-BNF is:
+
+\begin{verbatim}
+Woof         :=   Form(+)
+
+Form         :=   Special  |  Application  |  Symbol  |  
+                  String   |  Number
+
+Special      :=  '{'  ( Define  |  Lambda )  '}'
+
+Define       :=  'define'  String(?)  Symbol  Form
+
+Lambda       :=  'lambda'  '{'  Symbol(*)  '}'  Form(+)
+
+Application  :=  '('  Form(+)  ')'
+
+Number       :=  \d(+)
+
+Symbol       :=  ( \w  |  Schar)  ( \w  |  \d  |  Schar)(*)
+
+Schar        :=  (oneof "<>!@#$%^&*_-+=|:?")
+
+String       :=  '"'  ( Escape  |  (not  ( '"'  |  '\')))(*)  '"'
+
+Escape       :=  '\'  ( '"'  |  '\' )
+
+Whitespace   :=  \s+
+
+Comment      :=  ';'  (not '\n')(*)
+\end{verbatim}
+With the additional rule that whitespace and comments 
+may appear in any amount before any token.
+Tokens are:  {, }, (, ), Symbol, String, and Number.
+
 \section{Example 1: recognizing}
+
+\subsection{An Elementary Parser}
+The basic types of parsers will be something like
+\begin{verbatim}[t] -> m ([t], a)\end{verbatim}, or, a function that consumes
+part of a list of tokens of type t with computational effects m and
+result a.
+A very basic and fundamental parser is the one that succeeds whenever
+the list of tokens is not empty, consuming one token and presenting it
+as the result, and fails when the list is empty.  We'll call this item,
+and we can easily implement it in terms of the MonadState type class:
+\begin{verbatim}
+item :: (MonadState [t] m, Plus m) => m t
+item =
+    get >>= \xs -> case xs of
+                        (t:ts) -> put ts *> pure t;
+                        []     -> zero;
+\end{verbatim}
+
+\subsection{Token Parsers}
+Let's start with parsers for recognizing our most basic syntactic elements:  tokens.
+These include the seven listed above, as well as whitespace and comments.
 
@@
 \section{Example 3: whitespace/comment-logging}
-
-\section{Verbatim and code}
-
-Um wow this is code:
-\begin{spec}
-class Foo f where
-  foo :: f a -> f b -> f (a,b)
-  -- use for wibbling
-
-bar :: Int -> Bool
-bar  23  =  False
-bar  x   =  not (x `elem` [1,2,3])
-
-\end{spec}
-that was some pretty code.
 

### 2013-05-01 src/parse.hs
don't parse Tokens, parse Chars in basic parser, since no use is made of the position information

@@ -10,63 +10,61 @@ import Common
 import Instances
 
 
-character :: (MonadState [Token] m, Plus m) => Char -> m Token
-character c = satisfy ((==) c . chr)
-
-munch :: (MonadState [Token] m, Plus m, Switch m) => m a -> m a
-munch p = many0 (whitespace <+> comment) *> p
-
+character :: (MonadState [Char] m, Plus m) => Char -> m Char
+character c = satisfy ((==) c)
 
 ocurly = character '{'
-
 ccurly = character '}'
-
 oparen = character '('
-
 cparen = character ')'
 
-digit = fmap chr $ satisfy (flip elem ['0' .. '9'] . chr)
+digit = satisfy (flip elem ['0' .. '9'])
 
 number = munch $ many1 digit
 
 symbolOpens = ['a' .. 'z'] ++ ['A' .. 'Z'] ++ "!@#$%^&*_-+=:?<>/"
 
+
+whitespace :: (MonadState [Char] m, Plus m, Switch m) => m [Char]
+whitespace = many1 $ satisfy (flip elem " \n\t\r\f")
+
+comment :: (MonadState [Char] m, Plus m, Switch m) => m [Char]
+comment = character ';' *> many0 (not1 $ character '\n')
+
+munch :: (MonadState [Char] m, Plus m, Switch m) => m a -> m a
+munch p = many0 (whitespace <+> comment) *> p
+
+
 symbol = munch (fmap (:) first <*> rest)
   where
-    first = fmap chr $ satisfy (flip elem symbolOpens . chr)
+    first = satisfy (flip elem symbolOpens)
     rest = many0 (first <+> digit)
     
-string :: (MonadState [Token] m, Plus m, Switch m) => m String
+string :: (MonadState [Char] m, Plus m, Switch m) => m String
 string = munch (dq *> many0 char <* dq)
   where
     dq = character '"'
-    char = fmap chr (escape <+> normal)
+    char = escape <+> normal
     escape = character '\\' *> slash_or_dq
     normal = not1 slash_or_dq
     slash_or_dq = character '\\' <+> character '"'
 
-
-whitespace :: (MonadState [Token] m, Plus m, Switch m) => m [Token]
-whitespace = many1 $ satisfy (flip elem " \n\t\r\f" . chr)
-
-comment :: (MonadState [Token] m, Plus m, Switch m) => m [Token]
-comment = character ';' *> many0 (not1 $ character '\n')
 
 top = munch oparen
 tcp = munch cparen
[…]

### 2013-05-01 article.tex
continue write-up of basic parser

@@
 \begin{verbatim}
-[t] -> m ([t], a)
+[t] -> m (a, [t])
 \end{verbatim}
@@
 \begin{verbatim}
-    item :: [Char] -> Maybe ([Char], Char)
-    item (t:ts) = Just (ts, t)
+    item :: [Char] -> Maybe (Char, [Char])
+    item (t:ts) = Just (t, ts)
     item []     = Nothing
@@
     ghci$ item "abcde"
-    Just ("bcde", 'a')
+    Just ('a', "bcde")
 
@@
 \begin{verbatim}
-    check :: (a -> Bool) -> ([Char] -> Maybe ([Char], a)) -> [Char] -> Maybe ([Char], a)
-    check f p = p >>= \(ts, x) -> if (f x) then Just (ts, x) else Nothing
+    check :: (a -> Bool) -> ([Char] -> Maybe (a, [Char])) -> [Char] -> Maybe (a, [Char])
+    check f p = p >>= \(x, ts) -> if (f x) then Just (x, ts) else Nothing
 
@@
     ghci$ check (== 'a') item "abcde"
-    Just ("bcde", 'a')
+    Just ('a', "bcde")
 
@@
 \section{Monad transformers}
+Monad transformers are a technique for combining monads to form new monads.
+The motivation behind them is that while single monads are quite useful, oftentimes
+the semantics of multiple monads are required simultaneously; it is then desirable
+to have a modular approach for creating combined monads.
+
+There are multiple to approaches to combining monads and to monad transformers
+themselves, but we'll follow that of the standard transformers/mtl library, since
+it's easy to obtain, easy to use, and quite effective in practice.
+
+Along the way, we'll also explore some of the consequences of combining monads,
+both with respect to semantics and to the definition of the Haskell language.  We'll
+also bump in to some of the limitations, trade-offs and design decisions of the 
+transformer/mtl library.
 
 \section{Introducing Woof:  a Simple Lisp}
-
-The following examples will all use Woof, a simple dialect of Lisp,
+The simple language we'll use as a motivation for building parsers is 
+Woof, a simple dialect of Lisp.  We'll be 
 progressively adding features to a simplistic initial implementation
-to create a usable parser.
+on our way to creating a practical, feature-rich and usable parser.
 
@@
 
-\section{Example 1: recognizing}
+\section{Example 1: Recognition and Tree-Building}
+The first parser we build will be responsible for determining whether input
+text conforms to the language definition and simultaneously build an 
+Abstract Syntax Tree representing of the structure of the recognized input.
 
-\subsection{An Elementary Parser}
-The basic types of parsers will be something like
-\begin{verbatim}[t] -> m ([t], a)\end{verbatim}, or, a function that consumes
-part of a list of tokens of type t with computational effects m and
-result a.
-A very basic and fundamental parser is the one that succeeds whenever
-the list of tokens is not empty, consuming one token and presenting it
-as the result, and fails when the list is empty.  We'll call this item,
-and we can easily implement it in terms of the MonadState type class:
+Here's the AST definition we'll use:
+\begin{verbatim}
+data AST
+    = ANumber Integer
+    | ASymbol String
+    | AString String
+    | ALambda [String] [AST]
+    | ADefine (Maybe String) String AST
+    | AApp    AST  [AST]
+  deriving (Show, Eq)
+\end{verbatim}
+Plus we'll need a type for our parsers.  We'll use the basic type mentioned 
+above -- [t] -> m (a, [t]) -- as our starting point.  As luck would have it, that
+corresponds to our first monad transformer:  StateT (Control.Monad.State), so we
+can write:
+\begin{verbatim}
+type Parser a = StateT [Char] Maybe a
+\end{verbatim}
+Which means that our Parser is a State monad transformer, with the Maybe type
+as its underlying Monad parameter.
+
+Now let's get down to the business of building parsers, starting with parsers
+for our most basic syntactic elements:  tokens.
+These include the four braces, symbols, strings, numbers, whitespace, and comments.
+To recognize braces, we need to re-implement the item and check combinators that
+we did before.  Fortunately, using the StateT transformer makes it simpler
+to do so:
 \begin{verbatim}
@@
 \end{verbatim}
-
-\subsection{Token Parsers}
-Let's start with parsers for recognizing our most basic syntactic elements:  tokens.
-These include the seven listed above, as well as whitespace and comments.
+Although we now also need to discuss the Plus type class, why we need it instead of
+Alternative, and what its instances are.
 

### 2013-05-08 article.tex
massively simplify grammar, in order to allow examples to be much simpler and more focused on the important issues of parser combinators and monad transformers:
 - remove string, number tokens
 - simplify symbol definition

create new simple parser implementation, and describe it in write-up, as simplest version 1 parse tree and recognization demonstration

@@
 
-Form         :=   Special  |  Application  |  Symbol  |  
-                  String   |  Number
+Form         :=   Symbol  |  Special  |  Application
+
+Symbol       :=  [a-zA-Z](+)
 
@@
 
-Define       :=  'define'  String(?)  Symbol  Form
+Define       :=  'define'  Symbol  Form
 
@@
 
-Number       :=  \d(+)
-
-Symbol       :=  ( \w  |  Schar)  ( \w  |  \d  |  Schar)(*)
-
-Schar        :=  (oneof "<>!@#$%^&*_-+=|:?")
-
-String       :=  '"'  ( Escape  |  (not  ( '"'  |  '\')))(*)  '"'
-
-Escape       :=  '\'  ( '"'  |  '\' )
-
 Whitespace   :=  \s+
@@
 \end{verbatim}
-With the additional rule that whitespace and comments 
-may appear in any amount before any token.
-Tokens are:  {, }, (, ), Symbol, String, and Number.
+With the additional rule that whitespace and comments may appear in any 
+amount before any token.  Tokens are:  {, }, (, ), and Symbol.
 
@@
     | ALambda [String] [AST]
-    | ADefine (Maybe String) String AST
+    | ADefine String AST
     | AApp    AST  [AST]
@@
 Plus we'll need a type for our parsers.  We'll use the basic type mentioned 
-above -- [t] -> m (a, [t]) -- as our starting point.  As luck would have it, that
-corresponds to our first monad transformer:  StateT (Control.Monad.State), so we
-can write:
+above -- \begin{verbatim}[t] -> m (a, [t]) \end{verbatim}
+ -- as our starting point.  This type corresponds to our first monad transformer:  
+ StateT (from Control.Monad.State), so we can write:
 \begin{verbatim}
 type Parser a = StateT [Char] Maybe a
+\end{verbatim}
+and a function for running a parser:
+\begin{verbatim}
+runParser :: Parser a -> String -> Maybe (a, String)
+runParser = runStateT
 \end{verbatim}
@@
 
+The signficance of the Maybe parameter is two-fold:  it allows computational
+effects to express first success versus failure, using the Just and Nothing 
+constructors, respectively, and second backtracking/alternation.  We'll
+capture both of these characteristics with using a type class, Plus:
+\begin{verbatim}
+class Applicative f => Plus f where
+  zero :: f a
+  (<+>) :: f a -> f a -> f a
+\end{verbatim}
+This type class is identical in intent to the Alternative type class from
+Control.Applicative, with zero corresponding to empty and <+> to \verb+<|>+; 
+however, we aren't able to use Alternative directly because
+other of the monad transformers have unsuitable instances for our purposes,
+and as we'll see later, inability to swap out such instances is one of the 
+limitations of the monad transformer approach.
+
+To get around this limitation, we'll create instances for the Plus type class:
+\begin{verbatim}
+instance Plus Maybe where
+  zero = Nothing
+  Nothing <+> y = y
+  x       <+> _ = x
+
+instance (Monad m, Plus m) => Plus (StateT s m) where
+  zero = StateT (const zero)
+  StateT f <+> StateT g = StateT (\s -> f s <+> g s)
+\end{verbatim}
+Note that the StateT instance relies on its monad parameter, m, also being a 
+member of the Plus type class.  This works out for our parser, since m is Maybe
+and we do have a Plus instance for Maybe.
+
+\subsection{Token parsers}
 Now let's get down to the business of building parsers, starting with parsers
@@
 To recognize braces, we need to re-implement the item and check combinators that
-we did before.  Fortunately, using the StateT transformer makes it simpler
-to do so:
+we did before.  Using the MonadState and Plus type classes, 
+we can now implement the basic item parser as a stateful
+computation that fails if the state is empty:
 \begin{verbatim}
@@
 \end{verbatim}
-Although we now also need to discuss the Plus type class, why we need it instead of
-Alternative, and what its instances are.
+And we can easily implement a new version of the check combinator:
+\begin{verbatim}
+check :: (Monad m, Plus m) => (a -> Bool) -> m a -> m a
+check f p =
+    p >>= \x ->
+    if (f x) then return x else zero
+\end{verbatim}
+And use both of them to build a new combinator, satisfy, that checks whether
+the next character meets a condition:
+\begin{verbatim}
+satisfy :: (MonadState [t] m, Plus m) => (t -> Bool) -> m t
+satisfy = flip check item
+\end{verbatim}
+And finally a parser that checks that the next character is equal to a
+given character:
+\begin{verbatim}
+literal :: (Eq t, MonadState [t] m, Plus m) => t -> m t
+literal c = satisfy ((==) c)
+\end{verbatim}
+Now we're ready to build parsers for the four bracket tokens.  
+An exercise for the reader:  what is the inferred type of each parser?
+Let's first build a parser for open curly brackets:
+\begin{verbatim}
+ocurly = literal '{'
+\end{verbatim}
+And let's try it out in ghci to make sure it does what we expect:
+\begin{verbatim}
+ghci$ runParser ocurly "{abc"
+Just ('{', "abc")
+
+ghci$ map (runParser  ocurly) ["}abc", "(abc", "abc"]
+[Nothing, Nothing, Nothing]
+\end{verbatim}
+Good, it correctly accepts open curly brackets and rejects everything else.
+But what about the spec that said that comments and whitespace can occur 
+before any token -- will our parser be able to handle that?
+\begin{verbatim}
+ghci$ map (runParser  ocurly) ["}abc", "(abc", "abc"]
+Nothing
+\end{verbatim}
+Of course not -- we haven't told it how to skip comments and whitespace yet!
+Let's do that now, first by defining the whitespace and comment patterns, then
+by creating a combinator (which we'll call munch) that takes in a parser, 
+throws away all leading junk, then runs the parser:
+\begin{verbatim}
+whitespace = many1 $ satisfy (flip elem " \n\t\r\f")
+comment = pure (:) <*> literal ';' <*> many0 (not1 $ literal '\n')
+
+munch p = many0 (whitespace <+> comment) *> p
+\end{verbatim}
+The comment parser uses a new combinator, not1, which is built out of a new type class,
+Switch, whose intention is to model computations which can be switched from successful
+to failing and vice versa:
+\begin{verbatim}
+class Switch f where
+  switch :: f a -> f ()
+
+instance Switch Maybe where
+  switch (Just _) = Nothing
+  switch Nothing  = Just ()
+
+instance (Functor m, Switch m) => Switch (StateT s m) where
+  switch (StateT f) = StateT (\s -> fmap (const ((), s)) . switch $ f s)
+
+not1 :: (MonadState [t] m, Plus m, Switch m) => m a -> m t
+not1 p = switch p *> item
+\end{verbatim}
+And now let's fix the ocurly parser using munch and try it out:
+\begin{verbatim}
+ocurly = munch $ literal '{'
+
+ghci$ map (runParser ocurly) [";abc\n  {def", "{def", "(def", "}def"]
+[Just ('{',"def"),Just ('{',"def"),Nothing,Nothing]
+\end{verbatim}
+Excellent!  Now the open curly parser correctly throws away whitespace and comments,
+and succeeds, consuming one character, when the first non-junk character is an open 
+curly brace, and fails otherwise.
+We can now define parsers for the other three brace tokens:
+\begin{verbatim}
+ccurly = munch $ literal '}'
+oparen = munch $ literal '('
+cparen = munch $ literal ')'
+\end{verbatim}
+The parser for our last token, Symbol, is slightly more complicated because it allows
+multiple characters and multiple lengths.  We'll use the satisfy combinator along with 
+a function that checks whether a character is one of the legal symbol characters:
+\begin{verbatim}
+symbol = munch $ many1 char
+  where char = satisfy (flip elem (['a' .. 'z'] ++ ['A' .. 'Z']))
+\end{verbatim}
+Note the use of the many1 combinator to introduce repetition of any length of at least one,
+and we're again using the munch combinator to throw away leading junk.
+
+\subsection{Syntactic structures}
+Whereas our token parsers dealt with the syntactic primitives of the Woof grammar, 
+the remaining parsers will have to implement grammar rules that combine the tokens
+into syntactic structures.
+The first combining form is a function application.  The rule says that it's delimited
+by matching parentheses, in between which must appear one form as the operator, followed
+by any number of forms as the arguments to which the operator is applied.  We can
+say that like so:
+\begin{verbatim}
+application =
+    oparen      >>
+    form        >>= \op ->
+    many0 form  >>= \args ->
+    cparen      >>
+    return (AApp op args)
+\end{verbatim}
+Although this implementation uses the monadic operators >> and >>=, it could
+equivalently have been implemented using the applicative combinators *> and <*>:
+\begin{verbatim}
+application' =
+    oparen        *>
+    pure AApp    <*>
+    form         <*>
+    many0 form   <*
+    cparen
+\end{verbatim}
+Next, we move on to define, which is also a straightforward translation from the grammar:
+\begin{verbatim}
+define =
+    ocurly                       *>
+    check (== "define") symbol   *>
+    pure ADefine                <*>
+    symbol                      <*>
+    form                        <*
+    ccurly
+\end{verbatim}
+Lambda is more complicated -- not only does it have additional syntax with its parameter list,
+but we also need to ensure that the parameter names are unique.  We can do that using the 
+check combinator again:
+\begin{verbatim}
+lambda =
+    ocurly                          >>
+    check (== "lambda") symbol      >>
+    ocurly                          >>
+    check distinct (many0 symbol)   >>= \params ->
+    ccurly                          >>
+    many1 form                      >>= \bodies ->
+    ccurly                          >>
+    return (ALambda params bodies)
+  where
+    distinct names = length names == length (nub names)
+\end{verbatim}
+If the symbols are distinct, the parameter list subparser will succeed, and parsing will
+continue; if there's a repeated symbol, the parsing will stop with a failure.
+
+\subsection{Finishing touches}
+Now all we have left to do is to define the special, form (which we've already used to build
+the previous parsers), and woof parsers.  The special parser is simple:  it parses either 
+define or lambda:
+\begin{verbatim}
+special = define <+> lambda
+\end{verbatim}
+The form parser parsers either a symbol, application, or a special form.  We could just write:
+\begin{verbatim}
+form = symbol <+> application <+> special
+\end{verbatim}
+except that the types don't match -- symbol's type parameter is a String, whereas the other two
+have ASTs, so we can just fix that by mapping the correct function over symbol:
+\begin{verbatim}
+form = fmap ASymbol symbol <+> application <+> special
+\end{verbatim}
+The final parser, woof, needs to not only parse all the forms, but also make sure that the
+entire input is consumed.  The end of input check is done with the end parser:
+\begin{verbatim}
+end = switch item
+\end{verbatim}
+And we also need to munch any trailing comments and whitespace, so our final woof parser is:
+\begin{verbatim}
+woof = many1 form <* munch end
+\end{verbatim}
+
+\subsection{Examples}
+Let's look at a few examples of our parser in action:
 
 \section{Example 2: error-reporting}
+While our first attempt at a parser was useful, when it encountered malformed input, it wasn't
+helpful -- all it could do was throw up its hands in despair and give up.  However, we'd like
+an error message that tells us what the problem was and where it was encountered whenever bad
+input is encountered.  Otherwise, we'd waste time finding the problem by eye -- not a lot of 
+fun.  Here are some examples of bad input that we'd like to get specific error messages for,
+but currently can't:
+\begin{verbatim}
+unrecognized character:
+application missing operator:
+application missing close parentheses:
+special form missing close curly:
+invalid special form name:
+duplicate lambda parameters:
+lambda missing body forms:
+lambda missing parameter list:
+define missing symbol:
+define missing form:
+\end{verbatim}
 

### 2013-05-08 src/Common.hs

@@ -6,6 +6,7 @@ module Common (
   , line
   , col
   , countLineCol
+  , simpleExample
   , example
   
 ) where
@@ -16,7 +17,7 @@ data AST
     | ASymbol String
     | AString String
     | ALambda [String] [AST]
-    | ADefine (Maybe String) String AST
+    | ADefine String AST
     | AApp    AST  [AST]
   deriving (Show, Eq)
 
@@ -33,6 +34,18 @@ countLineCol = reverse . snd . foldl f ((1, 1), [])
     f ((line, col), ts)  c   = ((line, col + 1), (c, line, col):ts)
 
 
+simpleExample = "{define \n\
+\  f \n\
+\  {lambda {x y}\n\
+\    (plus x y)}}\n\
+\\n\
+\(a b (c d e))\n\
+\\n\
+\; here's a nice comment !!\n\
+\\n\
+\"
+
+
 example = "{define \n\
 \  f \n\
 \  {lambda {x y}\n\
@@ -40,11 +53,10 @@ example = "{define \n\
 \\n\
 \(* 3 (/ 4 1))\n\
 \\n\
-\; what?? no !!!\n\
+\; here's a nice comment !!\n\
 \\n\
 \{define\n\
-\  \"very important number\"\n\
 \  n\n\
-\  ; seriously, it's really important\n\
+\  ; a very important number\n\
 \  22}\n\
 \"

### 2013-05-08 src/Instances.hs

@@ -35,7 +35,7 @@ instance (Monad m, Plus m) => Plus (StateT s m) where
 instance Switch Maybe where
   switch (Just _) = Nothing
   switch Nothing  = Just ()
-  
+
 instance (Functor m, Switch m) => Switch (StateT s m) where
   switch (StateT f) = StateT (\s -> fmap (const ((), s)) . switch $ f s)
 

### 2013-05-08 src/Parser.hs

@@ -4,6 +4,7 @@ module Parser (
     item
   , check
   , satisfy
+  , literal
   , Switch(..)
   , not1
   , optionalM
@@ -31,6 +32,9 @@ check f p =
 
 satisfy :: (MonadState [t] m, Plus m) => (t -> Bool) -> m t
 satisfy = flip check item
+
+literal :: (Eq t, MonadState [t] m, Plus m) => t -> m t
+literal c = satisfy ((==) c)
 
 not1 :: (MonadState [t] m, Plus m, Switch m) => m a -> m t
 not1 p = switch p *> item

### 2013-05-08 src/eparse.hs

@@ -103,7 +103,6 @@ lambda =
 define open =
     check (== "define") symbol    *>
     pure ADefine                 <*>
-    optionalM string             <*>
     sym                          <*>
     for
   where
@@ -129,7 +128,7 @@ parser = many0 form <* endCheck
                             (t:_) -> throwError ("unparsed input", t)
                             []    -> pure ()
 
-runParse :: Parse e t a -> [t] -> Either' e (Maybe (a, [t]))
-runParse p xs = runMaybeT (runStateT p xs)
+runParser :: Parse e t a -> [t] -> Either' e (Maybe (a, [t]))
+runParser p xs = runMaybeT (runStateT p xs)
 
 type Parse e t a = StateT [t] (MaybeT (Either' e)) a

### 2013-05-08 src/lparse.hs

@@ -122,7 +122,6 @@ lambda =
 define open =
     check (== "define") symbol    *>
     pure ADefine                 <*>
-    optionalM string             <*>
     sym                          <*>
     for
   where

### 2013-05-08 src/parse.hs

@@ -10,30 +10,26 @@ import Common
 import Instances
 
 
-character :: (MonadState [Char] m, Plus m) => Char -> m Char
-character c = satisfy ((==) c)
+ocurly = literal '{'
+ccurly = literal '}'
+oparen = literal '('
+cparen = literal ')'
 
-ocurly = character '{'
-ccurly = character '}'
-oparen = character '('
-cparen = character ')'
+whitespace :: (MonadState [Char] m, Plus m, Switch m) => m [Char]
+whitespace = many1 $ satisfy (flip elem " \n\t\r\f")
+
+comment :: (MonadState [Char] m, Plus m, Switch m) => m [Char]
+comment = literal ';' *> many0 (not1 $ literal '\n')
+
+munch :: (MonadState [Char] m, Plus m, Switch m) => m a -> m a
+munch p = many0 (whitespace <+> comment) *> p
+
 
 digit = satisfy (flip elem ['0' .. '9'])
 
 number = munch $ many1 digit
 
 symbolOpens = ['a' .. 'z'] ++ ['A' .. 'Z'] ++ "!@#$%^&*_-+=:?<>/"
-
-
-whitespace :: (MonadState [Char] m, Plus m, Switch m) => m [Char]
-whitespace = many1 $ satisfy (flip elem " \n\t\r\f")
-
-comment :: (MonadState [Char] m, Plus m, Switch m) => m [Char]
-comment = character ';' *> many0 (not1 $ character '\n')
-
-munch :: (MonadState [Char] m, Plus m, Switch m) => m a -> m a
-munch p = many0 (whitespace <+> comment) *> p
-
 
 symbol = munch (fmap (:) first <*> rest)
   where
@@ -43,11 +39,11 @@ symbol = munch (fmap (:) first <*> rest)
 string :: (MonadState [Char] m, Plus m, Switch m) => m String
 string = munch (dq *> many0 char <* dq)
   where
-    dq = character '"'
+    dq = literal '"'
     char = escape <+> normal
-    escape = character '\\' *> slash_or_dq
+    escape = literal '\\' *> slash_or_dq
     normal = not1 slash_or_dq
-    slash_or_dq = character '\\' <+> character '"'
+    slash_or_dq = literal '\\' <+> literal '"'
 
 
 top = munch oparen
@@ -87,7 +83,6 @@ define :: (MonadState [Char] m, Plus m, Switch m) => m AST
 define =
     check (== "define") symbol   *>
     pure ADefine                <*>
[…]

### 2013-05-08 src/simpleparse.hs

@@ -1,0 +1,62 @@
+{-# LANGUAGE   NoMonomorphismRestriction
+             , FlexibleContexts           #-}
+
+import Control.Monad.State       (MonadState (..), StateT(..), State)
+import Control.Applicative       (Applicative(..))
+import Data.List                 (nub)
+import Parser
+import Classes
+import Common
+import Instances
+
+
+whitespace = many1 $ satisfy (flip elem " \n\t\r\f")
+comment = pure (:) <*> literal ';' <*> many0 (not1 $ literal '\n')
+
+munch p = many0 (whitespace <+> comment) *> p
+
+ocurly = munch $ literal '{'
+ccurly = munch $ literal '}'
+oparen = munch $ literal '('
+cparen = munch $ literal ')'
+symbol = munch $ many1 char
+  where char = satisfy (flip elem (['a' .. 'z'] ++ ['A' .. 'Z']))
+
+application =
+    oparen      >>
+    form        >>= \op ->
+    many0 form  >>= \args ->
+    cparen      >>
+    return (AApp op args)
+
+define =
+    ocurly                       *>
+    check (== "define") symbol   *>
+    pure ADefine                <*>
+    symbol                      <*>
+    form                        <*
+    ccurly
+    
+lambda =
+    ocurly                          >>
+    check (== "lambda") symbol      >>
+    ocurly                          >>
+    check distinct (many0 symbol)   >>= \params ->
+    ccurly                          >>
+    many1 form                      >>= \bodies ->
+    ccurly                          >>
+    return (ALambda params bodies)
+  where
+    distinct names = length names == length (nub names)
+
+special = define <+> lambda
+
+form = fmap ASymbol symbol <+> application <+> special
+
+woof = many1 form <* munch end
+
+
+type Parser a = StateT [Char] Maybe a 
+
+runParser :: Parser a -> String -> Maybe (a, String)
+runParser = runStateT

### 2013-05-08 src/parse.hs
remove old grammar and parser level 1, as they are too complicated to give a succinct demonstration of.  rename parsers to reflect order

@@ -1,106 +1,0 @@
-{-# LANGUAGE   NoMonomorphismRestriction
-             , FlexibleContexts           #-}
-
-import Control.Monad.State       (MonadState (..), StateT(..), State)
-import Control.Applicative       (Applicative(..))
-import Data.List                 (nub)
-import Parser
-import Classes
-import Common
-import Instances
-
-
-ocurly = literal '{'
-ccurly = literal '}'
-oparen = literal '('
-cparen = literal ')'
-
-whitespace :: (MonadState [Char] m, Plus m, Switch m) => m [Char]
-whitespace = many1 $ satisfy (flip elem " \n\t\r\f")
-
-comment :: (MonadState [Char] m, Plus m, Switch m) => m [Char]
-comment = literal ';' *> many0 (not1 $ literal '\n')
-
-munch :: (MonadState [Char] m, Plus m, Switch m) => m a -> m a
-munch p = many0 (whitespace <+> comment) *> p
-
-
-digit = satisfy (flip elem ['0' .. '9'])
-
-number = munch $ many1 digit
-
-symbolOpens = ['a' .. 'z'] ++ ['A' .. 'Z'] ++ "!@#$%^&*_-+=:?<>/"
-
-symbol = munch (fmap (:) first <*> rest)
-  where
-    first = satisfy (flip elem symbolOpens)
-    rest = many0 (first <+> digit)
-    
-string :: (MonadState [Char] m, Plus m, Switch m) => m String
-string = munch (dq *> many0 char <* dq)
-  where
-    dq = literal '"'
-    char = escape <+> normal
-    escape = literal '\\' *> slash_or_dq
-    normal = not1 slash_or_dq
-    slash_or_dq = literal '\\' <+> literal '"'
-
-
-top = munch oparen
-tcp = munch cparen
-toc = munch ocurly
-tcc = munch ccurly
-
-tsymbol :: (MonadState [Char] m, Plus m, Switch m) => m AST
-tsymbol = fmap ASymbol symbol
-
-tstring :: (MonadState [Char] m, Plus m, Switch m) => m AST
-tstring = fmap AString string
-
-tnumber :: (MonadState [Char] m, Plus m, Switch m) => m AST
-tnumber = fmap (ANumber . read) number
-
-application :: (MonadState [Char] m, Plus m, Switch m) => m AST
-application =
-    top        >>= \open ->
-    form       >>= \op ->
-    many0 form  >>= \args ->
-    tcp        >>
-    return (AApp op args)
-    
[…]

### 2013-05-08 src2/Instances.hs
possible backup Transformers and Instances in case MTL doesn't fit my needs

@@ -1,0 +1,146 @@
+-- transformers:
+--   functor, applicative, monad
+--   switch, alternative
+
+-- Id:  ???every type class???
+
+-- Switch:  ???every datatype???
+
+module Instances (
+  
+) where
+
+import Transformers
+import Control.Applicative  (Applicative(..), Alternative(..))
+import Control.Monad        (ap, liftM)
+
+
+class Switch m where
+  switch :: m a -> m ()
+
+instance Switch Maybe where
+  switch  (Just _)  =  Nothing
+  switch  Nothing   =  Just ()
+
+
+instance Monad m => Functor (StateT s m) where
+  fmap = liftM
+
+instance Monad m => Applicative (StateT s m) where
+  pure  = return
+  (<*>) = ap
+
+instance Monad m => Monad (StateT s m) where
+  return x = StateT (\s -> return (s, x))
+  (StateT m1) >>= f = StateT h
+    where
+      h s1 = 
+          m1 s1 >>= \(s2, x) ->
+          getStateT (f x) s2
+
+instance Alternative (StateT s m) where
+  empty = undefined
+  (<|>) = undefined
+{-
+-- Plus, Zero, Switch:  'lift' semantics
+instance Plus m => Plus (StateT s m) where
+  StateT f  <+>  StateT g  =  StateT (\s -> f s <+> g s)
+  
+instance Zero m => Zero (StateT s m) where
+  zero = StateT (const zero)
+
+instance Switch m => Switch (StateT s m) where
+  switch (StateT f) = StateT (\s -> fmap (const (s, ())) . switch $ f s)
+
+
+
+instance Functor m => Functor (ErrorT e m) where
+  fmap f (ErrorT m) = ErrorT (fmap (fmap f) m)
+
+instance Pointed m => Pointed (ErrorT e m) where
+  pure = ErrorT . pure . pure
+
+instance Applicative' m => Applicative' (ErrorT e m) where
+  ErrorT f <*> ErrorT x = ErrorT (pure (<*>) <*> f <*> x)
+
+instance Monad' m => Monad' (ErrorT e m) where
+  join =
+      ErrorT                 .
+      fmap join              .
+      join                   .
+      fmap commute           .
+      fmap (fmap getErrorT)  .
+      getErrorT
+
+-- Plus, Zero, Switch:  'lift' semantics
+instance Plus m => Plus (ErrorT e m) where
[…]

### 2013-05-08 src2/Transformers.hs

@@ -1,0 +1,54 @@
+{-# LANGUAGE FunctionalDependencies #-}
+module Transformers (
+  
+    Trans   (..)
+  , TMaybe  (..)
+  , TState  (..)
+  , TError  (..)
+  , TWriter (..)
+  
+  , Id      (..)
+  , MaybeT  (..)
+  , StateT  (..)
+  , ErrorT  (..)
+  , WriterT (..)
+  
+) where
+
+
+class Trans t m where
+  lift :: Monad m => m a -> t m a
+
+
+class Monad m => TMaybe m where
+  mzero :: m a
+
+class Monad m => TState s m | m -> s where
+  get :: m s
+  put :: s -> m ()
+
+class Monad m => TError e m | m -> e where
+  throwE :: e -> m a
+  catchE :: m a -> (e -> m a) -> m a
+  
+class Monad m => TWriter w m | m -> w where
+  tell :: w -> m ()
+
+
+  
+newtype Id a =
+    Id {getId :: a}
+  deriving (Show)
+
+newtype MaybeT m a =
+    MaybeT {getMaybeT :: m (Maybe a)}
+
+newtype StateT s m a =
+    StateT {getStateT :: s -> m (s, a)}
+
+newtype ErrorT e m a =
+    ErrorT {getErrorT :: m (Either e a)}
+    
+newtype WriterT w m a =
+    WriterT {getWriterT :: m (w, a)}
+  

### 2013-05-08 src2/Instances.hs
remove backup Transformers and Instances

@@ -1,146 +1,0 @@
--- transformers:
---   functor, applicative, monad
---   switch, alternative
-
--- Id:  ???every type class???
-
--- Switch:  ???every datatype???
-
-module Instances (
-  
-) where
-
-import Transformers
-import Control.Applicative  (Applicative(..), Alternative(..))
-import Control.Monad        (ap, liftM)
-
-
-class Switch m where
-  switch :: m a -> m ()
-
-instance Switch Maybe where
-  switch  (Just _)  =  Nothing
-  switch  Nothing   =  Just ()
-
-
-instance Monad m => Functor (StateT s m) where
-  fmap = liftM
-
-instance Monad m => Applicative (StateT s m) where
-  pure  = return
-  (<*>) = ap
-
-instance Monad m => Monad (StateT s m) where
-  return x = StateT (\s -> return (s, x))
-  (StateT m1) >>= f = StateT h
-    where
-      h s1 = 
-          m1 s1 >>= \(s2, x) ->
-          getStateT (f x) s2
-
-instance Alternative (StateT s m) where
-  empty = undefined
-  (<|>) = undefined
-{-
--- Plus, Zero, Switch:  'lift' semantics
-instance Plus m => Plus (StateT s m) where
-  StateT f  <+>  StateT g  =  StateT (\s -> f s <+> g s)
-  
-instance Zero m => Zero (StateT s m) where
-  zero = StateT (const zero)
-
-instance Switch m => Switch (StateT s m) where
-  switch (StateT f) = StateT (\s -> fmap (const (s, ())) . switch $ f s)
-
-
-
-instance Functor m => Functor (ErrorT e m) where
-  fmap f (ErrorT m) = ErrorT (fmap (fmap f) m)
-
-instance Pointed m => Pointed (ErrorT e m) where
-  pure = ErrorT . pure . pure
-
-instance Applicative' m => Applicative' (ErrorT e m) where
-  ErrorT f <*> ErrorT x = ErrorT (pure (<*>) <*> f <*> x)
-
-instance Monad' m => Monad' (ErrorT e m) where
-  join =
-      ErrorT                 .
-      fmap join              .
-      join                   .
-      fmap commute           .
-      fmap (fmap getErrorT)  .
-      getErrorT
-
--- Plus, Zero, Switch:  'lift' semantics
-instance Plus m => Plus (ErrorT e m) where
[…]

### 2013-05-08 src2/Transformers.hs

@@ -1,54 +1,0 @@
-{-# LANGUAGE FunctionalDependencies #-}
-module Transformers (
-  
-    Trans   (..)
-  , TMaybe  (..)
-  , TState  (..)
-  , TError  (..)
-  , TWriter (..)
-  
-  , Id      (..)
-  , MaybeT  (..)
-  , StateT  (..)
-  , ErrorT  (..)
-  , WriterT (..)
-  
-) where
-
-
-class Trans t m where
-  lift :: Monad m => m a -> t m a
-
-
-class Monad m => TMaybe m where
-  mzero :: m a
-
-class Monad m => TState s m | m -> s where
-  get :: m s
-  put :: s -> m ()
-
-class Monad m => TError e m | m -> e where
-  throwE :: e -> m a
-  catchE :: m a -> (e -> m a) -> m a
-  
-class Monad m => TWriter w m | m -> w where
-  tell :: w -> m ()
-
-
-  
-newtype Id a =
-    Id {getId :: a}
-  deriving (Show)
-
-newtype MaybeT m a =
-    MaybeT {getMaybeT :: m (Maybe a)}
-
-newtype StateT s m a =
-    StateT {getStateT :: s -> m (s, a)}
-
-newtype ErrorT e m a =
-    ErrorT {getErrorT :: m (Either e a)}
-    
-newtype WriterT w m a =
-    WriterT {getWriterT :: m (w, a)}
-  

### 2013-05-08 src/Common.hs
clean up error parser to implement simple grammar and consolidate error messages; clean up unused imports

@@ -6,7 +6,6 @@ module Common (
   , line
   , col
   , countLineCol
-  , simpleExample
   , example
   
 ) where
@@ -34,7 +33,7 @@ countLineCol = reverse . snd . foldl f ((1, 1), [])
     f ((line, col), ts)  c   = ((line, col + 1), (c, line, col):ts)
 
 
-simpleExample = "{define \n\
+example = "{define \n\
 \  f \n\
 \  {lambda {x y}\n\
 \    (plus x y)}}\n\
@@ -43,20 +42,4 @@ simpleExample = "{define \n\
 \\n\
 \; here's a nice comment !!\n\
 \\n\
-\"
-
-
-example = "{define \n\
-\  f \n\
-\  {lambda {x y}\n\
-\    (+ x y)}}\n\
-\\n\
-\(* 3 (/ 4 1))\n\
-\\n\
-\; here's a nice comment !!\n\
-\\n\
-\{define\n\
-\  n\n\
-\  ; a very important number\n\
-\  22}\n\
 \"

### 2013-05-08 src/parse1.hs

@@ -1,7 +1,7 @@
 {-# LANGUAGE   NoMonomorphismRestriction
              , FlexibleContexts           #-}
 
-import Control.Monad.State       (MonadState (..), StateT(..), State)
+import Control.Monad.State       (MonadState (..), StateT(..))
 import Control.Applicative       (Applicative(..))
 import Data.List                 (nub)
 import Parser

### 2013-05-08 src/parse2.hs

@@ -1,12 +1,11 @@
 {-# LANGUAGE   NoMonomorphismRestriction
              , FlexibleContexts           #-}
 
-import Control.Monad.State       (MonadState (..), StateT(..), State)
-import Control.Monad.Error       (MonadError (..), ErrorT(..))
+import Control.Monad.State       (MonadState (..), StateT(..))
+import Control.Monad.Error       (MonadError (..))
 import Control.Monad.Trans.Maybe (MaybeT(..))
 import Control.Applicative       (Applicative(..))
 import Data.List                 (nub)
-import Data.Functor.Identity     (Identity(..))
 import Parser
 import Classes
 import Instances
@@ -19,114 +18,72 @@ type Error = (String, Token)
 character :: (MonadState [Token] m, Plus m) => Char -> m Token
 character c = satisfy ((==) c . chr)
 
-ocurly = character '{'
-
-ccurly = character '}'
-
-oparen = character '('
-
-cparen = character ')'
-
-digit = fmap chr $ satisfy (flip elem ['0' .. '9'] . chr)
-
-number = munch $ many1 digit
-
-symbolOpens = ['a' .. 'z'] ++ ['A' .. 'Z'] ++ "!@#$%^&*_-+=:?<>/"
-
-symbol = munch (fmap (:) first <*> rest)
-  where
-    first = fmap chr $ satisfy (flip elem symbolOpens . chr)
-    rest = many0 (first <+> digit)
-    
--- string :: (MonadState [Token] m, Plus m, Switch m) => m String
-string = munch (dq          >>= \open ->
-                many0 char  <* 
-                commit ("unclosed string literal", open) dq)
-  where
-    dq = character '"'
-    char = fmap chr (escape <+> normal)
-    escape = character '\\' *> slash_or_dq
-    normal = not1 slash_or_dq
-    slash_or_dq = character '\\' <+> character '"'
-
-
--- whitespace :: (MonadState [Token] m, Plus m, Switch m) => m [Token]
 whitespace = many1 $ satisfy (flip elem " \n\t\r\f" . chr)
-
--- comment :: (MonadState [Token] m, Plus m, Switch m) => m [Token]
 comment = character ';' *> many0 (not1 $ character '\n')
 
--- munch :: (MonadState [Token] m, Plus m, Switch m) => m a -> m a
 munch p = many0 (whitespace <+> comment) *> p
 
-top = munch oparen
-tcp = munch cparen
-toc = munch ocurly
[…]

### 2013-05-08 src/Instances.hs
tidy up source: remove unused imports and instances, remove line-ending whitespace, alignment

@@ -6,7 +6,7 @@ module Instances (
 import Classes
 import Control.Monad              (liftM, ap)
 import Control.Applicative        (Applicative(..))
-import Control.Monad.Error        (ErrorT(..), MonadError(..))
+import Control.Monad.Error        (MonadError(..))
 import Control.Monad.State        (StateT(..))
 import Control.Monad.Trans.Maybe  (MaybeT(..))
 
@@ -23,10 +23,6 @@ instance (Functor m, Monad m) => Plus (MaybeT m) where
       x = l >>= \y -> case y of Nothing -> r;
                                 Just _  -> return y;
 
-instance (Monad m, Plus m) => Plus (ErrorT e m) where
-  zero = ErrorT zero
-  ErrorT a <+> ErrorT b = ErrorT (a <+> b)
-
 instance (Monad m, Plus m) => Plus (StateT s m) where
   zero = StateT (const zero)
   StateT f <+> StateT g = StateT (\s -> f s <+> g s)
@@ -42,20 +38,20 @@ instance (Functor m, Switch m) => Switch (StateT s m) where
 instance Functor m => Switch (MaybeT m) where
   switch (MaybeT m) = MaybeT (fmap switch m)
 
-  
+
 instance Functor (Either' e) where
   fmap = liftM
-    
+
 instance Applicative (Either' e) where
   pure = return
   (<*>) = ap
-  
+
 instance Monad (Either' e) where
-  return = Right'
+  return            =  Right'
   Left' e   >>=  _  =  Left' e
   Right' x  >>=  f  =  f x
 
 instance MonadError e (Either' e) where
-  throwError = Left'
-  catchError (Right' x) _ = Right' x
-  catchError (Left' e)  f = f e
+  throwError                =  Left'
+  catchError  (Right' x) _  =  Right' x
+  catchError  (Left' e)  f  =  f e

### 2013-05-08 src/parse2.hs

@@ -19,7 +19,7 @@ character :: (MonadState [Token] m, Plus m) => Char -> m Token
 character c = satisfy ((==) c . chr)
 
 whitespace = many1 $ satisfy (flip elem " \n\t\r\f" . chr)
-comment = character ';' *> many0 (not1 $ character '\n')
+comment = pure (:) <*> character ';' <*> many0 (not1 $ character '\n')
 
 munch p = many0 (whitespace <+> comment) *> p
 
@@ -28,7 +28,7 @@ ccurly = munch $ character '}'
 oparen = munch $ character '('
 cparen = munch $ character ')'
 symbol = munch $ many1 char
-  where char = fmap chr $ satisfy (\t -> elem (chr t) (['a' .. 'z'] ++ ['A' .. 'Z']))
+  where char = fmap chr $ satisfy (flip elem (['a' .. 'z'] ++ ['A' .. 'Z']) . chr)
 
 eaOp   =  "application: missing operator"
 eaCls  =  "application: missing close parenthesis"
@@ -40,8 +40,9 @@ elPrms =  "lambda: duplicate parameter names"
 elPCls =  "lambda: missing parameter list close curly"
 elBody =  "lambda: missing body form"
 elCls  =  "lambda: missing close curly"
-ewUnp  =  "woof: unparsed input"
 esName =  "special form: unable to parse"
+ewUnp  =  "woof: unparsed input"
+-- other possibilities:  non-symbol in parameter list
 
 application =
     oparen                       >>= \open ->
@@ -85,7 +86,8 @@ endCheck =
 
 woof = many0 form <* endCheck
 
-runParser :: Parse e t a -> [t] -> Either' e (Maybe (a, [t]))
-runParser p xs = runMaybeT (runStateT p xs)
 
 type Parse e t a = StateT [t] (MaybeT (Either' e)) a
+
+runParser :: Parse e t a -> [t] -> Either' e (Maybe (a, [t]))
+runParser p xs = runMaybeT (runStateT p xs)

### 2013-05-08 article.tex
work on error reporting section of article

@@
 
-The signficance of the Maybe parameter is two-fold:  it allows computational
+The significance of the Maybe parameter is two-fold:  it allows computational
 effects to express first success versus failure, using the Just and Nothing 
@@
 input is encountered.  Otherwise, we'd waste time finding the problem by eye -- not a lot of 
-fun.  Here are some examples of bad input that we'd like to get specific error messages for,
-but currently can't:
+fun.  Here are the examples of bad input that we'd like to get specific error messages for:
+\begin{itemize}
+\item application: missing operator:  \begin{verbatim}()\end{verbatim}
+\item application: missing close parenthesis:  \begin{verbatim}(a b\end{verbatim}
+\item define: missing symbol:  \begin{verbatim}{define (a b c)}\end{verbatim}
+\item define: missing form:  \begin{verbatim}{define a}\end{verbatim}
+\item define: missing close curly:  \begin{verbatim}{define a b\end{verbatim}
+\item lambda: missing parameter list:  \begin{verbatim}{lambda (a b c)}\end{verbatim}
+\item lambda: duplicate parameter names:  \begin{verbatim}{lambda {a b a} (c d)}\end{verbatim}
+\item lambda: missing parameter list close curly:  \begin{verbatim}{lambda {a b (c d)}\end{verbatim}
+\item lambda: missing body form:  \begin{verbatim}{lambda {a b}}\end{verbatim}
+\item lambda: missing close curly:  \begin{verbatim}{lambda {a b} (c d)\end{verbatim}
+\item special form: unable to parse:  \begin{verbatim}{defin x y}\end{verbatim}
+\item woof: unparsed input:  \begin{verbatim}a,b\end{verbatim}
+\end{itemize}
+To do this, we're going to bring in a second monad transformer type class -- MonadError -- and
+another monad transformer datatype -- MaybeT.  Our parser stack and runParser function are now
+going to be:
 \begin{verbatim}
-unrecognized character:
-application missing operator:
-application missing close parentheses:
-special form missing close curly:
-invalid special form name:
-duplicate lambda parameters:
-lambda missing body forms:
-lambda missing parameter list:
-define missing symbol:
-define missing form:
+type Parse e t a = StateT [t] (MaybeT (Either' e)) a
+
+runParser :: Parse e t a -> [t] -> Either' e (Maybe (a, [t]))
+runParser p xs = runMaybeT (runStateT p xs)
+\end{verbatim}
+No, that ' isn't accidental -- we again introduce a replacement data type to avoid a type class
+instance that doesn't fulfill our needs.  In this case, Either' is intended as a replacement for
+Either, with the only difference being its instance of MonadError.  All other instances for 
+Either' are analagous to those of Either:
+\begin{verbatim}
+data Either' a b
+    = Left' a
+    | Right' b
+  deriving (Show, Eq)
+
+instance MonadError e (Either' e) where
+  throwError                =  Left'
+  catchError  (Right' x) _  =  Right' x
+  catchError  (Left' e)  f  =  f e
+\end{verbatim}
+We'll also need an instance of Switch for MaybeT:
+\begin{verbatim}
+instance Functor m => Switch (MaybeT m) where
+  switch (MaybeT m) = MaybeT (fmap switch m)
+\end{verbatim}
+
+\subsection{Reporting position}
+To be able to report the position of the error, we'll pre-process the input string, calculating
+the line and column number of each character.  Then instead of our parsers operating on a list
+of characters, they'll operate on a list of Tokens, where a Token is a character tupled up with
+a line number and column number, which we'll express with a simple typedef and some accessor functions:
+\begin{verbatim}
+type Token = (Char, Int, Int)
+chr  (a, _, _)  =  a
+line (_, b, _)  =  b
+col  (_, _, c)  =  c
+\end{verbatim}
+The function responsible for calculating the position will simply transform a list of characters into
+a list of tokens:
+\begin{verbatim}
+countLineCol :: [Char] -> [Token]
+countLineCol = reverse . snd . foldl f ((1, 1), [])
+  where
+    f ((line, col), ts) '\n' = ((line + 1, 1), ('\n', line, col):ts)
+    f ((line, col), ts)  c   = ((line, col + 1), (c, line, col):ts)
+\end{verbatim}
+
+\subsection{Token parsers}
+Unfortunately, by adding the position to the input, we've created a new problem for ourselves: our 
+original token parsers assumed they would be operating on character lists, so we'll have to rewrite
+them to get rid of that assumption.  The cause of the problem is the \verb+literal+ combinator, which
+we can easily replace with this one that assumes token lists:
+\begin{verbatim}
+character :: (MonadState [Token] m, Plus m) => Char -> m Token
+character c = satisfy ((==) c . chr)
+\end{verbatim}
+And now our token parsers are:
+\begin{verbatim}
+whitespace = many1 $ satisfy (flip elem " \n\t\r\f" . chr)
+comment = pure (:) <*> character ';' <*> many0 (not1 $ character '\n')
+
+munch p = many0 (whitespace <+> comment) *> p
+
+ocurly = munch $ character '{'
+ccurly = munch $ character '}'
+oparen = munch $ character '('
+cparen = munch $ character ')'
+symbol = munch $ many1 char
+  where char = fmap chr $ satisfy (flip elem (['a' .. 'z'] ++ ['A' .. 'Z']) . chr)
+\end{verbatim}
+The only differences are that literal was replaced everywhere by character, and a couple uses
+of the chr accessor.  The munch combinator didn't have to change at all (although its inferred
+type did in fact change).
+
+\subsection{Error messages}
+Now we get to the meat of this section -- modifying the parsers to report accurate and useful errors.
+Let's start out with a list, in code, of all the errors that we're going to deal with:
+\begin{verbatim}
+eaOp   =  "application: missing operator"
+eaCls  =  "application: missing close parenthesis"
+edSym  =  "define: missing symbol"
+edForm =  "define: missing form"
+edCls  =  "define: missing close curly"
+elPL   =  "lambda: missing parameter list"
+elPrms =  "lambda: duplicate parameter names"
+elPCls =  "lambda: missing parameter list close curly"
+elBody =  "lambda: missing body form"
+elCls  =  "lambda: missing close curly"
+esName =  "special form: unable to parse"
+ewUnp  =  "woof: unparsed input"
+\end{verbatim}
+The MonadError type class provides two useful combinators, throwError for generating errors, and 
+catchError for dealing with them.  We can ignore catchError because we don't want to try to recover
+from errors, we just want them reported.  Now the question is, when should an error be created
+and what data should it include?  \verb+throwError+ is too basic for our needs, so we'll introduce
+the \verb+commit+ combinator:
+\begin{verbatim}
+commit :: (MonadError e m, Plus m) => e -> m a -> m a
+commit err p = p <+> throwError err
+\end{verbatim}
+This combinator takes two arguments: 1. an error value, and 2. a monadic computation.  It first tries
+to run the computation, and if the computation fails, generates an error whose contents are the first
+argument.  The definition of 'failure' is provided by the \verb+Plus+ instance of \verb+m+.
+So what data do we want to include in our errors?  A String describing the error and a token reporting
+the position should describe the error decently:
+\begin{verbatim}
+type Error = (String, Token)
+\end{verbatim}
+Now lets start with \verb+Application+ as our first modified, error-reporting parser:
+\begin{verbatim}
+application =
+    oparen                       >>= \open ->
+    commit (eaOp, open) form     >>= \op ->
+    many0 form                   >>= \args ->
+    commit (eaCls, open) cparen  >>
+    return (AApp op args)
+\end{verbatim}
+This parser works by running the open-parenthesis parser, capturing its value in \verb+open+, and continuing
+with the parsing; if either \verb+form+ or close-parenthesis fails, an appropriate error can be generated
+which includes a string and the location of the open parenthesis.  However, if the open parenthesis is not
+successfully parsed, it's a failure but not an error.  Here are some examples to demonstrate this in action:
+\begin{verbatim}
+ghci$ runParser application (countLineCol "oops")
+Right' Nothing
+
+ghci$ runParser application (countLineCol "(a b c)")
+Right' (Just (AApp (ASymbol "a") [ASymbol "b",ASymbol "c"],[]))
+
+ghci$ runParser application (countLineCol "\n()")
+Left' ("application: missing operator",('(',2,1))
+
+ghci$ runParser application (countLineCol "(a b")
+Left' ("application: missing close parenthesis",('(',1,1))
+\end{verbatim}
+Although the previous version of the \verb+application+ parser could be implemented with just applicative combinators, 
+this version certainly can't:  \verb+open+ is used to build the following parsers, making this decidedly monadic.
+
+The \verb+define+ and \verb+lambda+ parsers are upgraded similarly:
+\begin{verbatim}
+define =
+    ocurly                       >>= \open ->
+    check (== "define") symbol    *>
+    pure ADefine                 <*>
+    commit (edSym, open) symbol  <*>
+    commit (edForm, open) form   <*
+    commit (edCls, open) ccurly
+
+lambda =
+    ocurly                               >>= \open ->
+    check (== "lambda") symbol           >>
+    commit (elPL, open) ocurly           >>= \p_open ->
+    many0 symbol                         >>= \params ->
+    (if distinct params 
+        then return ()
+        else throwError (elPrms, open))  >>
+    commit (elPCls, open) ccurly         >>
+    commit (elBody, open) (many1 form)   >>= \bodies ->
+    commit (elCls, open) ccurly          >>
+    return (ALambda params bodies)
+  where
+    distinct names = length names == length (nub names)
+\end{verbatim}
+Note that \verb+commit+ isn't used until we can be sure that we're in the right rule:  if, in the \verb+define+
+parser, we committed to parsing the "define" keyword, we wouldn't be able to backtrack and try the \verb+lambda+
+parser if that failed.  Thus, if treating errors as unrecoverable, it's important not to place a \verb+commit+
+where backtracking might be necessary to parse a valid alternative.  However, this isn't the case with how
+we've used it, since once we see an open curly brace and a define, we're sure that we're in the \verb+define+
+rule, and allowing backtracking would only destroy our ability to produce clean and accurate error reports.
+
+\subsection{Error messages:  finishing touches}
+To complete the error-reporting parser, we need to change a couple more details.  First, we want an error 
+reported when an open curly brace is found, but no special form can be reported.  To accomplish this, we
+can extend the \verb+special+ parser with a third alternative, which results in an error if it can parse
+an open curly brace, but in a backtracking failure if not:
+\begin{verbatim}
+special = define <+> lambda <+> (ocurly >>= \o -> throwError (esName, o))
+\end{verbatim}
+The \verb+form+ parser doesn't have to be changed:
+\begin{verbatim}
+form = fmap ASymbol symbol <+> application <+> special
+\end{verbatim}
+When we can no longer parse any forms, instead of failing, we'd like to report an error if there's any input
+left, and otherwise succeed.  To implement this parser, we'll first munch any leading whitespace or comments, 
+as usual, and then we'll check the input:  if it's empty, great; if not, report an error at the location of the
+first remaining token, along with the appropriate error message that we defined earlier:
+\begin{verbatim}
+endCheck = 
+    many0 (whitespace <+> comment) *>
+    get >>= \xs -> case xs of
+                        (t:_) -> throwError (ewUnp, t)
+                        []    -> pure ()
+\end{verbatim}
+Now our \verb+woof+ parser uses this \verb+endCheck+ parser after it has consumed all the forms that it can:
+\begin{verbatim}
+woof = many0 form <* endCheck
+\end{verbatim}
+
+\subsection{Error messages:  examples}
+While correct input will be parsed in the same way by our new parsers, incorrect input will now (hopefully) be
+accurately reported as such.  Successful results will be slightly different, however -- there is now an extra
+constructor wrapper around the value:
+\begin{verbatim}
+runParser woof (countLineCol "{define fgh {lambda {f x y} {define a (b c)} (f y a (a x))}}")
+Right' (Just ([ADefine "fgh" 
+                       (ALambda ["f","x","y"] 
+                                [ADefine "a" (AApp (ASymbol "b") [ASymbol "c"]),
+                                 AApp (ASymbol "f") [ASymbol "y",
+                                                     ASymbol "a",
+                                                     AApp (ASymbol "a") [ASymbol "x"]]])],
+              []))
+\end{verbatim}
+Here are examples showing errors detected by the parser:
+\begin{verbatim}
+ghci$ map (runParser woof . countLineCol) ["()", "(a b", "{define (a b c)}",  
+                                     "{define a}", "{define a b", 
+                                     "{lambda (a b c)}", "{lambda {a b a} (c d)}", 
+                                     "{lambda {a b (c d)}", "{lambda {a b}}", 
+                                     "{lambda {a b} (c d)", "{defin x y}", "a,b"]
+Left' ("application: missing operator",('(',1,1))
+Left' ("application: missing close parenthesis",('(',1,1))
+Left' ("define: missing symbol",('{',1,1))
+Left' ("define: missing form",('{',1,1))
+Left' ("define: missing close curly",('{',1,1))
+Left' ("lambda: missing parameter list",('{',1,1))
+Left' ("lambda: duplicate parameter names",('{',1,1))
+Left' ("lambda: missing parameter list close curly",('{',1,1))
+Left' ("lambda: missing body form",('{',1,1))
+Left' ("lambda: missing close curly",('{',1,1))
+Left' ("special form: unable to parse",('{',1,1))
+Left' ("woof: unparsed input",(',',1,2))
 \end{verbatim}

### 2013-05-10 src/Classes.hs
move data types, instances, and combinators into specific parser files (yes, introducing duplication possibly) in order to make each parser totally self-contained and very clear exactly what instances and so on it needs

@@ -1,33 +1,0 @@
-module Classes (
-
-    Plus(..)
-  , many0
-  , many1
-  , Switch(..)
-  , Either'(..)
-  
-) where
-
-import Control.Applicative  (Applicative(..))
-
-
-class Applicative f => Plus f where
-  zero :: f a
-  (<+>) :: f a -> f a -> f a
-
-
-many0 :: Plus f => f a -> f [a]
-many0 p = many1 p <+> pure []
-
-many1 :: Plus f => f a -> f [a]
-many1 p = pure (:) <*> p <*> many0 p
-
-
-class Switch f where
-  switch :: f a -> f ()
-
-
-data Either' a b
-    = Left' a
-    | Right' b
-  deriving (Show, Eq)

### 2013-05-10 src/Common.hs

@@ -1,45 +1,0 @@
-module Common (
-
-    AST(..)
-  , Token
-  , chr
-  , line
-  , col
-  , countLineCol
-  , example
-  
-) where
-
-
-data AST
-    = ANumber Integer
-    | ASymbol String
-    | AString String
-    | ALambda [String] [AST]
-    | ADefine String AST
-    | AApp    AST  [AST]
-  deriving (Show, Eq)
-
-
-type Token = (Char, Int, Int)
-chr  (a, _, _)  =  a
-line (_, b, _)  =  b
-col  (_, _, c)  =  c
-
-countLineCol :: [Char] -> [Token]
-countLineCol = reverse . snd . foldl f ((1, 1), [])
-  where
-    f ((line, col), ts) '\n' = ((line + 1, 1), ('\n', line, col):ts)
-    f ((line, col), ts)  c   = ((line, col + 1), (c, line, col):ts)
-
-
-example = "{define \n\
-\  f \n\
-\  {lambda {x y}\n\
-\    (plus x y)}}\n\
-\\n\
-\(a b (c d e))\n\
-\\n\
-\; here's a nice comment !!\n\
-\\n\
-\"

### 2013-05-10 src/Instances.hs

@@ -1,57 +1,0 @@
-{-# LANGUAGE FlexibleInstances, MultiParamTypeClasses #-}
-module Instances (
-
-) where
-
-import Classes
-import Control.Monad              (liftM, ap)
-import Control.Applicative        (Applicative(..))
-import Control.Monad.Error        (MonadError(..))
-import Control.Monad.State        (StateT(..))
-import Control.Monad.Trans.Maybe  (MaybeT(..))
-
-
-instance Plus Maybe where
-  zero = Nothing
-  Nothing <+> y = y
-  x       <+> _ = x
-
-instance (Functor m, Monad m) => Plus (MaybeT m) where
-  zero = MaybeT (return Nothing)
-  MaybeT l  <+>  MaybeT r  =  MaybeT x
-    where
-      x = l >>= \y -> case y of Nothing -> r;
-                                Just _  -> return y;
-
-instance (Monad m, Plus m) => Plus (StateT s m) where
-  zero = StateT (const zero)
-  StateT f <+> StateT g = StateT (\s -> f s <+> g s)
-
-
-instance Switch Maybe where
-  switch (Just _) = Nothing
-  switch Nothing  = Just ()
-
-instance (Functor m, Switch m) => Switch (StateT s m) where
-  switch (StateT f) = StateT (\s -> fmap (const ((), s)) . switch $ f s)
-
-instance Functor m => Switch (MaybeT m) where
-  switch (MaybeT m) = MaybeT (fmap switch m)
-
-
-instance Functor (Either' e) where
-  fmap = liftM
-
-instance Applicative (Either' e) where
-  pure = return
-  (<*>) = ap
-
-instance Monad (Either' e) where
-  return            =  Right'
-  Left' e   >>=  _  =  Left' e
-  Right' x  >>=  f  =  f x
-
-instance MonadError e (Either' e) where
-  throwError                =  Left'
-  catchError  (Right' x) _  =  Right' x
-  catchError  (Left' e)  f  =  f e

### 2013-05-10 src/Parser.hs

@@ -1,49 +1,0 @@
-{-# LANGUAGE FlexibleContexts, NoMonomorphismRestriction #-}
-module Parser (
-    
-    item
-  , check
-  , satisfy
-  , literal
-  , Switch(..)
-  , not1
-  , optionalM
-  , end
-  , commit
-  
-) where
-
-import Control.Monad.State  (MonadState(..), StateT(..))
-import Control.Monad.Error  (MonadError(..))
-import Control.Applicative  (Applicative(..))
-import Classes
-
-
-item :: (MonadState [t] m, Plus m) => m t
-item =
-    get >>= \xs -> case xs of
-                        (t:ts) -> put ts *> pure t;
-                        []     -> zero;
-
-check :: (Monad m, Plus m) => (a -> Bool) -> m a -> m a
-check f p =
-    p >>= \x ->
-    if (f x) then return x else zero
-
-satisfy :: (MonadState [t] m, Plus m) => (t -> Bool) -> m t
-satisfy = flip check item
-
-literal :: (Eq t, MonadState [t] m, Plus m) => t -> m t
-literal c = satisfy ((==) c)
-
-not1 :: (MonadState [t] m, Plus m, Switch m) => m a -> m t
-not1 p = switch p *> item
-
-optionalM :: (Functor f, Plus f) => f a -> f (Maybe a)
-optionalM p = fmap Just p <+> pure Nothing
-
-end :: (MonadState [t] m, Plus m, Switch m) => m ()
-end = switch item
-
-commit :: (MonadError e m, Plus m) => e -> m a -> m a
-commit err p = p <+> throwError err

### 2013-05-10 src/parse1.hs

@@ -4,11 +4,84 @@
 import Control.Monad.State       (MonadState (..), StateT(..))
 import Control.Applicative       (Applicative(..))
 import Data.List                 (nub)
-import Parser
-import Classes
-import Common
-import Instances
 
+
+-- the prelude:  type classes, data types, instances, and general combinators
+
+class Applicative f => Plus f where
+  zero :: f a
+  (<+>) :: f a -> f a -> f a
+
+many0 :: Plus f => f a -> f [a]
+many0 p = many1 p <+> pure []
+
+many1 :: Plus f => f a -> f [a]
+many1 p = pure (:) <*> p <*> many0 p
+
+instance Plus Maybe where
+  zero = Nothing
+  Nothing <+> y = y
+  x       <+> _ = x
+
+instance (Monad m, Plus m) => Plus (StateT s m) where
+  zero = StateT (const zero)
+  StateT f <+> StateT g = StateT (\s -> f s <+> g s)
+
+
+class Switch f where
+  switch :: f a -> f ()
+
+instance Switch Maybe where
+  switch (Just _) = Nothing
+  switch Nothing  = Just ()
+
+instance (Functor m, Switch m) => Switch (StateT s m) where
+  switch (StateT f) = StateT (\s -> fmap (const ((), s)) . switch $ f s)
+
+data AST
+    = ANumber Integer
+    | ASymbol String
+    | AString String
+    | ALambda [String] [AST]
+    | ADefine String AST
+    | AApp    AST  [AST]
+  deriving (Show, Eq)
+
+item :: (MonadState [t] m, Plus m) => m t
+item =
+    get >>= \xs -> case xs of
+                        (t:ts) -> put ts *> pure t;
+                        []     -> zero;
+
+check :: (Monad m, Plus m) => (a -> Bool) -> m a -> m a
+check f p =
+    p >>= \x ->
+    if (f x) then return x else zero
+
+satisfy :: (MonadState [t] m, Plus m) => (t -> Bool) -> m t
+satisfy = flip check item
+
+literal :: (Eq t, MonadState [t] m, Plus m) => t -> m t
+literal c = satisfy ((==) c)
+
+not1 :: (MonadState [t] m, Plus m, Switch m) => m a -> m t
+not1 p = switch p *> item
+
+end :: (MonadState [t] m, Plus m, Switch m) => m ()
+end = switch item
+
[…]

### 2013-05-10 src/parse2.hs

@@ -1,15 +1,140 @@
 {-# LANGUAGE   NoMonomorphismRestriction
-             , FlexibleContexts           #-}
+             , FlexibleContexts       
+             , FlexibleInstances    
+             , MultiParamTypeClasses  #-}
 
+
+import Control.Monad             (liftM, ap)
 import Control.Monad.State       (MonadState (..), StateT(..))
 import Control.Monad.Error       (MonadError (..))
 import Control.Monad.Trans.Maybe (MaybeT(..))
 import Control.Applicative       (Applicative(..))
 import Data.List                 (nub)
-import Parser
-import Classes
-import Instances
-import Common
+
+-- the prelude
+
+class Applicative f => Plus f where
+  zero :: f a
+  (<+>) :: f a -> f a -> f a
+
+many0 :: Plus f => f a -> f [a]
+many0 p = many1 p <+> pure []
+
+many1 :: Plus f => f a -> f [a]
+many1 p = pure (:) <*> p <*> many0 p
+
+class Switch f where
+  switch :: f a -> f ()
+
+data Either' a b
+    = Left' a
+    | Right' b
+  deriving (Show, Eq)
+
+  
+instance Plus Maybe where
+  zero = Nothing
+  Nothing <+> y = y
+  x       <+> _ = x
+
+instance (Functor m, Monad m) => Plus (MaybeT m) where
+  zero = MaybeT (return Nothing)
+  MaybeT l  <+>  MaybeT r  =  MaybeT x
+    where
+      x = l >>= \y -> case y of Nothing -> r;
+                                Just _  -> return y;
+
+instance (Monad m, Plus m) => Plus (StateT s m) where
+  zero = StateT (const zero)
+  StateT f <+> StateT g = StateT (\s -> f s <+> g s)
+
+
+instance Switch Maybe where
+  switch (Just _) = Nothing
+  switch Nothing  = Just ()
+
+instance (Functor m, Switch m) => Switch (StateT s m) where
+  switch (StateT f) = StateT (\s -> fmap (const ((), s)) . switch $ f s)
+
+instance Functor m => Switch (MaybeT m) where
+  switch (MaybeT m) = MaybeT (fmap switch m)
+
+
+instance Functor (Either' e) where
+  fmap = liftM
+
+instance Applicative (Either' e) where
+  pure = return
+  (<*>) = ap
+
+instance Monad (Either' e) where
+  return            =  Right'
+  Left' e   >>=  _  =  Left' e
[…]

### 2013-05-10 src/parse3.hs
make 3rd parser self-sufficient and implement the new, simple grammar

@@ -1,66 +1,153 @@
 {-# LANGUAGE   NoMonomorphismRestriction
-             , FlexibleContexts           #-}
+             , FlexibleContexts       
+             , FlexibleInstances    
+             , MultiParamTypeClasses  #-}
 
-import Control.Monad.State       (MonadState (..), StateT(..), State)
-import Control.Monad.Error       (MonadError (..), ErrorT(..))
-import Control.Monad.Trans.Maybe (MaybeT(..))
+
+import Control.Monad             (liftM, ap)
+import Control.Monad.State       (MonadState (..), StateT(..))
+import Control.Monad.Error       (MonadError (..))
 import Control.Monad.Trans.Class (MonadTrans (..))
+import Control.Monad.Trans.Maybe (MaybeT(..))
 import Control.Applicative       (Applicative(..))
 import Data.List                 (nub)
-import Data.Functor.Identity     (Identity(..))
-import Parser
-import Classes
-import Instances
-import Common
 
-
--- other effects to try:
+-- should also try:
 --   partial results
---   whitespace/comment reporting
 
--- list of already seen forms,
--- list of text, line, column
---   for whitespace/comment
-type Log = [(String, Int, Int)]
+-- the prelude
+
+class Applicative f => Plus f where
+  zero :: f a
+  (<+>) :: f a -> f a -> f a
+
+many0 :: Plus f => f a -> f [a]
+many0 p = many1 p <+> pure []
+
+many1 :: Plus f => f a -> f [a]
+many1 p = pure (:) <*> p <*> many0 p
+
+class Switch f where
+  switch :: f a -> f ()
+
+data Either' a b
+    = Left' a
+    | Right' b
+  deriving (Show, Eq)
+
+  
+instance Plus Maybe where
+  zero = Nothing
+  Nothing <+> y = y
+  x       <+> _ = x
+
+instance (Functor m, Monad m) => Plus (MaybeT m) where
+  zero = MaybeT (return Nothing)
+  MaybeT l  <+>  MaybeT r  =  MaybeT x
+    where
+      x = l >>= \y -> case y of Nothing -> r;
+                                Just _  -> return y;
+
+instance (Monad m, Plus m) => Plus (StateT s m) where
+  zero = StateT (const zero)
+  StateT f <+> StateT g = StateT (\s -> f s <+> g s)
+
+
+instance Switch Maybe where
+  switch (Just _) = Nothing
+  switch Nothing  = Just ()
+
[…]

### 2013-05-11 src/parse2.hs
rework error parser to use own MonadError type class instead of own Either' data type.  this provides more flexibility in terms of monad stacking order while still avoiding the unwanted instances

@@ -1,12 +1,13 @@
 {-# LANGUAGE   NoMonomorphismRestriction
              , FlexibleContexts       
              , FlexibleInstances    
-             , MultiParamTypeClasses  #-}
+             , FunctionalDependencies
+             , UndecidableInstances      #-}
 
 
 import Control.Monad             (liftM, ap)
 import Control.Monad.State       (MonadState (..), StateT(..))
-import Control.Monad.Error       (MonadError (..))
+import Control.Monad.Trans.Class (lift)
 import Control.Monad.Trans.Maybe (MaybeT(..))
 import Control.Applicative       (Applicative(..))
 import Data.List                 (nub)
@@ -23,31 +24,25 @@ many0 p = many1 p <+> pure []
 many1 :: Plus f => f a -> f [a]
 many1 p = pure (:) <*> p <*> many0 p
 
-class Switch f where
-  switch :: f a -> f ()
-
-data Either' a b
-    = Left' a
-    | Right' b
-  deriving (Show, Eq)
-
-  
 instance Plus Maybe where
-  zero = Nothing
-  Nothing <+> y = y
-  x       <+> _ = x
+  zero             =  Nothing
+  Nothing  <+>  y  =  y
+  x        <+>  _  =  x
 
 instance (Functor m, Monad m) => Plus (MaybeT m) where
-  zero = MaybeT (return Nothing)
+  zero                     =  MaybeT (return Nothing)
   MaybeT l  <+>  MaybeT r  =  MaybeT x
     where
       x = l >>= \y -> case y of Nothing -> r;
                                 Just _  -> return y;
 
 instance (Monad m, Plus m) => Plus (StateT s m) where
-  zero = StateT (const zero)
-  StateT f <+> StateT g = StateT (\s -> f s <+> g s)
+  zero                   =  StateT (const zero)
+  StateT f <+> StateT g  =  StateT (\s -> f s <+> g s)
 
+
+class Switch f where
+  switch :: f a -> f ()
 
 instance Switch Maybe where
   switch (Just _) = Nothing
@@ -60,22 +55,22 @@ instance Functor m => Switch (MaybeT m) where
   switch (MaybeT m) = MaybeT (fmap switch m)
 
 
-instance Functor (Either' e) where
-  fmap = liftM
+class Monad m => MonadError e m | m -> e where
+  throwError :: e -> m a
+  catchError :: m a -> (e -> m a) -> m a
 
[…]

### 2013-05-11 src/parse3.hs
rework whitespace-logging parser to also use own MonadError instead of Either'

@@ -1,19 +1,20 @@
 {-# LANGUAGE   NoMonomorphismRestriction
              , FlexibleContexts       
              , FlexibleInstances    
-             , MultiParamTypeClasses  #-}
+             , FunctionalDependencies
+             , UndecidableInstances      #-}
 
 
 import Control.Monad             (liftM, ap)
 import Control.Monad.State       (MonadState (..), StateT(..))
-import Control.Monad.Error       (MonadError (..))
-import Control.Monad.Trans.Class (MonadTrans (..))
+import Control.Monad.Trans.Class (lift)
 import Control.Monad.Trans.Maybe (MaybeT(..))
 import Control.Applicative       (Applicative(..))
 import Data.List                 (nub)
 
 -- should also try:
 --   partial results
+
 
 -- the prelude
 
@@ -27,31 +28,25 @@ many0 p = many1 p <+> pure []
 many1 :: Plus f => f a -> f [a]
 many1 p = pure (:) <*> p <*> many0 p
 
-class Switch f where
-  switch :: f a -> f ()
-
-data Either' a b
-    = Left' a
-    | Right' b
-  deriving (Show, Eq)
-
-  
 instance Plus Maybe where
-  zero = Nothing
-  Nothing <+> y = y
-  x       <+> _ = x
+  zero             =  Nothing
+  Nothing  <+>  y  =  y
+  x        <+>  _  =  x
 
 instance (Functor m, Monad m) => Plus (MaybeT m) where
-  zero = MaybeT (return Nothing)
+  zero                     =  MaybeT (return Nothing)
   MaybeT l  <+>  MaybeT r  =  MaybeT x
     where
       x = l >>= \y -> case y of Nothing -> r;
                                 Just _  -> return y;
 
 instance (Monad m, Plus m) => Plus (StateT s m) where
-  zero = StateT (const zero)
-  StateT f <+> StateT g = StateT (\s -> f s <+> g s)
+  zero                   =  StateT (const zero)
+  StateT f <+> StateT g  =  StateT (\s -> f s <+> g s)
 
+
+class Switch f where
+  switch :: f a -> f ()
 
 instance Switch Maybe where
   switch (Just _) = Nothing
@@ -64,22 +59,22 @@ instance Functor m => Switch (MaybeT m) where
   switch (MaybeT m) = MaybeT (fmap switch m)
 
 
-instance Functor (Either' e) where
-  fmap = liftM
[…]

### 2013-05-13 gadt.tex
Add GADTs to repository.
Signed-off-by: Edward Z. Yang <ezyang@mit.edu>

\documentclass{tmr}

\usepackage{graphicx}

\renewcommand{\thetable}{\arabic{table}}

\usepackage{prettyref}
\newrefformat{fig}{Figure~\ref{#1}}
\newrefformat{tab}{Table~\ref{#1}}

\newcommand{\italic}[1]{\textit{#1}}

\title{Generalized Algebraic Data Types in Haskell}
\author{Anton Dergunov\email{anton.dergunov@gmail.com}}

\begin{document}

\begin{introduction}

Generalized algebraic data types (GADTs) is a feature of functional programming languages that generalizes ordinary algebraic data types by permitting value constructors to return specific types. This article is a tutorial about GADTs in Haskell programming language as they are implemented by the Glasgow Haskell Compiler (GHC) as a language extension. GADTs are widely used in practice: for domain-specific embedded languages, for generic programming, for ensuring program correctness and in other cases. The article describes these use cases with small GADT programs and also describes usage of GADTs in Yampa, a Haskell library.

\end{introduction}

\section{Introduction}

The usage of type systems is recognized as the most popular and best established lightweight formal method for ensuring that software behaves correctly~\cite{pierce_book}. It is used for detecting program errors, for documentation, to enforce abstraction and in other cases. The study of type systems is an active research area and Haskell is considered to be a kind of laboratory in which type-system extensions are designed, implemented and applied~\cite{haskell_history}.

This article is a tutorial about one such extension -- generalized algebraic data types (GADTs). The single idea of GADTs is to allow specific return types of value constructors. In this way they generalize ordinary algebraic data types. The theoretical foundation for GADTs is the notion of dependent types which is extensively described in literature on computer science and logic~\cite{pierce_book}.

GADTs are very useful in practice. Several examples of GADTs usage are described in this article. But these examples are not new. They were taken from many resources~\cite{fun_with_phantom_types,replib,singletons,okasaki_red_black_trees,weirich_dependently_typed,finally_tagless,yampa_gadts} and many examples are already a part of Haskell folklore.

The contribution of this article is presenting a concise introductory-level tutorial on the concept and popular use cases of GADTs. The following use cases were described in this article:

\begin{itemize}
\item domain-specific embedded languages;
\item generic programming;
\item ensuring program correctness.
\end{itemize}

We describe why type signatures are required for functions involving GADTs; the usage of GADTs in Yampa, a domain-specific language for functional reactive programming; an alternative implementation of one of the examples without GADTs usage. Additionally, several other Haskell extensions are covered in passing where necessary, such as type families and data kinds.

All examples in this article were tested with Glasgow Haskell Compiler, version 7.4.1.

\section{GADTs in a Nutshell}
\label{sec:nutshell}

\italic{Algebraic data types} (ADTs) are declared using the data keyword:

\begin{Verbatim}
data Test a = TI Int | TS String a
\end{Verbatim}

In the example above \verb}Test} is called a \italic{type constructor}. \verb|TI| and \verb|TS| are \italic{value constructors}. If a type has more than one value constructor, they are called \italic{alternatives}: one can use any of these alternatives to create a value of that type. Each alternative can specify zero or more components. For example, \verb|TS| specifies two components: one of them has type \verb|String| and another one -- type \verb|a|.

We can construct values of this type this way:

\begin{Verbatim}
ghci> let a = TI 10
ghci> :t a
a :: Test a
ghci> let b = TS "test" 'c'
ghci> :t b
b :: Test Char
\end{Verbatim}

Value constructors of the type \verb|Test| have the following types:

\begin{Verbatim}
ghci> :t TI
TI :: Int -> Test a
ghci> :t TS
TS :: String -> a -> Test a
\end{Verbatim}

Using GADT syntax we can define the \verb|Test| data type as:

\begin{Verbatim}
data Test a where
    TI :: Int -> Test a
    TS :: String -> a -> Test a
\end{Verbatim}

The GADTs feature is a Haskell language extension. Just like other extensions it can be enabled in GHC by:

\begin{longenum}

\item Using command line option \verb|-XGADTs|.

\item Using \verb|LANGUAGE| pragma in source files. This is the recommended way, because it enables this language extension per-file. This pragma must precede the module keyword in the source file and contain the following:
\begin{Verbatim}
{-# LANGUAGE GADTs #-}
\end{Verbatim}

\end{longenum}

The power of GADTs is not about syntax. In fact, \italic{the single idea of GADTs is to allow arbitrary return types of value constructors}. In this way they generalize ordinary algebraic data types. Of course, this return type must still be an instance of the more general data type that is defined. We can turn the \verb|Test| data type into a full-power GADT for example this way:

\begin{Verbatim}
data Test a where
    TI :: Int -> Test Int
    TS :: String -> a -> Test a
\end{Verbatim}

We have modified the \verb|TI| value constructor to return value of type \verb|Test Int| and we can test this:

\begin{Verbatim}
ghci> :t TI 10
TI 10 :: Test Int
\end{Verbatim}

The key feature of GADTs is that pattern matching causes type refinement. In the right-hand side of the following equation the type of \verb|a| is refined to \verb|Int|:

\begin{Verbatim}
f :: Test a -> a
f (TI i) = i + 10
\end{Verbatim}

Examples in the following sections show the real practical value of GADTs.

\section{Expression Evaluator}
\label{sec:expression_evaluator}

This section introduces GADTs with a canonical example of expression evaluator. At first, we attempt to implement it using ordinary algebraic data types. But as we will see, GADTs allow a more elegant implementation of the evaluator.

We start with the following type of expressions involving addition of integers:

\begin{Verbatim}
data IntExpr = IntVal Int
             | AddInt IntExpr IntExpr
\end{Verbatim}

\verb|IntVal| value constructor is used to wrap integer literal and \verb|AddInt| is used to represent an addition of two integer expressions. An example of such expression is:

\begin{Verbatim}
ghci> :t AddInt (IntVal 5) (IntVal 7)
AddInt (IntVal 5) (IntVal 7) :: IntExpr
\end{Verbatim}

Evaluation function for such expressions is easy to write:

\begin{Verbatim}
evaluate :: IntExpr -> Int
evaluate e = case e of
    IntVal i     -> i
    AddInt e1 e2 -> evaluate e1 + evaluate e2
\end{Verbatim}

Now we extend the type of expressions to support boolean values and add some operations on them:

\begin{Verbatim}
data ExtExpr = IntVal Int
             | BoolVal Bool
             | AddInt ExtExpr ExtExpr
             | IsZero ExtExpr
\end{Verbatim}

\verb|BoolVal| value constructor wraps Boolean literal. \verb|IsZero| is a unary function that takes an integer and returns a Boolean value. We immediately notice a problem with this type: it is possible to write incorrect expressions that type checker will accept. For example:

\begin{Verbatim}
ghci> :t IsZero (BoolVal True)
IsZero (BoolVal True) :: ExtExpr
ghci> :t AddInt (IntVal 5) (BoolVal True)
AddInt (IntVal 5) (BoolVal True) :: ExtExpr
\end{Verbatim}

Evaluation function for such expressions is also tricky. The result of evaluation can be either an integer or a Boolean value. The type \verb|ExtExpr| is not parametrized by return value type, so we have to use type \verb|Either Int Bool|. Also, evaluation will fail if the input expression is incorrect, so we have to use type \verb|Maybe|. Finally, type signature is the following:

\begin{Verbatim}
evaluate :: ExtExpr -> Maybe (Either Int Bool)
\end{Verbatim}

And implementation of this function is complicated. For example, processing \verb|AddInt| requires usage of a nested \verb|case|:

\begin{Verbatim}
evaluate e = case e of
    AddInt e1 e2 -> case (evaluate e1, evaluate e2) of
        (Just (Left i1), Just (Left i2)) -> Just $ Left $ i1 + i2
        _ -> error "AddInt takes two integers"
\end{Verbatim}

The conclusion is that we need to represent expressions using values of types parametrized by expression return value type. \italic{Phantom type} is a parametrized type whose parameters do not appear on the right-hand side of its definition. One can use them this way:

\begin{Verbatim}
data PhantomExpr t = IntVal Int
                   | BoolVal Bool
                   | AddInt (PhantomExpr Int) (PhantomExpr Int)
                   | IsZero (PhantomExpr Int)
\end{Verbatim}

Type \verb|t| in this type corresponds to the expression return value type. For example, integer expression has type \verb|PhantomExpr Int|. But this type definition alone is still not helpful, because it is still possible to write incorrect expressions that type checker will accept:

\begin{Verbatim}
ghci> :t IsZero (BoolVal True)
IsZero (BoolVal True) :: PhantomExpr t
\end{Verbatim}

The trick is to wrap value constructors with corresponding functions:

\begin{Verbatim}
intVal :: Int -> PhantomExpr Int
intVal = IntVal
boolVal :: Bool -> PhantomExpr Bool
boolVal = BoolVal
isZero :: PhantomExpr Int -> PhantomExpr Bool
isZero = IsZero
\end{Verbatim}

And now bad expressions are rejected by type checker:

\begin{Verbatim}
ghci> :t isZero (boolVal True)
Couldn't match expected type `Int' with actual type `Bool'...
ghci> :t isZero (intVal 5)
isZero (intVal 5) :: PhantomExpr Bool
\end{Verbatim}

Ideally we want the following type signature for \verb|evaluate| method:

\begin{Verbatim}
evaluate :: PhantomExpr t -> t
\end{Verbatim}

But we can't define such function. For example, the following line produces error ``\verb|Couldn't match type `t' with `Int'|'':

\begin{Verbatim}
evaluate (IntVal i) = i
\end{Verbatim}

The reason of this error is that return type of value constructor \verb|IntVal| is \verb|Phantom t| and \verb|t| can be refined to any type. For example:

\begin{Verbatim}
ghci> :t IntVal 5 :: PhantomExpr Bool
IntVal 5 :: PhantomExpr Bool :: PhantomExpr Bool
\end{Verbatim}

What is really needed here is to specify type signature of value constructors exactly. In this case pattern matching in \verb|evaluate| would cause type refinement for \verb|IntVal|. And this is exactly what GADTs do.

As described in the previous section, GADTs use a different syntax than ordinary algebraic data types. In fact, value constructors specified by the data type \verb|PhantomExpr| can be written as the following functions:

\begin{Verbatim}
IntVal  :: Int -> PhantomExpr t
BoolVal :: Bool -> PhantomExpr t
AddInt  :: PhantomExpr Int -> PhantomExpr Int -> PhantomExpr t
IsZero  :: PhantomExpr Int -> PhantomExpr t
\end{Verbatim}

Using GADT syntax the data type \verb|PhantomExpr| type can be declared this way:

\begin{Verbatim}
data PhantomExpr t where
    IntVal  :: Int -> PhantomExpr t
    BoolVal :: Bool -> PhantomExpr t
    AddInt  :: PhantomExpr Int -> PhantomExpr Int -> PhantomExpr t
    IsZero  :: PhantomExpr Int -> PhantomExpr t
\end{Verbatim}

All value constructors have \verb|PhantomExpr t| as their return type. As noted in the previous section, the distinctive feature of GADTs is the ability to return specific types in value constructors, for example \verb|PhantomExpr Int|. GADT for the expression language looks this way:

\begin{Verbatim}
data Expr t where
    IntVal  :: Int -> Expr Int
    BoolVal :: Bool -> Expr Bool
    AddInt  :: Expr Int -> Expr Int -> Expr Int
    IsZero  :: Expr Int -> Expr Bool
    If      :: Expr Bool -> Expr t -> Expr t -> Expr t
\end{Verbatim}

Note that value constructors of this data type have specific return types. Now bad expressions are rejected by the type checker:

\begin{Verbatim}
ghci> :t IsZero (BoolVal True)
Couldn't match expected type `Int' with actual type `Bool'...
ghci> :t IsZero (IntVal 5)
IsZero (IntVal 5) :: Expr Bool
\end{Verbatim}

Note that the type of value \verb|IsZero (IntVal 5)| is specific: \verb|Expr Bool|.

GADTs allow to write well-defined \verb|evaluate| function:

\begin{Verbatim}
evaluate :: Expr t -> t
evaluate (IntVal i)     = i
evaluate (BoolVal b)    = b
evaluate (AddInt e1 e2) = evaluate e1 + evaluate e2
evaluate (IsZero e)     = evaluate e == 0
evaluate (If e1 e2 e3)  = if evaluate e1 then
    evaluate e2 else evaluate e3
\end{Verbatim}

Pattern matching causes type refinement, so for example in the right-hand side of the following expression \verb|i| has type \verb|Int|:

\begin{Verbatim}
evaluate :: Expr t -> t
evaluate (IntVal i) = ...
\end{Verbatim}

The type of \verb|AddInt e1 e2| expression is \verb|Expr Int| and the types of \verb|e1| and \verb|e2| must also be \verb|Expr Int|, so we can evaluate recursively the individual expressions and then return the sum (value of type \verb|Int|).

At the end of this article we describe one more implementation of expression evaluator.

\section{Generic Programming with GADTs}
\label{sec:generic_programming}

In datatype-generic programming functions take a type as argument and their behavior depends on the structure of this type. There are several approaches to such kind of generic programming in Haskell. Paper by Hinze et al.~\cite{generic_programming} provides an overview of these approaches. In this section we present an approach which uses GADTs. Ideas for this section were taken from another paper by Hinze at al.~\cite{fun_with_phantom_types}.

Suppose we would like to write a function to encode data in binary form. This function must be able to work with values of several types. Functions like this one can be implemented using type classes. But GADTs offer an interesting alternative.

First we need to declare a \italic{representation type}~\cite{replib}, a type whose values represent types:

\begin{Verbatim}
data Type t where
    TInt  :: Type Int
    TChar :: Type Char
    TList :: Type t -> Type [t]
\end{Verbatim}

This is GADT with value constructors that create a representation of the corresponding type. For example:

\begin{Verbatim}
ghci> let a = TInt
ghci> :t a
a :: Type Int
ghci> let b = TList TInt
ghci> :t b
b :: Type [Int]
\end{Verbatim}

\verb|String| type is defined in Haskell as a list of \verb|Char| elements, so we can define a value constructor for string type representation this way:

\begin{Verbatim}
tString :: Type String
tString = TList TChar
\end{Verbatim}

The output of the encoding function is a list of bits where bits are represented using:

\begin{Verbatim}
data Bit = F | T deriving(Show)
\end{Verbatim}

The encoding function takes a representation of the type, the value of this type and returns a list of bits.

\begin{Verbatim}
encode :: Type t -> t -> [Bit]
encode TInt i = encodeInt i
encode TChar c = encodeChar c
encode (TList _) [] = F : []
encode (TList t) (x : xs) = T :
    (encode t x) ++ encode (TList t) xs
\end{Verbatim}

We can test this function:

\begin{Verbatim}
ghci> encode TInt 333
[T,F,T,...,F,F,F]
ghci> encode (TList TInt) [1,2,3]
[T,T,F,...,F,F,F]
ghci> encode tString "test"
[T,F,F,...,F,F,F]
\end{Verbatim}

If we pair the representation type and the value together, we get a \italic{universal data type}, the type \verb|Dynamic| (this code requires using \verb|ExistentialQuantification| extension):

\begin{Verbatim}
data Dynamic = forall t. Dyn (Type t) t
\end{Verbatim}

Above we have defined an \italic{existential data type} which can also be represented as a GADT:

\begin{Verbatim}
data Dynamic where
    Dyn :: Type t -> t -> Dynamic
\end{Verbatim}

Now we can declare a variant of \verb|encode| function which gets a \verb|Dynamic| type value as input:

\begin{Verbatim}
encode' :: Dynamic -> [Bit]
encode' (Dyn t v) = encode t v
\end{Verbatim}

The following session illustrates the usage of this type:

\begin{Verbatim}
ghci> let c = Dyn (TList TInt) [5,4,3]
ghci> :t c
c :: Dynamic
ghci> encode' c
[T,T,F,...,F,F,F]
\end{Verbatim}

We can now define heterogeneous lists using the \verb|Dynamic| type:

\begin{Verbatim}
ghci> let d = [Dyn TInt 10, Dyn tString "test"]
ghci> :t d
d :: [Dynamic]
\end{Verbatim}

However, we can't make this list a \verb|Dynamic| value itself. To fix this problem, we need to extend the representation type: add a value constructor for the \verb|Dynamic| data type.

\begin{Verbatim}
data Type t where
    ...
    TDyn :: Type Dynamic
\end{Verbatim}

We also need to update \verb|encode| function to handle the \verb|Dynamic| data type:

\begin{Verbatim}
encode :: Type t -> t -> [Bit]
...
encode TDyn (Dyn t v) = encode t v
\end{Verbatim}

Now we can represent a list of \verb|Dynamic| values as a \verb|Dynamic| value itself and encode it:

\begin{Verbatim}
ghci> let d = [Dyn TInt 10, Dyn tString "test"]
ghci> :t d
d :: [Dynamic]
ghci> let e = Dyn (TList TDyn) d
ghci> :t e
e :: Dynamic
ghic> encode' e
[T,F,T,...,F,F,F]
\end{Verbatim}

The \verb|Dynamic| data type is useful for communication with the environment when the actual type of the data is not known in advance. In this case it is required to use a type cast to get the useful data. A simple way to implement a type cast from \verb|Dynamic| data type to an integer is the following:

\begin{Verbatim}
castInt :: Dynamic -> Maybe Int
castInt (Dyn TInt i) = Just i
castInt (Dyn _ _) = Nothing
\end{Verbatim}

There is a more generic solution~\cite{fun_with_phantom_types} of this problem that works for all types, not just integer, but it is out of scope for this article.

While the presented approach is an important use case of GADTs, the disadvantage of this approach is that we have to extend the representation type whenever we define a new data type.

\section{Proving Correctness of List Operations}
\label{sec:list_operations}

An important role of type systems is to ensure that data is manipulated in appropriate ways (for example, to make sure that we pass a list to \verb|head| function). But types can be used to express more sophisticated properties. For example, we can define a type of lists of a particular length and then define \verb|headSafe| function that only accepts non-empty lists. The idea described in this section is to use types to express correctness properties and then use type checker of the programming language to ensure that we can express only those programs that have the desired properties. This idea was developed in $\Omega$mega system~\cite{curry_howard,omega}. In this section we GADTs in Haskell programming language to prove correctness of list operations and the next section describes proving correctness of insertion operation in red-black trees.

Lists can be represented using the following algebraic data type:

\begin{Verbatim}
data List t = Nil | Cons t (List t)
\end{Verbatim}

or using GADT syntax as:

\begin{Verbatim}
data List t where
    Nil :: List t
    Cons :: t -> List t  -> List t 
\end{Verbatim}

Now \verb|head| function can be implemented this way:

\begin{Verbatim}
listHead :: List t -> t
listHead (Cons a _) = a
listHead Nil = error "list is empty"
\end{Verbatim}

The disadvantage of this function is that it can fail: it fails when a list is empty and succeeds otherwise. To address this problem we define a type of non-empty lists. First we define two empty data types (this requires \verb|EmptyDataDecls| extension):

\begin{Verbatim}
data Empty
data NonEmpty
\end{Verbatim}

Now we define a safe list GADT:

\begin{Verbatim}
data SafeList t f where
    Nil :: SafeList t Empty
    Cons :: t -> SafeList t f -> SafeList t NonEmpty
\end{Verbatim}

Parameter \verb|f| takes type \verb|Empty| when the list is empty and \verb|NonEmpty| otherwise. The function \verb|headSafe| is a safe version of \verb|listHead| function that only accepts non-empty lists as parameter.

\begin{Verbatim}
headSafe :: SafeList t NonEmpty -> t
headSafe (Cons t _) = t
\end{Verbatim}

For example:

\begin{Verbatim}
ghci> headSafe Nil
Couldn't match expected type `NonEmpty' with actual type `Empty'
ghci> headSafe $ Cons 1 $ Cons 2 $ Cons 3 Nil
1
\end{Verbatim}

However, the implementation of a function to create a list containing an element repeated a given number of times using \verb|SafeList| data type is problematic: it is not possible to determine return value type of this function.

\begin{Verbatim}
repeatElem :: a -> Int -> SafeList a ???
repeatElem a 0 = Nil
repeatElem a n = Cons a (repeatElem a (n-1))
\end{Verbatim}

The problem is that empty and non-empty lists have completely different types. To fix this problem we can slightly relax \verb|Cons| value constructor:

\begin{Verbatim}
data SafeList t f where
    Nil :: SafeList t Empty
    Cons :: t -> SafeList t f -> SafeList t f'
\end{Verbatim}

Now \verb|SafeList t Empty| is a type of possibly empty lists, for example:

\begin{Verbatim}
ghci> :t Nil
Nil :: SafeList t Empty
ghci> :t Cons 'a' Nil
Cons 'a' Nil :: SafeList Char f'
ghci> :t Cons 'a' Nil :: SafeList Char Empty
Cons 'a' Nil :: SafeList Char Empty :: SafeList Char Empty
ghci> :t Cons 'a' Nil :: SafeList Char NonEmpty
Cons 'a' Nil :: SafeList Char NonEmpty :: SafeList Char NonEmpty
\end{Verbatim}

And we can define \verb|repeatElem| as a function returning possibly empty lists:

\begin{Verbatim}
repeatElem :: a -> Int -> SafeList a Empty
repeatElem a 0 = Nil
repeatElem a n = Cons a (repeatElem a (n-1))
\end{Verbatim}

Actually, with the current data type definition a term \verb|Cons 'a' Nil| can even be given type \verb|SafeList Char Int|. To fix this problem it is required to give the same kind for \verb|Empty| and \verb|NonEmpty| types. This is discussed later for \verb|Nat| data type.

But \verb|SafeList| data type does not have enough static information to prove list length invariants for list functions. For example, for the concatenation function we need to show that length of the concatenated list is a sum of source lists lengths. So it is not enough to just know if a list is empty or not. We need to encode the length of a list in its type.

The classical way to encode numbers at the type level is Peano numbers:

\begin{Verbatim}
data Zero
data Succ n
\end{Verbatim}

Zero is encoded as \verb|Zero|, one -- as \verb|Succ Zero|, two -- as \verb|Succ (Succ Zero)| and so on. Now list data type is defined as:

\begin{Verbatim}
data List a n where
    Nil :: List a Zero
    Cons :: a -> List a n -> List a (Succ n)
\end{Verbatim}

Function \verb|headSafe| can be defined as:

\begin{Verbatim}
headSafe :: List t (Succ n) -> t
headSafe (Cons t _) = t 
\end{Verbatim}

We can also show that the safe map function does not change the length of a list:

\begin{Verbatim}
mapSafe :: (a -> b) -> List a n -> List b n
mapSafe _ Nil = Nil
mapSafe f (Cons x xs) = Cons (f x) (mapSafe f xs) 
\end{Verbatim}

To implement the concatenation function we need a type-level function for addition of Peano numbers. A natural way to implement such function is to use \italic{type families} (which in context of this tutorial can be understood as type-level functions). First we need to declare type family \verb|Plus| (this requires \verb|TypeFamilies| extension):

\begin{Verbatim}
type family Plus a b
\end{Verbatim}

Then we need to declare type instances that implement addition of Peano numbers by induction:

\begin{Verbatim}
type instance Plus Zero n = n
type instance Plus (Succ m) n = Succ (Plus m n)
\end{Verbatim}

The type family \verb|Plus| that we have just defined can be used in the concatenation function signature:

\begin{Verbatim}
concatenate :: List a m -> List a n -> List a (Plus m n)
concatenate Nil ys = ys
concatenate (Cons x xs) ys = Cons x (concatenate xs ys)
\end{Verbatim}

% TODO describe other ways to implement Plus?

In fact, at the moment \verb|Succ| has type parameter of kind \verb|*|, so it possible to write nonsensical terms like \verb|Succ Int|, and they will be accepted by the type checker. This problem can be addressed using a new \italic{kind}. Just as types classify values, kinds classify types. We can declare the following data type:

\begin{Verbatim}
data Nat = Zero | Succ Nat
\end{Verbatim}

\begin{figure}
\centering
\includegraphics{promotion}
\vspace{5pt}
\caption{Promotion of a data type to a kind.}
\label{fig:promotion}
\end{figure}

Here \verb|Nat| is a type; \verb|Zero| and \verb|Succ| are value constructors. But due to promotion~\cite{promotion} \verb|Nat| also becomes a kind; \verb|Zero| and \verb|Succ| also become types (see \prettyref{fig:promotion}). Where necessary, a quote must be used to resolve ambiguity. For example, \verb|'Succ| refers to a type, not a value constructor. So, type-level representation of Peano number 2 can be written as:

\begin{Verbatim}
type T = 'Succ ('Succ 'Zero)
\end{Verbatim}

Quotes can be omitted in this case, because there is no ambiguity:

\begin{Verbatim}
type T2 = Succ (Succ Zero)
\end{Verbatim}

As a result, type checker now rejects wrong terms like \verb|Succ Int|.

The definition of the list data type can also be improved now to clearly specify that the type of its second parameter has kind \verb|Nat| (this requires \verb|DataKinds| extension):

\begin{Verbatim}
data List a (n::Nat) where
    Nil :: List a 'Zero
    Cons :: a -> List a n -> List a ('Succ n)
\end{Verbatim}

After the changes that we have made to the definition of the \verb|List| data type, the implementation of the \verb|repeatElem| function becomes more involved, because now we can't yet write its return type:

\begin{Verbatim}
repeatElem :: a -> Int -> List a ???
repeatElem a 0 = Nil
repeatElem a n = Cons a (repeatElem a (n-1))
\end{Verbatim}

On the one hand, the count parameter must be passed as a value to populate the list at run-time. On the other hand, we need a type-level representation of the same number for \verb|List| type. Haskell enforces a phase separation between run-time values and compile-time types.

The solution to this puzzle is the use of \italic{singleton types}. Singleton types are types that contain only one value (except, of course, the $\bot$ value, which is a member of every type in Haskell).

The singleton for Peano numbers type can be expressed using the following GADT:

\begin{Verbatim}
data NatSing (n::Nat) where
    ZeroSing :: NatSing 'Zero
    SuccSing :: NatSing n -> NatSing ('Succ n)
\end{Verbatim}

\begin{figure}
\centering
\includegraphics{singleton}
\vspace{5pt}
\caption{Singleton type for Peano numbers.}
\label{fig:singleton}
\end{figure}

The constructors of the singleton \verb|NatSing| mirror those of the kind \verb|Nat|. As a result, every type of kind \verb|Nat| corresponds to exactly one value (except $\bot$ value) of the singleton data type where parameter \verb|n| has exactly this type (see \prettyref{fig:singleton}). For example:

\begin{Verbatim}
ghci> :t ZeroSing
ZeroSing :: NatSing 'Zero
ghci> :t SuccSing $ SuccSing ZeroSing
SuccSing $ SuccSing ZeroSing :: NatSing ('Succ ('Succ 'Zero))
\end{Verbatim}

Now function \verb|repeatElem| can be defined this way:

\begin{Verbatim}
repeatElem :: a -> NatSing n -> List a n
repeatElem _ ZeroSing     = Nil
repeatElem x (SuccSing n) = Cons x (repeatElem x n)
\end{Verbatim}

In a function returning an element by index in the list we need to make sure that the index does not exceed the list length. This requires a type-level function to compute whether one number is less than the other. We define the following type family and instances (\verb|TypeOperators| extension is required to be able to define \verb|:<| operation for types):

\begin{Verbatim}
type family (m::Nat) :< (n::Nat) :: Bool
type instance m         :< 'Zero     = 'False
type instance 'Zero     :< ('Succ n) = 'True
type instance ('Succ m) :< ('Succ n) = m :< n
\end{Verbatim}

This type-level function is implemented using induction. It returns promoted type \verb|'True| of kind \verb|Bool| when first number is less than the second one.

Now the function can be defined this way:

\begin{Verbatim}
nthElem :: (n :< m) ~ 'True => List a m -> NatSing n -> a
nthElem (Cons x _)  ZeroSing     = x
nthElem (Cons _ xs) (SuccSing n) = nthElem xs n
\end{Verbatim}

The tilde operation is an \italic{equality constraint}. It asserts that two types are the same in the context. Thus, is it only possible to use this function when the index does not exceed the list length.

%The moral is that to verify a function it is required to carefully design data type.

\section{Proving Correctness of Red-Black Tree Insert Operation}
\label{sec:red_black}

A red-black tree is a binary search tree where every node has either red or black color. We use the implementation described by Okasaki~\cite{okasaki_red_black_trees}. The source code for the verified red-black tree was originally written by Stephanie Weirich~\cite{weirich_dependently_typed} for a university course.

\begin{Verbatim}
data Color = R | B
data Node a = E | N Color (Node a) a (Node a) 
type Tree a = Node a
\end{Verbatim}

\verb|N| is a value constructor of a regular node and \verb|E| is a value constructor for a leaf node. As in all binary search trees, for a particular node \verb|N c l x r| values less than \verb|x| are stored in left sub-tree (in \verb|l|) and values greater than \verb|x| are stored in right sub-tree (in \verb|r|). Membership function implements a recursive search:

\begin{Verbatim}
member :: Ord a => a -> Tree a -> Bool
member _ E  = False
member x (N _ l a r)
    | x < a  = member x l
    | x > a  = member x r
    | otherwise = True
\end{Verbatim}

Additionally red-black tree satisfies the following invariants:

\begin{longenum}
\item The root is black.
\item Every leaf is black. % (this is enforced by the tree data type).
\item Red nodes have black children.
\item For each node, all paths from that node to the leaf node contain the same number of black nodes. This number of black nodes is called the \italic{black height} of a node.
\end{longenum}

% excluding current node?

These invariants guarantee that tree is balanced. Indeed, the longest path from the root node (containing alternating red-black nodes) can only be twice as long as the shortest path (containing only red nodes). Thus basic operations (such as insertion) take $O(\log n)$ time in the worst case.

Insertion operation for red-black trees has the following structure:

\begin{Verbatim}
insert :: Ord a => Tree a -> a -> Tree a
insert t v = blacken (insertInternal t v) where
    insertInternal n@(N c l a r) x
        | x < a = leftBalance  (N c (insertInternal l x) a r)
        | x > a = rightBalance (N c l a (insertInternal r x))
        | otherwise = n
    insertInternal E x = N R E x E
    blacken (N _ l x r) = N B l x r
\end{Verbatim}

It has the same structure as insertion operation for regular binary search trees which is implemented by recursive descent (down to leaf nodes) until a suitable location for insertion is found. But additionally it must keep the invariants, so there are the following differences:

\begin{longitem}
\item The node is inserted with red color. This allows to maintain the 4th invariant, because the black height is not changed.
\item To maintain the 1st invariant we call \verb|blacken| at the end of insertion. Again, the 4th invariant remains valid.
\item To maintain the 3rd invariant we call \verb|leftBalance| and \verb|rightBalance|.
\end{longitem}

\begin{figure}
\centering
\includegraphics{balancing}
\vspace{5pt}
\caption{Possible cases of 3rd invariant violation after insertion in the left branch of the node.}
\label{fig:balancing}
\end{figure}

\prettyref{fig:balancing} shows 2 possible cases when the 3rd invariant is violated after insertion in the left branch of the node. To repair this invariant the tree must be restructured as shown on the figure. The following code uses pattern matching to implement the restructuring, otherwise the function returns the sub-tree as is:

\begin{Verbatim}
leftBalance :: Node a -> Node a
leftBalance (N B (N R (N R a x b) y c) z d) =
    N R (N B a x b) y (N B c z d)
leftBalance (N B (N R a x (N R b y c)) z d) =
    N R (N B a x b) y (N B c z d)
leftBalance n = n
\end{Verbatim}

The function \verb|rightBalance| is similar. Complete source code is in Appendix A.

\subsection{Proving that the 4th invariant is maintained by insert}

To show that the 4th invariant is maintained by the insertion operation we need to add black height as a parameter of the \verb|Node| data type.

First we define Peano numbers in same way as before:

\begin{Verbatim}
data Nat = Zero | Succ Nat
\end{Verbatim}

Then we turn Node into a GADT with a black height parameter:

\begin{Verbatim}
data Node a (bh::Nat) where
    E :: Node a 'Zero
    N :: Color -> Node a bh -> a -> Node a bh -> Node a ???
\end{Verbatim}

The leaf node has black height $0$. The definition of internal nodes requires that both children have the same black height. The black height of the node itself must be conditionally incremented based on its color. This is implemented using the following type family which computes the new height based on the color of the node and black height of its children. Both parameters are represented as types (of \verb|Color| and \verb|Nat| kinds correspondingly). This code requires \verb|TypeFamilies| and \verb|DataKinds| extensions.

\begin{Verbatim}
type family   IncBlackHeight (c::Color) (bh::Nat) :: Nat
type instance IncBlackHeight R bh = bh
type instance IncBlackHeight B bh = Succ bh
\end{Verbatim}

Now we see that color must be passed as a type (for \verb|IncBlackHeight| type family) and as a value (to the value constructor). Similarly as before, we need to use a singleton type as a bridge:

\begin{Verbatim}
data ColorSingleton (c::Color) where
    SR :: ColorSingleton R
    SB :: ColorSingleton B
\end{Verbatim}

The value of this singleton type is passed as a parameter to the node value constructor and the color type is used for type family:

\begin{Verbatim}
data Node a (bh::Nat) where
    E :: Node a 'Zero
    N :: ColorSingleton c -> Node a bh -> a
        -> Node a bh -> Node a (IncBlackHeight c bh)
\end{Verbatim}

After we have added a new parameter for the \verb|Node| data type, it is an error to write:

\begin{Verbatim}
type Tree a = Node a bh
\end{Verbatim}

Because normally when creating a new type in Haskell, every type variable that appears on the right-hand side of the definition must also appear on its left-hand side. One solution to this problem is usage of \italic{existential types} (this definition requires extension \verb|RankNTypes|):

\begin{Verbatim}
type Tree a = forall bh. Node a bh
\end{Verbatim}

It is also possible to do this with GADT:

\begin{Verbatim}
data Tree a where
    Root :: Node a bh -> Tree a
\end{Verbatim}

The implementation of insertion operation never violates the 4th invariant, so the remaining changes are adjustments of type annotations and so on. Complete source code is in Appendix B.

\subsection{Proving that the 3rd invariant is maintained by insert}

Proving the 3rd invariant is more involved. First we need to specify valid colors for a node on the type level. This can be done using type families as before or using type classes. We choose the latter and define a type class with 3 parameters corresponding to color of the parent and colors of the child nodes (this code requires \verb|MultiParamTypeClasses| extension):

\begin{Verbatim}
class ValidColors (parent::Color) (child1::Color) (child2::Color)
\end{Verbatim}

We do not need to define any functions in this type class, because our aim is just to declare instances with valid colors (this code requires \verb|FlexibleInstances| extension):

\begin{Verbatim}
instance ValidColors R B B
instance ValidColors B c1 c2
\end{Verbatim}

The allowed nodes are:

\begin{itemize}
\item red nodes with black child nodes;
\item black nodes with child nodes of any color.
\end{itemize}

We need to add color type as a parameter to the \verb|Node| data type and restrict it to have only correctly-colored nodes using the \verb|ValidColors| type class:

\begin{Verbatim}
data Node a (bh::Nat) (c::Color) where
    E :: Node a 'Zero B
    N :: ValidColors c c1 c1 => ColorSingleton c -> Node a bh c1
        -> a -> Node a bh c2 -> Node a (IncBlackHeight c bh) c
\end{Verbatim}

With this change we also statically ensure the 2nd invariant: leaf nodes have black color.

We also need to update the definition of the \verb|Tree| data type to specify that root node has black color (this way also ensuring the 1st invariant):

\begin{Verbatim}
data Tree a where
    Root :: Node a bh B -> Tree a
\end{Verbatim}

The implementation of the insertion operation can temporarily invalidate the 3rd invariant (see \prettyref{fig:balancing}), so during insertion we are not able to represent the tree using this data type. Thus it is required to declare a data type similar to \verb|Node|, but without the restriction on node colors:

\begin{Verbatim}
data IntNode a (n::Nat) where
    IntNode :: ColorSingleton c -> Node a n c1 -> a
        -> Node a n c2 -> IntNode a (IncBlackHeight c n)
\end{Verbatim}

As before we need to make changes in type annotations of the functions implementing insert operation. We also need to change the \verb|leftBalance| function type signature this way:

\begin{Verbatim}
leftBalance :: ColorSingleton c -> IntNode a n -> a
    -> Node a n c' -> IntNode a (IncBlackHeight c n)
\end{Verbatim}

Earlier we passed the whole node as a parameter. But we can't do it after the \verb|Node| data type was modified: the 3rd invariant could be violated due to insertion in the left branch of the node. So, we pass all parameters of the parameters of the node and left child is represented using \verb|IntNode| data type.

Previous cases should be rewritten using new types:

\begin{Verbatim}
leftBalance SB (IntNode SR (N SR a x b) y c) z d =
    IntNode SR (N SB a x b) y (N SB c z d)
leftBalance SB (IntNode SR a x (N SR b y c)) z d =
    IntNode SR (N SB a x b) y (N SB c z d)
\end{Verbatim}

However, now we can't write the same catch-all case as before:

\begin{Verbatim}
leftBalance c (IntNode c1 a x b) d n2 =
    IntNode c (N c1 a x b) d n2
\end{Verbatim}

This case does not type-check with the following error:

\begin{Verbatim}
Could not deduce (ValidColors c1 c2 c2) ...
\end{Verbatim}

The reason is that the type of the left node is \verb|IntNode|, so even though we have previously balanced the left sub-tree, technically this is not reflected in the type. We need to explicitly match against the correct cases and reconstruct the node. First, we match against the black nodes where children can have any color:

\begin{Verbatim}
leftBalance c (IntNode SB a x b) z d = IntNode c (N SB a x b) z d
\end{Verbatim}

Red nodes must have black children:

\begin{Verbatim}
leftBalance c (IntNode SR a@(N SB _ _ _) x b@(N SB _ _ _)) z d =
    IntNode c (N SR a x b) z d
leftBalance c (IntNode SR E x E) z d = IntNode c (N SR E x E) z d
\end{Verbatim}

Unfortunately, we haven't yet listed all cases. We know that the following cases can't happen, but we do not have enough information in the type to omit them. We can skip them, but this means producing ``Non-exhaustive patterns'' exception for these impossible cases.

\begin{Verbatim}
leftBalance _ (IntNode SR (N SR _ _ _) _ _) _ _ =
    error "can't happen"
leftBalance _ (IntNode SR _ _ (N SR _ _ _)) _ _ =
    error "can't happen"
\end{Verbatim}

Note that the case of one regular node and one leaf node is not valid, because these nodes must have different black heights.

Complete source code is in Appendix C.

The previous code illustrates a general problem with proofs. In fact, in Haskell $\bot$ (bottom) is a member of every type. As a result, we can write:

\begin{Verbatim}
concatenate :: List a m -> List a n -> List a (Plus m n)
concatenate = undefined
\end{Verbatim}

Of course, the implementation of the \verb|concatenate| function does not meet our expectations. But this code still type checks.

\section{Type Signatures for Functions Involving GADTs}
\label{sec:type_signatures}

Hindley-Milner (HM) is a classical type inference method~\cite{cardelli_typechecking}. One of the most important properties of HM is ability to always deduce the \italic{most general type} (\italic{principle type}) of every term.

However, GADTs pose a difficult problem for type inference, because programs with GADTs lose principle type property~\cite{type_inference_gadts}. For example, consider the following GADT program:

\begin{Verbatim}
data Test t where
    TInt :: Int -> Test Int
    TString :: String -> Test String

f (TString s) = s
\end{Verbatim}

There are two possible principal types of the function \verb|f|, but neither of them is an instance of the other:

\begin{Verbatim}
f :: Test t -> String
f :: Test t -> t
\end{Verbatim}

Also without type signature the following function fails to typecheck:

\begin{Verbatim}
f' (TString s) = s
f' (TInt i) = i
\end{Verbatim}

Adding type signature fixes the problem:

\begin{Verbatim}
f' :: Test t -> t
\end{Verbatim}

The paper by Schrijvers et al.~\cite{type_inference_gadts} provides more information on type inference for programs with GADTs.

\section{Usage of GADTs in Yampa}
\label{sec:yampa}

Yampa~\cite{yampa} is a domain-specific language for functional reactive programming (FRP). FRP is a programming paradigm of expressing data flows using the building blocks of functional programming. Based on the information from the paper by Nilsson~\cite{yampa_gadts}, this section describes how GADTs were used to improve performance of Yampa programs.

Signal function is a central abstraction in Yampa. It represents a simple synchronous process that maps an input signal to an output signal (see \prettyref{fig:signal_function}). The type of the signal function is \verb|SF a b| and it can be constructed from an ordinary function using the following function:

\begin{Verbatim}
arr :: (a -> b) -> SF a b
\end{Verbatim}

\begin{figure}
\centering
\includegraphics{signal_function}
\vspace{5pt}
\caption{Signal function \texttt{SF a b}.}
\label{fig:signal_function}
\end{figure}

The following function provides a composition of signal functions (as shown in \prettyref{fig:signal_composition}):

\begin{Verbatim}
(>>>) :: SF a b -> SF b c -> SF a c
\end{Verbatim}

\begin{figure}
\centering
\includegraphics{signal_composition}
\caption{Composition of signal functions.}
\label{fig:signal_composition}
\vspace{5pt}
\end{figure}

There is a natural requirement to eliminate the overhead of composition with identity function:

\begin{Verbatim}
arr id >>> f = f
f >>> arr id = f
\end{Verbatim}

As an attempt to implement this in Yampa we can imagine introducing a special value constructor to represent identity signal functions:

\begin{Verbatim}
data SF a b = ...
            | SFId -- Represents arr id.
\end{Verbatim}

But the return type of this value constructor is still \verb|SF a b|. We can use the same trick as before with phantom types. We can define a function to construct the value and restrict the type to \verb|SF a a|:

\begin{Verbatim}
identity :: SF a a
identity = SFId
\end{Verbatim}

Now we can try to use the new value constructor in the definition of the function \verb|>|\verb|>|\verb|>| this way:

\begin{Verbatim}
(>>>) :: SF a b -> SF b c -> SF a c
...
SFId >>> sf = sf
sf >>> SFId = sf
\end{Verbatim}

But this code does not type check, because when we pattern match using \verb|SFId| value constructor, the type is still \verb|SF a b|, not \verb|SF a a|. We have already seen this problem before when we attempted to use phantom types for expression evaluation. The solution is to use GADT to represent the signal function:

\begin{Verbatim}
data SF a b where
    ...
    SFId :: SF a a
\end{Verbatim}

After this change the function \verb|>|\verb|>|\verb|>| as presented above must type check due to type refinement in pattern matching.

\begin{table}[float,captionpos=b,belowcaptionskip=4pt]
\centering
\begin{tabular}{cll}
\hline\noalign{\smallskip}
Benchmark & $T_S$[s] & $T_G$[s]\\
\noalign{\smallskip}
\hline
\noalign{\smallskip}
1 & 0.41 & 0.00 \\
2 & 0.74 & 0.22 \\
3 & 0.45 & 0.22 \\
4 & 1.29 & 0.07 \\
5 & 1.95 & 0.08 \\
6 & 1.48 & 0.69 \\
7 & 2.85 & 0.72 \\
\hline
\end{tabular}
\label{tab:yampa_performance}
\caption{Performance improvements enabled by GADTs in Yampa programs}
\end{table}

There are other performance improvements that are enabled by GADTs in Yampa~\cite{yampa_gadts}. The results of performance improvements are shown in the table. It was taken from the paper by Nilsson~\cite{yampa_gadts}. The table shows execution time of several benchmarks using initial simply-optimized implementation ($T_S$) and the implementation with GADT-based optimizations ($T_G$). In addition to performance improvements, GADTs allowed to write a more concise and cleaner API without the need of pre-composed signal function (that were defined in Yampa only for performance reasons).

\section{GADTless Programming}
\label{sec:gadtless}

Previous sections should convince the reader that GADTs are a very powerful and helpful extension of the language. However, there are cases when this extension is not available (for example, this feature is not implemented in Hugs compiler). For this reason, there is an interest in replacing them with simpler features while not substantially changing programs and their meanings. This is called GADTless programming~\cite{gadtless}.

Earlier we discussed implementation of expression evaluator using GADT. It can also be implemented using type classes~\cite{finally_tagless}:

\begin{Verbatim}
class Expr e where
    intVal  :: Int -> e Int
    boolVal :: Bool -> e Bool
    add     :: e Int -> e Int -> e Int
    isZero  :: e Int -> e Bool
    if'     :: e Bool -> e t -> e t -> e t
\end{Verbatim}

Bad expressions are still rejected by the type checker:

\begin{Verbatim}
ghci> :t isZero $ boolVal True
Couldn't match expected type `Int' with actual type `Bool'...
ghci> :t isZero $ intVal 5
isZero $ intVal 5 :: Expr e => e Bool
\end{Verbatim}

Evaluation is implemented by defining a helper data type as an instance of \verb|Expr e| type class:

\begin{Verbatim}
newtype Eval a = Eval {runEval :: a}

instance Expr Eval where
    intVal x  = Eval x
    boolVal x = Eval x
    add x y   = Eval $ runEval x + runEval y
    isZero x  = Eval $ runEval x == 0
    if' x y z = if (runEval x) then y else z

t = runEval $ isZero $ intVal 5
\end{Verbatim}

Printing expressions is implemented in a similar way:

\begin{Verbatim}
newtype Print a = Print {printExpr :: String}

instance Expr Print where
    intVal x  = Print $ show x
    boolVal x = Print $ show x
    add x y   = Print $ printExpr x ++ "+" ++ printExpr y
    isZero x  = Print $ "isZero(" ++ printExpr x ++ ")"
    if' x y z = Print $ "if (" ++ printExpr x ++ ") then (" ++
        printExpr y ++ ") else (" ++ printExpr z ++ ")"

t' = printExpr $ isZero $ intVal 5
\end{Verbatim}

\section{Conclusion}
\label{sec:conclusion}

This article has demonstrated the use of GADTs in practice:

\begin{longitem}
\item We have shown that GADTs are useful for domain-specific embedded languages: they allow to statically type-check valid expressions.
\item GADTs allow to express datatype-generic functions using representation types and universal data types.
\item GADTs can be used as a lightweight way to ensure program correctness. They allow to encode domain-specific invariants in data type. The programmer can decide which parts of her or his program require verification and add only relevant invariants. Haskell enforces a phase separation between run-time values and compile-time types. Invariants are expressed using types, so there is no additional run-time cost. But on the other hand, we have shown the issue with the $\bot$ value.
\item We have also described how GADTs were used to improve performance of Yampa programs.
\end{longitem}

% Comparing to an approach to make verification by translating Haskell to other languages and back (such as xmonad), verification in Haskell itself...

As a result, we have shown that the notion of GADTs is a very valuable extension of the Haskell language.

\section*{Acknowledgments}

This tutorial article was inspired by Simon Peyton-Jones talk at LASER 2012 summer school. The source code for the red-black tree example was originally written by Stephanie Weirich for a university course. The author is thankful to have ability to use this great example of GADTs usage in this tutorial. Also the author would like to thank Simon Peyton-Jones, Owen Stephens, Alexander Batischev, Denis Kasak, Mann mit Hut, Alexei Alexandrov, Serge Le Huitouze, Eric Mullen, Egor Kazachkov and Vadim Zaytsev for the review of the article draft and providing valuable comments.

\bibliography{gadt}

\section*{Appendix A. Original Red-Black Tree Source Code}
\label{sec:appendix_a}

\begin{Verbatim}
module RBT1 where

data Color = R | B
data Node a = E | N Color (Node a) a (Node a) 
type Tree a = Node a

member :: Ord a => a -> Tree a -> Bool
member _ E  = False
member x (N _ l a r)
    | x < a  = member x l
    | x > a  = member x r
    | otherwise = True

insert :: Ord a => Tree a -> a -> Tree a
insert t v = blacken (insertInternal t v) where
    insertInternal n@(N c l a r) x
        | x < a = leftBalance  (N c (insertInternal l x) a r)
        | x > a = rightBalance (N c l a (insertInternal r x))
        | otherwise = n
    insertInternal E x = N R E x E
    blacken (N _ l x r) = N B l x r

leftBalance :: Node a -> Node a
leftBalance (N B (N R (N R a x b) y c) z d) =
    N R (N B a x b) y (N B c z d)
leftBalance (N B (N R a x (N R b y c)) z d) =
    N R (N B a x b) y (N B c z d)
leftBalance n = n

rightBalance :: Node a -> Node a
rightBalance (N B a x (N R b y (N R c z d))) =
    N R (N B a x b) y (N B c z d)
rightBalance (N B a x (N R (N R b y c) z d)) =
    N R (N B a x b) y (N B c z d)
rightBalance n = n
\end{Verbatim}

\section*{Appendix B. Red-Black Tree Source Code: Proving the Forth Invariant}
\label{sec:appendix_b}

\begin{Verbatim}
{-# LANGUAGE TypeFamilies, DataKinds, GADTs #-}

module RBT2 where

data Color = R | B
data ColorSingleton (c::Color) where
    SR :: ColorSingleton R
    SB :: ColorSingleton B

data Nat = Zero | Succ Nat

type family   IncBlackHeight (c::Color) (bh::Nat) :: Nat
type instance IncBlackHeight R bh = bh
type instance IncBlackHeight B bh = Succ bh

data Node a (bh::Nat) where
    E :: Node a 'Zero
    N :: ColorSingleton c -> Node a bh -> a
        -> Node a bh -> Node a (IncBlackHeight c bh)

data Tree a where
    Root :: Node a bh -> Tree a

insert :: Ord a => Tree a -> a -> Tree a
insert (Root t) v = blacken (insertInternal t v) where
    insertInternal :: Ord a => Node a n -> a -> Node a n
    insertInternal n@(N c l a r) x
        | x < a = leftBalance  (N c (insertInternal l x) a r)
        | x > a = rightBalance (N c l a (insertInternal r x))
        | otherwise = n
    insertInternal E x = N SR E x E
    blacken (N _ l x r) = Root (N SB l x r)

leftBalance :: Node a bh -> Node a bh
leftBalance (N SB (N SR (N SR a x b) y c) z d) =
    N SR (N SB a x b) y (N SB c z d)
leftBalance (N SB (N SR a x (N SR b y c)) z d) =
    N SR (N SB a x b) y (N SB c z d)
leftBalance n = n

rightBalance :: Node a bh -> Node a bh
rightBalance (N SB a x (N SR b y (N SR c z d))) =
    N SR (N SB a x b) y (N SB c z d)
rightBalance (N SB a x (N SR (N SR b y c) z d)) =
    N SR (N SB a x b) y (N SB c z d)
rightBalance n = n
\end{Verbatim}

\section*{Appendix C. Red-Black Tree Source Code: Proving the Third Invariant}
\label{sec:appendix_c}

\begin{Verbatim}
{-# LANGUAGE TypeFamilies, DataKinds, GADTs, RankNTypes #-}
{-# LANGUAGE MultiParamTypeClasses, FlexibleInstances #-}

module RBT3 where

data Color = R | B
data ColorSingleton (c::Color) where
    SR :: ColorSingleton R
    SB :: ColorSingleton B

data Nat = Zero | Succ Nat

type family   IncBlackHeight (c::Color) (bh::Nat) :: Nat
type instance IncBlackHeight R bh = bh
type instance IncBlackHeight B bh = Succ bh

class ValidColors (parent::Color) (child1::Color) (child2::Color)
instance ValidColors R B B
instance ValidColors B c1 c2

data Node a (bh::Nat) (c::Color) where
    E :: Node a 'Zero B
    N :: ValidColors c c1 c1 => ColorSingleton c -> Node a bh c1
        -> a -> Node a bh c2 -> Node a (IncBlackHeight c bh) c

data Tree a where
    Root :: Node a bh B -> Tree a

data IntNode a (n::Nat) where
    IntNode :: ColorSingleton c -> Node a n c1 -> a
        -> Node a n c2 -> IntNode a (IncBlackHeight c n)
\end{Verbatim}

\begin{Verbatim}
insert :: Ord a => Tree a -> a -> Tree a
insert (Root t) v = blacken (insertInternal t v) where
    insertInternal :: Ord a => Node a n c -> a -> IntNode a n
    insertInternal (N c l a r) x
        | x < a = leftBalance  c (insertInternal l x) a r
        | x > a = rightBalance c l a (insertInternal r x)
        | otherwise = IntNode c l a r
    insertInternal E x = IntNode SR E x E
    blacken (IntNode _ l x r) = Root (N SB l x r)

leftBalance :: ColorSingleton c -> IntNode a n -> a
    -> Node a n c' -> IntNode a (IncBlackHeight c n)
leftBalance SB (IntNode SR (N SR a x b) y c) z d =
    IntNode SR (N SB a x b) y (N SB c z d)
leftBalance SB (IntNode SR a x (N SR b y c)) z d =
    IntNode SR (N SB a x b) y (N SB c z d)
leftBalance c (IntNode SB a x b) z d = IntNode c (N SB a x b) z d
leftBalance c (IntNode SR a@(N SB _ _ _) x b@(N SB _ _ _)) z d =
    IntNode c (N SR a x b) z d
leftBalance c (IntNode SR E x E) z d = IntNode c (N SR E x E) z d
leftBalance _ (IntNode SR (N SR _ _ _) _ _) _ _ =
    error "can't happen"
leftBalance _ (IntNode SR _ _ (N SR _ _ _)) _ _ =
    error "can't happen"

rightBalance :: ColorSingleton c -> Node a n c' -> a
    -> IntNode a n -> IntNode a (IncBlackHeight c n)
rightBalance SB a x (IntNode SR b y (N SR c z d)) =
    IntNode SR (N SB a x b) y (N SB c z d)
rightBalance SB a x (IntNode SR (N SR b y c) z d) =
    IntNode SR (N SB a x b) y (N SB c z d)
rightBalance c a x (IntNode SB b y d) = IntNode c a x (N SB b y d)
rightBalance c a x (IntNode SR b@(N SB _ _ _) y d@(N SB _ _ _)) =
    IntNode c a x (N SR b y d)
rightBalance c a x (IntNode SR E y E) = IntNode c a x (N SR E y E)
rightBalance _ _ _ (IntNode SR (N SR _ _ _) _ _) =
    error "can't happen"
rightBalance _ _ _ (IntNode SR _ _ (N SR _ _ _)) =
    error "can't happen"
\end{Verbatim}

\end{document}

### 2013-05-13 article.tex
rework article to place focus on error-reporting using transformers drop example showing whitespace/comment logging as it's not as important as the error reporting

@@
 
-\title{Combinator Parsers:  a Monad Transformer Approach}
+\title{Error Reporting Parsers:  a Monad Transformer Approach}
 \author{Matt Fenwick\email{mfenwick100@gmail.com}}
@@
 \begin{introduction}
-Parsing, or "the process of analysing a 
-string of symbols according to the rules of a formal grammar" (Wikipedia),
-is one of the most fundamental and interesting problems in programming.
-The general goal of parsing is to build a tree representing the structure
-of the parsed input; further operations such as interpretation or
-code generations are then performed on the tree.
-
-However, in addition to recognizing a stream of tokens and building a parse
+Parsing has long been one of the most fascinating and challenging applications
+of programming, as evidenced by the large number of parsing libraries available
+today.  However, in addition to recognizing a stream of tokens and building a parse
 tree, a practical parser has many additional duties.  If parsing fails,
 the location and cause of the failure must be accurately reported so that
-the user can address the problem; specially formatted comments may be need
-to be capture in order to be used
-to build hyperlinked code documentation, instead of simply thrown away; 
-failed parses may be accompanied by partial results to save the cost
-of reparsing the input before the error.
-
-Monad transformers offer an elegant solution, both to the problem of 
-implementing parser combinators themselves, as well as to the issue of
-building clean and modular parsers that support multiple effects.
-We'll explore what they have to offer using a simple language -- Woof --
-and build a number of separate parsers for it!
+the user can address the problem
+Thus, building composable parsers which accurately report any errors
+is extremely difficult.  In this article, we'll see how building
+a parsing library around monad transformers allows to build clean and simple 
+parsers which nevertheless provide relevant and location-specific error messages.
+At the same time, we'll explore the design and some of the tradeoffs of the 
+popular Haskell library for monad transformers.
 \end{introduction}
@@
 \section{Example 1: Recognition and Tree-Building}
+
+Parsing is "the process of analysing a 
+string of symbols according to the rules of a formal grammar" (Wikipedia)
+The general goal of parsing is to build a tree representing the structure
+of the parsed input; further operations such as interpretation or
+code generations are then performed on the tree.
+
 The first parser we build will be responsible for determining whether input
@@
                                      "{lambda {a b} (c d)", "{defin x y}", "a,b"]
-Left' ("application: missing operator",('(',1,1))
-Left' ("application: missing close parenthesis",('(',1,1))
-Left' ("define: missing symbol",('{',1,1))
-Left' ("define: missing form",('{',1,1))
-Left' ("define: missing close curly",('{',1,1))
-Left' ("lambda: missing parameter list",('{',1,1))
-Left' ("lambda: duplicate parameter names",('{',1,1))
-Left' ("lambda: missing parameter list close curly",('{',1,1))
-Left' ("lambda: missing body form",('{',1,1))
-Left' ("lambda: missing close curly",('{',1,1))
-Left' ("special form: unable to parse",('{',1,1))
-Left' ("woof: unparsed input",(',',1,2))
+Left ("application: missing operator",('(',1,1))
+Left ("application: missing close parenthesis",('(',1,1))
+Left ("define: missing symbol",('{',1,1))
+Left ("define: missing form",('{',1,1))
+Left ("define: missing close curly",('{',1,1))
+Left ("lambda: missing parameter list",('{',1,1))
+Left ("lambda: duplicate parameter names",('{',1,1))
+Left ("lambda: missing parameter list close curly",('{',1,1))
+Left ("lambda: missing body form",('{',1,1))
+Left ("lambda: missing close curly",('{',1,1))
+Left ("special form: unable to parse",('{',1,1))
+Left ("woof: unparsed input",(',',1,2))
 \end{verbatim}
 
-\section{Example 3: whitespace/comment-logging}
-
-\section{Transformer instance semantics}
-
-Pass-through vs. deal-with, examples of each
-
-\section{Limitations of the Monad Transformer Approach}
+\section{Monad Transformer Stacks}
 
@@
 
+\section{Transformer instance semantics}
+
+Pass-through vs. deal-with, examples of each
+
 \section{Conclusion}
-
-Parser combinators and monad transformers FTW!
+Furthur applications of monad transformers to parsing:  junk logging, partial results, ???
 

### 2013-05-13 src/parse3.hs

@@ -1,231 +1,0 @@
-{-# LANGUAGE   NoMonomorphismRestriction
-             , FlexibleContexts       
-             , FlexibleInstances    
-             , FunctionalDependencies
-             , UndecidableInstances      #-}
-
-
-import Control.Monad             (liftM, ap)
-import Control.Monad.State       (MonadState (..), StateT(..))
-import Control.Monad.Trans.Class (lift)
-import Control.Monad.Trans.Maybe (MaybeT(..))
-import Control.Applicative       (Applicative(..))
-import Data.List                 (nub)
-
--- should also try:
---   partial results
-
-
--- the prelude
-
-class Applicative f => Plus f where
-  zero :: f a
-  (<+>) :: f a -> f a -> f a
-
-many0 :: Plus f => f a -> f [a]
-many0 p = many1 p <+> pure []
-
-many1 :: Plus f => f a -> f [a]
-many1 p = pure (:) <*> p <*> many0 p
-
-instance Plus Maybe where
-  zero             =  Nothing
-  Nothing  <+>  y  =  y
-  x        <+>  _  =  x
-
-instance (Functor m, Monad m) => Plus (MaybeT m) where
-  zero                     =  MaybeT (return Nothing)
-  MaybeT l  <+>  MaybeT r  =  MaybeT x
-    where
-      x = l >>= \y -> case y of Nothing -> r;
-                                Just _  -> return y;
-
-instance (Monad m, Plus m) => Plus (StateT s m) where
-  zero                   =  StateT (const zero)
-  StateT f <+> StateT g  =  StateT (\s -> f s <+> g s)
-
-
-class Switch f where
-  switch :: f a -> f ()
-
-instance Switch Maybe where
-  switch (Just _) = Nothing
-  switch Nothing  = Just ()
-
-instance (Functor m, Switch m) => Switch (StateT s m) where
-  switch (StateT f) = StateT (\s -> fmap (const ((), s)) . switch $ f s)
-
-instance Functor m => Switch (MaybeT m) where
-  switch (MaybeT m) = MaybeT (fmap switch m)
-
-
-class Monad m => MonadError e m | m -> e where
-  throwError :: e -> m a
-  catchError :: m a -> (e -> m a) -> m a
-
-instance MonadError e (Either e) where
-  throwError               =  Left
-  catchError  (Right x) _  =  Right x
-  catchError  (Left e)  f  =  f e
-  
[…]

### 2013-05-13 src/parse4.hs
another parser example for logging partial results

@@ -1,0 +1,262 @@
+{-# LANGUAGE   NoMonomorphismRestriction
+             , FlexibleContexts       
+             , FlexibleInstances    
+             , FunctionalDependencies
+             , UndecidableInstances      #-}
+
+
+import Control.Monad             (liftM, ap)
+import Control.Monad.State       (MonadState (..), StateT(..))
+import Control.Monad.Error       (ErrorT(..))
+import Control.Monad.Trans.Class (lift)
+import Control.Monad.Trans.Maybe (MaybeT(..))
+import Control.Applicative       (Applicative(..))
+import Data.List                 (nub)
+
+-- should also try:
+--   partial results
+
+
+-- the prelude
+
+class Applicative f => Plus f where
+  zero :: f a
+  (<+>) :: f a -> f a -> f a
+
+many0 :: Plus f => f a -> f [a]
+many0 p = many1 p <+> pure []
+
+many1 :: Plus f => f a -> f [a]
+many1 p = pure (:) <*> p <*> many0 p
+
+instance Plus Maybe where
+  zero             =  Nothing
+  Nothing  <+>  y  =  y
+  x        <+>  _  =  x
+
+instance (Functor m, Monad m) => Plus (MaybeT m) where
+  zero                     =  MaybeT (return Nothing)
+  MaybeT l  <+>  MaybeT r  =  MaybeT x
+    where
+      x = l >>= \y -> case y of Nothing -> r;
+                                Just _  -> return y;
+
+instance (Monad m, Plus m) => Plus (StateT s m) where
+  zero                   =  StateT (const zero)
+  StateT f <+> StateT g  =  StateT (\s -> f s <+> g s)
+
+instance (Monad m, Plus m) => Plus (ErrorT e m) where
+  zero  =  ErrorT zero
+  ErrorT l <+> ErrorT r  =  ErrorT (l <+> r)
+
+
+class Switch f where
+  switch :: f a -> f ()
+
+instance Switch Maybe where
+  switch (Just _) = Nothing
+  switch Nothing  = Just ()
+
+instance (Functor m, Switch m) => Switch (StateT s m) where
+  -- StateT s m a -> StateT s m ()
+  -- (s -> m (a, s)) -> (s -> m ((), s))
+  switch (StateT f) = StateT (\s -> fmap (const ((), s)) . switch $ f s)
+
+instance Functor m => Switch (MaybeT m) where
+  -- m (Maybe a) -> m (Maybe ())
[…]

### 2013-05-13 src/parse4.hs
ErrorT instance of Monad uses Error, which means I would need to replace all the ErrorT instances; just get rid of the partial results parser example instead

@@ -1,262 +1,0 @@
-{-# LANGUAGE   NoMonomorphismRestriction
-             , FlexibleContexts       
-             , FlexibleInstances    
-             , FunctionalDependencies
-             , UndecidableInstances      #-}
-
-
-import Control.Monad             (liftM, ap)
-import Control.Monad.State       (MonadState (..), StateT(..))
-import Control.Monad.Error       (ErrorT(..))
-import Control.Monad.Trans.Class (lift)
-import Control.Monad.Trans.Maybe (MaybeT(..))
-import Control.Applicative       (Applicative(..))
-import Data.List                 (nub)
-
--- should also try:
---   partial results
-
-
--- the prelude
-
-class Applicative f => Plus f where
-  zero :: f a
-  (<+>) :: f a -> f a -> f a
-
-many0 :: Plus f => f a -> f [a]
-many0 p = many1 p <+> pure []
-
-many1 :: Plus f => f a -> f [a]
-many1 p = pure (:) <*> p <*> many0 p
-
-instance Plus Maybe where
-  zero             =  Nothing
-  Nothing  <+>  y  =  y
-  x        <+>  _  =  x
-
-instance (Functor m, Monad m) => Plus (MaybeT m) where
-  zero                     =  MaybeT (return Nothing)
-  MaybeT l  <+>  MaybeT r  =  MaybeT x
-    where
-      x = l >>= \y -> case y of Nothing -> r;
-                                Just _  -> return y;
-
-instance (Monad m, Plus m) => Plus (StateT s m) where
-  zero                   =  StateT (const zero)
-  StateT f <+> StateT g  =  StateT (\s -> f s <+> g s)
-
-instance (Monad m, Plus m) => Plus (ErrorT e m) where
-  zero  =  ErrorT zero
-  ErrorT l <+> ErrorT r  =  ErrorT (l <+> r)
-
-
-class Switch f where
-  switch :: f a -> f ()
-
-instance Switch Maybe where
-  switch (Just _) = Nothing
-  switch Nothing  = Just ()
-
-instance (Functor m, Switch m) => Switch (StateT s m) where
-  -- StateT s m a -> StateT s m ()
-  -- (s -> m (a, s)) -> (s -> m ((), s))
-  switch (StateT f) = StateT (\s -> fmap (const ((), s)) . switch $ f s)
-
-instance Functor m => Switch (MaybeT m) where
-  -- m (Maybe a) -> m (Maybe ())
[…]

### 2013-05-13 src/parse0.hs
transformer-less example of parsers as introduction

@@ -1,0 +1,40 @@
+import Control.Applicative  (Applicative(..))
+
+class Applicative f => Plus f where
+  zero :: f a
+  (<+>) :: f a -> f a -> f a
+
+many0 :: Plus f => f a -> f [a]
+many0 p = many1 p <+> pure []
+
+many1 :: Plus f => f a -> f [a]
+many1 p = pure (:) <*> p <*> many0 p
+
+instance Plus Maybe where
+  zero = Nothing
+  Nothing <+> y = y
+  x       <+> _ = x
+
+
+-- newtype Parser t a = Parser {getParser :: [t] -> Maybe ([t], a)}
+type Parser t a = [t] -> Maybe ([t], a)
+
+item :: Parser t t
+item [] = Nothing
+item (x:xs) = Just (xs, x)
+
+check :: (a -> Bool) -> Parser t a -> Parser t a
+check f p xs =
+    p xs >>= \(xs', x) ->
+    if f x
+       then Just (xs', x)
+       else Nothing
+
+satisfy :: (t -> Bool) -> Parser t t
+satisfy f = check f item
+
+literal :: Eq t => t -> Parser t t
+literal x = satisfy (== x)
+
+(<|>) :: (a -> Maybe b) -> (a -> Maybe b) -> a -> Maybe b
+l <|> r = \x -> l x <+> r x

### 2013-05-13 article.tex
transformer-less example of parsers as introduction
simplify parse0 and parse1 code to use standard type classes more; work on refactoring article

@@
 \begin{introduction}
-Parsing has long been one of the most fascinating and challenging applications
-of programming, as evidenced by the large number of parsing libraries available
-today.  However, in addition to recognizing a stream of tokens and building a parse
-tree, a practical parser has many additional duties.  If parsing fails,
-the location and cause of the failure must be accurately reported so that
-the user can address the problem
-Thus, building composable parsers which accurately report any errors
-is extremely difficult.  In this article, we'll see how building
-a parsing library around monad transformers allows to build clean and simple 
-parsers which nevertheless provide relevant and location-specific error messages.
-At the same time, we'll explore the design and some of the tradeoffs of the 
-popular Haskell library for monad transformers.
+Parsing is "the process of analysing a string of symbols according to the rules
+of a formal grammar" (Wikipedia), with the general goal of building a tree
+representing the structure of the parsed input, upon which further operations
+such as interpretation or code generation are then performed.
+
+In addition to recognizing a stream of tokens and building a parse tree, a
+real-world parser must recognize faulty input as such, and accurately report the
+location and cause of the failure.  But how -- what generates a parse error,
+what information is included in the error, how do error-reporting parsers
+compose?
+
+Monad transformers provide a clean and simple answer to these questions.
+We'll use them to build a parser combinator library, then we'll use the library
+build a simple parser, and then we'll add in error reporting.
 \end{introduction}
@@
 \section{Parser Combinators}
-Parser combinators have long been used for parsing in the 
-functional programming community, represent a powerful and simple solution.
-Their main advantages are ability to deal with complicated formats (see
-Chomsky levels), language integration, and the declarativeness and
-succinctness of the resulting parsers.
-Parser combinator libraries typically include a set of basic parsers and a
-set of combinators for combining little parsers into big ones.  Parsers are
-typically functions whose type is something like:
+Parser combinators are an excellent approach for building parsers that are easy
+to maintain and expressive.  In addition, they benefit from host-language
+integration, allowing them to snarf features such as type systems.
+
+Let's make a quick pass through a minimal parser combinator library so that we
+can see them in action before we move on to the real stuff.  Without any 
+further ado, here's the code:
 \begin{verbatim}
-[t] -> m (a, [t])
+import Control.Applicative  (Applicative(..), Alternative(..))
+import Control.Monad        (liftM, ap)
+
+newtype Parser t a = 
+    Parser {getParser :: [t] -> Maybe ([t], a)}
+
+item :: Parser t t
+item = Parser f
+  where 
+    f [] = Nothing
+    f (x:xs) = Just (xs, x)
+
+instance Monad (Parser t) where
+  return x = Parser (\ts -> Just (ts, x))
+  Parser p >>= f = Parser (\ts ->
+                             p ts >>= \(ts', x) ->
+                             getParser (f x) ts')
+
+instance Functor (Parser t) where
+  fmap = liftM
+
+instance Applicative (Parser t) where
+  pure = return
+  (<*>) = ap
+
+instance Alternative (Parser t) where
+  empty = Parser (const empty)
+  Parser l <|> Parser r = Parser (\ts -> l ts <|> r ts)
+
+-- similar to mfilter, but for Alternative instead of MonadPlus
+check :: (Monad f, Alternative f) => (a -> Bool) -> f a -> f a
+check p m =
+    m >>= \x -> if p x then return x else empty
+
+literal :: Eq t => t -> Parser t t
+literal x = afilter (== x) item
 \end{verbatim}
-where t is the token type, a is the result type, and m represents some
-computational effects.
-A very simple parser succeeds, consuming a single token, when the token stream
-is not empty, and fails otherwise.  We can make the substitutions t -> Char,
-m -> Maybe, and a -> Char and implement the function:
+First, we need to define the Parser type, which says that a Parser is a function
+that operates on lists of tokens, producing a new list of tokens and a result
+value, and possibly failing.
+Next is the item parser, the simplest parser we'll deal with.  All it does is 
+consume a single token if available, and succeed otherwise.
+For running parsers in sequence, we'll use the combinators from the Monad and
+Applicative type classes.  There's a little bit of book-keeping involved in the
+Monad instance -- it has to make sure that the correct token list is used by
+each parser -- but once we have a Monad instance, the Applicative and Functor
+instances can be handled by imports ap and liftM from Control.Monad.
+For choosing between alternatives, the MonadPlus type class will do, so we create
+an instance that is defined partly in terms of the MonadPlus instance of the
+underlying result type, Maybe.
+Last are a couple of convenience combinators for creating specific token parsers.
+
+Now for some examples:  first, let's see item in action.  It fails on empty lists,
+and succeeds consuming one token otherwise:
 \begin{verbatim}
-    item :: [Char] -> Maybe (Char, [Char])
-    item (t:ts) = Just (t, ts)
-    item []     = Nothing
+*Main> getParser item ""
+Nothing
 
-    -- examples
-    ghci$ item "abcde"
-    Just ('a', "bcde")
+*Main> getParser item "abcde"
+Just ("bcde",'a')
+\end{verbatim}
+We can run parsers in sequence using the Applicative combinators.  First one parser
+is run, then the second.  Note that the token position is handled such that the
+second parser sees the token list output by the first parser, and that both parsers
+must succeed in order for the entire parser to succeed:
+\begin{verbatim}
+*Main> getParser (item *> item) "abcde"
+Just ("cde",'b')
 
-    ghci$ item []
+*Main> getParser (fmap (,) item <*> item) "abcde"
+Just ("cde",('a','b'))
+
+*Main> getParser (item *> item) "a"
 Nothing
 \end{verbatim}
+Parsers built out of literal only succeed when the next token matches the given
+one:
+\begin{verbatim}
+*Main> getParser (literal 'a') "abcde"
+Just ("bcde",'a')
 
-We can also come up with a simple combinator that takes a parser and a function
-as input, and succeeds if the parser succeeds and the function applied to the
-parser's result succeeds:
-\begin{verbatim}
-    check :: (a -> Bool) -> ([Char] -> Maybe (a, [Char])) -> [Char] -> Maybe (a, [Char])
-    check f p = p >>= \(x, ts) -> if (f x) then Just (x, ts) else Nothing
-
-    -- examples
-    ghci$ check (== 'a') item "abcde"
-    Just ('a', "bcde")
-
-    ghci$ check (== 'b') item "abcde"
+*Main> getParser (literal 'a') "bcde"
 Nothing
 \end{verbatim}
-However, in these examples our computational effects are limited to Maybe -- we're
-unable to do error reporting, logging, and so on.  Coming up with more general 
-parsers and combinators will be the topic of the rest of the article.
+Using the MonadPlus combinators, we can create parsers that succeed if any of
+their sub-parsers succeed.  This is what allows parsers to backtrack, trying
+various alternatives if one fails:
+\begin{verbatim}
+*Main> getParser (literal 'a' <|> literal 'b') "abcde"
+Just ("bcde",'a')
 
-\section{Monad transformers}
-Monad transformers are a technique for combining monads to form new monads.
-The motivation behind them is that while single monads are quite useful, oftentimes
-the semantics of multiple monads are required simultaneously; it is then desirable
-to have a modular approach for creating combined monads.
+*Main> getParser (literal 'a' <|> literal 'b') "bacde"
+Just ("acde",'b')
 
-There are multiple to approaches to combining monads and to monad transformers
-themselves, but we'll follow that of the standard transformers/mtl library, since
-it's easy to obtain, easy to use, and quite effective in practice.
-
-Along the way, we'll also explore some of the consequences of combining monads,
-both with respect to semantics and to the definition of the Haskell language.  We'll
-also bump in to some of the limitations, trade-offs and design decisions of the 
-transformer/mtl library.
+*Main> getParser (literal 'a' <|> literal 'b') "cdeab"
+Nothing
+\end{verbatim}
+In the rest of this article, we'll be reworking these basic parser combinators.
+We'll be simplifying the implementations, extending the types of the results,
+all while taking advantage of monad transformers to keep it modular.
 
@@
 
+\section{Monad transformers}
+Monad transformers allow monads to be modularly combined to form new monads
+which combine the semantics of their constituents.  That's exactly what we need
+to do to build parsers.
+We'll try to use the standard transformers/mtl libraries as much as possible, 
+since they're easy to obtain, easy to use, and quite effective in practice, but
+occasionally we'll have to make an extension to avoid certain type class
+instances.
+
+First, note that the parser type we used earlier -- [t] -> Maybe ([t], a) -- 
+can be built out of the StateT monad transformer applied to Maybe, so the new
+parser type is:
+\begin{verbatim}
+type Parser t a = StateT [t] Maybe a
+\end{verbatim}
+And now we can build the item parser using the MonadState combinators:
+\begin{verbatim}
+
+\end{verbatim}
+
+Along the way, we'll also explore some of the consequences of combining monads,
+both with respect to semantics and to the definition of the Haskell language.  We'll
+also bump in to some of the limitations, trade-offs and design decisions of the 
+transformer/mtl library.
+
 \section{Example 1: Recognition and Tree-Building}
-
-Parsing is "the process of analysing a 
-string of symbols according to the rules of a formal grammar" (Wikipedia)
-The general goal of parsing is to build a tree representing the structure
-of the parsed input; further operations such as interpretation or
-code generations are then performed on the tree.
 
@@
 \section{Conclusion}
-Furthur applications of monad transformers to parsing:  junk logging, partial results, ???
+Furthur applications of monad transformers to parsing:  junk logging, partial results, 
+ non-deterministic parsing using the ListT or LogicT or just list data type
 

### 2013-05-13 src/parse0.hs

@@ -1,40 +1,36 @@
-import Control.Applicative  (Applicative(..))
+import Control.Applicative  (Applicative(..), Alternative(..))
+import Control.Monad        (liftM, ap)
 
-class Applicative f => Plus f where
-  zero :: f a
-  (<+>) :: f a -> f a -> f a
-
-many0 :: Plus f => f a -> f [a]
-many0 p = many1 p <+> pure []
-
-many1 :: Plus f => f a -> f [a]
-many1 p = pure (:) <*> p <*> many0 p
-
-instance Plus Maybe where
-  zero = Nothing
-  Nothing <+> y = y
-  x       <+> _ = x
-
-
--- newtype Parser t a = Parser {getParser :: [t] -> Maybe ([t], a)}
-type Parser t a = [t] -> Maybe ([t], a)
+newtype Parser t a = 
+    Parser {getParser :: [t] -> Maybe ([t], a)}
 
 item :: Parser t t
-item [] = Nothing
-item (x:xs) = Just (xs, x)
+item = Parser f
+  where 
+    f [] = Nothing
+    f (x:xs) = Just (xs, x)
 
-check :: (a -> Bool) -> Parser t a -> Parser t a
-check f p xs =
-    p xs >>= \(xs', x) ->
-    if f x
-       then Just (xs', x)
-       else Nothing
+instance Monad (Parser t) where
+  return x = Parser (\ts -> Just (ts, x))
+  Parser p >>= f = Parser (\ts ->
+                             p ts >>= \(ts', x) ->
+                             getParser (f x) ts')
 
-satisfy :: (t -> Bool) -> Parser t t
-satisfy f = check f item
+instance Functor (Parser t) where
+  fmap = liftM
+  
+instance Applicative (Parser t) where
+  pure = return
+  (<*>) = ap          
+
+instance Alternative (Parser t) where
+  empty = Parser (const empty)
+  Parser l <|> Parser r = Parser (\ts -> l ts <|> r ts)
+
+-- similar to mfilter, but for Alternative instead of MonadPlus
+check :: (Monad f, Alternative f) => (a -> Bool) -> f a -> f a
+check p m =
+    m >>= \x -> if p x then return x else empty
 
 literal :: Eq t => t -> Parser t t
-literal x = satisfy (== x)
-
-(<|>) :: (a -> Maybe b) -> (a -> Maybe b) -> a -> Maybe b
-l <|> r = \x -> l x <+> r x
+literal x = check (== x) item

### 2013-05-13 src/parse1.hs

@@ -2,31 +2,11 @@
              , FlexibleContexts           #-}
 
 import Control.Monad.State       (MonadState (..), StateT(..))
-import Control.Applicative       (Applicative(..))
+import Control.Applicative       (Applicative(..), Alternative(..))
 import Data.List                 (nub)
 
 
 -- the prelude:  type classes, data types, instances, and general combinators
-
-class Applicative f => Plus f where
-  zero :: f a
-  (<+>) :: f a -> f a -> f a
-
-many0 :: Plus f => f a -> f [a]
-many0 p = many1 p <+> pure []
-
-many1 :: Plus f => f a -> f [a]
-many1 p = pure (:) <*> p <*> many0 p
-
-instance Plus Maybe where
-  zero = Nothing
-  Nothing <+> y = y
-  x       <+> _ = x
-
-instance (Monad m, Plus m) => Plus (StateT s m) where
-  zero = StateT (const zero)
-  StateT f <+> StateT g = StateT (\s -> f s <+> g s)
-
 
 class Switch f where
   switch :: f a -> f ()
@@ -47,27 +27,30 @@ data AST
     | AApp    AST  [AST]
   deriving (Show, Eq)
 
-item :: (MonadState [t] m, Plus m) => m t
+many0 = many
+many1 = some
+
+item :: (MonadState [t] m, Alternative m) => m t
 item =
     get >>= \xs -> case xs of
                         (t:ts) -> put ts *> pure t;
-                        []     -> zero;
+                        []     -> empty;
 
-check :: (Monad m, Plus m) => (a -> Bool) -> m a -> m a
+check :: (Monad m, Alternative m) => (a -> Bool) -> m a -> m a
 check f p =
     p >>= \x ->
-    if (f x) then return x else zero
+    if (f x) then return x else empty
 
-satisfy :: (MonadState [t] m, Plus m) => (t -> Bool) -> m t
+satisfy :: (MonadState [t] m, Alternative m) => (t -> Bool) -> m t
 satisfy = flip check item
 
-literal :: (Eq t, MonadState [t] m, Plus m) => t -> m t
+literal :: (Eq t, MonadState [t] m, Alternative m) => t -> m t
 literal c = satisfy ((==) c)
 
-not1 :: (MonadState [t] m, Plus m, Switch m) => m a -> m t
+not1 :: (MonadState [t] m, Alternative m, Switch m) => m a -> m t
 not1 p = switch p *> item
 
[…]

### 2013-05-14 src/parse1.hs
reorder parse1; replace oparen with openparen, etc. -- no sense in saving a few characters if it makes the name hard to understand

@@ -8,27 +8,15 @@ import Data.List                 (nub)
 
 -- the prelude:  type classes, data types, instances, and general combinators
 
-class Switch f where
-  switch :: f a -> f ()
+type Parser t a = StateT [t] Maybe a 
 
-instance Switch Maybe where
-  switch (Just _) = Nothing
-  switch Nothing  = Just ()
+runParser :: Parser t a -> [t] -> Maybe (a, [t])
+runParser = runStateT
 
-instance (Functor m, Switch m) => Switch (StateT s m) where
-  switch (StateT f) = StateT (\s -> fmap (const ((), s)) . switch $ f s)
-
-data AST
-    = ANumber Integer
-    | ASymbol String
-    | AString String
-    | ALambda [String] [AST]
-    | ADefine String AST
-    | AApp    AST  [AST]
-  deriving (Show, Eq)
 
 many0 = many
 many1 = some
+
 
 item :: (MonadState [t] m, Alternative m) => m t
 item =
@@ -47,11 +35,33 @@ satisfy = flip check item
 literal :: (Eq t, MonadState [t] m, Alternative m) => t -> m t
 literal c = satisfy ((==) c)
 
+
+class Switch f where
+  switch :: f a -> f ()
+
+instance Switch Maybe where
+  switch (Just _) = Nothing
+  switch Nothing  = Just ()
+
+instance (Functor m, Switch m) => Switch (StateT s m) where
+  switch (StateT f) = StateT (\s -> fmap (const ((), s)) . switch $ f s)
+
 not1 :: (MonadState [t] m, Alternative m, Switch m) => m a -> m t
 not1 p = switch p *> item
 
 end :: (MonadState [t] m, Alternative m, Switch m) => m ()
 end = switch item
+
+
+data AST
+    = ANumber Integer
+    | ASymbol String
+    | AString String
+    | ALambda [String] [AST]
+    | ADefine String AST
+    | AApp    AST  [AST]
+  deriving (Show, Eq)
+
 
 example = "{define \n\
 \  f \n\
@@ -71,36 +81,36 @@ comment = pure (:) <*> literal ';' <*> many0 (not1 $ literal '\n')
 
 munch p = many0 (whitespace <|> comment) *> p
 
-ocurly = munch $ literal '{'
-ccurly = munch $ literal '}'
-oparen = munch $ literal '('
-cparen = munch $ literal ')'
+opencurly  = munch $ literal '{'
[…]

### 2013-05-14 article.tex
more examples for parse1 writeup, fix up code, tidy up explanation

@@
 
+\section{Monad transformers}
+Monad transformers allow monads to be modularly combined to form new monads
+which combine the semantics of their constituents.  That's exactly what we need
+to do to build parsers.
+We'll try to use the standard transformers/mtl libraries as much as possible, 
+since they're easy to obtain, easy to use, and quite effective in practice, but
+occasionally we'll have to make an extension to avoid certain type class
+instances.
+
+First, note that the parser type we used earlier -- [t] -> Maybe ([t], a) -- 
+can be built out of the StateT monad transformer applied to Maybe, so the new
+parser type is:
+\begin{verbatim}
+type Parser t a = StateT [t] Maybe a
+\end{verbatim}
+Now we can build the item parser using the combinators from MonadState, since
+its token lists are the 'state' that it modifies.
+\begin{verbatim}
+{-# LANGUAGE   FlexibleContexts  #-}
+
+import Control.Monad.State       (MonadState (..), StateT(..))
+
+item :: (MonadState [t] m, Alternative m) => m t
+item =
+    get >>= \xs -> case xs of
+                        (t:ts) -> put ts *> pure t;
+                        []     -> empty;
+\end{verbatim}
+Note that we needed to use the FlexibleContexts language extension for the 
+MonadState constraints in the type signatures.  This formulation of monad
+transformers relies heavily on extensions to the type system, more of which 
+we'll have to use for some of the later code in this article.
+
 \section{Introducing Woof:  a Simple Lisp}
@@
 \end{verbatim}
+
 With the additional rule that whitespace and comments may appear in any 
-amount before any token.  Tokens are:  {, }, (, ), and Symbol.
-
-\section{Monad transformers}
-Monad transformers allow monads to be modularly combined to form new monads
-which combine the semantics of their constituents.  That's exactly what we need
-to do to build parsers.
-We'll try to use the standard transformers/mtl libraries as much as possible, 
-since they're easy to obtain, easy to use, and quite effective in practice, but
-occasionally we'll have to make an extension to avoid certain type class
-instances.
-
-First, note that the parser type we used earlier -- [t] -> Maybe ([t], a) -- 
-can be built out of the StateT monad transformer applied to Maybe, so the new
-parser type is:
-\begin{verbatim}
-type Parser t a = StateT [t] Maybe a
-\end{verbatim}
-And now we can build the item parser using the MonadState combinators:
-\begin{verbatim}
-
-\end{verbatim}
-
-Along the way, we'll also explore some of the consequences of combining monads,
-both with respect to semantics and to the definition of the Haskell language.  We'll
-also bump in to some of the limitations, trade-offs and design decisions of the 
-transformer/mtl library.
+amount before any token.  Tokens are:
+\begin{itemize}
+\item \verb+{+
+\item \verb+}+
+\item \verb+(+
+\item \verb+)+
+\item \verb+Symbol+
+\end{itemize}
 
@@
 
-The first parser we build will be responsible for determining whether input
-text conforms to the language definition and simultaneously build an 
+\subsection{Preliminaries}
+The first Woof parser we build will be responsible for determining whether input
+text conforms to the language definition and simultaneously building an 
 Abstract Syntax Tree representing of the structure of the recognized input.
-
 Here's the AST definition we'll use:
@@
 \end{verbatim}
-Plus we'll need a type for our parsers.  We'll use the basic type mentioned 
-above -- \begin{verbatim}[t] -> m (a, [t]) \end{verbatim}
- -- as our starting point.  This type corresponds to our first monad transformer:  
- StateT (from Control.Monad.State), so we can write:
+
+We'll also want a function for running our parsers, using the type given in the
+prevous section:
 \begin{verbatim}
-type Parser a = StateT [Char] Maybe a
-\end{verbatim}
-and a function for running a parser:
-\begin{verbatim}
-runParser :: Parser a -> String -> Maybe (a, String)
+runParser :: Parser [t] a -> [t] -> Maybe (a, [t])
 runParser = runStateT
 \end{verbatim}
-Which means that our Parser is a State monad transformer, with the Maybe type
-as its underlying Monad parameter.
-
-The significance of the Maybe parameter is two-fold:  it allows computational
-effects to express first success versus failure, using the Just and Nothing 
-constructors, respectively, and second backtracking/alternation.  We'll
-capture both of these characteristics with using a type class, Plus:
-\begin{verbatim}
-class Applicative f => Plus f where
-  zero :: f a
-  (<+>) :: f a -> f a -> f a
-\end{verbatim}
-This type class is identical in intent to the Alternative type class from
-Control.Applicative, with zero corresponding to empty and <+> to \verb+<|>+; 
-however, we aren't able to use Alternative directly because
-other of the monad transformers have unsuitable instances for our purposes,
-and as we'll see later, inability to swap out such instances is one of the 
-limitations of the monad transformer approach.
-
-To get around this limitation, we'll create instances for the Plus type class:
-\begin{verbatim}
-instance Plus Maybe where
-  zero = Nothing
-  Nothing <+> y = y
-  x       <+> _ = x
-
-instance (Monad m, Plus m) => Plus (StateT s m) where
-  zero = StateT (const zero)
-  StateT f <+> StateT g = StateT (\s -> f s <+> g s)
-\end{verbatim}
-Note that the StateT instance relies on its monad parameter, m, also being a 
-member of the Plus type class.  This works out for our parser, since m is Maybe
-and we do have a Plus instance for Maybe.
 
@@
 These include the four braces, symbols, strings, numbers, whitespace, and comments.
-To recognize braces, we need to re-implement the item and check combinators that
-we did before.  Using the MonadState and Plus type classes, 
-we can now implement the basic item parser as a stateful
-computation that fails if the state is empty:
+To recognize braces, we need to use the item and check combinators to build a new
+ combinator, satisfy, that checks whether the next character meets a condition:
 \begin{verbatim}
-item :: (MonadState [t] m, Plus m) => m t
-item =
-    get >>= \xs -> case xs of
-                        (t:ts) -> put ts *> pure t;
-                        []     -> zero;
-\end{verbatim}
-And we can easily implement a new version of the check combinator:
-\begin{verbatim}
-check :: (Monad m, Plus m) => (a -> Bool) -> m a -> m a
-check f p =
-    p >>= \x ->
-    if (f x) then return x else zero
-\end{verbatim}
-And use both of them to build a new combinator, satisfy, that checks whether
-the next character meets a condition:
-\begin{verbatim}
-satisfy :: (MonadState [t] m, Plus m) => (t -> Bool) -> m t
+satisfy :: (MonadState [t] m, Alternative m) => (t -> Bool) -> m t
 satisfy = flip check item
 \end{verbatim}
-And finally a parser that checks that the next character is equal to a
-given character:
+
+And a combinator that checks that the next character is equal to a given character:
 \begin{verbatim}
-literal :: (Eq t, MonadState [t] m, Plus m) => t -> m t
+literal :: (Eq t, MonadState [t] m, Alternative m) => t -> m t
 literal c = satisfy ((==) c)
 \end{verbatim}
+
 Now we're ready to build parsers for the four bracket tokens.  
-An exercise for the reader:  what is the inferred type of each parser?
-Let's first build a parser for open curly brackets:
+First, a simple bracket parser:
 \begin{verbatim}
-ocurly = literal '{'
+opencurly = literal '{'
 \end{verbatim}
-And let's try it out in ghci to make sure it does what we expect:
+
+And some examples to make sure it's behaving correctly:
 \begin{verbatim}
-ghci$ runParser ocurly "{abc"
-Just ('{', "abc")
+*Main> map (runParser  opencurly) ["{abc", "}abc", "(abc", "abc"]
+[Just ('{', "abc"), Nothing, Nothing, Nothing]
+\end{verbatim}
 
-ghci$ map (runParser  ocurly) ["}abc", "(abc", "abc"]
-[Nothing, Nothing, Nothing]
-\end{verbatim}
 Good, it correctly accepts open curly brackets and rejects everything else.
@@
 \begin{verbatim}
-ghci$ map (runParser  ocurly) ["}abc", "(abc", "abc"]
+*Main> map (runParser  opencurly) ["}abc", "(abc", "abc"]
 Nothing
 \end{verbatim}
+
 Of course not -- we haven't told it how to skip comments and whitespace yet!
 Let's do that now, first by defining the whitespace and comment patterns, then
-by creating a combinator (which we'll call munch) that takes in a parser, 
+by creating a \verb+munch+ combinator that takes in a parser, 
 throws away all leading junk, then runs the parser:
@@
 \end{verbatim}
-The comment parser uses a new combinator, not1, which is built out of a new type class,
-Switch, whose intention is to model computations which can be switched from successful
-to failing and vice versa:
+Here are these three parsers in action:
+\begin{verbatim}
+*Main> runParser whitespace "\n \t   \tabc"
+Just ("\n \t   \t","abc")
+
+*Main> runParser whitespace "abc"
+Nothing
+
+*Main> runParser comment ";123  \t\nabc"
+Just (";123  \t","\nabc")
+
+*Main> runParser (munch $ literal 'a') "abc"
+Just ('a',"bc")
+
+*Main> runParser (munch $ literal 'a') ";comment!\n   abc"
+Just ('a',"bc")
+
+*Main> runParser (munch $ literal 'a') ";comment!\n   bca"
+Nothing
+
+*Main> runParser (munch $ literal 'a') "bca"
+Nothing
+\end{verbatim}
+
+The comment parser uses a new combinator, not1, which is built out of a new
+type class, Switch, which models computations which can be switched from
+successful to failing and vice versa:
 \begin{verbatim}
@@
 
-not1 :: (MonadState [t] m, Plus m, Switch m) => m a -> m t
+not1 :: (MonadState [t] m, Alternative m, Switch m) => m a -> m t
 not1 p = switch p *> item
 \end{verbatim}
-And now let's fix the ocurly parser using munch and try it out:
+
+With munch in hand, we can now fix opencurly and try it out:
 \begin{verbatim}
-ocurly = munch $ literal '{'
+opencurly = munch $ literal '{'
 
-ghci$ map (runParser ocurly) [";abc\n  {def", "{def", "(def", "}def"]
+*Main> map (runParser opencurly) [";abc\n  {def", "{def", "(def", "}def"]
 [Just ('{',"def"),Just ('{',"def"),Nothing,Nothing]
@@
 curly brace, and fails otherwise.
-We can now define parsers for the other three brace tokens:
+It's straightforward to define parsers for the other three brace tokens:
 \begin{verbatim}
-ccurly = munch $ literal '}'
-oparen = munch $ literal '('
-cparen = munch $ literal ')'
+closecurly = munch $ literal '}'
+openparen  = munch $ literal '('
+closeparen = munch $ literal ')'
 \end{verbatim}
+
 The parser for our last token, Symbol, is slightly more complicated because it allows
@@
   where char = satisfy (flip elem (['a' .. 'z'] ++ ['A' .. 'Z']))
+
+*Main> runParser symbol "abc123"
+Just ("abc","123")
+
+*Main> runParser symbol ";sdfasdfsa\n  abc123"
+Just ("abc","123")
+
+*Main> runParser symbol ";sdfasdfsa\n  123abc"
+Nothing
 \end{verbatim}
-Note the use of the many1 combinator to introduce repetition of any length of at least one,
-and we're again using the munch combinator to throw away leading junk.
+Note the use of many1 to introduce repetition,
+and we're again using munch to throw away leading junk.
 
@@
 application =
-    oparen      >>
-    form        >>= \op ->
-    many0 form  >>= \args ->
-    cparen      >>
-    return (AApp op args)
-\end{verbatim}
-Although this implementation uses the monadic operators >> and >>=, it could
-equivalently have been implemented using the applicative combinators *> and <*>:
-\begin{verbatim}
-application' =
-    oparen        *>
+    openparen    *>
     pure AApp   <*>
@@
     many0 form  <*
-    cparen
+    closeparen
 \end{verbatim}
+Note that, for now, we're doing all parser sequencing using the Applicative
+combinators -- we don't need to use the monadic ones.  Also, we used pure to
+inject a value -- the function AApp -- into the parser.  pure has no effect on
+the tokens and always succeeds:
+\begin{verbatim}
+*Main> runParser (pure 3) ""
+Just (3,"")
+
+*Main> runParser (pure 3) "abc"
+Just (3,"abc")
+\end{verbatim}
+
 Next, we move on to define, which is also a straightforward translation from the grammar:
@@
 define =
-    ocurly                       *>
+    opencurly                    *>
     check (== "define") symbol   *>
@@
     form                        <*
-    ccurly
+    closecurly
 \end{verbatim}
+
 Lambda is more complicated -- not only does it have additional syntax with its parameter list,
@@
 lambda = 
-    ocurly                          >>
-    check (== "lambda") symbol      >>
-    ocurly                          >>
-    check distinct (many0 symbol)   >>= \params ->
-    ccurly                          >>
-    many1 form                      >>= \bodies ->
-    ccurly                          >>
-    return (ALambda params bodies)
+    opencurly                       *>
+    check (== "lambda") symbol      *>
+    opencurly                       *>
+    pure ALambda                   <*>
+    check distinct (many0 symbol)  <*>
+    (closecurly                     *>
+     many1 form                    <*
+     closecurly)
   where
@@
 \end{verbatim}
-The form parser parsers either a symbol, application, or a special form.  We could just write:
+
+A form is either a symbol, application, or a special form.  We could just write:
 \begin{verbatim}
@@
 \end{verbatim}
+
 except that the types don't match -- symbol's type parameter is a String, whereas the other two
-have ASTs, so we can just fix that by mapping the correct function over symbol:
+have ASTs, so we fix that by mapping the ASymbol function over symbol:
 \begin{verbatim}
@@
 \end{verbatim}
+
 The final parser, woof, needs to not only parse all the forms, but also make sure that the
@@
 \end{verbatim}
+
 And we also need to munch any trailing comments and whitespace, so our final woof parser is:
@@
 Let's look at a few examples of our parser in action:
+\begin{verbatim}
+*Main> runParser woof "; \n   {lambda {x y z} a b (c b a)}end"
+Just ([ALambda ["x","y","z"] 
+               [ASymbol "a",
+                ASymbol "b",
+                AApp (ASymbol "c") 
+                     [ASymbol "b",
+                      ASymbol "a"]],
+       ASymbol "end"],
+      "")
+
+*Main> runParser woof "a b c {define q (f x)}"
+Just ([ASymbol "a",
+       ASymbol "b",
+       ASymbol "c",
+       ADefine "q" (AApp (ASymbol "f") [ASymbol "x"])],"")
+
+*Main> runParser woof "a b c {define q (f x)},"
+Nothing
+\end{verbatim}
+Note that the first two parses consume the entire input, which the last one 
+fails because it can't recognize the trailing comma.
 
@@
 \begin{verbatim}
-ghci$ runParser application (countLineCol "oops")
+*Main> runParser application (countLineCol "oops")
 Right' Nothing
 
-ghci$ runParser application (countLineCol "(a b c)")
+*Main> runParser application (countLineCol "(a b c)")
 Right' (Just (AApp (ASymbol "a") [ASymbol "b",ASymbol "c"],[]))
 
-ghci$ runParser application (countLineCol "\n()")
+*Main> runParser application (countLineCol "\n()")
 Left' ("application: missing operator",('(',2,1))
 
-ghci$ runParser application (countLineCol "(a b")
+*Main> runParser application (countLineCol "(a b")
 Left' ("application: missing close parenthesis",('(',1,1))
@@
 \begin{verbatim}
-ghci$ map (runParser woof . countLineCol) ["()", "(a b", "{define (a b c)}",  
+*Main> map (runParser woof . countLineCol) ["()", "(a b", "{define (a b c)}",  
                                      "{define a}", "{define a b", 

### 2013-05-14 src/parse1.hs
don't use monadic operators where applicative ones suffice

@@ -89,11 +89,11 @@ symbol = munch $ many1 char
   where char = satisfy (flip elem (['a' .. 'z'] ++ ['A' .. 'Z']))
 
 application =
-    openparen   >>
-    form        >>= \op ->
-    many0 form  >>= \args ->
-    closeparen  >>
-    return (AApp op args)
+    openparen    *>
+    pure AApp   <*>
+    form        <*>
+    many0 form  <*
+    closeparen
 
 define =
     opencurly                    *>
@@ -102,16 +102,16 @@ define =
     symbol                      <*>
     form                        <*
     closecurly
-    
-lambda =
-    opencurly                       >>
-    check (== "lambda") symbol      >>
-    opencurly                       >>
-    check distinct (many0 symbol)   >>= \params ->
-    closecurly                      >>
-    many1 form                      >>= \bodies ->
-    closecurly                      >>
-    return (ALambda params bodies)
+
+lambda = 
+    opencurly                       *>
+    check (== "lambda") symbol      *>
+    opencurly                       *>
+    pure ALambda                   <*>
+    check distinct (many0 symbol)  <*>
+    (closecurly                     *>
+     many1 form                    <*
+     closecurly)
   where
     distinct names = length names == length (nub names)
 

### 2013-05-14 src/parse2.hs
clean up parse2: improve variable names and definition order

@@ -4,7 +4,6 @@
              , FunctionalDependencies
              , UndecidableInstances      #-}
 
-
 import Control.Monad             (liftM, ap)
 import Control.Monad.State       (MonadState (..), StateT(..))
 import Control.Monad.Trans.Class (lift)
@@ -13,6 +12,12 @@ import Control.Applicative       (Applicative(..))
 import Data.List                 (nub)
 
 -- the prelude
+
+type Parse e t a = StateT [t] (MaybeT (Either e)) a
+
+runParser :: Parse e t a -> [t] -> Either e (Maybe (a, [t]))
+runParser p xs = runMaybeT (runStateT p xs)
+
 
 class Applicative f => Plus f where
   zero :: f a
@@ -40,6 +45,20 @@ instance (Monad m, Plus m) => Plus (StateT s m) where
   zero                   =  StateT (const zero)
   StateT f <+> StateT g  =  StateT (\s -> f s <+> g s)
 
+item :: (MonadState [t] m, Plus m) => m t
+item =
+    get >>= \xs -> case xs of
+                        (t:ts) -> put ts *> pure t;
+                        []     -> zero;
+
+check :: (Monad m, Plus m) => (a -> Bool) -> m a -> m a
+check f p =
+    p >>= \x ->
+    if (f x) then return x else zero
+
+satisfy :: (MonadState [t] m, Plus m) => (t -> Bool) -> m t
+satisfy = flip check item
+
 
 class Switch f where
   switch :: f a -> f ()
@@ -53,6 +72,12 @@ instance (Functor m, Switch m) => Switch (StateT s m) where
 
 instance Functor m => Switch (MaybeT m) where
   switch (MaybeT m) = MaybeT (fmap switch m)
+
+not1 :: (MonadState [t] m, Plus m, Switch m) => m a -> m t
+not1 p = switch p *> item
+
+end :: (MonadState [t] m, Plus m, Switch m) => m ()
+end = switch item
 
 
 class Monad m => MonadError e m | m -> e where
@@ -71,6 +96,9 @@ instance MonadError e m => MonadError e (StateT s m) where
 instance MonadError e m => MonadError e (MaybeT m) where
   throwError      =  lift . throwError
   catchError m f  =  MaybeT $ catchError (runMaybeT m) (runMaybeT . f)
+
+commit :: (MonadError e m, Plus m) => e -> m a -> m a
[…]

### 2013-05-14 src/parse2.hs
remove Plus type class from parse2 -- Alternative will do just fine.  it was the ErrorT instances, and some of the Either instances, that needed to be avoided

@@ -4,11 +4,10 @@
              , FunctionalDependencies
              , UndecidableInstances      #-}
 
-import Control.Monad             (liftM, ap)
 import Control.Monad.State       (MonadState (..), StateT(..))
 import Control.Monad.Trans.Class (lift)
 import Control.Monad.Trans.Maybe (MaybeT(..))
-import Control.Applicative       (Applicative(..))
+import Control.Applicative       (Applicative(..), Alternative(..))
 import Data.List                 (nub)
 
 -- the prelude
@@ -18,45 +17,21 @@ type Parse e t a = StateT [t] (MaybeT (Either e)) a
 runParser :: Parse e t a -> [t] -> Either e (Maybe (a, [t]))
 runParser p xs = runMaybeT (runStateT p xs)
 
+many0 = many
+many1 = some
 
-class Applicative f => Plus f where
-  zero :: f a
-  (<+>) :: f a -> f a -> f a
-
-many0 :: Plus f => f a -> f [a]
-many0 p = many1 p <+> pure []
-
-many1 :: Plus f => f a -> f [a]
-many1 p = pure (:) <*> p <*> many0 p
-
-instance Plus Maybe where
-  zero             =  Nothing
-  Nothing  <+>  y  =  y
-  x        <+>  _  =  x
-
-instance (Functor m, Monad m) => Plus (MaybeT m) where
-  zero                     =  MaybeT (return Nothing)
-  MaybeT l  <+>  MaybeT r  =  MaybeT x
-    where
-      x = l >>= \y -> case y of Nothing -> r;
-                                Just _  -> return y;
-
-instance (Monad m, Plus m) => Plus (StateT s m) where
-  zero                   =  StateT (const zero)
-  StateT f <+> StateT g  =  StateT (\s -> f s <+> g s)
-
-item :: (MonadState [t] m, Plus m) => m t
+item :: (MonadState [t] m, Alternative m) => m t
 item =
     get >>= \xs -> case xs of
                         (t:ts) -> put ts *> pure t;
-                        []     -> zero;
+                        []     -> empty;
 
-check :: (Monad m, Plus m) => (a -> Bool) -> m a -> m a
+check :: (Monad m, Alternative m) => (a -> Bool) -> m a -> m a
 check f p =
     p >>= \x ->
-    if (f x) then return x else zero
+    if (f x) then return x else empty
 
[…]

### 2013-05-14 article.tex
clean up article section on error reporting parser

@@
 \section{Example 2: error-reporting}
-While our first attempt at a parser was useful, when it encountered malformed input, it wasn't
-helpful -- all it could do was throw up its hands in despair and give up.  However, we'd like
-an error message that tells us what the problem was and where it was encountered whenever bad
-input is encountered.  Otherwise, we'd waste time finding the problem by eye -- not a lot of 
-fun.  Here are the examples of bad input that we'd like to get specific error messages for:
+While our first parser worked great when the input was valid, it wasn't helpful
+at all when the input was malformed.  We'd like it to produce an error message
+that tells us what the problem was and where it was encountered whenever bad
+input is encountered.  Otherwise, we'd waste time finding the problem by eye 
+-- not a lot of fun.
+
+To solve this problem, let's start with an informal spec outlining the error
+messages that we expect to get, and the malformed input that causes each of them: 
 \begin{itemize}
@@
 To do this, we're going to bring in a second monad transformer type class -- MonadError -- and
-another monad transformer datatype -- MaybeT.  Our parser stack and runParser function are now
-going to be:
+another monad transformer datatype -- MaybeT.  Our parser stack and runParser function are now:
 \begin{verbatim}
-type Parse e t a = StateT [t] (MaybeT (Either' e)) a
+type Parse e t a = StateT [t] (MaybeT (Either e)) a
 
-runParser :: Parse e t a -> [t] -> Either' e (Maybe (a, [t]))
+runParser :: Parse e t a -> [t] -> Either e (Maybe (a, [t]))
 runParser p xs = runMaybeT (runStateT p xs)
 \end{verbatim}
-No, that ' isn't accidental -- we again introduce a replacement data type to avoid a type class
-instance that doesn't fulfill our needs.  In this case, Either' is intended as a replacement for
-Either, with the only difference being its instance of MonadError.  All other instances for 
-Either' are analagous to those of Either:
+Now we have three different possible results:  success, failure, and an error:
 \begin{verbatim}
-data Either' a b
-    = Left' a
-    | Right' b
-  deriving (Show, Eq)
+*Main> runParser item "abc"
+Right (Just ('a',"bc"))
 
-instance MonadError e (Either' e) where
-  throwError                =  Left'
-  catchError  (Right' x) _  =  Right' x
-  catchError  (Left' e)  f  =  f e
+*Main> runParser item ""
+Right Nothing
+
+*Main> runParser (throwError "oops!") ""
+Left "oops!"
 \end{verbatim}
+Although the choice combinator can recover from failures, it can't recover from
+errors.  Compare:
+\begin{verbatim}
+*Main> runParser (satisfy (== 'a') <|> satisfy (== 'b')) "babc"
+Right (Just ('b',"abc"))
+
+*Main> runParser (throwError "no backtracking!" <|> satisfy (== 'b')) "babc"
+Left "no backtracking!"
+\end{verbatim}
+We'll take advantage of this backtracking restriction to produce accurate error
+messages.
+
+\subsection{Type Class Preliminaries}
+Unfortunately, the standard MonadError instance for the Either datatype does
+not fit our needs -- it requires an Error constraint that we do not want.  To
+solve this problem, we'll implement our own MonadError type class, based on the
+standard one, and instances of it.  We'll need an instance for each member of
+our stack -- Either, StateT, and MaybeT: 
+\begin{verbatim}
+class Monad m => MonadError e m | m -> e where
+  throwError :: e -> m a
+  catchError :: m a -> (e -> m a) -> m a
+
+instance MonadError e (Either e) where
+  throwError               =  Left
+  catchError  (Right x) _  =  Right x
+  catchError  (Left e)  f  =  f e
+
+instance MonadError e m => MonadError e (StateT s m) where
+  throwError      =  lift . throwError
+  catchError m f  =  StateT (\s -> catchError (runStateT m s) 
+                                              (\e -> runStateT (f e) s))
+
+instance MonadError e m => MonadError e (MaybeT m) where
+  throwError      =  lift . throwError
+  catchError m f  =  MaybeT $ catchError (runMaybeT m) (runMaybeT . f)
+\end{verbatim}
+Note how the StateT and MaybeT instances don't really do anything with the
+errors; they just pass the error on through to the next level.  That's why they
+both require that the next lower level supports MonadError.  Also, this code 
+requires some additional extensions:
+\begin{itemize}
+\item FlexibleInstances
+\item FunctionalDependencies
+\item UndecidableInstances 
+\end{itemize}
+
 We'll also need an instance of Switch for MaybeT:
@@
 \subsection{Reporting position}
-To be able to report the position of the error, we'll pre-process the input string, calculating
-the line and column number of each character.  Then instead of our parsers operating on a list
-of characters, they'll operate on a list of Tokens, where a Token is a character tupled up with
-a line number and column number, which we'll express with a simple typedef and some accessor functions:
+To be able to report the position of the error, we'll pre-process the input
+string, calculating the line and column number of each character.  Then instead 
+of our parsers operating on a list of characters, they'll operate on a list of 
+Tokens, where a Token is a character tupled up with a line number and column 
+number, which we'll express with a simple typedef and some accessor functions:
 \begin{verbatim}
@@
 \end{verbatim}
-The function responsible for calculating the position will simply transform a list of characters into
-a list of tokens:
+The function responsible for calculating the position will simply transform a 
+list of characters into a list of tokens:
 \begin{verbatim}
@@
 \subsection{Token parsers}
-Unfortunately, by adding the position to the input, we've created a new problem for ourselves: our 
-original token parsers assumed they would be operating on character lists, so we'll have to rewrite
-them to get rid of that assumption.  The cause of the problem is the \verb+literal+ combinator, which
-we can easily replace with this one that assumes token lists:
+By adding the position to the input, we've created a new problem 
+for ourselves: our original token parsers operated on 
+character lists, but we now need them to work on token lists.  
+The cause of the problem is the \verb+literal+ combinator, so we
+replace it with one for token lists:
 \begin{verbatim}
@@
 \end{verbatim}
-The only differences are that literal was replaced everywhere by character, and a couple uses
-of the chr accessor.  The munch combinator didn't have to change at all (although its inferred
-type did in fact change).
+The only differences were the replacement of literal with character and a couple
+uses of the chr accessor.  The munch combinator didn't have to change at all 
+(although its inferred type did in fact change).
 
 \subsection{Error messages}
-Now we get to the meat of this section -- modifying the parsers to report accurate and useful errors.
-Let's start out with a list, in code, of all the errors that we're going to deal with:
+Now we modify the remaining parsers to report accurate and useful errors.
+Here's the error messages that we'll use to report each error:
 \begin{verbatim}
-eaOp   =  "application: missing operator"
-eaCls  =  "application: missing close parenthesis"
-edSym  =  "define: missing symbol"
-edForm =  "define: missing form"
-edCls  =  "define: missing close curly"
-elPL   =  "lambda: missing parameter list"
-elPrms =  "lambda: duplicate parameter names"
-elPCls =  "lambda: missing parameter list close curly"
-elBody =  "lambda: missing body form"
-elCls  =  "lambda: missing close curly"
-esName =  "special form: unable to parse"
-ewUnp  =  "woof: unparsed input"
+eAppOper    =  "application: missing operator"
+eAppClose   =  "application: missing close parenthesis"
+eDefSym     =  "define: missing symbol"
+eDefForm    =  "define: missing form"
+eDefClose   =  "define: missing close curly"
+eLamParam   =  "lambda: missing parameter list"
+eLamDupe    =  "lambda: duplicate parameter names"
+eLamPClose  =  "lambda: missing parameter list close curly"
+eLamBody    =  "lambda: missing body form"
+eLamClose   =  "lambda: missing close curly"
+eSpecial    =  "special form: unable to parse"
+eWoof       =  "woof: unparsed input"
 \end{verbatim}
-The MonadError type class provides two useful combinators, throwError for generating errors, and 
-catchError for dealing with them.  We can ignore catchError because we don't want to try to recover
-from errors, we just want them reported.  Now the question is, when should an error be created
-and what data should it include?  \verb+throwError+ is too basic for our needs, so we'll introduce
-the \verb+commit+ combinator:
+
+The MonadError type class provides two useful combinators, throwError for 
+generating errors, and catchError for dealing with them.  We can ignore 
+catchError because we don't need to recover from errors, just report them. 
+When should an error be generated and what 
+data should it include?  \verb+throwError+ is too basic for our needs, so we'll 
+introduce the \verb+commit+ combinator:
 \begin{verbatim}
@@
 \end{verbatim}
-This combinator takes two arguments: 1. an error value, and 2. a monadic computation.  It first tries
-to run the computation, and if the computation fails, generates an error whose contents are the first
-argument.  The definition of 'failure' is provided by the \verb+Plus+ instance of \verb+m+.
-So what data do we want to include in our errors?  A String describing the error and a token reporting
-the position should describe the error decently:
+This combinator takes two arguments: 1. an error value, and 2. a monadic 
+computation.  It first tries to run the computation, and if the computation 
+fails, generates an error whose contents are the first argument.  The 
+definition of 'failure' is provided by the \verb+Plus+ instance of \verb+m+. So 
+what data do we want to include in our errors?  A String describing the error 
+and a token reporting the position should describe the error decently:
 \begin{verbatim}
@@
 application =
-    oparen                       >>= \open ->
-    commit (eaOp, open) form     >>= \op ->
+    openparen                             >>= \open ->
+    commit (eAppOper, open) form          >>= \op ->
     many0 form                            >>= \args ->
-    commit (eaCls, open) cparen  >>
+    commit (eAppClose, open) closeparen   >>
     return (AApp op args)
 \end{verbatim}
-This parser works by running the open-parenthesis parser, capturing its value in \verb+open+, and continuing
-with the parsing; if either \verb+form+ or close-parenthesis fails, an appropriate error can be generated
-which includes a string and the location of the open parenthesis.  However, if the open parenthesis is not
-successfully parsed, it's a failure but not an error.  Here are some examples to demonstrate this in action:
+This parser works by running the open-parenthesis parser, capturing its value 
+in \verb+open+, and continuing with the parsing; if either \verb+form+ or 
+close-parenthesis fails, an appropriate error can be generated which includes 
+a string and the location of the open parenthesis.  However, if the open 
+parenthesis is not successfully parsed, it's a failure but not an error.  
+Here are some examples to demonstrate this in action:
 \begin{verbatim}
 *Main> runParser application (countLineCol "oops")
-Right' Nothing
+Right Nothing
 
 *Main> runParser application (countLineCol "(a b c)")
-Right' (Just (AApp (ASymbol "a") [ASymbol "b",ASymbol "c"],[]))
+Right (Just (AApp (ASymbol "a") [ASymbol "b",ASymbol "c"],[]))
 
 *Main> runParser application (countLineCol "\n()")
-Left' ("application: missing operator",('(',2,1))
+Left ("application: missing operator",('(',2,1))
 
 *Main> runParser application (countLineCol "(a b")
-Left' ("application: missing close parenthesis",('(',1,1))
+Left ("application: missing close parenthesis",('(',1,1))
 \end{verbatim}
-Although the previous version of the \verb+application+ parser could be implemented with just applicative combinators, 
-this version certainly can't:  \verb+open+ is used to build the following parsers, making this decidedly monadic.
+Although the previous version of the \verb+application+ parser was implemented 
+with just applicative combinators, this version can't be:  \verb+open+ 
+is used to build the following parsers, making this decidedly monadic.
 
@@
 define =
-    ocurly                       >>= \open ->
+    opencurly                            >>= \open ->
     check (== "define") symbol            *>
     pure ADefine                         <*>
-    commit (edSym, open) symbol  <*>
-    commit (edForm, open) form   <*
-    commit (edCls, open) ccurly
+    commit (eDefSym, open) symbol        <*>
+    commit (eDefForm, open) form         <*
+    commit (eDefClose, open) closecurly
 
 lambda =
-    ocurly                               >>= \open ->
+    opencurly                                >>= \open ->
     check (== "lambda") symbol               >>
-    commit (elPL, open) ocurly           >>= \p_open ->
+    commit (eLamParam, open) opencurly       >>= \p_open ->
     many0 symbol                             >>= \params ->
@@
         then return ()
-        else throwError (elPrms, open))  >>
-    commit (elPCls, open) ccurly         >>
-    commit (elBody, open) (many1 form)   >>= \bodies ->
-    commit (elCls, open) ccurly          >>
+        else throwError (eLamDupe, p_open))  >>
+    commit (eLamPClose, p_open) closecurly   >>
+    commit (eLamBody, open) (many1 form)     >>= \bodies ->
+    commit (eLamClose, open) closecurly      >>
     return (ALambda params bodies)
@@
 \end{verbatim}
-Note that \verb+commit+ isn't used until we can be sure that we're in the right rule:  if, in the \verb+define+
-parser, we committed to parsing the "define" keyword, we wouldn't be able to backtrack and try the \verb+lambda+
-parser if that failed.  Thus, if treating errors as unrecoverable, it's important not to place a \verb+commit+
-where backtracking might be necessary to parse a valid alternative.  However, this isn't the case with how
-we've used it, since once we see an open curly brace and a define, we're sure that we're in the \verb+define+
-rule, and allowing backtracking would only destroy our ability to produce clean and accurate error reports.
+Note that \verb+commit+ isn't used until we can be sure that we're in the right 
+rule:  if, in the \verb+define+ parser, we committed to parsing the "define" 
+keyword, we wouldn't be able to backtrack and try the \verb+lambda+ parser if 
+that failed.  Thus, if treating errors as unrecoverable, it's important not to 
+place a \verb+commit+ where backtracking might be necessary to parse a valid 
+alternative.  However, this isn't the case with how we've used it, since once 
+we see an open curly brace and a define, we're sure that we're in the \verb+define+
+rule, and allowing backtracking would only destroy our ability to produce clean 
+and accurate error reports.
 
 \subsection{Error messages:  finishing touches}
-To complete the error-reporting parser, we need to change a couple more details.  First, we want an error 
-reported when an open curly brace is found, but no special form can be reported.  To accomplish this, we
-can extend the \verb+special+ parser with a third alternative, which results in an error if it can parse
+To complete the error-reporting parser, we need to change a couple more details. 
+First, we want an error reported when an open curly brace is found, but no 
+special form can be reported.  To accomplish this, we can extend the \verb+special+ 
+parser with a third alternative, which results in an error if it can parse
 an open curly brace, but in a backtracking failure if not:
@@
 \begin{verbatim}
-runParser woof (countLineCol "{define fgh {lambda {f x y} {define a (b c)} (f y a (a x))}}")
-Right' (Just ([ADefine "fgh" 
-                       (ALambda ["f","x","y"] 
-                                [ADefine "a" (AApp (ASymbol "b") [ASymbol "c"]),
-                                 AApp (ASymbol "f") [ASymbol "y",
-                                                     ASymbol "a",
-                                                     AApp (ASymbol "a") [ASymbol "x"]]])],
+*Main> runParser woof (countLineCol "{define fgh {lambda {f x} (f x)}}")
+Right (Just ([ADefine "fgh" (ALambda ["f","x"] 
+                                     [AApp (ASymbol "f") [ASymbol "x"]])],
        []))
 \end{verbatim}
-Here are examples showing errors detected by the parser:
+Here's the proof showing that the parser correctly report all of the errors 
+covered by our spec:
 \begin{verbatim}
-*Main> map (runParser woof . countLineCol) ["()", "(a b", "{define (a b c)}",  
-                                     "{define a}", "{define a b", 
-                                     "{lambda (a b c)}", "{lambda {a b a} (c d)}", 
-                                     "{lambda {a b (c d)}", "{lambda {a b}}", 
-                                     "{lambda {a b} (c d)", "{defin x y}", "a,b"]
+*Main> let p = runParser woof . countLineCol
+
+*Main> p "()"
 Left ("application: missing operator",('(',1,1))
+
+*Main> p "(a b"
 Left ("application: missing close parenthesis",('(',1,1))
+
+*Main> p "{define (a b c)}"
 Left ("define: missing symbol",('{',1,1))
+
+*Main> p "{define a}"
 Left ("define: missing form",('{',1,1))
+
+*Main> p "{define a b"
 Left ("define: missing close curly",('{',1,1))
+
+*Main> p "{lambda (a b c)}"
 Left ("lambda: missing parameter list",('{',1,1))
-Left ("lambda: duplicate parameter names",('{',1,1))
-Left ("lambda: missing parameter list close curly",('{',1,1))
+
+*Main> p "{lambda {a b a} (a b c)}"
+Left ("lambda: duplicate parameter names",('{',1,9))
+
+*Main> p "{lambda {a b (c d)}"
+Left ("lambda: missing parameter list close curly",('{',1,9))
+
+*Main> p "{lambda {a b}}"
 Left ("lambda: missing body form",('{',1,1))
+
+*Main> p "{lambda {a b} (c d)"
 Left ("lambda: missing close curly",('{',1,1))
+
+*Main> p "{defin x y}"
 Left ("special form: unable to parse",('{',1,1))
+
+*Main> p "a,b"
 Left ("woof: unparsed input",(',',1,2))

### 2013-05-14 article.tex
stack order:  code and write-up; lame conclusion

@@
 
-\section{Monad Transformer Stacks}
+\section{Stack Order}
+Not only are the contents of a transformer stack important, but the order of the
+layers as well.  Let's revisit our first woof parser -- what if our stack had
+instead been Maybe/State instead of State/Maybe?
+\begin{verbatim}
+State/Maybe:   s -> Maybe (a, s)
 
-1. Exact vs. relative layer order; 
-2. one instance per stack, even when more than one is applicable --
-how to specify which one?
-3. an inappropriate instance higher-up in the stack can mask a 
-desired instance lower down in the stack -- that's why we had to
-create a replacement for the Alternative type class and provide
-some new instances for it -- to be able to use the right instances
+Maybe/State:   s -> (Maybe a, s)
+\end{verbatim}
 
-\section{Transformer instance semantics}
+given:
+\begin{verbatim}
+import Control.Monad.Trans.Maybe (MaybeT(..))
+import Data.Functor.Identity     (Identity(..))
 
-Pass-through vs. deal-with, examples of each
+type Parser' t a = MaybeT (StateT [t] Identity) a
+
+runParser' :: Parser' t a -> [t] -> (Maybe a, [t])
+runParser' p xs = runIdentity $ runStateT (runMaybeT p) xs
+\end{verbatim}
+We can compare it to the earlier parser:
+\begin{verbatim}
+*Main> runParser' (literal 'a') "aqrs"
+(Just 'a',"qrs")
+
+*Main> runParser' (literal 'a') "qrs"
+(Nothing,"rs")
+\end{verbatim}
+Whereas:
+\begin{verbatim}
+*Main> runParser (literal 'a') "qrs"
+Nothing
+
+*Main> runParser (literal 'a') "aqrs"
+Just ('a',"qrs")
+\end{verbatim}
+The difference is that Maybe/State always returns a modified state, regardless
+of the success of the computation, whereas State/Maybe only returns a modified
+state if the computation is successful.  This means that backtracking doesn't
+work right with Maybe/State -- tokens are consumed even when the parsers
+don't match, as the example shows.
+
+On the other hand, the order of Error and Maybe relative to each other isn't 
+important.  Try it for yourself!
+
+Another important point to note is that our parsers didn't actually specify
+which order -- Maybe/State or State/Maybe -- they prefer, and in fact they will 
+work with either order.  Since our parsers don't specify the stack order, the 
+semantics are ultimately determined by the actual stack we use.
 
 \section{Conclusion}
-Furthur applications of monad transformers to parsing:  junk logging, partial results, 
- non-deterministic parsing using the ListT or LogicT or just list data type
+Monad transformers provide an elegant solution to error reporting in parser
+combinators.  Better yet, they are extensible -- we could go back and add more
+features to our parsers using additional transformers, perhaps extending our
+parser to save partial results, so that if there's an error, it reports the
+progress that it had made before the error.  We could do that using a Writer
+transformer.  Or we check to make sure that all variables are in scope when
+they're used -- using a Reader transformer.  We could even capture whitespace
+and comment tokens, instead of just throwing them away, in case they contained
+valuable information.  And the best part of it is, most of the parsers we already
+wrote would continue to work fine without needing any modifications.  We'd just
+have to set up our transformer stack and modify the relevant parsers!
 

### 2013-05-15 article.tex
fix typo
article: clean up references to code variable names:  surround with \verb+...+; remove references to no-longer existent Plus type class

@@
 \title{Error Reporting Parsers:  a Monad Transformer Approach}
-\author{Matt Fenwick\email{mfenwick100@gmail.com}}
+\author{Matt Fenwick\email{mfenwick100@@gmail.com}}
 
@@
 \end{verbatim}
-First, we need to define the Parser type, which says that a Parser is a function
-that operates on lists of tokens, producing a new list of tokens and a result
+First, we define the Parser type, which says that a Parser is a function
+of a list of tokens, producing a new list of tokens and a result
 value, and possibly failing.
-Next is the item parser, the simplest parser we'll deal with.  All it does is 
+Next is the \verb+item+ parser, the simplest parser we'll deal with.  All it does is 
 consume a single token if available, and succeed otherwise.
@@
 each parser -- but once we have a Monad instance, the Applicative and Functor
-instances can be handled by imports ap and liftM from Control.Monad.
-For choosing between alternatives, the MonadPlus type class will do, so we create
-an instance that is defined partly in terms of the MonadPlus instance of the
-underlying result type, Maybe.
+instances can be handled by imports \verb+ap+ and \verb+liftM+ from Control.Monad.
+For choosing between alternatives, the Alternative type class will do, so we create
+an instance that is defined partly in terms of the Alternative instance of the
+underlying result type, \verb+Maybe+.
 Last are a couple of convenience combinators for creating specific token parsers.
 
-Now for some examples:  first, let's see item in action.  It fails on empty lists,
+Now for some examples:  first, let's see \verb+item+ in action.  It fails on empty lists,
 and succeeds consuming one token otherwise:
@@
 \end{verbatim}
+
 We can run parsers in sequence using the Applicative combinators.  First one parser
@@
 \end{verbatim}
-Parsers built out of literal only succeed when the next token matches the given
+
+Parsers built out of \verb+literal+ only succeed when the next token matches the given
 one:
@@
 \end{verbatim}
-Using the MonadPlus combinators, we can create parsers that succeed if any of
+
+Using the Alternative combinators, we can create parsers that succeed if any of
 their sub-parsers succeed.  This is what allows parsers to backtrack, trying
@@
 
-First, note that the parser type we used earlier -- [t] -> Maybe ([t], a) -- 
-can be built out of the StateT monad transformer applied to Maybe, so the new
+First, note that the parser type we used earlier -- \verb+[t] -> Maybe ([t], a)+
+-- can be built out of the StateT monad transformer applied to Maybe, so the new
 parser type is:
@@
 \end{verbatim}
-Now we can build the item parser using the combinators from MonadState, since
-its token lists are the 'state' that it modifies.
+
+Now we can build the \verb+item+ parser using the combinators from MonadState, 
+since its token lists are the 'state' that it modifies.
 \begin{verbatim}
@@
 
-And a combinator that checks that the next character is equal to a given character:
+And a the \verb+literal+ combinator:
 \begin{verbatim}
@@
 \end{verbatim}
+
 Here are these three parsers in action:
@@
 
-The comment parser uses a new combinator, not1, which is built out of a new
-type class, Switch, which models computations which can be switched from
+The comment parser uses a new combinator, \verb+not1+, which is built out of a 
+new type class, Switch, which models computations which can be switched from
 successful to failing and vice versa:
@@
 
-With munch in hand, we can now fix opencurly and try it out:
+With \verb+munch+ in hand, we can now fix \verb+opencurly+ and try it out:
 \begin{verbatim}
@@
 \end{verbatim}
-Excellent!  Now the open curly parser correctly throws away whitespace and comments,
-and succeeds, consuming one character, when the first non-junk character is an open 
-curly brace, and fails otherwise.
+
+Excellent!  Now the \verb+opencurly+ parser correctly throws away whitespace 
+and comments and succeeds, consuming one character, when the first non-junk 
+character is an open curly brace, failing otherwise.
 It's straightforward to define parsers for the other three brace tokens:
@@
 \end{verbatim}
-Note the use of many1 to introduce repetition,
-and we're again using munch to throw away leading junk.
+
+Note the use of \verb+many1+ to introduce repetition,
+and we're again using \verb+munch+ to throw away leading junk.
 
@@
 \end{verbatim}
+
 Note that, for now, we're doing all parser sequencing using the Applicative
-combinators -- we don't need to use the monadic ones.  Also, we used pure to
-inject a value -- the function AApp -- into the parser.  pure has no effect on
-the tokens and always succeeds:
+combinators -- we don't need to use the monadic ones.  Also, we used \verb+pure+ 
+to inject a value -- the function \verb+AApp+ -- into the parser.  \verb+pure+ 
+has no effect on the tokens and always succeeds:
 \begin{verbatim}
@@
 
-Next, we move on to define, which is also a straightforward translation from the grammar:
+Next, we move on to \verb+define+, which is also a straightforward translation 
+from the grammar:
 \begin{verbatim}
@@
 but we also need to ensure that the parameter names are unique.  We can do that using the 
-check combinator again:
+\verb+check+ combinator again:
 \begin{verbatim}
@@
 \subsection{Finishing touches}
-Now all we have left to do is to define the special, form (which we've already used to build
-the previous parsers), and woof parsers.  The special parser is simple:  it parses either 
-define or lambda:
+Now all we have left to do is to define the \verb+special+, \verb+form+ (which 
+we've already used to build the previous parsers), and \verb+woof+ parsers.  
+The special parser is simple -- it parses either define or lambda:
 \begin{verbatim}
@@
 
-except that the types don't match -- symbol's type parameter is a String, whereas the other two
-have ASTs, so we fix that by mapping the ASymbol function over symbol:
+except that the types don't match -- \verb+symbol+'s type parameter is a String, 
+whereas the other two have ASTs, so we fix that by mapping the \verb+ASymbol+ 
+function over \verb+symbol+:
 \begin{verbatim}
@@
 
-The final parser, woof, needs to not only parse all the forms, but also make sure that the
-entire input is consumed.  The end of input check is done with the end parser:
+The final parser, \verb+woof+, needs to not only parse all the forms, but also 
+make sure that the entire input is consumed.  The end of input check is done 
+with the end parser:
 \begin{verbatim}
@@
 
-And we also need to munch any trailing comments and whitespace, so our final woof parser is:
+And we also need to munch any trailing comments and whitespace, so our final 
+woof parser is:
 \begin{verbatim}
@@
 \end{verbatim}
-Note that the first two parses consume the entire input, which the last one 
+Note that the first two parses consume the entire input, while the last one 
 fails because it can't recognize the trailing comma.
@@
 \end{verbatim}
-Now we have three different possible results:  success, failure, and an error:
+
+A parser has three different possible results:  success, failure, and error:
 \begin{verbatim}
@@
 \end{verbatim}
-Although the choice combinator can recover from failures, it can't recover from
+Although the \verb+<|>+ combinator can recover from failures, it can't recover from
 errors.  Compare:
@@
 \subsection{Type Class Preliminaries}
-Unfortunately, the standard MonadError instance for the Either datatype does
-not fit our needs -- it requires an Error constraint that we do not want.  To
+Unfortunately, the standard MonadError instance for the \verb+Either+ datatype does
+not fit our needs -- it requires an \verb+Error+ constraint that we do not want.  To
 solve this problem, we'll implement our own MonadError type class, based on the
@@
 \begin{verbatim}
-character :: (MonadState [Token] m, Plus m) => Char -> m Token
+character :: (MonadState [Token] m, Alternative m) => Char -> m Token
 character c = satisfy ((==) c . chr)
 \end{verbatim}
+
 And now our token parsers are:
@@
 \end{verbatim}
-The only differences were the replacement of literal with character and a couple
-uses of the chr accessor.  The munch combinator didn't have to change at all 
-(although its inferred type did in fact change).
+
+The only differences were the replacement of \verb+literal+ with \verb+character+ 
+and a couple uses of the \verb+chr+ accessor.  The \verb+munch+ combinator didn't 
+have to change at all (although its inferred type did in fact change).
 
@@
 Now we modify the remaining parsers to report accurate and useful errors.
-Here's the error messages that we'll use to report each error:
+Here are the error messages that we'll use to report each error:
 \begin{verbatim}
@@
 
-The MonadError type class provides two useful combinators, throwError for 
-generating errors, and catchError for dealing with them.  We can ignore 
-catchError because we don't need to recover from errors, just report them. 
-When should an error be generated and what 
+The MonadError type class provides two useful combinators, \verb+throwError+ for 
+generating errors, and \verb+catchError+ for dealing with them.  We can ignore 
+the second because we don't need to recover from errors, just report them, 
+but when should an error be generated and what 
 data should it include?  \verb+throwError+ is too basic for our needs, so we'll 
@@
 \begin{verbatim}
-commit :: (MonadError e m, Plus m) => e -> m a -> m a
-commit err p = p <+> throwError err
+commit :: (MonadError e m, Alternative m) => e -> m a -> m a
+commit err p = p <|> throwError err
 \end{verbatim}
+
 This combinator takes two arguments: 1. an error value, and 2. a monadic 
@@
 fails, generates an error whose contents are the first argument.  The 
-definition of 'failure' is provided by the \verb+Plus+ instance of \verb+m+. So 
+definition of 'failure' is provided by the Alternative instance of \verb+m+. So 
 what data do we want to include in our errors?  A String describing the error 
@@
 \end{verbatim}
-Now lets start with \verb+Application+ as our first modified, error-reporting parser:
+
+Now lets start with \verb+application+ as our first modified, error-reporting parser:
 \begin{verbatim}
@@
 parenthesis is not successfully parsed, it's a failure but not an error.  
-Here are some examples to demonstrate this in action:
+Examples:
 \begin{verbatim}
@@
 \end{verbatim}
+
 Although the previous version of the \verb+application+ parser was implemented 
@@
 \end{verbatim}
+
 Note that \verb+commit+ isn't used until we can be sure that we're in the right 
@@
 \end{verbatim}
+
 The \verb+form+ parser doesn't have to be changed:
@@
 \end{verbatim}
+
 When we can no longer parse any forms, instead of failing, we'd like to report an error if there's any input
@@
 \end{verbatim}
-Here's the proof showing that the parser correctly report all of the errors 
+Here's the proof showing that the parser correctly reports all of the errors 
 covered by our spec:
@@
 \end{verbatim}
+
 We can compare it to the earlier parser:
@@
 \end{verbatim}
+
 The difference is that Maybe/State always returns a modified state, regardless
@@
 On the other hand, the order of Error and Maybe relative to each other isn't 
-important.  Try it for yourself!
+important.  But don't just take my word for it -- try it for yourself!
 

### 2013-05-16 article.tex
bibliography:  add references, cite them, create simple build script to run pdflatex and bibtex

@@
 Parsing is "the process of analysing a string of symbols according to the rules
-of a formal grammar" (Wikipedia), with the general goal of building a tree
+of a formal grammar" \cite{wikipedia_parsing}, with the general goal of building a tree
 representing the structure of the parsed input, upon which further operations
@@
 integration, allowing them to snarf features such as type systems.
+
+There are such many excellent papers for reading about parser combinators, such
+as this \cite{wadler}, this \cite{hutton}, and this \cite{leijen}.  In fact, the
+inspiration and terminology for much of this article is drawn directly from
+those references.
 
@@
 \section{Monad transformers}
-Monad transformers allow monads to be modularly combined to form new monads
+Monad transformers \cite{liang} allow monads to be modularly combined to form new monads
 which combine the semantics of their constituents.  That's exactly what we need
 to do to build parsers.
-We'll try to use the standard transformers/mtl libraries as much as possible, 
+We'll try to use the standard transformers/mtl \cite{mtl} libraries as much as possible, 
 since they're easy to obtain, easy to use, and quite effective in practice, but
@@
 
-The language definition in pseudo-BNF is:
+The language definition in pseudo-BNF \cite{bnf} is:
 
@@
 Not only are the contents of a transformer stack important, but the order of the
-layers as well.  Let's revisit our first woof parser -- what if our stack had
+layers as well \cite{stack}.  Let's revisit our first woof parser -- what if our stack had
 instead been Maybe/State instead of State/Maybe?
@@
 
-\bibliography{Author}
+\bibliography{article}
 

### 2013-05-16 makeme.sh

@@ -1,0 +1,5 @@
+pdflatex article
+bibtex article
+pdflatex article
+pdflatex article
+firefox article.pdf &

### 2013-05-17 article.tex
format code lines to prevent them running off right edge of article -- add linebreaks in convenient places

@@
 instance (Functor m, Switch m) => Switch (StateT s m) where
-  switch (StateT f) = StateT (\s -> fmap (const ((), s)) . switch $ f s)
+  switch (StateT f) = StateT g
+    where 
+      g s = fmap (const ((), s)) . switch $ f s
 
@@
 
-*Main> map (runParser opencurly) [";abc\n  {def", "{def", "(def", "}def"]
-[Just ('{',"def"),Just ('{',"def"),Nothing,Nothing]
+*Main> runParser opencurly ";abc\n{def"
+Just ('{', "def")
+
+*Main> runParser opencurly "{def"
+Just ('{', "def")
+
+*Main> runParser opencurly "(def"
+Nothing
+
+*Main> runParser opencurly "}def"
+Nothing
 \end{verbatim}
@@
 
-*Main> runParser (throwError "no backtracking!" <|> satisfy (== 'b')) "babc"
-Left "no backtracking!"
+*Main> runParser (throwError "no!" <|> satisfy (== 'b')) "babc"
+Left "no!"
 \end{verbatim}
@@
   throwError      =  lift . throwError
-  catchError m f  =  StateT (\s -> catchError (runStateT m s) 
-                                              (\e -> runStateT (f e) s))
+  catchError m f  =  StateT g
+    where
+      g s = catchError (runStateT m s) 
+                       (\e -> runStateT (f e) s)
 
@@
   throwError      =  lift . throwError
-  catchError m f  =  MaybeT $ catchError (runMaybeT m) (runMaybeT . f)
+  catchError m f  =  MaybeT $ catchError (runMaybeT m) 
+                                         (runMaybeT . f)
 \end{verbatim}
@@
 \begin{verbatim}
-character :: (MonadState [Token] m, Alternative m) => Char -> m Token
+character :: (MonadState [Token] m, 
+              Alternative m) => Char -> m Token
 character c = satisfy ((==) c . chr)
@@
 whitespace = many1 $ satisfy (flip elem " \n\t\r\f" . chr)
-comment = pure (:) <*> character ';' <*> many0 (not1 $ character '\n')
+comment = pure (:) <*> character ';' <*> many0 chars
+  where chars = not1 $ character '\n'
 
@@
 symbol = munch $ many1 char
-  where char = fmap chr $ satisfy (flip elem (['a' .. 'z'] ++ ['A' .. 'Z']) . chr)
+  where char = fmap chr $ satisfy (f . chr)
+        f = flip elem (['a' .. 'z'] ++ ['A' .. 'Z'])
 \end{verbatim}
@@
 \begin{verbatim}
-special = define <+> lambda <+> (ocurly >>= \o -> throwError (esName, o))
+special = define   <+> 
+          lambda   <+> 
+          (ocurly >>= \o -> throwError (esName, o))
 \end{verbatim}
@@
 \begin{verbatim}
-*Main> runParser woof (countLineCol "{define fgh {lambda {f x} (f x)}}")
+*Main> let p = runParser woof . countLineCol
+
+*Main> p "{define fgh {lambda {f x} (f x)}}"
 Right (Just ([ADefine "fgh" (ALambda ["f","x"] 
-                                     [AApp (ASymbol "f") [ASymbol "x"]])],
+                                     [AApp (ASymbol "f") 
+                                           [ASymbol "x"]])],
        []))
@@
 \begin{verbatim}
-*Main> let p = runParser woof . countLineCol
-
 *Main> p "()"

### 2013-05-17 src/parse2.hs

@@ -66,7 +66,10 @@ instance MonadError e (Either e) where
   
 instance MonadError e m => MonadError e (StateT s m) where
   throwError      =  lift . throwError
-  catchError m f  =  StateT (\s -> catchError (runStateT m s) (\e -> runStateT (f e) s))
+  catchError m f  =  StateT g
+    where
+      g s = catchError (runStateT m s) 
+                       (\e -> runStateT (f e) s)
 
 instance MonadError e m => MonadError e (MaybeT m) where
   throwError      =  lift . throwError

### 2013-05-17 article.tex
improve wording, grammar on first four pages of article

@@
 In addition to recognizing a stream of tokens and building a parse tree, a
-real-world parser must recognize faulty input as such, and accurately report the
-location and cause of the failure.  But how -- what generates a parse error,
+real-world parser must recognize faulty input, and accurately report the
+location and cause of the problem.  But how -- what generates a parse error,
 what information is included in the error, how do error-reporting parsers
@@
 We'll use them to build a parser combinator library, then we'll use the library
-build a simple parser, and then we'll add in error reporting.
+to build a simple parser, and finally we'll add in error reporting.
 \end{introduction}
@@
 \section{Parser Combinators}
-Parser combinators are an excellent approach for building parsers that are easy
-to maintain and expressive.  In addition, they benefit from host-language
-integration, allowing them to snarf features such as type systems.
+Parser combinators are an excellent approach for building expressive parsers 
+that are easy to maintain.  In addition, they benefit from host-language
+integration, allowing them to snarf features such as type systems and test 
+frameworks.
 
-There are such many excellent papers for reading about parser combinators, such
-as this \cite{wadler}, this \cite{hutton}, and this \cite{leijen}.  In fact, the
-inspiration and terminology for much of this article is drawn directly from
-those references.
+There are many excellent papers on parser combinators, such
+as this \cite{wadler} classic Wadler paper on non-deterministic parser combinators, 
+Hutton's paper which covers factoring parsers into smaller pieces \cite{hutton}, 
+and Leijen's Parsec paper \cite{leijen}.  
+The inspiration and and much of the names used in this article are drawn 
+from those references.
 
 Let's make a quick pass through a minimal parser combinator library so that we
-can see them in action before we move on to the real stuff.  Without any 
+can see them in action before we move on to the hard stuff.  Without any 
 further ado, here's the code:
@@
 Next is the \verb+item+ parser, the simplest parser we'll deal with.  All it does is 
-consume a single token if available, and succeed otherwise.
+consume a single token if available, and fail otherwise.
 For running parsers in sequence, we'll use the combinators from the Monad and
@@
 We can run parsers in sequence using the Applicative combinators.  First one parser
-is run, then the second.  Note that the token position is handled such that the
-second parser sees the token list output by the first parser, and that both parsers
-must succeed in order for the entire parser to succeed:
+is run, then the second, and the token position is threaded between the two.
+Note that both parsers must succeed in order for the combined parser to succeed:
 \begin{verbatim}
@@
 
-Parsers built out of \verb+literal+ only succeed when the next token matches the given
-one:
+Parsers built out of \verb+literal+ only succeed when the next token matches 
+the given one:
 \begin{verbatim}
@@
 \section{Monad transformers}
-Monad transformers \cite{liang} allow monads to be modularly combined to form new monads
-which combine the semantics of their constituents.  That's exactly what we need
-to do to build parsers.
+Monad transformers \cite{liang} allow monads to be modularly combined to form 
+new, more complicated monads; the combined monads support all the operations of
+their constituents.  Doing just that -- combining multiple monads into a stack --
+we can build semantically rich yet modular parsers.
 We'll try to use the standard transformers/mtl \cite{mtl} libraries as much as possible, 
 since they're easy to obtain, easy to use, and quite effective in practice, but
-occasionally we'll have to make an extension to avoid certain type class
-instances.
+occasionally we'll have to make a slight modification or extensions.
 
 First, note that the parser type we used earlier -- \verb+[t] -> Maybe ([t], a)+
--- can be built out of the StateT monad transformer applied to Maybe, so the new
-parser type is:
+-- can be built out of the StateT monad transformer applied to Maybe.  We can 
+now define the new parser type as:
 \begin{verbatim}
@@
 
-Now we can build the \verb+item+ parser using the combinators from MonadState, 
-since its token lists are the 'state' that it modifies.
+We can build the \verb+item+ parser using the combinators from MonadState --
+token lists are the 'state' that \verb+item+ must inspect and modify:
 \begin{verbatim}

### 2013-05-17 src/parse1.hs
oops -- why were there unused terms in the AST ADTs?

@@ -54,9 +54,7 @@ end = switch item
 
 
 data AST
-    = ANumber Integer
-    | ASymbol String
-    | AString String
+    = ASymbol String
     | ALambda [String] [AST]
     | ADefine String AST
     | AApp    AST  [AST]

### 2013-05-17 src/parse2.hs

@@ -80,9 +80,7 @@ commit err p = p <|> throwError err
 
 
 data AST
-    = ANumber Integer
-    | ASymbol String
-    | AString String
+    = ASymbol String
     | ALambda [String] [AST]
     | ADefine String AST
     | AApp    AST  [AST]

### 2013-05-17 article.tex
improve wording, grammar through end of first woof parser section

@@
 \section{Introducing Woof:  a Simple Lisp}
-The simple language we'll use as a motivation for building parsers is 
-Woof, a simple dialect of Lisp.  We'll be 
-progressively adding features to a simplistic initial implementation
-on our way to creating a practical, feature-rich and usable parser.
+The simple language we'll use as a motivation for building parsers is Woof, a 
+simple dialect of Lisp.  We'll be progressively adding features to a simple 
+initial implementation on our way to creating an error-reporting parser.
 
@@
 \subsection{Preliminaries}
-The first Woof parser we build will be responsible for determining whether input
-text conforms to the language definition and simultaneously building an 
-Abstract Syntax Tree representing of the structure of the recognized input.
+Our first Woof parser will be responsible for: 1) determining whether input
+text conforms to the language definition, and 2) building an 
+Abstract Syntax Tree representing the structure of the parsed input.
 Here's the AST definition we'll use:
@@
 data AST
-    = ANumber Integer
-    | ASymbol String
-    | AString String
+    = ASymbol String
     | ALambda [String] [AST]
@@
 These include the four braces, symbols, strings, numbers, whitespace, and comments.
-To recognize braces, we need to use the item and check combinators to build a new
- combinator, satisfy, that checks whether the next character meets a condition:
+To recognize braces, we need to use the \verb+item+ and \verb+check+ 
+combinators to build a new combinator, \verb+satisfy+, that checks whether 
+the next character meets a condition:
 \begin{verbatim}
@@
 
-And a the \verb+literal+ combinator:
+And here's the slightly modified \verb+literal+ combinator:
 \begin{verbatim}
@@
 by creating a \verb+munch+ combinator that takes in a parser, 
-throws away all leading junk, then runs the parser:
+throws away all leading junk, and finally runs the parser:
 \begin{verbatim}
@@
 
-The comment parser uses a new combinator, \verb+not1+, which is built out of a 
-new type class, Switch, which models computations which can be switched from
+The comment parser uses a new combinator, \verb+not1+, built out of a 
+type class, Switch, modeling computations which can be switched from
 successful to failing and vice versa:
@@
 into syntactic structures.
-The first combining form is a function application.  The rule says that it's delimited
+The first combining form is function application.  The rule says that it's delimited
 by matching parentheses, in between which must appear one form as the operator, followed
@@
 
-Note that, for now, we're doing all parser sequencing using the Applicative
+Note that all the parser sequencing is done using Applicative
 combinators -- we don't need to use the monadic ones.  Also, we used \verb+pure+ 
@@
 
-Next, we move on to \verb+define+, which is also a straightforward translation 
-from the grammar:
+Next, we move on to \verb+define+.
+A straightforward translation from the grammar produces:
 \begin{verbatim}
@@
 
-Lambda is more complicated -- not only does it have additional syntax with its parameter list,
-but we also need to ensure that the parameter names are unique.  We can do that using the 
-\verb+check+ combinator again:
+On to Lambda.  Not only does Lambda have additional syntax with its parameter 
+list, but we also need to ensure that the parameter names are unique.  We can 
+do that using the \verb+check+ combinator again:
 \begin{verbatim}
@@
 \end{verbatim}
-If the symbols are distinct, the parameter list subparser will succeed, and parsing will
-continue; if there's a repeated symbol, the parsing will stop with a failure.
+If the symbols are distinct, the parameter list subparser will succeed,
+whereas if there's a repeated symbol, it will fail.
 
 \subsection{Finishing touches}
-Now all we have left to do is to define the \verb+special+, \verb+form+ (which 
+All we have still to do is to define the \verb+special+, \verb+form+ (which 
 we've already used to build the previous parsers), and \verb+woof+ parsers.  
@@
 except that the types don't match -- \verb+symbol+'s type parameter is a String, 
-whereas the other two have ASTs, so we fix that by mapping the \verb+ASymbol+ 
+whereas the other two have ASTs.  We fix that by mapping the \verb+ASymbol+ 
 function over \verb+symbol+:
@@
 make sure that the entire input is consumed.  The end of input check is done 
-with the end parser:
+with the \verb+end+ parser:
 \begin{verbatim}
@@
 
-And we also need to munch any trailing comments and whitespace, so our final 
-woof parser is:
+We also need to munch any trailing comments and whitespace -- otherwise \verb+end+
+would fail -- so the final woof parser is:
 \begin{verbatim}
@@
 Note that the first two parses consume the entire input, while the last one 
-fails because it can't recognize the trailing comma.
+fails because it doesn't recognize the trailing comma.
 

### 2013-05-17 article.tex
work on grammar and wording of error-reporting section and end

@@
 \section{Example 2: error-reporting}
-While our first parser worked great when the input was valid, it wasn't helpful
-at all when the input was malformed.  We'd like it to produce an error message
-that tells us what the problem was and where it was encountered whenever bad
-input is encountered.  Otherwise, we'd waste time finding the problem by eye 
--- not a lot of fun.
+While our first parser worked great on valid input, it wasn't helpful
+when the input was malformed.  When the parser finds bad input, we'd like it to 
+produce an informative error, indicating what and where the problem was.  
+This allows a user to quickly find and correct problems.
 
-To solve this problem, let's start with an informal spec outlining the error
-messages that we expect to get, and the malformed input that causes each of them: 
+Let's start with an informal spec of different cases of malformed input, along 
+with the error messages that should be reported: 
 \begin{itemize}
@@
 \end{itemize}
-To do this, we're going to bring in a second monad transformer type class -- MonadError -- and
-another monad transformer datatype -- MaybeT.  Our parser stack and runParser function are now:
+To allow error reporting in our parsers, we first need to extend our monad stack, 
+adding in a layer for errors.  We'll use a second monad transformer type class 
+-- MonadError, replace the Maybe layer with its corresponding transformer datatype 
+MaybeT, and add in an Either layer on the bottom of the stack.  Our parser stack 
+and \verb+runParser+ function are now:
 \begin{verbatim}
@@
 
-A parser has three different possible results:  success, failure, and error:
+While our previous parsers had two different results -- Nothing and Just ... --
+our new parsers have three possible results:  success, failure, and error.  Note
+that successful and failing parses are wrapped in an additional \verb+Right+
+constructor:
 \begin{verbatim}
@@
 \end{verbatim}
+
 Although the \verb+<|>+ combinator can recover from failures, it can't recover from
@@
 Unfortunately, the standard MonadError instance for the \verb+Either+ datatype does
-not fit our needs -- it requires an \verb+Error+ constraint that we do not want.  To
-solve this problem, we'll implement our own MonadError type class, based on the
+not fit our needs -- it requires an \verb+Error+ constraint that we do not want.
+We'll implement our own MonadError type class, based on the
 standard one, and instances of it.  We'll need an instance for each member of
@@
 \subsection{Reporting position}
-To be able to report the position of the error, we'll pre-process the input
-string, calculating the line and column number of each character.  Then instead 
+To report error positionr, we'll pre-process the input
+string to calculate the line and column number of each character.  Then instead 
 of our parsers operating on a list of characters, they'll operate on a list of 
@@
 The cause of the problem is the \verb+literal+ combinator, so we
-replace it with one for token lists:
+replace it with a similar one for token lists:
 \begin{verbatim}
@@
 fails, generates an error whose contents are the first argument.  The 
-definition of 'failure' is provided by the Alternative instance of \verb+m+. So 
-what data do we want to include in our errors?  A String describing the error 
-and a token reporting the position should describe the error decently:
+definition of 'failure' is provided by the Alternative instance of \verb+m+.
+
+What data do we want to include in our errors?  A String describing the error 
+and a token reporting the position will describe the error decently:
 \begin{verbatim}
@@
 
-Now lets start with \verb+application+ as our first modified, error-reporting parser:
+Now let's start with \verb+application+ as our first modified, error-reporting 
+parser:
 \begin{verbatim}
@@
 in \verb+open+, and continuing with the parsing; if either \verb+form+ or 
-close-parenthesis fails, an appropriate error can be generated which includes 
-a string and the location of the open parenthesis.  However, if the open 
+close-parenthesis fails, an appropriate error is generated with a message and 
+the location of the open parenthesis.  However, if the open 
 parenthesis is not successfully parsed, it's a failure but not an error.  
@@
 with just applicative combinators, this version can't be:  \verb+open+ 
-is used to build the following parsers, making this decidedly monadic.
+is used to build the following parsers, making this monadic.
 
@@
 keyword, we wouldn't be able to backtrack and try the \verb+lambda+ parser if 
-that failed.  Thus, if treating errors as unrecoverable, it's important not to 
+that failed.  Thus, if errors are treated as unrecoverable, it's important not to 
 place a \verb+commit+ where backtracking might be necessary to parse a valid 
-alternative.  However, this isn't the case with how we've used it, since once 
+alternative.  However, in our example, once 
 we see an open curly brace and a define, we're sure that we're in the \verb+define+
-rule, and allowing backtracking would only destroy our ability to produce clean 
-and accurate error reports.
+rule, and allowing useless backtracking would only destroy our ability to report 
+errors cleanly and accurately.
 
@@
 First, we want an error reported when an open curly brace is found, but no 
-special form can be reported.  To accomplish this, we can extend the \verb+special+ 
-parser with a third alternative, which results in an error if it can parse
-an open curly brace, but in a backtracking failure if not:
+special form can be parsed.  We extend the \verb+special+ 
+parser with a third alternative, which throws an error if it can parse
+an open curly brace:
 \begin{verbatim}
@@
 
-When we can no longer parse any forms, instead of failing, we'd like to report an error if there's any input
-left, and otherwise succeed.  To implement this parser, we'll first munch any leading whitespace or comments, 
-as usual, and then we'll check the input:  if it's empty, great; if not, report an error at the location of the
-first remaining token, along with the appropriate error message that we defined earlier:
+When we can no longer parse any forms, instead of failing, we'd like to report 
+an error if there's any input left, and otherwise succeed.  Our ending parser
+must first munch any leading whitespace or comments, as usual, and then check 
+whether there are any tokens left.  If there are, it should report an error at 
+the location of the first remaining token:
 \begin{verbatim}
@@
 \end{verbatim}
-Now our \verb+woof+ parser uses this \verb+endCheck+ parser after it has consumed all the forms that it can:
+The \verb+woof+ parser uses \verb+endCheck+ after consuming all possible forms:
 \begin{verbatim}
@@
 \subsection{Error messages:  examples}
-While correct input will be parsed in the same way by our new parsers, incorrect input will now (hopefully) be
-accurately reported as such.  Successful results will be slightly different, however -- there is now an extra
-constructor wrapper around the value:
+While correct input is still parsed successfully:
 \begin{verbatim}
@@
 \end{verbatim}
-Here's the proof showing that the parser correctly reports all of the errors 
-covered by our spec:
+
+Incorrect input is reported as an error.  For each of the items in our error spec, 
+we can show that our parser correctly reports an error: 
 \begin{verbatim}
@@
 Not only are the contents of a transformer stack important, but the order of the
-layers as well \cite{stack}.  Let's revisit our first woof parser -- what if our stack had
-instead been Maybe/State instead of State/Maybe?
+layers as well \cite{stack}.  Let's revisit our first woof parser -- what if our 
+stack had been Maybe/State instead of State/Maybe?
 \begin{verbatim}
@@
 
-Another important point to note is that our parsers didn't actually specify
-which order -- Maybe/State or State/Maybe -- they prefer, and in fact they will 
-work with either order.  Since our parsers don't specify the stack order, the 
-semantics are ultimately determined by the actual stack we use.
+Another important point to note is that our parsers don't specify
+which order -- Maybe/State or State/Maybe -- they prefer, and will 
+work with either.  This means that the semantics are ultimately determined by 
+the actual stack we use.
 

### 2013-05-17 article.tex
rearrange code in parser level 0 writeup:  intersperse explanations between code chunks

@@
 Let's make a quick pass through a minimal parser combinator library so that we
-can see them in action before we move on to the hard stuff.  Without any 
-further ado, here's the code:
+can see them in action before we move on to the hard stuff.  There are three
+main pieces of the code:
+\begin{itemize}
+ \item a Parser datatype
+ \item parser-specific combinators
+ \item type class instances
+\end{itemize}
+
+\subsection{Parser datatype}
 \begin{verbatim}
-import Control.Applicative  (Applicative(..), Alternative(..))
-import Control.Monad        (liftM, ap)
-
 newtype Parser t a = 
     Parser {getParser :: [t] -> Maybe ([t], a)}
+\end{verbatim}
+This type definition says that a Parser is a function
+of a list of tokens, producing a new list of tokens and a result
+value, and possibly failing.
 
+\subsection{Parser-specific combinators}
+\begin{verbatim}
 item :: Parser t t
@@
     f (x:xs) = Just (xs, x)
+
+literal :: Eq t => t -> Parser t t
+literal x = afilter (== x) item
+\end{verbatim}
+Next is the \verb+item+ parser, the simplest parser we'll deal with.  All it does is 
+consume a single token if available, and fail otherwise.
+
+\subsection{Type class instances}
+For running parsers in sequence, we'll use the combinators from the Monad and
+Applicative type classes.  There's a little bit of book-keeping involved in the
+Monad instance -- it has to make sure that the correct token list is used by
+each parser -- but once we have a Monad instance, the Applicative and Functor
+instances can be handled by imports \verb+ap+ and \verb+liftM+ from Control.Monad.
+\begin{verbatim}
+import Control.Applicative  (Applicative(..), Alternative(..))
+import Control.Monad        (liftM, ap)
 
@@
   (<*>) = ap 
+\end{verbatim}
 
+For choosing between alternatives, the Alternative type class will do, so we 
+create an instance that is defined in terms of the Alternative instance of the
+underlying result type, \verb+Maybe+.
+\begin{verbatim}
 instance Alternative (Parser t) where
@@
     m >>= \x -> if p x then return x else empty
+\end{verbatim}
 
-literal :: Eq t => t -> Parser t t
-literal x = afilter (== x) item
-\end{verbatim}
-First, we define the Parser type, which says that a Parser is a function
-of a list of tokens, producing a new list of tokens and a result
-value, and possibly failing.
-Next is the \verb+item+ parser, the simplest parser we'll deal with.  All it does is 
-consume a single token if available, and fail otherwise.
-For running parsers in sequence, we'll use the combinators from the Monad and
-Applicative type classes.  There's a little bit of book-keeping involved in the
-Monad instance -- it has to make sure that the correct token list is used by
-each parser -- but once we have a Monad instance, the Applicative and Functor
-instances can be handled by imports \verb+ap+ and \verb+liftM+ from Control.Monad.
-For choosing between alternatives, the Alternative type class will do, so we create
-an instance that is defined partly in terms of the Alternative instance of the
-underlying result type, \verb+Maybe+.
-Last are a couple of convenience combinators for creating specific token parsers.
-
+\subsection{Examples}
 Now for some examples:  first, let's see \verb+item+ in action.  It fails on empty lists,
@@
 \end{verbatim}
-In the rest of this article, we'll be reworking these basic parser combinators.
-We'll be simplifying the implementations, extending the types of the results,
+In the rest of this article, we'll be reworking these basic parser combinators,
+by improving the implementations and extending the result types,
 all while taking advantage of monad transformers to keep it modular.

### 2013-05-18 article.tex
improve article section introducing parsers

@@
 Parsing is "the process of analysing a string of symbols according to the rules
-of a formal grammar" \cite{wikipedia_parsing}, with the general goal of building a tree
-representing the structure of the parsed input, upon which further operations
-such as interpretation or code generation are then performed.
+of a formal grammar" \cite{wikipedia_parsing}, with the general goal of building 
+a tree representing the structure of the parsed input, upon which further 
+operations such as interpretation or code generation are then performed.
 
@@
 
-\subsection{Parser-specific combinators}
-\begin{verbatim}
-item :: Parser t t
-item = Parser f
-  where 
-    f [] = Nothing
-    f (x:xs) = Just (xs, x)
-
-literal :: Eq t => t -> Parser t t
-literal x = afilter (== x) item
-\end{verbatim}
-Next is the \verb+item+ parser, the simplest parser we'll deal with.  All it does is 
-consume a single token if available, and fail otherwise.
-
 \subsection{Type class instances}
@@
 
-For choosing between alternatives, the Alternative type class will do, so we 
-create an instance that is defined in terms of the Alternative instance of the
+We use the Alternative type class for expressing choice; the Parser instance is  
+is defined in terms of the Alternative instance of the
 underlying result type, \verb+Maybe+.
@@
   Parser l <|> Parser r = Parser (\ts -> l ts <|> r ts)
+\end{verbatim}
 
--- similar to mfilter, but for Alternative instead of MonadPlus
+\subsection{Parser-specific combinators}
+\verb+item+ is a very simple parser; it 
+consumes a single token if available, and fails otherwise:
+\begin{verbatim}
+item :: Parser t t
+item = Parser f
+  where 
+    f [] = Nothing
+    f (x:xs) = Just (xs, x)
+\end{verbatim}
+
+Analagous to \verb+mfilter+ is the \verb+check+ combinator, which is used to
+validate a parse result:
+\begin{verbatim}
 check :: (Monad f, Alternative f) => (a -> Bool) -> f a -> f a
@@
     m >>= \x -> if p x then return x else empty
+\end{verbatim}
+
+Using \verb+check+, we can build a very convenient combinator to parse specific
+matching tokens: 
+\begin{verbatim}
+literal :: Eq t => t -> Parser t t
+literal x = check (== x) item
 \end{verbatim}

### 2013-05-18 article.tex
Jay is an author too
oops, remove double-@ in emails
extend intro parser:  nested () example, and also add explanation of it to article, and re-work explanation

@@
 \title{Error Reporting Parsers:  a Monad Transformer Approach}
-\author{Matt Fenwick\email{mfenwick100@@gmail.com}}
+\author{Matt Fenwick\email{mfenwick100@gmail.com}}
+\author{Jay Vyas\email{jayunit100@gmail.com}}
 
@@
 \section{Parser Combinators}
-Parser combinators are an excellent approach for building expressive parsers 
-that are easy to maintain.  In addition, they benefit from host-language
+Parser combinators are an excellent approach for building expressive, 
+declarative, and easy-to-maintain parsers.  They also benefit from host-language
 integration, allowing them to snarf features such as type systems and test 
@@
 There are many excellent papers on parser combinators, such
-as this \cite{wadler} classic Wadler paper on non-deterministic parser combinators, 
+as Wadler's classic paper \cite{wadler} on non-deterministic parser combinators, 
 Hutton's paper which covers factoring parsers into smaller pieces \cite{hutton}, 
@@
 
-Let's make a quick pass through a minimal parser combinator library so that we
-can see them in action before we move on to the hard stuff.  There are three
-main pieces of the code:
+To get a feel for what parser combinators are and how they're used to build
+parsers, let's make a quick pass through a minimal parser combinator library 
+so that we can see them in action before we move on to the hard stuff.  We'll 
+divide the library into three main pieces of the code:
 \begin{itemize}
@@
 Nothing
+\end{verbatim}
+Finally, let's parse arbitrarily deeply nested parentheses.  This example 
+demonstrates the use of \verb+fmap+ to apply a function to a parse result, 
+\verb+check+ to make sure that \verb+char+ doesn't consume parentheses,
+\verb+literal+ to match specific tokens exactly, \verb+many+ to introduce
+repetition, \verb+<|>+ to express choice, and \verb+*>+ 
+and \verb+<*+ to sequence parsers:
+\begin{verbatim}
+data Nesting
+    = One Char
+    | Many [Nesting]
+  deriving (Show, Eq)
+
+char :: Parser Char Nesting
+char = fmap One $ check (not . flip elem "()") item
+
+level :: Parser Char Nesting
+level = literal '(' *> (fmap Many $ many element) <* literal ')'
+
+element :: Parser Char Nesting
+element = char <|> level
+
+parseNest :: Parser Char [Nesting]
+parseNest = many element
+\end{verbatim}
+Examples:
+\begin{verbatim}
+*Main> getParser parseNest "(((()))))"
+Just (")",[Many [Many [Many [Many []]]]])
+
+*Main> getParser parseNest "(()abc(def)())zy"
+Just ("",[Many [Many [],One 'a',One 'b',One 'c',
+                Many [One 'd',One 'e',One 'f'],Many []],
+          One 'z',
+          One 'y'])
+
+*Main> getParser parseNest "(()abc(def)()"
+Just ("(()abc(def)()",[])
 \end{verbatim}

### 2013-05-18 src/parse0.hs

@@ -34,3 +34,21 @@ check p m =
 
 literal :: Eq t => t -> Parser t t
 literal x = check (== x) item
+
+
+data Nesting
+    = One Char
+    | Many [Nesting]
+  deriving (Show, Eq)
+  
+char :: Parser Char Nesting
+char = fmap One $ check (not . flip elem "()") item
+
+level :: Parser Char Nesting
+level = literal '(' *> (fmap Many $ many element) <* literal ')'
+
+element :: Parser Char Nesting
+element = char <|> level
+
+parseNest :: Parser Char [Nesting]
+parseNest = many element

### 2013-05-19 article.tex
parse 0 and monad transformer article sections: continue working on English of explanation, and put more imports by the code that uses them

@@
 
-Analagous to \verb+mfilter+ is the \verb+check+ combinator, which is used to
+Analagous to \verb+Control.Monad.mfilter+ is the \verb+check+ combinator, which is used to
 validate a parse result:
@@
 \end{verbatim}
-In the rest of this article, we'll be reworking these basic parser combinators,
-by improving the implementations and extending the result types,
-all while taking advantage of monad transformers to keep it modular.
+In the rest of this article, we'll rework and augment these basic parser 
+combinators by genericizing the implementations and extending the result types,
+while taking advantage of monad transformers to keep them modular.
 
@@
 Monad transformers \cite{liang} allow monads to be modularly combined to form 
-new, more complicated monads; the combined monads support all the operations of
-their constituents.  Doing just that -- combining multiple monads into a stack --
-we can build semantically rich yet modular parsers.
-We'll try to use the standard transformers/mtl \cite{mtl} libraries as much as possible, 
-since they're easy to obtain, easy to use, and quite effective in practice, but
-occasionally we'll have to make a slight modification or extensions.
+new, combined monads which support operations of each of their constituents.  
+There are several useful monads for parsing, and by applying monad transformers, 
+we can build semantically rich parsers without sacrificing modularity.
+We'll use the standard transformers/mtl \cite{mtl} libraries -- 
+they're easy to get, easy to use, and very useful.
 
@@
 \begin{verbatim}
+import Control.Monad.State       (StateT(..))
+
 type Parser t a = StateT [t] Maybe a
@@
 
-We can build the \verb+item+ parser using the combinators from MonadState --
+The \verb+item+ parser is reimplemented using the combinators from MonadState --
 token lists are the 'state' that \verb+item+ must inspect and modify:
@@
 
-import Control.Monad.State       (MonadState (..), StateT(..))
+import Control.Monad.State       (MonadState (..))
 
@@
 \end{verbatim}
-Note that we needed to use the FlexibleContexts language extension for the 
-MonadState constraints in the type signatures.  This formulation of monad
+Note that the FlexibleContexts language extension is required by the 
+MonadState constraint in the type signature.  This formulation of monad
 transformers relies heavily on extensions to the type system, more of which 

### 2013-05-19 article.tex
continue improving article by working on English, removing unnecessary words, etc.

@@
 \subsection{Token parsers}
-Now let's get down to the business of building parsers, starting with parsers
-for our most basic syntactic elements:  tokens.
+Our first parsers are for the most basic syntactic elements:  tokens.
 These include the four braces, symbols, strings, numbers, whitespace, and comments.
-To recognize braces, we need to use the \verb+item+ and \verb+check+ 
+To recognize braces, we use the \verb+item+ and \verb+check+ 
 combinators to build a new combinator, \verb+satisfy+, that checks whether 
@@
 
-And here's the slightly modified \verb+literal+ combinator:
+Because we used MonadState's combinators to build \verb+item+, \verb+literal+ 
+has a slightly different type; we also take advantage of \verb+satisfy+ to 
+simplify its implementation:
 \begin{verbatim}
@@
 by creating a \verb+munch+ combinator that takes in a parser, 
-throws away all leading junk, and finally runs the parser:
+throws away all leading junk, and finally runs the parser.  Also, we rename
+the \verb+some+ and \verb+many+ functions from Control.Alternative to more
+clearly indicate their semantics:
 \begin{verbatim}
+many0 = many
+many1 = some
+
 whitespace = many1 $ satisfy (flip elem " \n\t\r\f")
@@
 
-With \verb+munch+ in hand, we can now fix \verb+opencurly+ and try it out:
+With \verb+munch+ in hand, we fix \verb+opencurly+ and try it out:
 \begin{verbatim}
@@
 
-Excellent!  Now the \verb+opencurly+ parser correctly throws away whitespace 
+Excellent!  The \verb+opencurly+ parser correctly throws away whitespace 
 and comments and succeeds, consuming one character, when the first non-junk 
 character is an open curly brace, failing otherwise.
+
 It's straightforward to define parsers for the other three brace tokens:
@@
 
-The parser for our last token, Symbol, is slightly more complicated because it allows
-multiple characters and multiple lengths.  We'll use the satisfy combinator along with 
-a function that checks whether a character is one of the legal symbol characters:
+The parser for our last token, Symbol, is more complicated because any number of
+alphabetical characters is valid.  We'll use \verb+satisfy+ and an appropriate
+predicate to check that a character is valid, and \verb+many1+ to 
+match any positive number of valid characters:
 \begin{verbatim}
@@
 
-Note the use of \verb+many1+ to introduce repetition,
-and we're again using \verb+munch+ to throw away leading junk.
+Note that we're again using \verb+munch+ to throw away leading junk.
 
@@
 Whereas our token parsers dealt with the syntactic primitives of the Woof grammar, 
-the remaining parsers will have to implement grammar rules that combine the tokens
+the remaining parsers will implement grammar rules that combine the tokens
 into syntactic structures.
+
 The first combining form is function application.  The rule says that it's delimited
@@
 \subsection{Finishing touches}
-All we have still to do is to define the \verb+special+, \verb+form+ (which 
+All we have left to do are the \verb+special+, \verb+form+ (which 
 we've already used to build the previous parsers), and \verb+woof+ parsers.  

### 2013-05-21 gadt.tex
Minor edits to make it build
Signed-off-by: Edward Z. Yang <ezyang@mit.edu>

@@
 
-Generalized algebraic data types (GADTs) is a feature of functional programming languages that generalizes ordinary algebraic data types by permitting value constructors to return specific types.
+Generalized algebraic data types (GADTs) are a feature of functional programming languages which generalize ordinary algebraic data types by permitting value constructors to return specific types.
 This article is a tutorial about GADTs in Haskell programming language as they are implemented by the Glasgow Haskell Compiler (GHC) as a language extension.

### 2013-05-21 makeme.sh
Integrate with rest of build.
Signed-off-by: Edward Z. Yang <ezyang@mit.edu>

@@ -1,5 +1,0 @@
-pdflatex article
-bibtex article
-pdflatex article
-pdflatex article
-firefox article.pdf &

### 2013-05-26 article.tex
Fix quote
Signed-off-by: Edward Z. Yang <ezyang@mit.edu>

@@
 \begin{introduction}
-Parsing is "the process of analysing a string of symbols according to the rules
+Parsing is ``the process of analysing a string of symbols according to the rules
 of a formal grammar" \cite{wikipedia_parsing}, with the general goal of building 

### 2013-05-27 article.tex
Edits and comments
Signed-off-by: Edward Z. Yang <ezyang@mit.edu>

--- 2013-05-19 article.tex
@@
 \begin{introduction}
-Parsing is "the process of analysing a string of symbols according to the rules
-of a formal grammar" \cite{wikipedia_parsing}, with the general goal of building 
-a tree representing the structure of the parsed input, upon which further 
-operations such as interpretation or code generation are then performed.
 
-In addition to recognizing a stream of tokens and building a parse tree, a
-real-world parser must recognize faulty input, and accurately report the
-location and cause of the problem.  But how -- what generates a parse error,
-what information is included in the error, how do error-reporting parsers
-compose?
+\EZY{I rewrote the abstract, let me know if you like it.}
+
+The goal of parsing, in abstract, is to take a string and build a tree
+representing the structure of this input according to a formal grammar.
+However, real-world parsers must also recognize faulty input and
+accurately report the location and cause of the problem.  What generates
+a parse error, how is that information included in the error, and how do
+error-reporting parsers compose?
 
 Monad transformers provide a clean and simple answer to these questions.
-We'll use them to build a parser combinator library, then we'll use the library
-to build a simple parser, and finally we'll add in error reporting.
+We'll show how monad transformers can be used to build a parser combinator
+and then add error reporting in later in a modular way.
+
 \end{introduction}
@@
 \section{Parser Combinators}
+
+\EZY{It's probably worth rephrasing what is stated in the abstract here,
+for those of us who are in the habit of skipping abstracts; it also helps
+build continuity.}
+
 Parser combinators are an excellent approach for building expressive, 
@@
 and Leijen's Parsec paper \cite{leijen}.
-The inspiration and and much of the names used in this article are drawn 
+The inspiration and many of the names used in this article are drawn 
 from those references.
 
-To get a feel for what parser combinators are and how they're used to build
-parsers, let's make a quick pass through a minimal parser combinator library 
-so that we can see them in action before we move on to the hard stuff.  We'll 
-divide the library into three main pieces of the code:
+To get a feel for what parser combinators are and how they're used to
+build parsers, we'll quickly create a minimal parser combinator library to see
+the basic principles of their design before adding monad transformers to
+the mix.  We'll divide the library into three main pieces of the code:
+
 \begin{itemize}
@@
 \subsection{Parser datatype}
+\EZY{These sections are very terse.  If the reader is expected to already be familiar with
+most of the moving bits of a parsing library, perhaps this section can be boiled down some more.
+It is a very curious audience the current presentation targets: a reader who knows how to
+implement a parser library ``from scratch'', but doesn't know how to use
+monad transformers to build them.  Perhaps the components of this
+presentation which are key to understanding how the monad transformer works should be
+fleshed out more.}
+
 \begin{verbatim}
@@
 \end{verbatim}
-This type definition says that a Parser is a function
-of a list of tokens, producing a new list of tokens and a result
-value, and possibly failing.
+This type definition defines Parser to be a function
+on a list of tokens, producing a new list of tokens and a result
+value, possibly failing.
 
@@
 instances can be handled by imports \verb+ap+ and \verb+liftM+ from Control.Monad.
+\EZY{Fun fact: it is possible to create parsers which only have an applicative instance,
+and are more efficient but less expressive than the style you have here!}
 \begin{verbatim}
@@
 Woof         :=   Form(+)
-
 Form         :=   Symbol  |  Special  |  Application
-
 Symbol       :=   [a-zA-Z](+)
-
 Special      :=   '{'  ( Define  |  Lambda )  '}'
-
 Define       :=   'define'  Symbol  Form
-
 Lambda       :=   'lambda'  '{'  Symbol(*)  '}'  Form(+)
-
 Application  :=   '('  Form(+)  ')'
-
 Whitespace   :=   \s+
-
 Comment      :=   ';'  (not '\n')(*)

### 2013-05-27 article.tex
Comments
Signed-off-by: Edward Z. Yang <ezyang@mit.edu>

@@
 \verb+item+ is a very simple parser; it 
-consumes a single token if available, and fails otherwise:
+consumes a single token if available and fails otherwise:
 \begin{verbatim}
@@
 
-Analagous to \verb+Control.Monad.mfilter+ is the \verb+check+ combinator, which is used to
+The \verb+check+ combinator is analogous to \verb+Control.Monad.mfilter+, which is used to
 validate a parse result:
@@
 Parsers built out of \verb+literal+ only succeed when the next token matches 
-the given one:
+the specified one:
 \begin{verbatim}
@@
 In the rest of this article, we'll rework and augment these basic parser 
-combinators by genericizing the implementations and extending the result types,
+combinators by generalizing the implementations and extending the result types,
 while taking advantage of monad transformers to keep them modular.
@@
 \section{Monad transformers}
-Monad transformers \cite{liang} allow monads to be modularly combined to form 
-new, combined monads which support operations of each of their constituents.  
-There are several useful monads for parsing, and by applying monad transformers, 
-we can build semantically rich parsers without sacrificing modularity.
+Monad transformers~\cite{liang} provide a method of combining monads into
+new monads which support operations of each of their constituents.  
+By applying monad transformers, we can build semantically rich parsers without sacrificing modularity.
 We'll use the standard transformers/mtl \cite{mtl} libraries -- 
-they're easy to get, easy to use, and very useful.
+they're easy to install, easy to use, and very useful.
 
 First, note that the parser type we used earlier -- \verb+[t] -> Maybe ([t], a)+
--- can be built out of the StateT monad transformer applied to Maybe.  We can 
+-- can be built out of the StateT monad transformer applied to Maybe.  \EZY{Do you maybe want to say why this is the case?}  We can 
 now define the new parser type as:
@@
 The \verb+item+ parser is reimplemented using the combinators from MonadState --
-token lists are the 'state' that \verb+item+ must inspect and modify:
+token lists are the `state' that \verb+item+ must inspect and modify:
 \begin{verbatim}
@@
 Note that the FlexibleContexts language extension is required by the 
-MonadState constraint in the type signature.  This formulation of monad
+MonadState constraint in the type signature.  This \EZY{which?} formulation of monad
 transformers relies heavily on extensions to the type system, more of which 
@@
 \section{Introducing Woof:  a Simple Lisp}
+
+\EZY{An alternate structure you can use here is to introduce Woof first, before talking about monad transformers.
+After all, the example parser doesn't require you to know what is going on in transformer-land.}
+
 The simple language we'll use as a motivation for building parsers is Woof, a 
@@
 \subsection{Preliminaries}
-Our first Woof parser will be responsible for: 1) determining whether input
-text conforms to the language definition, and 2) building an 
-Abstract Syntax Tree representing the structure of the parsed input.
+Our first Woof parser will be responsible for determining whether input
+text conforms to the language definition and building an 
+Abstract Syntax Tree (AST) representing the structure of the parsed input.
 Here's the AST definition we'll use:
@@
 
-And some examples to make sure it's behaving correctly:
+And some examples to make sure it's behaving correctly: \EZY{An example?! Why not a QuickCheck property? \verb|;-)|}
 \begin{verbatim}
@@
 the \verb+some+ and \verb+many+ functions from Control.Alternative to more
-clearly indicate their semantics:
+clearly indicate their semantics: \EZY{I would suspect many compiler authors would be of the opinion that the proper way to remove comments and whitespace is to lex first, and parse later.}
 \begin{verbatim}
@@
 
-On to Lambda.  Not only does Lambda have additional syntax with its parameter 
+On to Lambda!  Not only does Lambda have additional syntax with its parameter 
 list, but we also need to ensure that the parameter names are unique.  We can 
@@
 We also need to munch any trailing comments and whitespace -- otherwise \verb+end+
-would fail -- so the final woof parser is:
+would fail -- so the final Woof parser is:
 \begin{verbatim}
@@
 
-While our previous parsers had two different results -- Nothing and Just ... --
-our new parsers have three possible results:  success, failure, and error.  Note
+While our previous parsers had two different results -- Nothing and Just --
+our new parsers have three possible results:  success, failure, and error.
+\EZY{Question a reader might have: what is the difference between failure and error?
+You give a little guidance later but give some intuition now!}  Note
 that successful and failing parses are wrapped in an additional \verb+Right+
@@
 \subsection{Reporting position}
-To report error positionr, we'll pre-process the input
-string to calculate the line and column number of each character.  Then instead 
+To report error position, we'll pre-process the input
+string to calculate the line and column number of each character.  Then, instead 
 of our parsers operating on a list of characters, they'll operate on a list of 
@@
 number, which we'll express with a simple typedef and some accessor functions:
+\EZY{I don't think this is how most real parser libraries do it.  Why not do it
+with some extra state on the parser monad; it's yet another application of the
+transformer approach?}
 \begin{verbatim}
@@
 \subsection{Token parsers}
+\EZY{In contrast, this section doesn't seem very enlightening, so maybe
+it could be glossed over or relegated to an appendix.}
 By adding the position to the input, we've created a new problem 
@@
 
-And now our token parsers are:
+Now our token parsers are:
 \begin{verbatim}
@@
 \subsection{Error messages}
-Now we modify the remaining parsers to report accurate and useful errors.
+Now, we modify the remaining parsers to report accurate and useful errors.
 Here are the error messages that we'll use to report each error:
@@
 
-The MonadError type class provides two useful combinators, \verb+throwError+ for 
-generating errors, and \verb+catchError+ for dealing with them.  We can ignore 
+The MonadError type class provides two useful combinators: \verb+throwError+ for 
+generating errors and \verb+catchError+ for dealing with them.  We can ignore 
 the second because we don't need to recover from errors, just report them, 
@@
 
-This combinator takes two arguments: 1. an error value, and 2. a monadic 
-computation.  It first tries to run the computation, and if the computation 
-fails, generates an error whose contents are the first argument.  The 
+This combinator takes two arguments: an error value and a monadic 
+computation.  It first tries to run the computation; if the computation 
+fails, it generates an error.  The 
 definition of 'failure' is provided by the Alternative instance of \verb+m+.
+\EZY{Give a little intuition on why the operator is called commit earlier;
+you don't get around it until later.}
 
 What data do we want to include in our errors?  A String describing the error 
-and a token reporting the position will describe the error decently:
+and a token reporting the position will do:
 \begin{verbatim}
@@
 Note that \verb+commit+ isn't used until we can be sure that we're in the right 
-rule:  if, in the \verb+define+ parser, we committed to parsing the "define" 
+rule: \EZY{Here it is}  if, in the \verb+define+ parser, we committed to parsing the ``define" 
 keyword, we wouldn't be able to backtrack and try the \verb+lambda+ parser if 
@@
 place a \verb+commit+ where backtracking might be necessary to parse a valid 
-alternative.  However, in our example, once 
+alternative.  In our example, however, once 
 we see an open curly brace and a define, we're sure that we're in the \verb+define+
-rule, and allowing useless backtracking would only destroy our ability to report 
+rule; allowing useless backtracking would only destroy our ability to report 
 errors cleanly and accurately.
@@
 
-Incorrect input is reported as an error.  For each of the items in our error spec, 
+Incorrect input results in an error.  For each of the items in our error spec, 
 we can show that our parser correctly reports an error: 
@@
 
+\Red{It's worth noting that up until this point, we haven't talked about how transformers contribute to the design of the parser library.  Maybe give the juicy bits earlier.}
+
 \section{Stack Order}
-Not only are the contents of a transformer stack important, but the order of the
-layers as well \cite{stack}.  Let's revisit our first woof parser -- what if our 
+While the contents of a transformer stack is important, the order of the
+layers is important as well~\cite{stack}.  Let's revisit our first Woof parser -- what if our 
 stack had been Maybe/State instead of State/Maybe?
@@
 important.  But don't just take my word for it -- try it for yourself!
+\EZY{It's worth mentioning why, formally speaking, this is the case, which has
+to do with the relevant structures being isomorphic to each other.}
 
@@
 work with either.  This means that the semantics are ultimately determined by 
-the actual stack we use.
+the actual monad stack we use.
 
@@
 valuable information.  And the best part of it is, most of the parsers we already
-wrote would continue to work fine without needing any modifications.  We'd just
-have to set up our transformer stack and modify the relevant parsers!
+wrote would continue to work fine without needing any modifications.  That
+is the power of monad transformers!
 

### 2013-05-27 gadt.tex
Comments
Signed-off-by: Edward Z. Yang <ezyang@mit.edu>

@@
 The single idea of GADTs is to allow specific return types of value constructors.
-In this way they generalize ordinary algebraic data types.
+\EZY{I think this explanation doesn't mean anything to someone who doesn't already know what GADTs are.} In this way they generalize ordinary algebraic data types.
 The theoretical foundation for GADTs is the notion of dependent types which is extensively described in literature on computer science and logic~\cite{pierce_book}.
@@
 
+\EZY{I think I'd add something like, ``While creating a values with the TI results in a value of type Test Int, the converse holds too: when a pattern-match succeeds on the TI constructor, we can now conclude that a is Int; that is to say, pattern matching causes type refinement.''}
 The key feature of GADTs is that pattern matching causes type refinement. In the right-hand side of the following equation the type of \verb|a| is refined to \verb|Int|:
@@
 
-Examples in the following sections show the real practical value of GADTs.
+Examples in the following sections show some practical applications of GADTs.
 
@@
 
-This section introduces GADTs with a canonical example of expression evaluator.
+This section introduces GADTs with the canonical example of an expression evaluator.
 At first, we attempt to implement it using ordinary algebraic data types.
@@
 The type \verb|ExtExpr| is not parametrized by return value type, so we have to use type \verb|Either Int Bool|.
-Also, evaluation will fail if the input expression is incorrect, so we have to use type \verb|Maybe|.
-Finally, type signature is the following:
+Furthermore, evaluation will fail if the input expression is incorrect, so we have to use type \verb|Maybe|.
+Finally, the type signature of \verb|evaluate| is the following:
 
@@
 
-And implementation of this function is complicated. For example, processing \verb|AddInt| requires usage of a nested \verb|case|:
+Implementing this function is complicated. For example, processing \verb|AddInt| requires usage of a nested \verb|case|:
 
@@
 
-Ideally we want the following type signature for \verb|evaluate| method:
+Ideally, we want the following type signature for \verb|evaluate| method:
 
@@
 
-Using GADT syntax the data type \verb|PhantomExpr| type can be declared this way:
+Using GADT syntax, the data type \verb|PhantomExpr| type can be declared this way:
 
@@
 As noted in the previous section, the distinctive feature of GADTs is the ability to return specific types in value constructors, for example \verb|PhantomExpr Int|.
-GADT for the expression language looks this way:
+The expression language can be written with GADTs in this fashion:
 
@@
 
-Pattern matching causes type refinement, so for example in the right-hand side of the following expression \verb|i| has type \verb|Int|:
+Pattern matching causes type refinement: the right-hand side of the following expression \verb|i| has type \verb|Int|:
 
@@
 
-At the end of this article we describe one more implementation of expression evaluator.
+At the end of this article, we describe one more implementation of expression evaluator.
 
@@
 
-In datatype-generic programming functions take a type as argument and their behavior depends on the structure of this type.
+In datatype-generic programming, functions take a type as an argument and change behavior depending on the structure of this type.
 There are several approaches to such kind of generic programming in Haskell.
-Paper by Hinze et al.~\cite{generic_programming} provides an overview of these approaches.
-In this section we present an approach which uses GADTs.
+A paper by Hinze et al.~\cite{generic_programming} provides an overview of these approaches.
+In this section, we present an approach which uses GADTs.
 Ideas for this section were taken from another paper by Hinze at al.~\cite{fun_with_phantom_types}.
@@
 Functions like this one can be implemented using type classes.
-But GADTs offer an interesting alternative.
+However, GADTs offer an interesting alternative.
 
@@
 \begin{Verbatim}
-ghci> let d = [Dyn TInt 10, Dyn tString "test"]
+ghci> let d = [Dyn TInt 10, Dyn TString "test"]
 ghci> :t d
@@
 
-However, we can't make this list a \verb|Dynamic| value itself. To fix this problem, we need to extend the representation type: add a value constructor for the \verb|Dynamic| data type.
+However, we can't make this list a \verb|Dynamic| value itself. To fix this problem, we need to extend the representation type, adding a value constructor for the \verb|Dynamic| data type.
 
@@
 The \verb|Dynamic| data type is useful for communication with the environment when the actual type of the data is not known in advance.
-In this case it is required to use a type cast to get the useful data.
+In this case a type cast is required to get useful data.
 A simple way to implement a type cast from \verb|Dynamic| data type to an integer is the following:
@@
 
-There is a more generic solution~\cite{fun_with_phantom_types} of this problem that works for all types, not just integer, but it is out of scope for this article.
+There is a more generic solution~\cite{fun_with_phantom_types} for this problem that works for all types, not just integer, but it is out of scope for this article.
 
@@
 
-An important role of type systems is to ensure that data is manipulated in appropriate ways (for example, to make sure that we pass a list to \verb|head| function).
+An important role of type systems is to ensure that data is manipulated in appropriate ways (for example, to ensure we pass a list to \verb|head| function).
 But types can be used to express more sophisticated properties.
-For example, we can define a type of lists of a particular length and then define \verb|headSafe| function that only accepts non-empty lists.
+For example, we can define a type of lists of a particular length and then define a \verb|headSafe| function that only accepts non-empty lists.
 The idea described in this section is to use types to express correctness properties and then use type checker of the programming language to ensure that we can express only those programs that have the desired properties.
 This idea was developed in $\Omega$mega system~\cite{curry_howard,omega}.
-In this section we GADTs in Haskell programming language to prove correctness of list operations and the next section describes proving correctness of insertion operation in red-black trees.
+In this section, we use GADTs in the Haskell programming language to prove the correctness of list operations; in the next section, we describe how to use GADTs to prove correctness of the insertion operation in red-black trees.
 
@@
 
-Actually, with the current data type definition a term \verb|Cons 'a' Nil| can even be given type \verb|SafeList Char Int|.
-To fix this problem it is required to give the same kind for \verb|Empty| and \verb|NonEmpty| types.
+It's worth noting that with the current data type definition, a term \verb|Cons 'a' Nil| can even be given the type \verb|SafeList Char Int|.
+To fix this problem, we need to somehow give the \verb|Empty| and \verb|NonEmpty| types the same kind.
 This is discussed later for \verb|Nat| data type.
 
-But \verb|SafeList| data type does not have enough static information to prove list length invariants for list functions.
+Regardless, \verb|SafeList| data type does not have enough static information to prove stronger list length invariants for many list functions.
 For example, for the concatenation function we need to show that length of the concatenated list is a sum of source lists lengths.
-So it is not enough to just know if a list is empty or not.
-We need to encode the length of a list in its type.
+It is not enough to just know if a list is empty or not: we need to encode the length of a list in its type.
 
@@
 
-Zero is encoded as \verb|Zero|, one -- as \verb|Succ Zero|, two -- as \verb|Succ (Succ Zero)| and so on. Now list data type is defined as:
+Zero is encoded as \verb|Zero|, one -- as \verb|Succ Zero|, two -- as \verb|Succ (Succ Zero)| and so on. Now the list data type is defined as:
 
@@
 
-To implement the concatenation function we need a type-level function for addition of Peano numbers.
+To implement the concatenation function, we need a type-level function for addition of Peano numbers.
 A natural way to implement such function is to use \italic{type families} (which in context of this tutorial can be understood as type-level functions).
-First we need to declare type family \verb|Plus| (this requires \verb|TypeFamilies| extension):
+First, we need to declare type family \verb|Plus| (this requires \verb|TypeFamilies| extension):
 
@@
 
-Then we need to declare type instances that implement addition of Peano numbers by induction:
+Then, we need to declare type instances that implement addition of Peano numbers by induction:
 
@@
 
-In fact, at the moment \verb|Succ| has type parameter of kind \verb|*|, so it possible to write nonsensical terms like \verb|Succ Int|, and they will be accepted by the type checker.
+As mentioned previously, \verb|Succ| has a type parameter of kind \verb|*|, so it possible to write nonsensical terms like \verb|Succ Int| which will be accepted by the type checker.
 This problem can be addressed using a new \italic{kind}.
@@
 For example, \verb|'Succ| refers to a type, not a value constructor.
-So, type-level representation of Peano number 2 can be written as:
+So, type-level representation of the number two can be written as:
 
@@
 On the other hand, we need a type-level representation of the same number for \verb|List| type.
-Haskell enforces a phase separation between run-time values and compile-time types.
+Haskell enforces a phase separation between run-time values and compile-time types \EZY{so there is a problem.
+Please state what the problem is.}.
 
@@
 This requires a type-level function to compute whether one number is less than the other.
-We define the following type family and instances (\verb|TypeOperators| extension is required to be able to define \verb|:<| operation for types):
+We define the following type family and instances (the \verb|TypeOperators| extension is required to be able to define \verb|:<| operation for types):
 
@@
 
+\EZY{Yes, please write a conclusion here!}
+
 %The moral is that to verify a function it is required to carefully design data type.
@@
 The source code for the verified red-black tree was originally written by Stephanie Weirich~\cite{weirich_dependently_typed} for a university course.
+
+\EZY{Before you dive into all of the hoary details of the construction, it is well worth motivating the example. What's the point? Answer this question!}
 
@@
 
-Now we see that color must be passed as a type (for \verb|IncBlackHeight| type family) and as a value (to the value constructor). Similarly as before, we need to use a singleton type as a bridge:
+Now we see that a color must be passed as a type (for \verb|IncBlackHeight| type family) and as a value (to the value constructor). Similarly as before, we need to use a singleton type as a bridge:
 
@@
 
-Because normally when creating a new type in Haskell, every type variable that appears on the right-hand side of the definition must also appear on its left-hand side.
+since normally when creating a new type in Haskell, every type variable that appears on the right-hand side of the definition must also appear on its left-hand side.
 One solution to this problem is usage of \italic{existential types} (this definition requires extension \verb|RankNTypes|):
@@
 
-It is also possible to do this with GADT:
+It is also possible to do this with a GADT:
 
@@
 Proving the 3rd invariant is more involved.
-First we need to specify valid colors for a node on the type level.
+First, we need to specify valid colors for a node on the type level.
 This can be done using type families as before or using type classes.
@@
 The implementation of the insertion operation can temporarily invalidate the 3rd invariant (see \prettyref{fig:balancing}), so during insertion we are not able to represent the tree using this data type.
-Thus it is required to declare a data type similar to \verb|Node|, but without the restriction on node colors:
+Thus, we need to declare a data type similar to \verb|Node|, but without the restriction on node colors:
 
@@
 
-As before we need to make changes in type annotations of the functions implementing insert operation. We also need to change the \verb|leftBalance| function type signature this way:
+As before, we need to make changes in type annotations of the functions implementing insert operation. We also need to change the \verb|leftBalance| function type signature this way:
 
@@
 
-Earlier we passed the whole node as a parameter.
-But we can't do it after the \verb|Node| data type was modified: the 3rd invariant could be violated due to insertion in the left branch of the node.
+Earlier, we passed the whole node as a parameter.
+But we can't do this after the \verb|Node| data type was modified: the 3rd invariant could be violated due to insertion in the left branch of the node.
 So, we pass all parameters of the parameters of the node and left child is represented using \verb|IntNode| data type.
@@
 
+\EZY{Conclusion???}
+
 \section{Type Signatures for Functions Involving GADTs}
@@
 
-Hindley-Milner (HM) is a classical type inference method~\cite{cardelli_typechecking}.
+\EZY{This section is something of a digression from the litany of examples that have been parading down this article; so it's probably worth pointing it out and mentioning why this is an important thing to know about.}
+
+Hindley-Milner (HM) is the classic type inference method~\cite{cardelli_typechecking}.
 One of the most important properties of HM is ability to always deduce the \italic{most general type} (\italic{principle type}) of every term.
@@
 Yampa~\cite{yampa} is a domain-specific language for functional reactive programming (FRP).
-FRP is a programming paradigm of expressing data flows using the building blocks of functional programming.
-Based on the information from the paper by Nilsson~\cite{yampa_gadts}, this section describes how GADTs were used to improve performance of Yampa programs.
-
-Signal function is a central abstraction in Yampa.
+FRP is a programming paradigm which involves expressing data flows using the building blocks of functional programming.
+\EZY{What are the building blocks of functional programming?} Based on the information from the paper by Nilsson~\cite{yampa_gadts}, this section describes how GADTs were used to improve performance of Yampa programs.
+
+The signal function is a central abstraction in Yampa.
 It represents a simple synchronous process that maps an input signal to an output signal (see \prettyref{fig:signal_function}).
@@
 There are other performance improvements that are enabled by GADTs in Yampa~\cite{yampa_gadts}.
-The results of performance improvements are shown in the table.
-It was taken from the paper by Nilsson~\cite{yampa_gadts}.
+The results of performance improvements described by the paper by Nilsson~\cite{yampa_gadts} are shown in the table.
 The table shows execution time of several benchmarks using initial simply-optimized implementation ($T_S$) and the implementation with GADT-based optimizations ($T_G$).
-In addition to performance improvements, GADTs allowed to write a more concise and cleaner API without the need of pre-composed signal function (that were defined in Yampa only for performance reasons).
+In addition to performance improvements, GADTs allowed the authors of the paper to write a more concise and cleaner API without the need of pre-composed signal functions (which were defined in Yampa only for performance reasons).
 
@@
 \end{Verbatim}
+
+\EZY{What are the limitations of finally tagless in this context?}
 
@@
 As a result, we have shown that the notion of GADTs is a very valuable extension of the Haskell language.
+
+\EZY{I think you can write a better concluding sentence then that!}
 

### 2013-05-27 article.tex
addressed easy problems, still working on hard ones

@@
 \EZY{I rewrote the abstract, let me know if you like it.}
+\Matt{LGTM, thanks!}
 
@@
 First, note that the parser type we used earlier -- \verb+[t] -> Maybe ([t], a)+
--- can be built out of the StateT monad transformer applied to Maybe.  \EZY{Do you maybe want to say why this is the case?}  We can 
-now define the new parser type as:
+-- can be built out of the StateT monad transformer applied to Maybe.
+\EZY{Do you maybe want to say why this is the case?}
+From the implementation:
+\begin{verbatim}
+newtype StateT s m a = StateT { runStateT :: s -> m (a,s) }
+\end{verbatim}
+Substituting in [s] for s and Maybe for m in the right-hand side produces
+(note that the type variable m is no longer needed):
+\begin{verbatim}
+newtype StateT s a = StateT { runStateT :: [s] -> Maybe (a,[s]) }
+\end{verbatim}
+Which is identical to the Parser type given before, except for the order 
+of the result tuple.
+This allows us to define the new parser type as:
 \begin{verbatim}
@@
 Note that the FlexibleContexts language extension is required by the 
-MonadState constraint in the type signature.  This \EZY{which?} formulation of monad
+MonadState constraint in the type signature.  The mtl \EZY{which?} \Matt{I'm not sure what the library's called -- is it mtl or transformers?} formulation of monad
 transformers relies heavily on extensions to the type system, more of which 
@@
 clearly indicate their semantics: \EZY{I would suspect many compiler authors would be of the opinion that the proper way to remove comments and whitespace is to lex first, and parse later.}
+\Matt{That's a good point, I toyed with doing it that way but it was just too much trouble for no benefit that I could see.  It ended up being a lot more code.
+But this approach is equivalent, since the token parsers get the first crack at the input string, and the non-token parsers don't touch the input string directly.}
 \begin{verbatim}
@@
 our new parsers have three possible results:  success, failure, and error. 
-\EZY{Question a reader might have: what is the difference between failure and error?
-You give a little guidance later but give some intuition now!}  Note
-that successful and failing parses are wrapped in an additional \verb+Right+
+A parse failure means that a parser wasn't able to match the input, but
+nothing bad happened; failures can be recovered from if there is an alternative
+parse that succeeds.  However, an error means that an unrecoverable problem
+has been found and that parsing should immediately stop; the parser will
+not be able to recover, even if there are alternatives since backtracking
+will not be performed.
+\EZY{Question a reader might have: what is the difference between failure and error?  You give a little guidance later but give some intuition now!}
+Note that successful and failing parses are wrapped in an additional \verb+Right+
 constructor:
@@
 transformer approach?}
+\Matt{Good point, I thought about doing it this way but decided not to because
+the article is already long and this would add a lot, and the other approach 
+seems to work fine and is simple.
+But I'm not opposed to it at all.  Do you think I should switch it?}
 \begin{verbatim}
@@
 you don't get around it until later.}
+The reason for its name is that it restricts backtracking and alternatives;
+it ``commits" the parser to a do-or-die situation -- either it successfully
+parses the input, or the parsing stops.
 
@@
 to do with the relevant structures being isomorphic to each other.}
+\Matt{Uh-oh!  I don't really know anything formal about monad transformers,
+I really have no idea how to go about showing something like this!
+I guess I'll have to rethink my goals for this section.}
 

### 2013-05-27 article.tex
forgot reference to other Hutton parsing paper
Responses.
Signed-off-by: Edward Z. Yang <ezyang@mit.edu>

@@
 \begin{introduction}
-
-\EZY{I rewrote the abstract, let me know if you like it.}
-\Matt{LGTM, thanks!}
 
@@
 Note that the FlexibleContexts language extension is required by the 
-MonadState constraint in the type signature.  The mtl \EZY{which?} \Matt{I'm not sure what the library's called -- is it mtl or transformers?} formulation of monad
+MonadState constraint in the type signature.
+The mtl \EZY{which?} \Matt{I'm not sure what the library's called -- is it mtl or transformers?} \EZY{Well, both export the necessary interfaces.
+Just pick one and be consistent.} formulation of monad
 transformers relies heavily on extensions to the type system, more of which 
@@
 the \verb+some+ and \verb+many+ functions from Control.Alternative to more
-clearly indicate their semantics: \EZY{I would suspect many compiler authors would be of the opinion that the proper way to remove comments and whitespace is to lex first, and parse later.}
-\Matt{That's a good point, I toyed with doing it that way but it was just too much trouble for no benefit that I could see.  It ended up being a lot more code.
-But this approach is equivalent, since the token parsers get the first crack at the input string, and the non-token parsers don't touch the input string directly.}
+clearly indicate their semantics:
 \begin{verbatim}
@@
 it ``commits" the parser to a do-or-die situation -- either it successfully
-parses the input, or the parsing stops.
+parses the input, or the parsing stops.  It's based on the \verb+nofail+
+combinator of \cite{hutton1992higher}.
 
@@
 I guess I'll have to rethink my goals for this section.}
+\EZY{It's pretty simple, and just has to do with the fact that Maybe (Either a b) is
+equivalent (modulo laziness) with Either a (Maybe b).  To show this is a case, you have
+to define functions which convert from one to the other, and then show that the roundtrip
+is identity.}
 

### 2013-06-01 gadt.tex
Addressing comments from Edward Z. Yang

@@
 The study of type systems is an active research area and Haskell is considered to be a kind of laboratory in which type-system extensions are designed, implemented and applied~\cite{haskell_history}.
-
 This article is a tutorial about one such extension -- generalized algebraic data types (GADTs).
-The single idea of GADTs is to allow specific return types of value constructors.
-\EZY{I think this explanation doesn't mean anything to someone who doesn't already know what GADTs are.} In this way they generalize ordinary algebraic data types.
 The theoretical foundation for GADTs is the notion of dependent types which is extensively described in literature on computer science and logic~\cite{pierce_book}.
@@
 
-\EZY{I think I'd add something like, ``While creating a values with the TI results in a value of type Test Int, the converse holds too: when a pattern-match succeeds on the TI constructor, we can now conclude that a is Int; that is to say, pattern matching causes type refinement.''}
-The key feature of GADTs is that pattern matching causes type refinement. In the right-hand side of the following equation the type of \verb|a| is refined to \verb|Int|:
+While creating a values with the \verb|TI| results in a value of type \verb|Test Int|, the converse holds too: when a pattern-match succeeds on the \verb|TI| constructor in the example below, we can now conclude that the type of \verb|a| is \verb|Int|.
+That is to say, pattern matching causes type refinement, which is the key feature of GADTs.
 
@@
 
-% TODO describe other ways to implement Plus?
-
 As mentioned previously, \verb|Succ| has a type parameter of kind \verb|*|, so it possible to write nonsensical terms like \verb|Succ Int| which will be accepted by the type checker.
@@
 On the other hand, we need a type-level representation of the same number for \verb|List| type.
-Haskell enforces a phase separation between run-time values and compile-time types \EZY{so there is a problem.
-Please state what the problem is.}.
+Haskell enforces a phase separation between run-time values and compile-time types.
+This causes a problem with type signature of function \verb|repeatElem|: we can not specify type-level representation of list elements count in the return value type of the function matching the count passed as a value.
 
@@
 
-\EZY{Yes, please write a conclusion here!}
-
-%The moral is that to verify a function it is required to carefully design data type.
+We have shown that GADTs provide a way to use Haskell type checker to verify correctness of list operations.
+To do this, we need to specify necessary properties in the data type.
+The set of such properties is motivated by the operations that we want to implement.
+At first we only made a distinction between empty and non-empty lists.
+This was suitable for the \verb|listHead| function.
+To implement \verb|repeatElem|, \verb|concatenate| and other functions we extended the list data type with the count of elements that it contains.
 
@@
 
-A red-black tree is a binary search tree where every node has either red or black color.
-We use the implementation described by Okasaki~\cite{okasaki_red_black_trees}.
+This section describes another, more involved example of using GADTs to verify correctness of programs: proving correctness of insertion in red-black trees.
+A red-black tree is a binary search tree where every node has either red or black color and which additionally satisfies several invariants which we describe later.
+The presense of invariants guarantees that the tree is balanced and thus searching takes $O (\log n)$ time (where $n$ is total number of elements).
+The insertion operation must maintain these properties, so it is interesting to see how GADTs can be used to enforse these invariants.
+
+We use the implementation of red-black trees described by Okasaki~\cite{okasaki_red_black_trees}.
 The source code for the verified red-black tree was originally written by Stephanie Weirich~\cite{weirich_dependently_typed} for a university course.
-
-\EZY{Before you dive into all of the hoary details of the construction, it is well worth motivating the example. What's the point? Answer this question!}
 
@@
 \item The root is black.
-\item Every leaf is black. % (this is enforced by the tree data type).
+\item Every leaf is black.
 \item Red nodes have black children.
@@
 
-% excluding current node?
-
 These invariants guarantee that tree is balanced.
 Indeed, the longest path from the root node (containing alternating red-black nodes) can only be twice as long as the shortest path (containing only red nodes).
-Thus basic operations (such as insertion) take $O(\log n)$ time in the worst case.
+Thus basic operations (such as insertion and search) take $O(\log n)$ time in the worst case.
 
@@
 
-The previous code illustrates a general problem with proofs. In fact, in Haskell $\bot$ (bottom) is a member of every type. As a result, we can write:
+We have shown that insertion operation maintains all invariants in red-black trees by adding necessary properties to the tree data type and then modifying the functions accordingly.
+However, the previous code illustrates a general problem with proofs.
+In fact, in Haskell $\bot$ (bottom) is a member of every type.
+As a result, we can write:
 
@@
 
-Of course, the implementation of the \verb|concatenate| function does not meet our expectations. But this code still type checks.
-
-\EZY{Conclusion???}
+This code type checks, but, of course, the implementation of the \verb|concatenate| function does not meet our expectations.
 
@@
 
-\EZY{This section is something of a digression from the litany of examples that have been parading down this article; so it's probably worth pointing it out and mentioning why this is an important thing to know about.}
+Type inference for programs with GADTs is out of scope for this article.
+But when writing programs we are faced with the problem: typically Haskell functions do not require writing type signatures, but many functions involving GADTs do not compile without specifying type signatures explicitly, because Haskell compiler can not automatically infer the type.
+This section describes the reasons of this situation.
 
@@
 Yampa~\cite{yampa} is a domain-specific language for functional reactive programming (FRP).
-FRP is a programming paradigm which involves expressing data flows using the building blocks of functional programming.
-\EZY{What are the building blocks of functional programming?} Based on the information from the paper by Nilsson~\cite{yampa_gadts}, this section describes how GADTs were used to improve performance of Yampa programs.
+FRP is a programming paradigm which involves expressing data flows using functional programming languages.
+Based on the information from the paper by Nilsson~\cite{yampa_gadts}, this section describes how GADTs were used to improve performance of Yampa programs.
 
@@
 
-\EZY{What are the limitations of finally tagless in this context?}
+Detailed discussion of GADTless programming and comparing advantages of these approaches is out of scope of this paper.
+However, it is worth pointing out that it is much easier to do deep pattern matching with data types (including GADTs) than with class instances.
+For example, earlier we have used deep pattern matching in \verb|leftBalance| function:
+
+\begin{Verbatim}
+leftBalance :: Node a -> Node a
+leftBalance (N B (N R (N R a x b) y c) z d) =
+    N R (N B a x b) y (N B c z d)
+leftBalance (N B (N R a x (N R b y c)) z d) =
+    N R (N B a x b) y (N B c z d)
+leftBalance n = n
+\end{Verbatim}
+
+Another important aspect is that GADTs, just like all data types in Haskell, are closed: once they are declared, they can not be extended in the rest of the program.
+In contrast, type class instances are open.
+For example, we could split \verb|Expr e| type class into two separate type classes (this way we improve modularity):
+
+\begin{Verbatim}
+class ArithExpr e where
+    intVal  :: Int -> e Int
+    add     :: e Int -> e Int -> e Int
+    isZero  :: e Int -> e Bool
+
+class LogExpr e where
+    boolVal :: Bool -> e Bool
+    if'     :: e Bool -> e t -> e t -> e t
+
+class (ArithExpr e, LogExpr e) => Expr e
+\end{Verbatim}
 
@@
 \end{longitem}
-
-% Comparing to an approach to make verification by translating Haskell to other languages and back (such as xmonad), verification in Haskell itself...
-
-As a result, we have shown that the notion of GADTs is a very valuable extension of the Haskell language.
-
-\EZY{I think you can write a better concluding sentence then that!}
 

### 2013-06-04 gadt.tex
Minor tweaks and two new comments.
Signed-off-by: Edward Z. Yang <ezyang@mit.edu>

@@
 Haskell enforces a phase separation between run-time values and compile-time types.
-This causes a problem with type signature of function \verb|repeatElem|: we can not specify type-level representation of list elements count in the return value type of the function matching the count passed as a value.
+This causes a problem with the type signature of the function \verb|repeatElem|: there is no direct way to specify a type-level representation of list's count in the return value type of the function which will match the count passed in as a value.
 
@@
 
-We have shown that insertion operation maintains all invariants in red-black trees by adding necessary properties to the tree data type and then modifying the functions accordingly.
+\EZY{The new sentence you added here is really awkward} We have shown that the insertion operation maintains all invariants in red-black trees by adding necessary properties to the tree data type and then modifying the functions accordingly.
 However, the previous code illustrates a general problem with proofs.
@@
 Type inference for programs with GADTs is out of scope for this article.
-But when writing programs we are faced with the problem: typically Haskell functions do not require writing type signatures, but many functions involving GADTs do not compile without specifying type signatures explicitly, because Haskell compiler can not automatically infer the type.
+But when writing programs we are faced with a problem: typically Haskell functions do not require writing type signatures, but many functions involving GADTs do not compile without specifying type signatures explicitly, as the Haskell compiler can not automatically infer the type.
 This section describes the reasons of this situation.
@@
 
+\EZY{You should probably cite Data Types a la Carte here}
+
 \begin{Verbatim}

### 2013-06-08 monoids.tex
Add monoids
Signed-off-by: Edward Z. Yang <ezyang@mit.edu>

\documentclass[tikz]{tmr}

\title{Two Monoids for Approximating \np-Complete Problems}
\author{Mike Izbicki\email{mike@izbicki.me}}

\newcommand\h{\lstinline}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% misc commands

\newcommand{\prob}[1]{{\sc {#1}}}
\newcommand{\np}{{\ensuremath{\mathbb{NP}}}}

\newcommand\doubleplus{+\kern-1.3ex+\kern0.8ex}
\newcommand\+{\mdoubleplus}
\newcommand\mdoubleplus{\ensuremath{\mathbin{+\mkern-10mu+}}}

\begin{document}

\begin{introduction}
As a TA I was confronted with a real life instance of the \np-complete \prob{Scheduling} problem.
To solve the problem, I turned to the classic Least Processing Time First (LPTF) approximation algorithm.
In this article, we'll see that because LPTF is a monoid homomorphism, we can implement it using HLearn's {\upshape\h{HomTrainer}} type class.
This gives us parallel and online versions of LPTF ``for free.''
We'll also be able to use these same techniques to solve a related problem called \prob{BinPacking}.
Hopefully, at the end of the article you'll understand when the {\upshape\h{HomTrainer}} class might be a useful tool, and how to use and build your own instances.
% What's more, LPTF is a particularly illustrative \h{HomTrainer} instance for three reasons.
% First, the monoid operation for \prob{Scheduling} takes non-constant time, and the \h{HomTrainer} class makes reasoning about this run time straightforward.
% Second, the monoid uses lazy semantics to improve performance.  
% Finally, it demonstrates the versatility of the \h{HomTrainer} to domains outside of machine learning.
\end{introduction}

\section{Framing The Problem}
I enjoy TAing the introduction to C++ course at my university.
Teaching pointer arithmetic can be immensely frustrating, but it's worth it to see the students when it all finally clicks.  
% (
% (Not to mention the subtle indoctrination against these low level languages.)
% Of course, watching them struggle for hours to hunt down a segfault reminds me why Haskell is so awesome.
% I enjoy teaching.  
% It's really fun helping my intro to C++ students struggle through pointer arithmetic.  
% (I doubt the students feel the same way after spending five hours tracking down a segfault!)
%(It's only fun because \emph{I} am not the one spending hours tracking down the segfaults!)
% Ah, the joys of segfaults... 
%(to be young and segfaulting...)
Teaching is even better when it causes you to stumble onto an interesting problem.  
Oddly enough, I found this cool Haskell problem because of my C++ teaching assistanceship.

The professor wanted to assign a group project, and I had to pick the groups.
There had to be exactly five groups, and the groups needed to be as fair as possible.
% The only requirement was that the groups should be as fair as possible.
%He didn't want to let students pick their own groups because he wanted the groups to be as fair as possible.
In other words, I was supposed to evenly distribute the best and worst students.
% Because of this fairness criterion, I couldn't let the students pick their own groups.
% Because of these requirements, the students couldn't pick their own groups.

After a little thought, I realized this was an instance of the \np-complete \prob{Scheduling} problem in disguise.
This problem was first formulated in the context of concurrent computing.
In the textbook example of \prob{Scheduling}, we are given $p$ processors and $n$ tasks.
Each task $t_i$ has some associated time it will take to complete it.
The goal is to assign the tasks to processors so as to complete the tasks as quickly as possible.

The \prob{Scheduling} problem is shown graphically in Figure \ref{fig:scheduling}.
Processors are drawn as bins (labeled $b_1 ... b_4$), and tasks are drawn as green blocks (labeled $t_1 ... t_{10}$) and placed inside a bin.
The height of the task $t_i$ represents the length of time required to process that task.
Our goal is to minimize the \textcolor{red}{red} processing time.

\begin{figure}[H]
\centering
\caption{The \prob{Scheduling} problem}
\label{fig:scheduling}
\resizebox{!}{2in}{
\begin{tikzpicture}
\definecolor{hlearn_bgbox}{RGB}{127,255,127}
\node[shape=rectangle,draw,fill=hlearn_bgbox,minimum width=2cm,minimum height=2.1cm] at (3,1.05) { $t_{2}$ };
\node[shape=rectangle,draw,fill=hlearn_bgbox,minimum width=2cm,minimum height=0.7cm] at (3,2.45) { $t_{6}$ };
\node[shape=rectangle,draw,fill=hlearn_bgbox,minimum width=2cm,minimum height=2.0cm] at (5,1.0) { $t_{1}$ };
\node[shape=rectangle,draw,fill=hlearn_bgbox,minimum width=2cm,minimum height=0.8cm] at (5,2.4) { $t_{7}$ };
\node[shape=rectangle,draw,fill=hlearn_bgbox,minimum width=2cm,minimum height=0.6cm] at (5,3.0999999999999996) { $t_{5}$ };
\node[shape=rectangle,draw,fill=hlearn_bgbox,minimum width=2cm,minimum height=1.7cm] at (7,0.85) { $t_{10}$ };
\node[shape=rectangle,draw,fill=hlearn_bgbox,minimum width=2cm,minimum height=0.9cm] at (7,2.15) { $t_{8}$ };
\node[shape=rectangle,draw,fill=hlearn_bgbox,minimum width=2cm,minimum height=0.65cm] at (7,2.9250000000000003) { $t_{4}$ };
\node[shape=rectangle,draw,fill=hlearn_bgbox,minimum width=2cm,minimum height=1.5cm] at (9,0.75) { $t_{3}$ };
\node[shape=rectangle,draw,fill=hlearn_bgbox,minimum width=2cm,minimum height=1.3cm] at (9,2.15) { $t_{9}$ };
\draw[line width=0.1cm] (2,4) to (2,0) to node[below] {$b_{1}$} (2+2,0) to (2+2,4);
\draw[line width=0.1cm] (4,4) to (4,0) to node[below] {$b_{2}$} (4+2,0) to (4+2,4);
\draw[line width=0.1cm] (6,4) to (6,0) to node[below] {$b_{3}$} (6+2,0) to (6+2,4);
\draw[line width=0.1cm] (8,4) to (8,0) to node[below] {$b_{4}$} (8+2,0) to (8+2,4);

\draw[dashed,color=red,line width=0.05cm] (1,3.4) to (10,3.4);
\draw[dashed,color=red,line width=0.05cm] (1,0) to (2,0);
\draw[<->,color=red,line width=0.05cm] (1.5,3.4) to  (1.5,0);
\node[color=red] at (1,1.7) {\rotatebox{90}{
    \begin{tabular}{c}
%     variable\\
    processing time
    \end{tabular}}};
\draw[dashed,color=blue,line width=0.05cm] (2,4) to (2,5);
\draw[dashed,color=blue,line width=0.05cm] (10,4) to (10,5);
\draw[<->,color=blue,line width = 0.05cm] (2,4.5) to (10,4.5);
\node[color=blue] at (6,5) {fixed number of processors};
\end{tikzpicture}
}
\vspace{0.1in}
\end{figure}

What does this have to do with the problem my professor gave me?
Well, we can think of the number of groups as the number of processors, each student is a task, and the student's current grade is the task's processing time.
Then, the problem is to find a way to divvy up the students so that the sum of grades for the ``best'' group is as small as possible.
We'll see some code for solving this problem in a bit.

There is another closely related problem called \prob{BinPacking} that is easily confused with \prob{Scheduling}.
In \prob{BinPacking}, instead of fixing the number of bins and minimizing the amount in the bins, we fix the capacity of each bin and minimize the total number of bins used.
Compare Figure \ref{fig:binpacking} below and Figure \ref{fig:scheduling} above to see the difference.
The \prob{BinPacking} problem was originally studied by shipping companies, although like \prob{Sheduling} it occurs in many domains.
% The original motivation for the \prob{BinPacking} problem was in shipping.
% Every container shipped costs the same amount, no matter how much is stored in side of it.
% Therefore, we would like to reduce the number of containers we use to ship our products.

\begin{figure}[H]
\caption{The \prob{BinPacking} problem}
\label{fig:binpacking}
\centering
\resizebox{!}{1.5in}{
\begin{tikzpicture}
\definecolor{hlearn_bgbox}{RGB}{127,255,127}
\node[shape=rectangle,draw,fill=hlearn_bgbox,minimum width=2cm,minimum height=2.1cm] at (3,1.05) { $t_{2}$ };
\node[shape=rectangle,draw,fill=hlearn_bgbox,minimum width=2cm,minimum height=0.7cm] at (3,2.45) { $t_{6}$ };
\node[shape=rectangle,draw,fill=hlearn_bgbox,minimum width=2cm,minimum height=2.0cm] at (5,1.0) { $t_{1}$ };
\node[shape=rectangle,draw,fill=hlearn_bgbox,minimum width=2cm,minimum height=0.8cm] at (5,2.4) { $t_{7}$ };
\node[shape=rectangle,draw,fill=hlearn_bgbox,minimum width=2cm,minimum height=1.7cm] at (7,0.85) { $t_{10}$ };
\node[shape=rectangle,draw,fill=hlearn_bgbox,minimum width=2cm,minimum height=0.9cm] at (7,2.15) { $t_{8}$ };
\node[shape=rectangle,draw,fill=hlearn_bgbox,minimum width=2cm,minimum height=1.5cm] at (9,0.75) { $t_{3}$ };
\node[shape=rectangle,draw,fill=hlearn_bgbox,minimum width=2cm,minimum height=1.3cm] at (9,2.15) { $t_{9}$ };
\node[shape=rectangle,draw,fill=hlearn_bgbox,minimum width=2cm,minimum height=0.65cm] at (11,0.325) { $t_{4}$ };
\node[shape=rectangle,draw,fill=hlearn_bgbox,minimum width=2cm,minimum height=0.6cm] at (11,0.95) { $t_{5}$ };
\draw[line width=0.1cm] (2,3.0) to (2,0) to node[below] {$b_{1}$} (2+2,0) to (2+2,3.0);
\draw[line width=0.1cm] (4,3.0) to (4,0) to node[below] {$b_{2}$} (4+2,0) to (4+2,3.0);
\draw[line width=0.1cm] (6,3.0) to (6,0) to node[below] {$b_{3}$} (6+2,0) to (6+2,3.0);
\draw[line width=0.1cm] (8,3.0) to (8,0) to node[below] {$b_{4}$} (8+2,0) to (8+2,3.0);
\draw[line width=0.1cm] (10,3.0) to (10,0) to node[below] {$b_{5}$} (10+2,0) to (10+2,3.0);
\draw[line width=0.1cm] (12,3.0) to (12,0) to node[below] {$b_{6}$} (12+2,0) to (12+2,3.0);
% \draw[line width=0.1cm] (14,3.0) to (14,0) to node[below] {$b_{5}$} (14+2,0) to (14+2,3.0);
% \draw[line width=0.1cm] (16,3.0) to (16,0) to node[below] {$b_{5}$} (16+2,0) to (16+2,3.0);
\draw[line width=0.1cm] (2,3.0) to (15,3.0);
\draw[line width=0.1cm] (2,0) to (15,0);
\draw[dashed,color=red,line width=0.05cm] (2,3) to (2,4);
\draw[dashed,color=red,line width=0.05cm] (12,3) to (12,4);
\draw[<->,line width=0.05cm,color=red] (2,3.5) to node[above] {number of bins} (12,3.5);
\draw[<->,line width=0.05cm,color=blue] (1.5,0) to (1.5,3);
\draw[dashed,color=blue,line width=0.05cm] (1,0) to (2,0);
\draw[dashed,color=blue,line width=0.05cm] (1,3) to (2,3);
\node[color=blue] at (1,1.5) {\rotatebox{90}{fixed bin size}};
\node at (15,1.5) {\textbf{...}};
\end{tikzpicture}
}
\end{figure}

Both \prob{Scheduling} and \prob{BinPacking} are \np-complete.
% Since \prob{Scheduling} is \np-complete, finding an exact solution takes exponential time.  
Therefore, I had no chance of creating an optimal grouping for my professor---the class had 100 students, and $2^{100}$ is a \textit{big} number!
So I turned to approximation algorithms.
% These algorithms give us solutions that are guaranteed to be within a certain factor of optimal.
One popular approximation for \prob{Scheduling} is called Longest Processing Time First (LPTF).
When analyzing these approximation algorithms, it is customary to compare the quality of their result with that of the theoretical optimal result.
% Even when it is impractical to actually calculate this optimal, we can often bound it to within a certain range.
In this case, we denote the total processing time of the schedule returned by LPTF as \textit{LPTF}, and the processing time of the optimal solution as \textit{OPT}.
It can be shown that the following bound holds:
$$
\textit{LPTF} \le \left( {4\over3} - {1\over{3n}} \right) \textit{OPT}
$$
This bound was proven in the late 1960's, but the original paper remains quite readable today \cite{graham69}.
% For a survey on variants of the Scheduling problem, see Chapter 1 of \textit{Approximation Algorithms for NP Hard Problems} \cite{hochbaum96scheduling}.
By making this small sacrifice in accuracy, we get an algorithm that runs in time $\Theta(n\log n)$ instead of $\Theta(2^n)$.  Much better!  In the rest of this article, we'll take a detailed look at a Haskell implementation of the LPTF algorithm, and then briefly use similar techniques to solve the \prob{BinPacking} problem.

\section{The Scheduling HomTrainer}

When implementing an algorithm in Haskell, you always start with the type signature.
% I set out to implement the LPTF algorithm, and like all good Haskellers I started with the type signature.
LPTF takes a collection of tasks and produces a schedule, so it's type might look something like:
\begin{spec}
:: [Task] -> Schedule
\end{spec}
Anytime I see see a function of this form, I ask myself, ``Can it be implemented using the \h{HomTrainer} type class?''
\h{HomTrainer}s are useful because the compiler automatically derives online and parallel algorithms for all instances.
For LPTF, the answer is ``yes.''
This section will cover what exactly that means.

We start by looking at the format of a \h{HomTrainer} instance as shown graphically in Figure \ref{fig:HomTrainer} below.

\begin{figure}[H]
\caption{Basic requirements of the HomTrainer type class}
\label{fig:HomTrainer}
\centering
\tikzstyle{blackbox}=[shape=rectangle, draw, fill=black, text=white, minimum width=0.8in]
\definecolor{darkgreen}{RGB}{0,127,0}
\begin{tikzpicture}%[node distance=2.5in, auto, >=stealth]
\node[blackbox] (dataset) [minimum height=0.3in] {\textbf{data set}};
\node[blackbox] (model)   [minimum height=0.3in, node distance=2in,right of = dataset] {\textbf{model}};
\node[blackbox] (answer)  [minimum height=0.3in, node distance=2in,right of = model] {\textbf{answer}};
\draw[->,ultra thick] (dataset) to node[above] {train} (model);
\draw[->,ultra thick] (model) to node[above] {question} (answer);

\node (l1) [color=blue] at (-0.1,-1.3) {free monoid};
\node (l2) [color=blue] at (2.5,1.5) {homomorphism};
\node (l3) [color=blue] at (5.5,-1.3) {monoid};
\node (l4) [color=darkgreen] at (5.5,1.3) {\h{HomTrainer}};

\draw[->,color=blue,ultra thick] (l1) to (dataset);
\draw[->,color=blue,ultra thick] (l2) to (2.5,0.5);
\draw[->,color=blue,ultra thick] (l3) to (model);
\draw[->,color=darkgreen,ultra thick] (l4) to (model);
\end{tikzpicture}
\vspace{0.15in}
\end{figure}

There's a lot going on in Figure \ref{fig:HomTrainer}, but we'll look at things piece-by-piece.
The black boxes represent data types and the black arrows represent functions.
In our case, the data set is the collection of tasks we want scheduled and the model is the schedule.
The train function is the LPTF algorithm, which generates a model from the data points.
Finally, our model is only useful if we can ask it questions.
In this case, we might want to ask, ``Which processor is assigned to task $t_{10}$?'' or ``What are all the tasks assigned to processor 2?''
% The answer type would be a processor.
% This framework was originally meant for use with machine learning algorithms, however it is general enough to be applied in many contexts.

The blue arrows in the diagram are a little trickier.
They specify that our training algorithm must be a \emph{monoid homomorphism} from the \emph{free monoid}.
% This sounds a lot scarier than it actually is.
These are scary sounding words, but they're pretty simple once you're familiar with them.
We'll define them and see some examples.

In Haskell, the \h{Monoid} type class is defined as having an identity called \h{mempty} and a binary operation called \h{mappend}:
\begin{spec}
class Monoid m where
    mempty  :: m
    mappend :: m -> m -> m
\end{spec}
Sometimes we use the infix operation \h{(<>) = mappend} to make our code easier to read.
All instances must obey the identity and associativity laws:
\begin{spec}
mempty <> m = m <> mempty = m
(m1 <> m2) <> m3 = m1 <> (m2 <> m3)
\end{spec}
% \end{spec}
% mappend mempty m = mappend m mempty = m
% and associativity law:
% \begin{spec}
% $$
% \epsilon \diamond m = m\diamond \epsilon = m
% $$
Lists are one of the simplest examples of monoids.
Their identity element is the empty list, and their binary operation is concatenation:
\begin{spec}
instance Monoid [a] where
    mempty = []
    mappend = ++
\end{spec}
Vectors have a similar definition, and both lists and vectors are examples of free monoids.
Free monoids are monoids that can be generated by some underlying set.
In our monoid instance for lists above, the underlying set is the type variable \h{a}.
For the \h{HomTrainer}, when we say that our data set must be a free monoid, all we mean is that it is a collection of some data points
(\eg\ a list of tasks).

A homomorphism from the free monoid is a function that ``preserves the free monoid's structure.''
More formally, if the function is called \h{train}, then it obeys the law that for all \lstinline{xs} and \lstinline{ys} of type \lstinline{[a]}:
\begin{spec}
train (xs ++ ys) = (train xs) <> (train ys)
\end{spec}
The LPTF algorithm turns out to have this property.
Figure \ref{fig:LPTF-monoid} shows this in picture form with a commutative diagram.
This means that it doesn't matter whether we take the orange path (first train schedules from our data sets, then combine the schedules with \h{mappend}) or the purple path (first concatenate our data sets, then train a schedule on the result).
Either way, we get the exact same answer.

\begin{figure}[H]
\caption{The LPTF is a monoid homomorphism because this diagram commutes}
\label{fig:LPTF-monoid}
% \centering
\hspace{-0.3in}
\begin{tikzpicture}

\draw[->,line width=0.05cm,color=orange] (-2in,3.3) to (-2in,2.5);
\draw[->,line width=0.05cm,color=orange] (0in,3.3) to (0in,2.5);
\draw[dotted,line width=0.025cm,color=orange] (-2in,2.5) to (-2in,-2.5);
\draw[dotted,line width=0.025cm,color=orange] (0in,2.5) to (0in,-2.5);
\draw[->,line width=0.05cm,color=orange] (0in,-2.5) to (0in,-3) to (2.5in,-3) to (2.5in,-2.5);
\draw[->,line width=0.05cm,color=orange] (-2in,-2.5) to (-2in,-3) to (2.5in,-3) to (2.5in,-2.5);
\draw[->,line width=0.05cm,color=purple]  (2.5in,3.3) to (2.5in,2.5);
\draw[->,line width=0.05cm,color=purple] (0in,4.8) to (0in,5.3) to (2.5in,5.3) to (2.5in,4.8);
\draw[->,line width=0.05cm,color=purple] (-2in,4.8) to (-2in,5.3) to (2.5in,5.3) to (2.5in,4.8);
\draw[dotted,line width=0.025cm,color=purple] (2.5in,4.8) to (2.5in,3.3);

\node at (-2in,4) {
    \resizebox{!}{0.25in}{
    \begin{tikzpicture}
    \definecolor{hlearn_bgbox}{RGB}{127,255,127}
    \definecolor{hlearn_bgboxB}{RGB}{224,224,255}
    \node[shape=rectangle,draw,fill=hlearn_bgbox,minimum width=2cm,minimum height=2.1cm]   at (1,1.05) { $t_{a,1}$ };
    \node[shape=rectangle,draw,fill=hlearn_bgbox,minimum width=2cm,minimum height=2.7cm]   at (3,1.35) { $t_{a,2}$ };
    \node[shape=rectangle,draw,fill=hlearn_bgbox,minimum width=2cm,minimum height=3.1cm]   at (5,1.55) { $t_{a,3}$ };;
    \node[shape=rectangle,draw,fill=hlearn_bgbox,minimum width=2cm,minimum height=1.7cm]   at (7,0.85) { $t_{a,4}$ };
    \node[shape=rectangle,draw,fill=hlearn_bgbox,minimum width=2cm,minimum height=1.6cm]   at (9,0.8) { $t_{a,5}$ };
    \node[shape=rectangle,draw,fill=hlearn_bgbox,minimum width=2cm,minimum height=0.6cm]   at (11,0.3) { $t_{a,6}$ };
    \end{tikzpicture}
    }};

\node at (-2in,0) {
    \resizebox{1.5in}{!}{
    \begin{tikzpicture}
    \definecolor{hlearn_bgbox}{RGB}{127,255,127}
    \node[shape=rectangle,draw,fill=hlearn_bgbox,minimum width=2cm,minimum height=3.1cm] at (3,1.55) { $t_{a,3}$ };
    \node[shape=rectangle,draw,fill=hlearn_bgbox,minimum width=2cm,minimum height=0.6cm] at (3,3.4) { $t_{a,6}$ };
    \node[shape=rectangle,draw,fill=hlearn_bgbox,minimum width=2cm,minimum height=2.7cm] at (5,1.35) { $t_{a,2}$ };
    \node[shape=rectangle,draw,fill=hlearn_bgbox,minimum width=2cm,minimum height=1.6cm] at (5,3.5) { $t_{a,5}$ };
    \node[shape=rectangle,draw,fill=hlearn_bgbox,minimum width=2cm,minimum height=2.1cm] at (7,1.05) { $t_{a,1}$ };
    \node[shape=rectangle,draw,fill=hlearn_bgbox,minimum width=2cm,minimum height=1.7cm] at (7,2.95) { $t_{a,4}$ };
    \draw[line width=0.1cm] (2,7.0) to (2,0) to (2+2,0) to (2+2,7.0);
    \draw[line width=0.1cm] (4,7.0) to (4,0) to (4+2,0) to (4+2,7.0);
    \draw[line width=0.1cm] (6,7.0) to (6,0) to (6+2,0) to (6+2,7.0);
    \node at (3,-0.5) {$b_{1}$};
    \node at (5,-0.5) {$b_{2}$};
    \node at (7,-0.5) {$b_{3}$};
    \end{tikzpicture}
    }};

\node at (0,4) {
    \resizebox{!}{0.25in}{
%     \resizebox{1.75in}{!}{
    \begin{tikzpicture}
    \definecolor{hlearn_bgbox}{RGB}{127,255,127}
    \definecolor{hlearn_bgboxB}{RGB}{224,224,255}
    \node[shape=rectangle,draw,fill=hlearn_bgboxB,minimum width=2cm,minimum height=0.6cm]  at (1,0.3) { $t_{b,1}$ };
    \node[shape=rectangle,draw,fill=hlearn_bgboxB,minimum width=2cm,minimum height=2.5cm]  at (3,1.25) { $t_{b,2}$ };
    \node[shape=rectangle,draw,fill=hlearn_bgboxB,minimum width=2cm,minimum height=1.0cm]  at (5,0.5) { $t_{b,3}$ };
    \node[shape=rectangle,draw,fill=hlearn_bgboxB,minimum width=2cm,minimum height=0.7cm]  at (7,0.35) { $t_{b,4}$ };
    \node[shape=rectangle,draw,fill=hlearn_bgboxB,minimum width=2cm,minimum height=0.65cm] at (9,0.325) { $t_{b,5}$ };
    \node[shape=rectangle,draw,fill=hlearn_bgboxB,minimum width=2cm,minimum height=0.7cm]  at (11,0.35) { $t_{b,6}$ };
    \node[shape=rectangle,draw,fill=hlearn_bgboxB,minimum width=2cm,minimum height=0.9cm]  at (13,0.45) { $t_{b,7}$ };
    \node[shape=rectangle,draw,fill=hlearn_bgboxB,minimum width=2cm,minimum height=0.85cm] at (15,0.425) { $t_{b,8}$ };
    \end{tikzpicture}
    }};

\node at (0,0) {
    \resizebox{1.5in}{!}{
    \begin{tikzpicture}
    \definecolor{hlearn_bgboxB}{RGB}{224,224,255}
    \node[shape=rectangle,draw,fill=hlearn_bgboxB,minimum width=2cm,minimum height=2.5cm] at (3,1.25) { $t_{b,2}$ };
    \node[shape=rectangle,draw,fill=hlearn_bgboxB,minimum width=2cm,minimum height=1.0cm] at (5,0.5) { $t_{b,3}$ };
    \node[shape=rectangle,draw,fill=hlearn_bgboxB,minimum width=2cm,minimum height=0.7cm] at (5,1.35) { $t_{b,4}$ };
    \node[shape=rectangle,draw,fill=hlearn_bgboxB,minimum width=2cm,minimum height=0.7cm] at (5,2.05) { $t_{b,6}$ };
    \node[shape=rectangle,draw,fill=hlearn_bgboxB,minimum width=2cm,minimum height=0.6cm] at (5,2.6999999999999997) { $t_{b,1}$ };
    \node[shape=rectangle,draw,fill=hlearn_bgboxB,minimum width=2cm,minimum height=0.9cm] at (7,0.45) { $t_{b,7}$ };
    \node[shape=rectangle,draw,fill=hlearn_bgboxB,minimum width=2cm,minimum height=0.85cm] at (7,1.325) { $t_{b,8}$ };
    \node[shape=rectangle,draw,fill=hlearn_bgboxB,minimum width=2cm,minimum height=0.65cm] at (7,2.075) { $t_{b,5}$ };
    \draw[line width=0.1cm] (2,7.0) to (2,0) to (2+2,0) to (2+2,7.0);
    \draw[line width=0.1cm] (4,7.0) to (4,0) to (4+2,0) to (4+2,7.0);
    \draw[line width=0.1cm] (6,7.0) to (6,0) to (6+2,0) to (6+2,7.0);
    \node at (3,-0.5) {$b_{1}$};
    \node at (5,-0.5) {$b_{2}$};
    \node at (7,-0.5) {$b_{3}$};
    \end{tikzpicture}
    }};

\node at (2.5in,4) {
    \resizebox{!}{0.25in}{
    \begin{tikzpicture}
    \definecolor{hlearn_bgbox}{RGB}{127,255,127}
    \definecolor{hlearn_bgboxB}{RGB}{224,224,255}
    \node[shape=rectangle,draw,fill=hlearn_bgbox,minimum width=2cm,minimum height=2.1cm]   at (1,1.05) { $t_{a,1}$ };
    \node[shape=rectangle,draw,fill=hlearn_bgbox,minimum width=2cm,minimum height=2.7cm]   at (3,1.35) { $t_{a,2}$ };
    \node[shape=rectangle,draw,fill=hlearn_bgbox,minimum width=2cm,minimum height=3.1cm]   at (5,1.55) { $t_{a,3}$ };;
    \node[shape=rectangle,draw,fill=hlearn_bgbox,minimum width=2cm,minimum height=1.7cm]   at (7,0.85) { $t_{a,4}$ };
    \node[shape=rectangle,draw,fill=hlearn_bgbox,minimum width=2cm,minimum height=1.6cm]   at (9,0.8) { $t_{a,5}$ };
    \node[shape=rectangle,draw,fill=hlearn_bgbox,minimum width=2cm,minimum height=0.6cm]   at (11,0.3) { $t_{a,6}$ };
    \node[shape=rectangle,draw,fill=hlearn_bgboxB,minimum width=2cm,minimum height=0.6cm]  at (13,0.3) { $t_{b,1}$ };
    \node[shape=rectangle,draw,fill=hlearn_bgboxB,minimum width=2cm,minimum height=2.5cm]  at (15,1.25) { $t_{b,2}$ };
    \node[shape=rectangle,draw,fill=hlearn_bgboxB,minimum width=2cm,minimum height=1.0cm]  at (17,0.5) { $t_{b,3}$ };
    \node[shape=rectangle,draw,fill=hlearn_bgboxB,minimum width=2cm,minimum height=0.7cm]  at (19,0.35) { $t_{b,4}$ };
    \node[shape=rectangle,draw,fill=hlearn_bgboxB,minimum width=2cm,minimum height=0.65cm] at (21,0.325) { $t_{b,5}$ };
    \node[shape=rectangle,draw,fill=hlearn_bgboxB,minimum width=2cm,minimum height=0.7cm]  at (23,0.35) { $t_{b,6}$ };
    \node[shape=rectangle,draw,fill=hlearn_bgboxB,minimum width=2cm,minimum height=0.9cm]  at (25,0.45) { $t_{b,7}$ };
    \node[shape=rectangle,draw,fill=hlearn_bgboxB,minimum width=2cm,minimum height=0.85cm] at (27,0.425) { $t_{b,8}$ };
    \end{tikzpicture}
    }};

\node at (2.5in,0) {
    \resizebox{1.5in}{!}{
    \begin{tikzpicture}
    \definecolor{hlearn_bgbox}{RGB}{127,255,127}
    \definecolor{hlearn_bgboxB}{RGB}{224,224,255}
    \node[shape=rectangle,draw,fill=hlearn_bgbox,minimum width=2cm,minimum height=3.1cm] at (3,1.55) { $t_{a,3}$ };
    \node[shape=rectangle,draw,fill=hlearn_bgbox,minimum width=2cm,minimum height=1.6cm] at (3,3.9000000000000004) { $t_{a,5}$ };
    \node[shape=rectangle,draw,fill=hlearn_bgboxB,minimum width=2cm,minimum height=0.85cm] at (3,5.125) { $t_{b,8}$ };
    \node[shape=rectangle,draw,fill=hlearn_bgboxB,minimum width=2cm,minimum height=0.65cm] at (3,5.875) { $t_{b,5}$ };
    \node[shape=rectangle,draw,fill=hlearn_bgbox,minimum width=2cm,minimum height=2.7cm] at (5,1.35) { $t_{a,2}$ };
    \node[shape=rectangle,draw,fill=hlearn_bgbox,minimum width=2cm,minimum height=1.7cm] at (5,3.5500000000000003) { $t_{a,4}$ };
    \node[shape=rectangle,draw,fill=hlearn_bgboxB,minimum width=2cm,minimum height=1.0cm] at (5,4.9) { $t_{b,3}$ };
    \node[shape=rectangle,draw,fill=hlearn_bgboxB,minimum width=2cm,minimum height=0.7cm] at (5,5.75) { $t_{b,4}$ };
    \node[shape=rectangle,draw,fill=hlearn_bgbox,minimum width=2cm,minimum height=0.6cm] at (5,6.4) { $t_{a,6}$ };
    \node[shape=rectangle,draw,fill=hlearn_bgboxB,minimum width=2cm,minimum height=2.5cm] at (7,1.25) { $t_{b,2}$ };
    \node[shape=rectangle,draw,fill=hlearn_bgbox,minimum width=2cm,minimum height=2.1cm] at (7,3.55) { $t_{a,1}$ };
    \node[shape=rectangle,draw,fill=hlearn_bgboxB,minimum width=2cm,minimum height=0.9cm] at (7,5.05) { $t_{b,7}$ };
    \node[shape=rectangle,draw,fill=hlearn_bgboxB,minimum width=2cm,minimum height=0.7cm] at (7,5.85) { $t_{b,6}$ };
    \node[shape=rectangle,draw,fill=hlearn_bgboxB,minimum width=2cm,minimum height=0.6cm] at (7,6.5) { $t_{b,1}$ };
    \draw[line width=0.1cm] (2,7.0) to (2,0) to (2+2,0) to (2+2,7.0);
    \draw[line width=0.1cm] (4,7.0) to (4,0) to (4+2,0) to (4+2,7.0);
    \draw[line width=0.1cm] (6,7.0) to (6,0) to (6+2,0) to (6+2,7.0);
    \node at (3,-0.5) {$b_{1}$};
    \node at (5,-0.5) {$b_{2}$};
    \node at (7,-0.5) {$b_{3}$};
    \end{tikzpicture}
    }};

\node at (-1in,0) {$\diamond$};
\node at (1.125in,0) {$=$};
\node at (-1in,4) {$\+$};
\node at (1.125in,4) {$=$};

\node at (-3in,4.25) {\rotatebox{90}{data sets}};
\node at (-3in,-0.5) {\rotatebox{90}{models}};

\draw[dotted] (-3in,3) to (3in,3);
\end{tikzpicture}
\vspace{0.00in}
\end{figure}

Now that we understand what \h{HomTrainer} is, we can look at what it gives us.
Most importantly, it gives us a simple interface for interacting with our models.
This interface is shown in Figure \ref{alg:HomTrainer}.
In the class, we associate a specific \h{Datapoint} type to our model, and get four functions for training functions.
The most important training function is the batch trainer, called \h{train}.
This is the homomorphism that converts the data set into a model.
In our case, it will be the LPTF algorithm.
The second most important function is the online trainer \h{add1dp}.
This function takes a model and a datapoint as input, and ``adds'' the data point to the model.
Developing new online functions is an important research area in approximation algorithms.
As we will see later, the compiler generates these two functions automatically for all instances of \h{HomTrainer}.

Finally, HLearn comes with a higher order function for making all batch trainers run efficiently on multiple cores.
The function
\begin{spec}
parallel :: (...) => 
    (container datapoint -> model) -> (container datapoint -> model)
\end{spec}
takes a batch trainer as input and returns a parallelized one as output.
In the next section, we'll see an example of its use in practice.

\begin{figure}
\caption{The \h{HomTrainer} type class}
\label{alg:HomTrainer}
\begin{code}
class (Monoid model) => HomTrainer model where
    type Datapoint model

    -- The singleton trainer
    train1dp :: Datapoint model -> model

    -- The batch trainer
    train :: (Functor container, Foldable container) => 
        container (Datapoint model) -> model

    -- The online trainer
    add1dp :: model -> Datapoint model -> model

    -- The online batch trainer
    addBatch :: (Functor container, Foldable container) =>  
        model -> container (Datapoint model) -> model
\end{code}
\end{figure}

\section {Using the Scheduling HomTrainer}

Before we look at implementing a \h{HomTrainer} to solve \prob{Scheduling}, we'll take a look at how it's used.
In particular, we'll look at the Haskell code I used to solve the problem of grouping my students.
In order to run the code, you'll need to download the latest \h{HLearn-approximations} library:

\begin{lstlisting}
cabal install HLearn-approximations-1.0.0
\end{lstlisting}

% As always, we'll need to load some extensions and import some modules.
% Every program that uses the HLearn library is going to require at least the DataKinds and TypeFamilies extensions.
Let's begin by doing some experiments in GHCI to get ourselves oriented.
The \h{Scheduling} type is our model, and here we ask GHCI for it's kind signature:
\begin{lstlisting}
ghci> import HLearn.NPHard.Scheduling
ghci> :kind Scheduling
Scheduling :: Nat -> * -> *
\end{lstlisting}
\h{Scheduling} takes two type parameters.
The first is a type-level natural number that specifies the number of processors in our schedule.
In HLearn, any parameters to our training functions must be specified in the model's type signature.
In this case, the type-level numbers require the \h{DataKinds} extension to be enabled.
The second parameter to \h{Scheduling} is the type of the task we are trying to schedule.
% We can't infer much about these tasks from the kind signature alone.

Next, let's find out about \h{Scheduling}'s \h{HomTrainer} instance:
\begin{lstlisting}
ghci> :info Scheduling
...
instance (Norm a,...) => HomTrainer (Scheduling n a) where
...
\end{lstlisting}
We have a constraint on our task parameter specifying that it must be an instance of \h{Norm}.
What does that mean?
In mathematics, a type has a norm if it has a ``size'' of some sort.
In HLearn, the \h{Norm} type class is defined in two parts.
First we associate a specific number type with our model using the \h{HasRing} type class.
Then, the \h{Norm} type class defines a function \h{magnitude} that converts a model into the number type.
\begin{lstlisting}
class (Num (Ring m)) => HasRing m where
    type Ring m

class (HasRing m, Ord (Ring m)) => Norm m where
    magnitude :: m -> Ring m
\end{lstlisting}
Usually the associated ring will be a \h{Double}.
But we might make it a \h{Rational} if we need more accuracy or an \h{Int} if we don't want fractions.

Figure \ref{fig:main} shows the code for solving the student groups problem.
In this case, I defined a new data type called \h{Student} to be the data points, and defined the \h{magnitude} of a \h{Student} to be its \h{grade}.
Notice that since I am deriving an instance of \h{Ord} automatically, I must define \h{grade} before \h{name} and \h{section}.
This ensures that the ordering over students will be determined by their \h{grade}s, and so be the same as the ordering over their \h{magnitude}s.

The \h{main} function is divided up into three logical units.
First, we load a CSV file into a variable
\h{allStudents::[Student]}
% \begin{spec}
% allStudents::[Student]
% \end{spec}
using the \h{Cassava} package.
The details of how this works aren't too important.

In the middle section, we divide up the students according to which section they are in, and then \h{train} a \h{Schedule} model for each section.
We use the function:
\begin{spec}
getSchedules :: Scheduling n a -> [[a]]
\end{spec}
to extract a list of schedules from our \h{Schedule} type, then print them to the terminal.
Just for illustration, we train the third section's \h{Scheduling} model in parallel.
With only about 30 students in the section, we don't notice any improvement.
But as the data sets grow, more processors provide drastic improvements, as shown in Table \ref{table:rt}.

In the last section, we combine our section specific models together to get a combined model factoring in all of the sections.
Because \h{Scheduling} is an instance of \h{HomTrainer}, we don't have to retrain our model from scratch.
We can reuse the work we did in training our original models, resulting in a faster computation.

% Finally, if a student had been added to the class late, I could have easily accounted for this using the online trainer provided by the \h{HomTrainer} type class:
% \begin{spec}
% add1dp :: HomTrainer model => model -> Datapoint model -> model
% \end{spec}

\begin{figure}
\begin{code}             
{-# LANGUAGE TypeFamilies, DataKinds #-}

import Data.Csv
import qualified Data.ByteString.Lazy.Char8  as BS
import qualified Data.Vector  as V
import HLearn.Algebra
import HLearn.NPHard.Scheduling

---------------------------------------------------------------------

data Student = Student
   { grade   :: Double
   , name    :: String
   , section :: Int
   }
   deriving (Read,Show,Eq,Ord)

instance HasRing Student where
   type Ring (Student) = Double

instance Norm Student where
   magnitude = grade

---------------------------------------------------------------------

main = do
   Right allStudents <- 
      fmap (fmap (fmap (\(n,s,g) -> Student g n s) . V.toList) . decode True) 
      $ BS.readFile "students.csv" :: IO (Either String [Student])

   let section1 = filter (\s -> 1 == section s) allStudents
   let section2 = filter (\s -> 2 == section s) allStudents
   let section3 = filter (\s -> 3 == section s) allStudents
   let solution1 = train section1 :: Scheduling 5 Student
   let solution2 = train section2 :: Scheduling 5 Student
   let solution3 = parallel train section3 :: Scheduling 5 Student
   print $ map (map name) $ getSchedules solution1

   let solutionAll = solution1 <> solution2 <> solution3
   print $ map (map name) $ getSchedules solutionAll
\end{code}
\caption{Solution to my professor's problem}
\label{fig:main}
\end{figure}

\section{Implementing the LPTF HomTrainer}

Now we're ready to dive into the details of how \h{Scheduling} works under the hood.
% We've already looked at the type parameters.
% Let's take a more detailed look at the internals.
Here's the declaration:
\begin{spec}
data Scheduling (p::Nat) a = Scheduling
    { vector   :: !(SortedVector a)
    , schedule :: Map Bin [a]
    }
\end{spec}
Scheduling has two member variables.
The first is a \h{SortedVector}.
This custom data type is a wrapper around the \h{vector} package's \h{Vector} type that maintains the invariant that items are always sorted.
This vector will be used as an intermediate data structure while performing the LPTF computations.
The second member is the actual schedule.
It is represented as a \h{Map} with a \h{Bin} as the key and a list of tasks as the value.
\h{Bin} is just a type synonym for \h{Int}:
\begin{spec}
type Bin = Int
\end{spec}
and it represents the index of the processor in the range of 1 to $p$.

Before we look at \h{Scheduling}'s \h{Monoid} and \h{HomTrainer} instances, we need to take a more detailed look at the LPTF algorithm.
Traditionally, LPTF is described as a two step process.
First, sort the list of tasks in descending order.
Then, iterate through the sorted list.
On each iteration, assign the next task to the processor with the least amount of work.
Figure \ref{fig:lptf} shows a single iteration of this procedure.

\begin{figure}[H]
\centering
\caption{A single iteration of the LPTF algorithm}
\label{fig:lptf}
\resizebox{!}{3in}{
\begin{tikzpicture}
\definecolor{hlearn_bgbox}{RGB}{127,255,127}
\node[shape=rectangle,draw,fill=hlearn_bgbox,minimum width=2cm,minimum height=2.1cm] at (3,1.05) { $t_{2}$ };
\node[shape=rectangle,draw,fill=hlearn_bgbox,minimum width=2cm,minimum height=2.0cm] at (5,1) { $t_{1}$ };
\node[shape=rectangle,draw,fill=hlearn_bgbox,minimum width=2cm,minimum height=1.7cm] at (7,0.85) { $t_{10}$ };
\node[shape=rectangle,draw,fill=hlearn_bgbox,minimum width=2cm,minimum height=1.5cm] at (9,0.75) { $t_{3}$ };

% \node[shape=rectangle,draw,pattern=north east lines,pattern color=hlearn_bgbox,minimum width=2cm,minimum height=1.3cm] at (3,2.75) { $t_{9}$ };
% \node[shape=rectangle,draw,pattern=north east lines,pattern color=hlearn_bgbox,minimum width=2cm,minimum height=1.3cm] at (5,2.65) { $t_{9}$ };
% \node[shape=rectangle,draw,pattern=north east lines,pattern color=hlearn_bgbox,minimum width=2cm,minimum height=1.3cm] at (7,2.35) { $t_{9}$ };
\node[shape=rectangle,draw,pattern=north east lines,pattern color=hlearn_bgbox,minimum width=2cm,minimum height=1.3cm] at (9,2.15) { $t_{9}$ };

\node[shape=rectangle,draw,fill=hlearn_bgbox,minimum width=2cm,minimum height=1.3cm] at (1,0.5+5.65) { $t_{9}$ };
\node[shape=rectangle,draw,fill=hlearn_bgbox,minimum width=2cm,minimum height=0.9cm] at (3,0.5+5.45) { $t_{8}$ };
\node[shape=rectangle,draw,fill=hlearn_bgbox,minimum width=2cm,minimum height=0.8cm] at (5,0.5+5.4) { $t_{7}$ };
\node[shape=rectangle,draw,fill=hlearn_bgbox,minimum width=2cm,minimum height=0.7cm] at (7,0.5+5.35) { $t_{6}$ };
\node[shape=rectangle,draw,fill=hlearn_bgbox,minimum width=2cm,minimum height=0.65cm] at (9,0.5+5.325) { $t_{4}$ };
\node[shape=rectangle,draw,fill=hlearn_bgbox,minimum width=2cm,minimum height=0.6cm] at (11,0.5+5.3) { $t_{5}$ };

% \node at (6,4.5) {\emph{partial schedule}};
% \node at (6,7) {\emph{sorted list}};
\node at (-1,2) {
    \rotatebox{90}{
    \begin{tabular}{c}
    \emph{partial schedule}
%     \emph{partial} \\ \emph{bin packing}
    \end{tabular}
    }};
\node at (-1,6.25) {
    \rotatebox{90}{
    \begin{tabular}{c}
    \emph{sorted} \\ \emph{vector}
    \end{tabular}
%     \emph{sorted vector}
    }};

\draw[dotted] (-1,5) to (13,5);

\draw[line width=0.1cm] (2,4) to (2,0) to (2+2,0) to (2+2,4);
\draw[line width=0.1cm] (4,4) to (4,0) to (4+2,0) to (4+2,4);
\draw[line width=0.1cm] (6,4) to (6,0) to (6+2,0) to (6+2,4);
\draw[line width=0.1cm] (8,4) to (8,0) to (8+2,0) to (8+2,4);

\draw[->,line width=0.05in,color=red] (1,5.5) to [in=110,out=-70] (9,2.7);

\node at (3,-0.25) {$b_1$};
\node at (5,-0.25) {$b_2$};
\node at (7,-0.25) {$b_3$};
\node at (9,-0.25) {$b_4$};
\end{tikzpicture}
\vspace{0.1in}
}
\end{figure}

This conversion process is implemented with the internal function \h{vector2schedule}, whose code is shown in Figure \ref{code:vector2schedule} below.
The details of this function aren't particularly important.
What is important is that \h{vector2schedule} runs in time $\Theta(n\log p)$.
This will be important when determining the run times of the \h{mappend} and \h{train} functions.

\begin{figure}[H]
\caption{The \h{vector2schedule} helper function for LPTF}
\label{code:vector2schedule}
\begin{lstlisting}
vector2schedule :: (Norm a) => Int -> SortedVector a -> Map.Map Int [a]
vector2schedule p vector = snd $ F.foldr cata (emptyheap p,Map.empty) vector
    where
        emptyheap p = Heap.fromAscList [(0,i) | i<-[1..p]]
        cata x (heap,map) = 
            let Just top = Heap.viewHead heap
                set = snd top
                prio = (fst top)+magnitude x
                heap' = Heap.insert (prio,set) (Heap.drop 1 heap)
                map' = Map.insertWith (++) set [x] map
            in (heap',map')

\end{lstlisting}
\end{figure}

Our \h{mappend} operation will implement the LPTF algorithm internally in a way that reuses the results from the input \h{Schedule}s.
We won't be able to reuse the actual schedules, but we can reuse the sorting of the vectors.
We do this by taking advantage of the \h{HomTrainer} instance of the \h{SortedVector} type.
It turns out that merge sort is a monoid homomorphism, and so \h{SortedVector} can be made an instance of the \h{HomTrainer} type class.
The commutative diagram for \h{SortedVector} is shown in Figure \ref{fig:SortedVector} below.

\begin{figure}[H]
\caption{Constructing a \h{SortedVector} is a monoid homomorphism}
\label{fig:SortedVector}
\centering
\begin{tikzpicture}
\draw[->,line width=0.05cm,color=orange] (-1.7in,5.5) to (-1.7in,4.5);
\draw[->,line width=0.05cm,color=orange] (0in,5.5) to (0in,4.5);
\draw[dotted,line width=0.025cm,color=orange] (-1.7in,4.5) to (-1.7in,3.5);
\draw[dotted,line width=0.025cm,color=orange] (0in,4.5) to (0in,3.5);
\draw[->,line width=0.05cm,color=orange] (0in,3.5) to (0in,3) to (2.3in,3) to (2.3in,3.5);
\draw[->,line width=0.05cm,color=orange] (-1.7in,3.5) to (-1.7in,3) to (2.3in,3) to (2.3in,3.5);
\draw[->,line width=0.05cm,color=purple]  (2.3in,5.5) to (2.3in,4.5);
\draw[->,line width=0.05cm,color=purple] (0in,6.5) to (0in,7) to (2.3in,7) to (2.3in,6.5);
\draw[->,line width=0.05cm,color=purple] (-1.7in,6.5) to (-1.7in,7) to (2.3in,7) to (2.3in,6.5);
\draw[dotted,line width=0.025cm,color=purple] (2.3in,6.5) to (2.3in,5.5);

\node at (-1.7in,6) {
    \resizebox{!}{0.25in}{
    \begin{tikzpicture}
    \definecolor{hlearn_bgbox}{RGB}{127,255,127}
    \definecolor{hlearn_bgboxB}{RGB}{224,224,255}
    \node[shape=rectangle,draw,fill=hlearn_bgbox,minimum width=2cm,minimum height=2.1cm]   at (1,1.05) { $t_{a,1}$ };
    \node[shape=rectangle,draw,fill=hlearn_bgbox,minimum width=2cm,minimum height=2.7cm]   at (3,1.35) { $t_{a,2}$ };
    \node[shape=rectangle,draw,fill=hlearn_bgbox,minimum width=2cm,minimum height=3.1cm]   at (5,1.55) { $t_{a,3}$ };;
    \node[shape=rectangle,draw,fill=hlearn_bgbox,minimum width=2cm,minimum height=1.7cm]   at (7,0.85) { $t_{a,4}$ };
    \node[shape=rectangle,draw,fill=hlearn_bgbox,minimum width=2cm,minimum height=1.6cm]   at (9,0.8) { $t_{a,5}$ };
    \node[shape=rectangle,draw,fill=hlearn_bgbox,minimum width=2cm,minimum height=0.6cm]   at (11,0.3) { $t_{a,6}$ };
    \end{tikzpicture}
    }};

\node at (0,6) {
    \resizebox{!}{0.25in}{
%     \resizebox{1.75in}{!}{
    \begin{tikzpicture}
    \definecolor{hlearn_bgbox}{RGB}{127,255,127}
    \definecolor{hlearn_bgboxB}{RGB}{224,224,255}
    \node[shape=rectangle,draw,fill=hlearn_bgboxB,minimum width=2cm,minimum height=0.6cm]  at (1,0.3) { $t_{b,1}$ };
    \node[shape=rectangle,draw,fill=hlearn_bgboxB,minimum width=2cm,minimum height=2.5cm]  at (3,1.25) { $t_{b,2}$ };
    \node[shape=rectangle,draw,fill=hlearn_bgboxB,minimum width=2cm,minimum height=1.0cm]  at (5,0.5) { $t_{b,3}$ };
    \node[shape=rectangle,draw,fill=hlearn_bgboxB,minimum width=2cm,minimum height=0.7cm]  at (7,0.35) { $t_{b,4}$ };
    \node[shape=rectangle,draw,fill=hlearn_bgboxB,minimum width=2cm,minimum height=0.65cm] at (9,0.325) { $t_{b,5}$ };
    \node[shape=rectangle,draw,fill=hlearn_bgboxB,minimum width=2cm,minimum height=0.7cm]  at (11,0.35) { $t_{b,6}$ };
    \node[shape=rectangle,draw,fill=hlearn_bgboxB,minimum width=2cm,minimum height=0.9cm]  at (13,0.45) { $t_{b,7}$ };
    \node[shape=rectangle,draw,fill=hlearn_bgboxB,minimum width=2cm,minimum height=0.85cm] at (15,0.425) { $t_{b,8}$ };
    \end{tikzpicture}
    }};

\node at (2.3in,6) {
    \resizebox{!}{0.25in}{
    \begin{tikzpicture}
    \definecolor{hlearn_bgbox}{RGB}{127,255,127}
    \definecolor{hlearn_bgboxB}{RGB}{224,224,255}
    \node[shape=rectangle,draw,fill=hlearn_bgbox,minimum width=2cm,minimum height=2.1cm]   at (1,1.05) { $t_{a,1}$ };
    \node[shape=rectangle,draw,fill=hlearn_bgbox,minimum width=2cm,minimum height=2.7cm]   at (3,1.35) { $t_{a,2}$ };
    \node[shape=rectangle,draw,fill=hlearn_bgbox,minimum width=2cm,minimum height=3.1cm]   at (5,1.55) { $t_{a,3}$ };;
    \node[shape=rectangle,draw,fill=hlearn_bgbox,minimum width=2cm,minimum height=1.7cm]   at (7,0.85) { $t_{a,4}$ };
    \node[shape=rectangle,draw,fill=hlearn_bgbox,minimum width=2cm,minimum height=1.6cm]   at (9,0.8) { $t_{a,5}$ };
    \node[shape=rectangle,draw,fill=hlearn_bgbox,minimum width=2cm,minimum height=0.6cm]   at (11,0.3) { $t_{a,6}$ };
    \node[shape=rectangle,draw,fill=hlearn_bgboxB,minimum width=2cm,minimum height=0.6cm]  at (13,0.3) { $t_{b,1}$ };
    \node[shape=rectangle,draw,fill=hlearn_bgboxB,minimum width=2cm,minimum height=2.5cm]  at (15,1.25) { $t_{b,2}$ };
    \node[shape=rectangle,draw,fill=hlearn_bgboxB,minimum width=2cm,minimum height=1.0cm]  at (17,0.5) { $t_{b,3}$ };
    \node[shape=rectangle,draw,fill=hlearn_bgboxB,minimum width=2cm,minimum height=0.7cm]  at (19,0.35) { $t_{b,4}$ };
    \node[shape=rectangle,draw,fill=hlearn_bgboxB,minimum width=2cm,minimum height=0.65cm] at (21,0.325) { $t_{b,5}$ };
    \node[shape=rectangle,draw,fill=hlearn_bgboxB,minimum width=2cm,minimum height=0.7cm]  at (23,0.35) { $t_{b,6}$ };
    \node[shape=rectangle,draw,fill=hlearn_bgboxB,minimum width=2cm,minimum height=0.9cm]  at (25,0.45) { $t_{b,7}$ };
    \node[shape=rectangle,draw,fill=hlearn_bgboxB,minimum width=2cm,minimum height=0.85cm] at (27,0.425) { $t_{b,8}$ };
    \end{tikzpicture}
    }};

\node at (-1.7in,4) {
    \resizebox{!}{0.25in}{
    \begin{tikzpicture}
    \definecolor{hlearn_bgbox}{RGB}{127,255,127}
    \definecolor{hlearn_bgboxB}{RGB}{224,224,255}
    \node[shape=rectangle,draw,fill=hlearn_bgbox,minimum width=2cm,minimum height=3.1cm]   at (1,1.55) { $t_{a,3}$ };;
    \node[shape=rectangle,draw,fill=hlearn_bgbox,minimum width=2cm,minimum height=2.7cm]   at (3,1.35) { $t_{a,2}$ };
    \node[shape=rectangle,draw,fill=hlearn_bgbox,minimum width=2cm,minimum height=2.1cm]   at (5,1.05) { $t_{a,1}$ };
    \node[shape=rectangle,draw,fill=hlearn_bgbox,minimum width=2cm,minimum height=1.7cm]   at (7,0.85) { $t_{a,4}$ };
    \node[shape=rectangle,draw,fill=hlearn_bgbox,minimum width=2cm,minimum height=1.6cm]   at (9,0.8) { $t_{a,5}$ };
    \node[shape=rectangle,draw,fill=hlearn_bgbox,minimum width=2cm,minimum height=0.6cm]   at (11,0.3) { $t_{a,6}$ };
    \end{tikzpicture}
    }};

% \node at (-2.5in,0) {
%     \resizebox{1.5in}{!}{
%     \begin{tikzpicture}
%     \definecolor{hlearn_bgbox}{RGB}{127,255,127}
%     \node[shape=rectangle,draw,fill=hlearn_bgbox,minimum width=2cm,minimum height=3.1cm] at (3,1.55) { $t_{a,3}$ };
%     \node[shape=rectangle,draw,fill=hlearn_bgbox,minimum width=2cm,minimum height=0.6cm] at (3,3.4) { $t_{a,6}$ };
%     \node[shape=rectangle,draw,fill=hlearn_bgbox,minimum width=2cm,minimum height=2.7cm] at (5,1.35) { $t_{a,2}$ };
%     \node[shape=rectangle,draw,fill=hlearn_bgbox,minimum width=2cm,minimum height=1.6cm] at (5,3.5) { $t_{a,5}$ };
%     \node[shape=rectangle,draw,fill=hlearn_bgbox,minimum width=2cm,minimum height=2.1cm] at (7,1.05) { $t_{a,1}$ };
%     \node[shape=rectangle,draw,fill=hlearn_bgbox,minimum width=2cm,minimum height=1.7cm] at (7,2.95) { $t_{a,4}$ };
%     \draw[line width=0.1cm] (2,7.0) to (2,0) to (2+2,0) to (2+2,7.0);
%     \draw[line width=0.1cm] (4,7.0) to (4,0) to (4+2,0) to (4+2,7.0);
%     \draw[line width=0.1cm] (6,7.0) to (6,0) to (6+2,0) to (6+2,7.0);
%     \node at (3,-0.5) {$b_{1}$};
%     \node at (5,-0.5) {$b_{2}$};
%     \node at (7,-0.5) {$b_{3}$};
%     \end{tikzpicture}
%     }};

\node at (0,4) {
    \resizebox{!}{0.25in}{
    \begin{tikzpicture}
    \definecolor{hlearn_bgbox}{RGB}{127,255,127}
    \definecolor{hlearn_bgboxB}{RGB}{224,224,255}
    \node[shape=rectangle,draw,fill=hlearn_bgboxB,minimum width=2cm,minimum height=2.5cm]  at (1,1.25) { $t_{b,2}$ };
    \node[shape=rectangle,draw,fill=hlearn_bgboxB,minimum width=2cm,minimum height=1.0cm]  at (3,0.5) { $t_{b,3}$ };
    \node[shape=rectangle,draw,fill=hlearn_bgboxB,minimum width=2cm,minimum height=0.9cm]  at (5,0.45) { $t_{b,7}$ };
    \node[shape=rectangle,draw,fill=hlearn_bgboxB,minimum width=2cm,minimum height=0.85cm] at (7,0.425) { $t_{b,8}$ };
    \node[shape=rectangle,draw,fill=hlearn_bgboxB,minimum width=2cm,minimum height=0.7cm]  at (9,0.35) { $t_{b,6}$ };
    \node[shape=rectangle,draw,fill=hlearn_bgboxB,minimum width=2cm,minimum height=0.7cm]  at (11,0.35) { $t_{b,4}$ };
    \node[shape=rectangle,draw,fill=hlearn_bgboxB,minimum width=2cm,minimum height=0.65cm] at (13,0.325) { $t_{b,5}$ };
    \node[shape=rectangle,draw,fill=hlearn_bgboxB,minimum width=2cm,minimum height=0.6cm]  at (15,0.3) { $t_{b,1}$ };
    \end{tikzpicture}
    }};

% \node at (0.5,0) {
%     \resizebox{1.5in}{!}{
%     \begin{tikzpicture}
%     \definecolor{hlearn_bgboxB}{RGB}{224,224,255}
%     \node[shape=rectangle,draw,fill=hlearn_bgboxB,minimum width=2cm,minimum height=2.5cm] at (3,1.25) { $t_{b,2}$ };
%     \node[shape=rectangle,draw,fill=hlearn_bgboxB,minimum width=2cm,minimum height=1.0cm] at (5,0.5) { $t_{b,3}$ };
%     \node[shape=rectangle,draw,fill=hlearn_bgboxB,minimum width=2cm,minimum height=0.7cm] at (5,1.35) { $t_{b,4}$ };
%     \node[shape=rectangle,draw,fill=hlearn_bgboxB,minimum width=2cm,minimum height=0.7cm] at (5,2.05) { $t_{b,6}$ };
%     \node[shape=rectangle,draw,fill=hlearn_bgboxB,minimum width=2cm,minimum height=0.6cm] at (5,2.6999999999999997) { $t_{b,1}$ };
%     \node[shape=rectangle,draw,fill=hlearn_bgboxB,minimum width=2cm,minimum height=0.9cm] at (7,0.45) { $t_{b,7}$ };
%     \node[shape=rectangle,draw,fill=hlearn_bgboxB,minimum width=2cm,minimum height=0.85cm] at (7,1.325) { $t_{b,8}$ };
%     \node[shape=rectangle,draw,fill=hlearn_bgboxB,minimum width=2cm,minimum height=0.65cm] at (7,2.075) { $t_{b,5}$ };
%     \draw[line width=0.1cm] (2,7.0) to (2,0) to (2+2,0) to (2+2,7.0);
%     \draw[line width=0.1cm] (4,7.0) to (4,0) to (4+2,0) to (4+2,7.0);
%     \draw[line width=0.1cm] (6,7.0) to (6,0) to (6+2,0) to (6+2,7.0);
%     \node at (3,-0.5) {$b_{1}$};
%     \node at (5,-0.5) {$b_{2}$};
%     \node at (7,-0.5) {$b_{3}$};
%     \end{tikzpicture}
%     }};

\node at (2.3in,4) {
    \resizebox{!}{0.25in}{
    \begin{tikzpicture}
    \definecolor{hlearn_bgbox}{RGB}{127,255,127}
    \definecolor{hlearn_bgboxB}{RGB}{224,224,255}
    \node[shape=rectangle,draw,fill=hlearn_bgbox,minimum width=2cm,minimum height=3.1cm]   at (1,1.55) { $t_{a,3}$ };;
    \node[shape=rectangle,draw,fill=hlearn_bgbox,minimum width=2cm,minimum height=2.7cm]   at (3,1.35) { $t_{a,2}$ };
    \node[shape=rectangle,draw,fill=hlearn_bgboxB,minimum width=2cm,minimum height=2.5cm]  at (5,1.25) { $t_{b,2}$ };
    \node[shape=rectangle,draw,fill=hlearn_bgbox,minimum width=2cm,minimum height=2.1cm]   at (7,1.05) { $t_{a,1}$ };
    \node[shape=rectangle,draw,fill=hlearn_bgbox,minimum width=2cm,minimum height=1.7cm]   at (9,0.85) { $t_{a,4}$ };
    \node[shape=rectangle,draw,fill=hlearn_bgbox,minimum width=2cm,minimum height=1.6cm]   at (11,0.8) { $t_{a,5}$ };
    \node[shape=rectangle,draw,fill=hlearn_bgboxB,minimum width=2cm,minimum height=1.0cm]  at (13,0.5) { $t_{b,3}$ };
    \node[shape=rectangle,draw,fill=hlearn_bgboxB,minimum width=2cm,minimum height=0.9cm]  at (15,0.45) { $t_{b,7}$ };
    \node[shape=rectangle,draw,fill=hlearn_bgboxB,minimum width=2cm,minimum height=0.85cm] at (17,0.425) { $t_{b,8}$ };
    \node[shape=rectangle,draw,fill=hlearn_bgboxB,minimum width=2cm,minimum height=0.7cm]  at (19,0.35) { $t_{b,6}$ };
    \node[shape=rectangle,draw,fill=hlearn_bgboxB,minimum width=2cm,minimum height=0.7cm]  at (21,0.35) { $t_{b,4}$ };
    \node[shape=rectangle,draw,fill=hlearn_bgboxB,minimum width=2cm,minimum height=0.65cm] at (23,0.325) { $t_{b,5}$ };
    \node[shape=rectangle,draw,fill=hlearn_bgbox,minimum width=2cm,minimum height=0.6cm]   at (25,0.3) { $t_{a,6}$ };
    \node[shape=rectangle,draw,fill=hlearn_bgboxB,minimum width=2cm,minimum height=0.6cm]  at (27,0.3) { $t_{b,1}$ };
    \end{tikzpicture}
    }};

% \node at (-1in,-8) {
%     \resizebox{1.5in}{!}{
%     \begin{tikzpicture}
%     \definecolor{hlearn_bgbox}{RGB}{127,255,127}
%     \definecolor{hlearn_bgboxB}{RGB}{224,224,255}
%     \node[shape=rectangle,draw,fill=hlearn_bgbox,minimum width=2cm,minimum height=3.1cm] at (3,1.55) { $t_{a,3}$ };
%     \node[shape=rectangle,draw,fill=hlearn_bgbox,minimum width=2cm,minimum height=1.6cm] at (3,3.9000000000000004) { $t_{a,5}$ };
%     \node[shape=rectangle,draw,fill=hlearn_bgboxB,minimum width=2cm,minimum height=0.85cm] at (3,5.125) { $t_{b,8}$ };
%     \node[shape=rectangle,draw,fill=hlearn_bgboxB,minimum width=2cm,minimum height=0.65cm] at (3,5.875) { $t_{b,5}$ };
%     \node[shape=rectangle,draw,fill=hlearn_bgbox,minimum width=2cm,minimum height=2.7cm] at (5,1.35) { $t_{a,2}$ };
%     \node[shape=rectangle,draw,fill=hlearn_bgbox,minimum width=2cm,minimum height=1.7cm] at (5,3.5500000000000003) { $t_{a,4}$ };
%     \node[shape=rectangle,draw,fill=hlearn_bgboxB,minimum width=2cm,minimum height=1.0cm] at (5,4.9) { $t_{b,3}$ };
%     \node[shape=rectangle,draw,fill=hlearn_bgboxB,minimum width=2cm,minimum height=0.7cm] at (5,5.75) { $t_{b,4}$ };
%     \node[shape=rectangle,draw,fill=hlearn_bgbox,minimum width=2cm,minimum height=0.6cm] at (5,6.4) { $t_{a,6}$ };
%     \node[shape=rectangle,draw,fill=hlearn_bgboxB,minimum width=2cm,minimum height=2.5cm] at (7,1.25) { $t_{b,2}$ };
%     \node[shape=rectangle,draw,fill=hlearn_bgbox,minimum width=2cm,minimum height=2.1cm] at (7,3.55) { $t_{a,1}$ };
%     \node[shape=rectangle,draw,fill=hlearn_bgboxB,minimum width=2cm,minimum height=0.9cm] at (7,5.05) { $t_{b,7}$ };
%     \node[shape=rectangle,draw,fill=hlearn_bgboxB,minimum width=2cm,minimum height=0.7cm] at (7,5.85) { $t_{b,6}$ };
%     \node[shape=rectangle,draw,fill=hlearn_bgboxB,minimum width=2cm,minimum height=0.6cm] at (7,6.5) { $t_{b,1}$ };
%     \draw[line width=0.1cm] (2,7.0) to (2,0) to (2+2,0) to (2+2,7.0);
%     \draw[line width=0.1cm] (4,7.0) to (4,0) to (4+2,0) to (4+2,7.0);
%     \draw[line width=0.1cm] (6,7.0) to (6,0) to (6+2,0) to (6+2,7.0);
%     \node at (3,-0.5) {$b_{1}$};
%     \node at (5,-0.5) {$b_{2}$};
%     \node at (7,-0.5) {$b_{3}$};
%     \end{tikzpicture}
%     }};

\node at (-1in,4) {$\diamond$};
\node at (-1in,6) {$\diamond$};
\node at (1in,4) {$=$};
\node at (1in,6) {$=$};

\draw[dotted] (-2in,5) to (3in,5);

\end{tikzpicture}
\vspace{.15in}
\end{figure}

It is important to note that \h{SortedVector}'s \h{mappend} operation does not take constant time.
In fact, it takes $\Theta(n)$ time, where $n$ is the size of both input vectors put together.
The \h{HomTrainer} type class can handle these non-constant \h{mappend} operations elegantly.
By looking up in Table \ref{table:rt}, we can see the run times of the derived algorithms.
Notice that if the monoid operation takes time $\Theta(n)$, then our batch trainer will take time $\Theta(n\log n)$, and this is exactly what we would expect for a sorting.
Details of how these numbers were derived can be found in my TFP13 submission on the HLearn library \cite{me_tfp13}.

\begin{table}[H]
\caption{Given a run time for \h{mappend}, you can calculate the run time of the automatically generated functions using this table.}
\label{table:rt}
% \centering
\hspace{-0.3in}
\begin{tabular}{ c c c c }
\hline
Monoid operation & Sequential batch trainer & \ \ Parallel batch trainer\ \ & Online trainer\\
\mbox{(\h{mapend})} & \mbox{(\h{train})} & \mbox{(\h{parallel train})} & \mbox{(\h{add1dp})}\\
\hline \hline
$\Theta(1)$ & $\Theta(n)$ & $\Theta\left(\frac{n}{p}+\log p\right)$ & $\Theta(1)$ \\
$\Theta(\log^a n)$, $a>0$ & $\Theta(n)$ & $\Theta\left(\frac{n}{p}+(\log^a n)(\log p)\right)$ & $\Theta(\log^a n)$ \\
$\Theta(n)$ & $\Theta(n\log n)$ & $\Theta\left(\frac{n}{p}\log\frac{n}{p}+n\right)$ & $\Theta(n)$ \\
$\Theta(n^b)$, $b>1$ & $\Theta(n^b)$ & no improvement & no improvement\\
\hline
\end{tabular}
\end{table}

% We're now ready to look at the \h{Monoid} instance for \h{Scheduling}.
With all of these building blocks in place, the \h{Monoid} instance for \h{Scheduling} is relatively simple.
The \h{mempty} and \h{mappend} operations are exactly the same as they are for \h{SortedVector}, except that we also call the helper function \h{lptf}.
This function just packages the \h{SortedVector} into a \h{Scheduling} type using the \h{vector2schedule} function we saw earlier.
\begin{figure}[H]
\begin{lstlisting}
instance (Ord a, Norm a, SingI n) => Monoid (Scheduling n a) where
    mempty = lptf mempty
    p1 `mappend` p2 = lptf $ (vector p1) <> (vector p2)

lptf :: forall a p. (Norm a, SingI p) => SortedVector a -> Scheduling p a
lptf vector = Scheduling
    { vector = vector
    , schedule = vector2schedule p vector
    }
    where p = fromIntegral $ fromSing (sing :: Sing n))
\end{lstlisting}
\caption{The \h{Monoid} instance for the \h{Scheduling} model}
\end{figure}
Since \h{vector2schedule} runs in linear time and \h{SortedVector}'s \h{mappend} runs in linear time, the \h{Scheduling}'s \h{mappend} runs in linear time as well.
By Table \ref{table:rt} again, we have that the automatically derived batch trainer will take time $\Theta(n\log n)$.
This is exactly what the traditional LPTF algorithm takes.

Of course, we still have to implement the \h{HomTrainer} instance.
But this is easy.
Inside the \h{HomTrainer} class is a function called the ``singleton trainer'':
\begin{spec}
train1dp :: HomTrainer model => Datapoint model -> model
\end{spec}
All this function does is create a model from a single data point.\footnote{The \h{train1dp} function is analogous to the \h{pure} function in the \h{Applicative} class, or the \h{return} function in the \h{Monad} class.}
In practice, such a singleton model is rarely useful by itself.
But if we define it, then the compiler can then use this function and \h{mappend} to build the other functions within the \h{HomTrainer} class automatically.
This is how we get the online and parallel functions ``for free.''

The resulting \h{Scheduling} instance looks like:
\begin{figure}[H]
\begin{lstlisting}
instance (Norm a, SingI n) => HomTrainer (Scheduling n a) where
    type Datapoint (Scheduling n a) = a
    train1dp dp = lptf $ train1dp dp
\end{lstlisting}
\caption{The \h{HomTrainer} instance is quite short and mechanical to write}
\end{figure}
That's all we \textit{need} to do to guarantee correct asymptotic performance, but we've got one last trick that will speed up our \h{train} function by a constant factor.
Recall that when performing the \h{mappend} operation on \h{Scheduling} variables, we can only reuse the work contained inside of \h{vector}.
The old \h{schedule}s must be completely discarded.
Since \h{mappend} is called many times in our automatically generated functions, calculating all of these intermediate schedules would give us no benefit but result in a lot of extra work.
That is why in the \h{Scheduling} type, the \h{vector} member was declared strict, whereas the \h{schedule} member was declared lazy.
The \h{schedule}s won't actually be calculated until someone demands them, and since no one will ever demand a schedule from the intermediate steps, we never calculate them.

\section{Back to Bin Packing}

Since \prob{BinPacking} and \prob{Scheduling} were such similar problems, it's not too surprising that a similar technique can be used to implement a \h{BinPacking} model.
% We can solving \prob{BinPacking} will look similar to how we solved \prob{Scheduling}.
% We can approximately solve a \prob{BinPacking} problem similarly to how we derived a Scheduling instance.
The main difference is that we'll replace the LPTF algorithm with another simple algorithm called Best Fit Decreasing (BFD).  This gives us the performance guarantee of:
$$
\textit{BFD} \le {11\over9}\textit{OPT} + 1
$$
There are some slightly better approximations for \prob{BinPacking}, but we won't look at them here because they are much more complicated.
Chapter 2 of \textit{Approximation Algorithms for NP Hard Problems} gives a good overview on the considerable amount of literature for bin packing \cite{hochbaum96binpacking}.

BFD is a two stage algorithm in the same vein as LPTF.
First, we sort the data points by size.
% This step dominates our run time, so our batch trainer will run in time $\Theta(n\log n)$ and our monoid operation in time $\Theta(n)$.
Then, we iteratively take the largest item and find the ``best'' bin to place it in.
The best bin is defined as the bin with the least amount of space that can still hold the item.
If no bins can hold the item, then we create a new bin and add the item there.
This is shown graphically in Figure \ref{fig:bfd} below.

\begin{figure}[H]
\caption{One iteration of the Best First Decreasing (BFD) algorithm}
\label{fig:bfd}
\centering
\begin{tikzpicture}
    \definecolor{hlearn_bgbox}{RGB}{127,255,127}
%     \node[shape=rectangle,draw,fill=hlearn_bgbox,minimum width=2cm,minimum height=2.1cm] at (1,1.05) { $t_{2}$ };
%     \node[shape=rectangle,draw,fill=hlearn_bgbox,minimum width=2cm,minimum height=2.0cm] at (3,1.0) { $t_{1}$ };
%     \node[shape=rectangle,draw,fill=hlearn_bgbox,minimum width=2cm,minimum height=1.7cm] at (5,0.85) { $t_{10}$ };
%     \node[shape=rectangle,draw,fill=hlearn_bgbox,minimum width=2cm,minimum height=1.5cm] at (7,0.75) { $t_{3}$ };
    \node[shape=rectangle,draw,fill=hlearn_bgbox,minimum width=2cm,minimum height=1.3cm] at (3.5,0.5+4.65) { $t_{9}$ };
    \node[shape=rectangle,draw,fill=hlearn_bgbox,minimum width=2cm,minimum height=0.9cm] at (5.5,0.5+4.45) { $t_{8}$ };
    \node[shape=rectangle,draw,fill=hlearn_bgbox,minimum width=2cm,minimum height=0.8cm] at (7.5,0.5+4.4) { $t_{7}$ };
    \node[shape=rectangle,draw,fill=hlearn_bgbox,minimum width=2cm,minimum height=0.7cm] at (9.5,0.5+4.35) { $t_{6}$ };
    \node[shape=rectangle,draw,fill=hlearn_bgbox,minimum width=2cm,minimum height=0.65cm] at (11.5,0.5+4.325) { $t_{4}$ };
    \node[shape=rectangle,draw,fill=hlearn_bgbox,minimum width=2cm,minimum height=0.6cm] at (13.5,0.5+4.3) { $t_{5}$ };

    \definecolor{hlearn_bgbox}{RGB}{127,255,127}
    \definecolor{hlearn_bgbox_tmp}{RGB}{223,255,223}
%     \definecolor{hlearn_bgbox_tmp}{RGB}{127,255,127}
    \node[shape=rectangle,draw,fill=hlearn_bgbox,minimum width=2cm,minimum height=2.1cm] at (3,1.05) { $t_{2}$ };
%     \node[shape=rectangle,draw,fill=hlearn_bgbox,minimum width=2cm,minimum height=0.7cm] at (3,2.45) { $t_{6}$ };
    \node[shape=rectangle,draw,fill=hlearn_bgbox,minimum width=2cm,minimum height=2.0cm] at (5,1.0) { $t_{1}$ };
%     \node[shape=rectangle,draw,fill=hlearn_bgbox,minimum width=2cm,minimum height=0.8cm] at (5,2.4) { $t_{7}$ };
    \node[shape=rectangle,draw,fill=hlearn_bgbox,minimum width=2cm,minimum height=1.7cm] at (7,0.85) { $t_{10}$ };
%     \node[shape=rectangle,draw,fill=hlearn_bgbox,minimum width=2cm,minimum height=0.9cm] at (7,2.15) { $t_{8}$ };
    \node[shape=rectangle,draw,fill=hlearn_bgbox,minimum width=2cm,minimum height=1.5cm] at (9,0.75) { $t_{3}$ };
%     \node[shape=rectangle,draw,fill=hlearn_bgbox,minimum width=2cm,minimum height=0.65cm] at (11,0.325) { $t_{4}$ };
%     \node[shape=rectangle,draw,fill=hlearn_bgbox,minimum width=2cm,minimum height=0.6cm] at (11,0.95) { $t_{5}$ };

%     \node[shape=rectangle,pattern=north east lines,pattern color=hlearn_bgbox,draw,minimum width=2cm,minimum height=1.3cm] at (3,2.75) { $t_{9}$ };
%     \node[shape=rectangle,pattern=north east lines,pattern color=hlearn_bgbox,draw,minimum width=2cm,minimum height=1.3cm] at (5,2.65) { $t_{9}$ };
    \node[shape=rectangle,pattern=north east lines,pattern color=hlearn_bgbox,draw,minimum width=2cm,minimum height=1.3cm] at (7,2.35) { $t_{9}$ };
%     \node[shape=rectangle,pattern=north east lines,pattern color=hlearn_bgbox,draw,minimum width=2cm,minimum height=1.3cm] at (9,2.15) { $t_{9}$ };
%     \node[shape=rectangle,pattern=north east lines,pattern color=hlearn_bgbox,draw,minimum width=2cm,minimum height=1.3cm] at (11,0.65) { $t_{9}$ };

    \draw[line width=0.1cm] (2,3.0) to (2,0) to (2+2,0) to (2+2,3.0);
    \draw[line width=0.1cm] (4,3.0) to (4,0) to (4+2,0) to (4+2,3.0);
    \draw[line width=0.1cm] (6,3.0) to (6,0) to (6+2,0) to (6+2,3.0);
    \draw[line width=0.1cm] (8,3.0) to (8,0) to (8+2,0) to (8+2,3.0);
    \draw[line width=0.1cm] (10,3.0) to (10,0) to (10+2,0) to (10+2,3.0);
    \draw[line width=0.1cm] (12,3.0) to (12,0) to (12+2,0) to (12+2,3.0);

    \node at (3,-0.5) {$b_{1}$};
    \node at (5,-0.5) {$b_{2}$};
    \node at (7,-0.5) {$b_{3}$};
    \node at (9,-0.5) {$b_{4}$};
    \node at (11,-0.5) {$b_{5}$};
    \node at (13,-0.5) {$b_{6}$};
    % \draw[line width=0.1cm] (14,3.0) to (14,0) to node[below] {$b_{5}$} (14+2,0) to (14+2,3.0);
    % \draw[line width=0.1cm] (16,3.0) to (16,0) to node[below] {$b_{5}$} (16+2,0) to (16+2,3.0);
    \draw[line width=0.1cm] (2,0) to (2,3.0) to (15,3.0);
    \draw[line width=0.1cm] (2,0) to (15,0);
    % \draw[dashed,color=red,line width=0.05cm] (2,3) to (2,4);
    % \draw[dashed,color=red,line width=0.05cm] (12,3) to (12,4);
    % \draw[<->,line width=0.05cm,color=red] (2,3.5) to node[above] {number of bins} (12,3.5);
    % \draw[<->,line width=0.05cm,color=blue] (1.5,0) to (1.5,3);
    % \draw[dashed,color=blue,line width=0.05cm] (1,0) to (2,0);
    % \draw[dashed,color=blue,line width=0.05cm] (1,3) to (2,3);
    % \node[color=blue] at (1,1.5) {\rotatebox{90}{fixed bin size}};
    \node at (15,1.5) {\textbf{...}};

    \draw[->,line width=0.05in,color=red] (3.5,4.5) to [in=110,out=-70] (7,2.9);

    \draw[dotted] (0.5,4) to (15,4);

\node at (1,1.5) {
    \rotatebox{90}{
    \begin{tabular}{c}
    \emph{partial} \\ \emph{bin packing}
    \end{tabular}
    }};
% \node[text centered] at (0,1.5) {\emph{partial \\ bin packing}};
\node at (1,5.25) {
    \rotatebox{90}{
    \begin{tabular}{c}
    \emph{sorted} \\ \emph{vector}
    \end{tabular}
%     \emph{sorted vector}
    }};

\end{tikzpicture}
\vspace{0.05in}
\end{figure}

The data type for bin packing is:
\begin{spec}
data BinPacking (n::Nat) a = BinPacking
    { vector  :: !(SortedVector a)
    , packing :: Map.Map Bin [a]
    }
\end{spec}
This is the exact same form as the Scheduling type had.
The only difference is that we will use the BFD strategy to generate our \h{Map}.
Therefore, by similar reasoning, the \h{BinPacking}'s \h{mappend} function takes time $\Theta(n)$ and its \h{train} function takes time $\Theta(n\log n)$.
Again, this is exactly what the traditional description of the BFD algorithm requires.

\section{Takeaways}

Most instances of the \h{HomTrainer} type class are related to statistics or machine learning, but the class is much more general than that.
For example, we've just seen how to use \h{HomTrainer}s to approximate two \np-complete problems.
So from now on, whenever you see a function that has type:
\begin{spec}
 :: [datapoint] -> model
\end{spec}
ask yourself, ``Could this algorithm be implemented using a \h{HomTrainer}?''
If yes, you'll get online and parallel versions for free.

Finally, we've looked at the monoid structure for \h{Scheduling} and \h{BinPacking}, but these types also have Abelian group and $\mathbb{Z}$-module structure as well.
I'll let you explore the documentation available on the github repository \cite{hlearn} 
(pull requests are always welcome!)
to find creative ways to exploit these structures.
If you have any questions or feedback, I'd love to hear it.

% \bibliographystyle{plain}
\bibliography{monoids}

\end{document}

### 2013-06-10 article.tex
Add monoids
Signed-off-by: Edward Z. Yang <ezyang@mit.edu>
example code showing that ErrorT/MaybeT transformer order doesn't matter

@@
 %include polycode.fmt
+
+\newcommand{\Matt}[1]{\Red{Matt: #1}}
 
@@
 On the other hand, the order of Error and Maybe relative to each other isn't 
-important.  But don't just take my word for it -- try it for yourself!
-\EZY{It's worth mentioning why, formally speaking, this is the case, which has
-to do with the relevant structures being isomorphic to each other.}
-\Matt{Uh-oh!  I don't really know anything formal about monad transformers,
-I really have no idea how to go about showing something like this!
-I guess I'll have to rethink my goals for this section.}
-\EZY{It's pretty simple, and just has to do with the fact that Maybe (Either a b) is
-equivalent (modulo laziness) with Either a (Maybe b).  To show this is a case, you have
-to define functions which convert from one to the other, and then show that the roundtrip
-is identity.}
-
-Another important point to note is that our parsers don't specify
+important.  We can show this by losslessly converting values from one stack to 
+the other and back:
+\begin{verbatim}
+type Stack1 e a = Either e (Maybe a)
+type Stack2 e a = Maybe (Either e a)
+
+forward :: Stack1 e a -> Stack2 e a
+forward (Left e)          =  Just $ Left e
+forward (Right Nothing)   =  Nothing
+forward (Right (Just x))  =  Just $ Right x
+
+backward :: Stack2 e a -> Stack1 e a
+backward Nothing           =  Right Nothing
+backward (Just (Left e))   =  Left e
+backward (Just (Right x))  =  Right $ Just x
+
+identityForward :: Stack1 e a -> Stack1 e a
+identityForward = backward . forward
+
+identityBackward :: Stack2 e a -> Stack2 e a
+identityBackward = forward . backward
+\end{verbatim}
+
+No partial functions were used.  Now we can use these functions to
+convert between the two stacks.  Note the final result is the same as the input:
+\begin{verbatim}
+vals = [Left "an error", Right Nothing, Right $ Just "success"]
+
+*Main> map forward vals
+[Just (Left "an error"),Nothing,Just (Right "success")]
+
+*Main> map identityForward vals
+[Left "an error",Right Nothing,Right (Just "success")]
+\end{verbatim}
+
+Another important point is that our parsers don't specify
 which order -- Maybe/State or State/Maybe -- they prefer, and will 

### 2013-06-10 src/stackorder.hs

@@ -29,3 +29,23 @@ satisfy = flip check item
 
 literal :: (Eq t, MonadState [t] m, Alternative m) => t -> m t
 literal c = satisfy ((==) c)
+
+
+type Stack1 e a = Either e (Maybe a)
+type Stack2 e a = Maybe (Either e a)
+
+forward :: Stack1 e a -> Stack2 e a
+forward (Left e)          =  Just $ Left e
+forward (Right Nothing)   =  Nothing
+forward (Right (Just x))  =  Just $ Right x
+
+backward :: Stack2 e a -> Stack1 e a
+backward Nothing           =  Right Nothing
+backward (Just (Left e))   =  Left e
+backward (Just (Right x))  =  Right $ Just x
+
+identityForward :: Stack1 e a -> Stack1 e a
+identityForward = backward . forward
+
+identityBackward :: Stack2 e a -> Stack2 e a
+identityBackward = forward . backward

### 2013-06-10 article.tex
a couple more questions about the comments

@@
 build continuity.}
+\Matt{Could you give me any pointers on how to do this?  I can't figure out
+an effective way of doing it without just repeating the abstract verbatim.}
 
@@
 fleshed out more.}
+\Matt{Good point, I did not have a specific audience in mind.  
+The article requires familiarity with monad transformers, but I don't think it should 
+require familiarity with parser combinators.  I'll work on improving this section with
+that in mind.}
 
@@
 Note that the FlexibleContexts language extension is required by the 
-MonadState constraint in the type signature.
-The mtl \EZY{which?} \Matt{I'm not sure what the library's called -- is it mtl or transformers?} \EZY{Well, both export the necessary interfaces.
-Just pick one and be consistent.} formulation of monad
+MonadState constraint in the type signature.  The mtl 
+\EZY{which?} 
+\Matt{I'm not sure what the library's called -- is it mtl or transformers?} 
+\EZY{Well, both export the necessary interfaces. Just pick one and be consistent.} 
+formulation of monad
 transformers relies heavily on extensions to the type system, more of which 
@@
 it could be glossed over or relegated to an appendix.}
+\Matt{I don't quite understand -- I'm missing the context.
+What does "in constrast" refer to?"}
 By adding the position to the input, we've created a new problem 

### 2013-06-16 gadt.tex
Rewritten conclusion of section about proving correctness of insertion in red-black trees.

@@
 We can skip them, but this means producing ``Non-exhaustive patterns'' exception for these impossible cases.
+To workaround the corresponding warning message from Haskell compiler we are using the \verb|error| function:
 
@@
 
-Note that the case of one regular node and one leaf node is not valid, because these nodes must have different black heights.
-
-Complete source code is in Appendix C.
-
-\EZY{The new sentence you added here is really awkward} We have shown that the insertion operation maintains all invariants in red-black trees by adding necessary properties to the tree data type and then modifying the functions accordingly.
-However, the previous code illustrates a general problem with proofs.
-In fact, in Haskell $\bot$ (bottom) is a member of every type.
-As a result, we can write:
+However, the previous code illustrates a general problem with proofs. In fact, in Haskell $\bot$ (bottom) is a member of every type. As a result, we can write:
 
@@
 This code type checks, but, of course, the implementation of the \verb|concatenate| function does not meet our expectations.
+
+By now we have specified all cases of the \verb|leftBalance| function (note that the case of one regular node and one leaf node is not valid, because these nodes must have different black heights; so, we do not need to take care of this case).
+
+Complete source code of this example is in Appendix C.
+
+In this section we have implemented and verified insertion operation in red-black trees.
+We have used GADTs to express invariants for this data type.
+This in turn guarantees that the tree is balanced and searching in such tree will take $O (\log n)$ time (where $n$ is total number of elements in the tree).
+Like in the previous example we extended the original data types with necessary properties to prove the invariants.
 

### 2013-06-16 gadt.tex
Wibble
Signed-off-by: Edward Z. Yang <ezyang@mit.edu>

--- 2013-06-04 gadt.tex
@@
 This code type checks, but, of course, the implementation of the \verb|concatenate| function does not meet our expectations.
+
+At this point, we have specified all cases of the \verb|leftBalance| function (note that the case of one regular node and one leaf node is not valid, because these nodes must have different black heights; so, we do not need to take care of this case).
+The complete source code of this example is in Appendix C.
+
+To sum up, we have implemented and verified the insertion operation on red-black trees.
+By using GADTs to express invariants for this data type, we guarantee that the tree is balanced and that searching in such tree will take $O (\log n)$ time (where $n$ is total number of elements in the tree).
 

### 2013-06-16 article.tex
Updates
Signed-off-by: Edward Z. Yang <ezyang@mit.edu>

@@
 an effective way of doing it without just repeating the abstract verbatim.}
+\EZY{Hmm, I'd probably convert the abstract into the intro, and write a new,
+shorter abstract, like ``Monad transformers can be used to ...''}
 
@@
 But I'm not opposed to it at all.  Do you think I should switch it?}
+\EZY{I think that it is a worthwhile change to make, if it doesn't add a lot
+of extra stuff to the article.  I think it could be done that way, but if
+you don't think so I'm happy to let this lie.}
 \begin{verbatim}
@@
 What does "in constrast" refer to?"}
+\EZY{Sorry, I was referring to the terseness of the first section. So there
+you give very little explanation, while here you spend a lot of time explaining
+what is going on.}
 By adding the position to the input, we've created a new problem 

### 2013-06-17 monoids.tex
edits
Signed-off-by: Edward Z. Yang <ezyang@mit.edu>

@@
 \begin{introduction}
-As a TA I was confronted with a real life instance of the \np-complete \prob{Scheduling} problem.
+As a TA, I was confronted with a real life instance of the \np-complete \prob{Scheduling} problem.
 To solve the problem, I turned to the classic Least Processing Time First (LPTF) approximation algorithm.
@@
 Therefore, I had no chance of creating an optimal grouping for my professor---the class had 100 students, and $2^{100}$ is a \textit{big} number!
-So I turned to approximation algorithms.
+\EZY{Well, it's not at all clear that a SAT solver wouldn't be able to find some sort of interesting structure in your problem domain which lets it find the optimal grouping!
+But yes, the brute force solution is infeasible.}
+As a result, I turned to approximation algorithms.
 % These algorithms give us solutions that are guaranteed to be within a certain factor of optimal.
@@
 \end{spec}
-Anytime I see see a function of this form, I ask myself, ``Can it be implemented using the \h{HomTrainer} type class?''
+Any time I see see a function of this form, I ask myself, ``Can it be implemented using the \h{HomTrainer} type class?'' \EZY{HomTrainer is not in the base library, so it's worth mentioning where you can get it from.
+You did say so in the abstract (it's HLearn), but it's worth repeating here, with a cite.}
 \h{HomTrainer}s are useful because the compiler automatically derives online and parallel algorithms for all instances.
@@
 
-The blue arrows in the diagram are a little trickier.
-They specify that our training algorithm must be a \emph{monoid homomorphism} from the \emph{free monoid}.
+The blue arrows in the diagram impose some constraints on the data types and functions. These requirements
+are a little trickier:
+they specify that our training algorithm must be a \emph{monoid homomorphism} from the \emph{free monoid}.
 % This sounds a lot scarier than it actually is.
@@
 \end{spec}
-Sometimes we use the infix operation \h{(<>) = mappend} to make our code easier to read.
+Sometimes, we use the infix operation \h{(<>) = mappend} to make our code easier to read.
 All instances must obey the identity and associativity laws:
@@
 Vectors have a similar definition, and both lists and vectors are examples of free monoids.
+\EZY{This is a funny claim to make. The conventional definition of a vector is an nat-indexed
+data type, i.e. the fixed sized lists.  So if you try to define a Monoid instance on them,
+you'll find that you can't get the types to line up, because the size changes when you append.
+You could say, ``Well, these are vectors where the size is unknown'', but that is isomorphic
+to list, so there's no point in mentioning them.}
 Free monoids are monoids that can be generated by some underlying set.
@@
 
-A homomorphism from the free monoid is a function that ``preserves the free monoid's structure.''
+A homomorphism from the free monoid is a function that ``preserves the free monoid's structure.'' \EZY{Well, you can have a monoid homomorphism between two monoids which are not free; the point of being a free monoid is that there \emph{always} exists a monoid homomorphism to any other monoid.}
 More formally, if the function is called \h{train}, then it obeys the law that for all \lstinline{xs} and \lstinline{ys} of type \lstinline{[a]}:
@@
 This interface is shown in Figure \ref{alg:HomTrainer}.
-In the class, we associate a specific \h{Datapoint} type to our model, and get four functions for training functions.
+In the class, we associate a specific \h{Datapoint} type to our model and get four functions for training functions.
 The most important training function is the batch trainer, called \h{train}.
@@
 
+\EZY{At the beginning of this section, you said ``Anytime I see see a function of this form, I ask myself, “Can it be implemented
+using the HomTrainer type class?” HomTrainers are useful because the compiler
+automatically derives online and parallel algorithms for all instances. For LPTF,
+the answer is “yes.” This section will cover what exactly that means.''  I have reached the end of the section. Is the question answered? Maybe you should recapitulate.}
+
 \begin{figure}
@@
 
+\EZY{I think it would be clearer if you split up the code fragment and interspersed it with the commentary. That way, the reader doesn't have to continually flip back and forth.}
+
 Before we look at implementing a \h{HomTrainer} to solve \prob{Scheduling}, we'll take a look at how it's used.
@@
 % Every program that uses the HLearn library is going to require at least the DataKinds and TypeFamilies extensions.
-Let's begin by doing some experiments in GHCI to get ourselves oriented.
-The \h{Scheduling} type is our model, and here we ask GHCI for it's kind signature:
+Let's begin by doing some experiments in GHCi to get ourselves oriented.
+The \h{Scheduling} type is our model, and here we ask GHCi for it's kind signature:
 \begin{lstlisting}
@@
 What does that mean?
-In mathematics, a type has a norm if it has a ``size'' of some sort.
+In mathematics, a type has a norm if it has a ``size'' of some sort. \EZY{The laws the norm must obey are not too hard to state, maybe mention them briefly?}
 In HLearn, the \h{Norm} type class is defined in two parts.
@@
 Figure \ref{fig:main} shows the code for solving the student groups problem.
-In this case, I defined a new data type called \h{Student} to be the data points, and defined the \h{magnitude} of a \h{Student} to be its \h{grade}.
+In this case, I defined a new data type called \h{Student} to be the data points and defined the \h{magnitude} of a \h{Student} to be its \h{grade}.
 Notice that since I am deriving an instance of \h{Ord} automatically, I must define \h{grade} before \h{name} and \h{section}.
@@
 % \end{spec}
-using the \h{Cassava} package.
+using the \h{Cassava} package.\EZY{cite?}
 The details of how this works aren't too important.
@@
 
-Now we're ready to dive into the details of how \h{Scheduling} works under the hood.
+Now we're ready to dive into the details of how our model, \h{Scheduling} works under the hood.
 % We've already looked at the type parameters.
 % Let's take a more detailed look at the internals.
-Here's the declaration:
+\h{Scheduling} is defined as:
 \begin{spec}
@@
 Scheduling has two member variables.
-The first is a \h{SortedVector}.
+The first is a \h{SortedVector}. \EZY{Is this why you said vectors form a free monoid? This is not the monoid you are looking.}
 This custom data type is a wrapper around the \h{vector} package's \h{Vector} type that maintains the invariant that items are always sorted.
@@
 The \h{HomTrainer} type class can handle these non-constant \h{mappend} operations elegantly.
+\EZY{I do not know what you mean here. Do you mean, using standard techniques for solving
+recurrences, we can still calculate the new runtime?}
 By looking up in Table \ref{table:rt}, we can see the run times of the derived algorithms.
@@
 Monoid operation & Sequential batch trainer & \ \ Parallel batch trainer\ \ & Online trainer\\
-\mbox{(\h{mapend})} & \mbox{(\h{train})} & \mbox{(\h{parallel train})} & \mbox{(\h{add1dp})}\\
+\mbox{(\h{mappend})} & \mbox{(\h{train})} & \mbox{(\h{parallel train})} & \mbox{(\h{add1dp})}\\
 \hline \hline
@@
 Finally, we've looked at the monoid structure for \h{Scheduling} and \h{BinPacking}, but these types also have Abelian group and $\mathbb{Z}$-module structure as well.
-I'll let you explore the documentation available on the github repository \cite{hlearn} 
+I'll let you explore the documentation available on the GitHub repository \cite{hlearn} 
 (pull requests are always welcome!)

### 2013-06-18 article.tex
Parsers:  remove resolved comments

@@
 -- can be built out of the StateT monad transformer applied to Maybe.
-\EZY{Do you maybe want to say why this is the case?}
 From the implementation:
@@
 Note that the FlexibleContexts language extension is required by the 
-MonadState constraint in the type signature.  The mtl 
-\EZY{which?} 
-\Matt{I'm not sure what the library's called -- is it mtl or transformers?} 
-\EZY{Well, both export the necessary interfaces. Just pick one and be consistent.} 
-formulation of monad
+MonadState constraint in the type signature.  The mtl formulation of monad
 transformers relies heavily on extensions to the type system, more of which 

### 2013-06-18 article.tex
Parsers: abstract: shorter; parser combinator introduction: work with a more specific audience in mind

@@
 
-The goal of parsing, in abstract, is to take a string and build a tree
-representing the structure of this input according to a formal grammar.
-However, real-world parsers must also recognize faulty input and
-accurately report the location and cause of the problem.  What generates
-a parse error, how is that information included in the error, and how do
-error-reporting parsers compose?
-
-Monad transformers provide a clean and simple answer to these questions.
-We'll show how monad transformers can be used to build a parser combinator
-and then add error reporting in later in a modular way.
+Monad transformers are used to deal with combinations of computational
+effects, including backtracking, errors, and state, in a modular and
+composable way \cite{liang}.
+
+Real-world parsers require such computational effects in order to provide
+key features such as error reporting and context-sensitive results.
+This article will explore how monad transformers contribute to the design
+and implementation of parsers.
 
@@
 
+The goals of parsing are to: 1) decide whether a string is part of the language
+in question, and 2) build a meaningful representation of the structure of the
+parsed string.  In addition, practical parsers must recognize faulty input and
+accurately report the location and cause of the problem.  Thus, a parser must
+also decide what generates an error, what information is in the error, and how
+error-reporting parsers compose.
+
 Parser combinators are an excellent approach for building expressive, 
-declarative, and easy-to-maintain parsers.  They also benefit from host-language
+composable and declarative parsers.  They also benefit from host-language
 integration, allowing them to snarf features such as type systems and test 
-frameworks.
-
-There are many excellent papers on parser combinators, such
+frameworks.  There are many excellent papers on parser combinators, such
 as Wadler's classic paper \cite{wadler} on non-deterministic parser combinators, 
 Hutton's paper which covers factoring parsers into smaller pieces \cite{hutton}, 
-and Leijen's Parsec paper \cite{leijen}.
-The inspiration and many of the names used in this article are drawn 
-from those references.
+and Leijen's Parsec paper \cite{leijen}.  The inspiration and many of the names 
+used in this article are drawn from those references.
 
@@
 \begin{itemize}
- \item a Parser datatype
- \item parser-specific combinators
+ \item Parser datatype
+ \item a primitive parser
  \item type class instances
+ \item combinators
 \end{itemize}
@@
 
+Parsers operate on token sequences, consuming tokens from the front of the sequence.
+We could model this with the function type:
+\begin{verbatim}
+newtype Parser t = 
+    Parser {getParser :: [t] -> [t]}
+\end{verbatim}
+
+But this ignores the possibility of failure.  To allow for failure, the return value needs to
+be wrapped in a Maybe:
+\begin{verbatim}
+newtype Parser t = 
+    Parser {getParser :: [t] -> Maybe [t]}
+\end{verbatim}
+
+But we also want parsers to return a value, if they succeed.  So let's extend the return type
+again to allow for a value, which also requires a second type parameter:
 \begin{verbatim}
@@
 \end{verbatim}
-This type definition defines Parser to be a function
-on a list of tokens, producing a new list of tokens and a result
-value, possibly failing.
-
-\subsection{Type class instances}
-For running parsers in sequence, we'll use the combinators from the Monad and
-Applicative type classes.  There's a little bit of book-keeping involved in the
-Monad instance -- it has to make sure that the correct token list is used by
-each parser -- but once we have a Monad instance, the Applicative and Functor
-instances can be handled by imports \verb+ap+ and \verb+liftM+ from Control.Monad.
-\EZY{Fun fact: it is possible to create parsers which only have an applicative instance,
-and are more efficient but less expressive than the style you have here!}
-\begin{verbatim}
-import Control.Applicative  (Applicative(..), Alternative(..))
-import Control.Monad        (liftM, ap)
-
-instance Monad (Parser t) where
-  return x = Parser (\ts -> Just (ts, x))
-  Parser p >>= f = Parser (\ts ->
-                             p ts >>= \(ts', x) ->
-                             getParser (f x) ts')
-
-instance Functor (Parser t) where
-  fmap = liftM
-
-instance Applicative (Parser t) where
-  pure = return
-  (<*>) = ap
-\end{verbatim}
-
-We use the Alternative type class for expressing choice; the Parser instance is
-is defined in terms of the Alternative instance of the
-underlying result type, \verb+Maybe+.
-\begin{verbatim}
-instance Alternative (Parser t) where
-  empty = Parser (const empty)
-  Parser l <|> Parser r = Parser (\ts -> l ts <|> r ts)
-\end{verbatim}
-
-\subsection{Parser-specific combinators}
-\verb+item+ is a very simple parser; it 
+
+We've used \verb+newtype+ instead of \verb+type+ so that we can create type class instances
+for our parsers later on.
+
+\subsection{A primitive parser}
+The simplest parser is \verb+item+; it 
 consumes a single token if available and fails otherwise:
@@
 
-The \verb+check+ combinator is analogous to \verb+Control.Monad.mfilter+, which is used to
-validate a parse result:
+This captures the basic concept that parsers are inherently sequential --
+they consume tokens left-to-right from a list.
+
+\subsection{Type class instances}
+For running multiple parsers in sequence, the combinators from the Monad and
+Applicative type classes are great -- which means we need instances for
+them.
+
+First is the Functor instance, which we need since it's a superclass
+of Applicative.  It just maps a function over the result value, and
+relies on the Functor instances of Maybe and pairs:
+\begin{verbatim}
+instance Functor (Parser t) where
+  fmap f (Parser p) = Parser (fmap (fmap f) . p)
+\end{verbatim}
+
+The \verb+<*>+ operator of Applicative runs parsers sequentially, 
+and applies the result of the first to the result of the second; it
+has to make sure that the (possibly modified) output token stream from
+the first parser is passed to the second:
+\begin{verbatim}  
+import Control.Applicative  (Applicative(..))
+
+instance Applicative (Parser t) where
+  pure x = Parser (\xs -> pure (xs, x))
+  Parser p1 <*> Parser p2 = Parser p3
+    where p3 xs = case (p1 xs) of
+                       Nothing      -> Nothing;
+                       Just (ys, f) -> fmap (fmap f) (p2 ys)
+\end{verbatim}
+
+The Monad instance is similar to the Applicative instance, except that 
+the actual result value of the first parser is used to generate the 
+second parser; similarly to the Functor instance, this depends on 
+instances of its underlying types:
+\begin{verbatim}
+instance Monad (Parser t) where
+  return x = Parser (\ts -> return (ts, x))
+  Parser p >>= f = Parser (\ts ->
+                             p ts >>= \(ts', x) ->
+                             getParser (f x) ts')
+\end{verbatim}
+
+The Alternative type class expresses choice; the instance is
+is defined in terms of the Alternative instance of the
+underlying result type, \verb+Maybe+.
+\begin{verbatim}
+import Control.Applicative  (Alternative(..))
+
+instance Alternative (Parser t) where
+  empty = Parser (const empty)
+  Parser l <|> Parser r = Parser (\ts -> l ts <|> r ts)
+\end{verbatim}
+
+\subsection{Combinators}
+We can capture common patterns of parser construction in combinators in 
+order to make it more convenient.  The \verb+check+ combinator, analogous 
+to \verb+Control.Monad.mfilter+, is used to validate a parse result:
 \begin{verbatim}
@@
 
-Using \verb+check+, we can build a very convenient combinator to parse specific
+Using \verb+check+, we can build a convenient combinator to parse specific
 matching tokens: 
@@
 \subsection{Examples}
-Now for some examples:  first, let's see \verb+item+ in action.  It fails on empty lists,
+First, let's see \verb+item+ in action.  It fails on empty lists,
 and succeeds consuming one token otherwise:

### 2013-06-18 src/parse0.hs

@@ -1,5 +1,4 @@
 import Control.Applicative  (Applicative(..), Alternative(..))
-import Control.Monad        (liftM, ap)
 
 newtype Parser t a = 
     Parser {getParser :: [t] -> Maybe ([t], a)}
@@ -10,18 +9,21 @@ item = Parser f
     f [] = Nothing
     f (x:xs) = Just (xs, x)
 
+instance Functor (Parser t) where
+  fmap f (Parser p) = Parser (fmap (fmap f) . p)
+  
+instance Applicative (Parser t) where
+  pure x = Parser (\xs -> pure (xs, x))
+  Parser p1 <*> Parser p2 = Parser p3
+    where p3 xs = case (p1 xs) of
+                       Nothing      -> Nothing;
+                       Just (ys, f) -> fmap (fmap f) (p2 ys)
+
 instance Monad (Parser t) where
-  return x = Parser (\ts -> Just (ts, x))
+  return x = Parser (\ts -> return (ts, x))
   Parser p >>= f = Parser (\ts ->
                              p ts >>= \(ts', x) ->
                              getParser (f x) ts')
-
-instance Functor (Parser t) where
-  fmap = liftM
-  
-instance Applicative (Parser t) where
-  pure = return
-  (<*>) = ap          
 
 instance Alternative (Parser t) where
   empty = Parser (const empty)

### 2013-06-18 article.tex
Parsers:  explain how monad transformers help to build parser combinators

@@
 \section{Monad transformers}
-Monad transformers~\cite{liang} provide a method of combining monads into
-new monads which support operations of each of their constituents.  
-By applying monad transformers, we can build semantically rich parsers without sacrificing modularity.
-We'll use the standard transformers/mtl \cite{mtl} libraries -- 
-they're easy to install, easy to use, and very useful.
-
-First, note that the parser type we used earlier -- \verb+[t] -> Maybe ([t], a)+
--- can be built out of the StateT monad transformer applied to Maybe.
-From the implementation:
+
+How can we use monad transformers to help us create parsers and parser combinators?
+The parser type we used earlier -- \verb+[t] -> Maybe ([t], a)+
+-- can be built out of the StateT monad transformer (from the standard 
+mtl \cite{mtl} library) applied to Maybe.  Here's the StateT implementation:
 \begin{verbatim}
@@
 \end{verbatim}
+
 Substituting in [s] for s and Maybe for m in the right-hand side produces
@@
 \end{verbatim}
-Which is identical to the Parser type given before, except for the order 
-of the result tuple.
-This allows us to define the new parser type as:
+
+This is identical to the Parser type given before, except for the order 
+of the result tuple, and so we can redefine the parser type as:
 \begin{verbatim}
@@
 \end{verbatim}
-Note that the FlexibleContexts language extension is required by the 
-MonadState constraint in the type signature.  The mtl formulation of monad
-transformers relies heavily on extensions to the type system, more of which 
-we'll have to use for some of the later code in this article.
+
+There are two major benefits to our parsers from using monad transformers:
+\begin{itemize}
+ \item semantic building blocks -- the mtl library provides implementations
+    for Maybe, Error, and State building blocks; using these, we can describe
+    describe the semantics of our parsers
+ \item type class instances -- in addition to providing the building blocks,
+    the mtl provides instances for common type classes, so we only have to 
+    worry about parser-specific operations
+\end{itemize}
 

### 2013-06-21 src/parse3.hs
Parser source: count line/column using an extra state layer instead of pre-processing input; split the error-reporting section in 2: one for error messages, the second for position as well

@@ -10,31 +10,54 @@ import Control.Monad.Trans.Maybe (MaybeT(..))
 import Control.Applicative       (Applicative(..), Alternative(..))
 import Data.List                 (nub)
 
+
+-- parser definition
+
+type Pos = (Int, Int)
+type Err = (String, Pos)
+
+type Parser t a = StateT [t] (StateT Pos (MaybeT (Either Err))) a 
+
+runParser :: Parser t a -> [t] -> Either Err (Maybe ((a, [t]), Pos))
+runParser p ts = runMaybeT $ runStateT (runStateT p ts) (1, 1)
+
+-- basic parser:  item
+
+getState :: Parser Char Pos
+getState = lift get
+
+putState :: Pos -> Parser Char ()
+putState = lift . put
+
+updateState :: (Pos -> Pos) -> Parser Char ()
+updateState f = getState >>= (putState . f)
+
+basic_item :: (MonadState [t] m, Alternative m) => m t
+basic_item =
+    get >>= \xs -> case xs of
+                        (t:ts) -> put ts *> pure t;
+                        []     -> empty;
+
+item :: Parser Char Char
+item = basic_item >>= \x -> updateState (f x) >> pure x
+  where f '\n' (ln, c) = (ln + 1, 1)
+        f _    (ln, c) = (ln, c + 1)
+
 -- the prelude
-
-type Parser e t a = StateT [t] (MaybeT (Either e)) a
-
-runParser :: Parser e t a -> [t] -> Either e (Maybe (a, [t]))
-runParser p xs = runMaybeT (runStateT p xs)
 
 many0 = many
 many1 = some
 
-item :: (MonadState [t] m, Alternative m) => m t
-item =
-    get >>= \xs -> case xs of
-                        (t:ts) -> put ts *> pure t;
-                        []     -> empty;
 
 check :: (Monad m, Alternative m) => (a -> Bool) -> m a -> m a
 check f p =
     p >>= \x ->
     if (f x) then return x else empty
 
-satisfy :: (MonadState [t] m, Alternative m) => (t -> Bool) -> m t
+satisfy :: (Char -> Bool) -> Parser Char Char
 satisfy = flip check item
 
-literal :: (Eq t, MonadState [t] m, Alternative m) => t -> m t
+literal :: Char -> Parser Char Char
 literal c = satisfy ((==) c)
 
 
[…]

### 2013-06-21 article.tex
Parser article: split error reporting into two separate sections

@@
 
-\subsection{Reporting position}
-To report error position, we'll pre-process the input
-string to calculate the line and column number of each character.  Then, instead 
-of our parsers operating on a list of characters, they'll operate on a list of 
-Tokens, where a Token is a character tupled up with a line number and column 
-number, which we'll express with a simple typedef and some accessor functions:
-\EZY{I don't think this is how most real parser libraries do it.  Why not do it
-with some extra state on the parser monad; it's yet another application of the
-transformer approach?}
-\Matt{Good point, I thought about doing it this way but decided not to because
-the article is already long and this would add a lot, and the other approach 
-seems to work fine and is simple.
-But I'm not opposed to it at all.  Do you think I should switch it?}
-\EZY{I think that it is a worthwhile change to make, if it doesn't add a lot
-of extra stuff to the article.  I think it could be done that way, but if
-you don't think so I'm happy to let this lie.}
-\begin{verbatim}
-type Token = (Char, Int, Int)
-chr  (a, _, _)  =  a
-line (_, b, _)  =  b
-col  (_, _, c)  =  c
-\end{verbatim}
-The function responsible for calculating the position will simply transform a 
-list of characters into a list of tokens:
-\begin{verbatim}
-countLineCol :: [Char] -> [Token]
-countLineCol = reverse . snd . foldl f ((1, 1), [])
-  where
-    f ((line, col), ts) '\n' = ((line + 1, 1), ('\n', line, col):ts)
-    f ((line, col), ts)  c   = ((line, col + 1), (c, line, col):ts)
-\end{verbatim}
-
-\subsection{Token parsers}
-\EZY{In contrast, this section doesn't seem very enlightening, so maybe
-it could be glossed over or relegated to an appendix.}
-\Matt{I don't quite understand -- I'm missing the context.
-What does "in constrast" refer to?"}
-\EZY{Sorry, I was referring to the terseness of the first section. So there
-you give very little explanation, while here you spend a lot of time explaining
-what is going on.}
-By adding the position to the input, we've created a new problem 
-for ourselves: our original token parsers operated on 
-character lists, but we now need them to work on token lists.  
-The cause of the problem is the \verb+literal+ combinator, so we
-replace it with a similar one for token lists:
-\begin{verbatim}
-character :: (MonadState [Token] m, 
-              Alternative m) => Char -> m Token
-character c = satisfy ((==) c . chr)
-\end{verbatim}
-
-Now our token parsers are:
-\begin{verbatim}
-whitespace = many1 $ satisfy (flip elem " \n\t\r\f" . chr)
-comment = pure (:) <*> character ';' <*> many0 chars
-  where chars = not1 $ character '\n'
-
-munch p = many0 (whitespace <+> comment) *> p
-
-ocurly = munch $ character '{'
-ccurly = munch $ character '}'
-oparen = munch $ character '('
-cparen = munch $ character ')'
-symbol = munch $ many1 char
-  where char = fmap chr $ satisfy (f . chr)
-        f = flip elem (['a' .. 'z'] ++ ['A' .. 'Z'])
-\end{verbatim}
-
-The only differences were the replacement of \verb+literal+ with \verb+character+ 
-and a couple uses of the \verb+chr+ accessor.  The \verb+munch+ combinator didn't 
-have to change at all (although its inferred type did in fact change).
-
 \subsection{Error messages}
-Now, we modify the remaining parsers to report accurate and useful errors.
-Here are the error messages that we'll use to report each error:
+We must modify the remaining parsers to report meaningful error messages; let's
+define the messages that will be reported in each case:
 \begin{verbatim}
@@
 application =
-    openparen                             >>= \open ->
-    commit (eAppOper, open) form          >>= \op ->
+    openparen                     >>
+    commit eAppOper form          >>= \op ->
     many0 form                    >>= \args ->
-    commit (eAppClose, open) closeparen   >>
+    commit eAppClose closeparen   >>
     return (AApp op args)
 \end{verbatim}
-This parser works by running the open-parenthesis parser, capturing its value 
-in \verb+open+, and continuing with the parsing; if either \verb+form+ or 
-close-parenthesis fails, an appropriate error is generated with a message and 
-the location of the open parenthesis.  However, if the open 
+
+This parser works by first running the open-parenthesis parser; then, if either 
+\verb+form+ or close-parenthesis fails, an appropriate error is generated with 
+a message describing the problem.  However, if the open 
 parenthesis is not successfully parsed, it's a failure but not an error.  
@@
 \begin{verbatim}
-*Main> runParser application (countLineCol "oops")
+*Main> runParser application "oops"
 Right Nothing
 
-*Main> runParser application (countLineCol "(a b c)")
-Right (Just (AApp (ASymbol "a") [ASymbol "b",ASymbol "c"],[]))
-
-*Main> runParser application (countLineCol "\n()")
-Left ("application: missing operator",('(',2,1))
-
-*Main> runParser application (countLineCol "(a b")
-Left ("application: missing close parenthesis",('(',1,1))
-\end{verbatim}
-
-Although the previous version of the \verb+application+ parser was implemented 
-with just applicative combinators, this version can't be:  \verb+open+ 
-is used to build the following parsers, making this monadic.
+*Main> runParser application "(a b c)"
+Right (Just (AApp (ASymbol "a") [ASymbol "b",ASymbol "c"],""))
+
+*Main> runParser application "\n()"
+Left "application: missing operator"
+
+*Main> runParser application "(a b"
+Left "application: missing close parenthesis"
+\end{verbatim}
 
@@
 define =
-    opencurly                            >>= \open ->
     check (== "define") symbol    *>
     pure ADefine                 <*>
-    commit (eDefSym, open) symbol        <*>
-    commit (eDefForm, open) form         <*
-    commit (eDefClose, open) closecurly
+    commit eDefSym symbol        <*>
+    commit eDefForm form         <*
+    commit eDefClose closecurly
 
 lambda =
-    opencurly                                >>= \open ->
     check (== "lambda") symbol       >>
-    commit (eLamParam, open) opencurly       >>= \p_open ->
+    commit eLamParam opencurly       >>
     many0 symbol                     >>= \params ->
@@
         then return ()
-        else throwError (eLamDupe, p_open))  >>
-    commit (eLamPClose, p_open) closecurly   >>
-    commit (eLamBody, open) (many1 form)     >>= \bodies ->
-    commit (eLamClose, open) closecurly      >>
-    return (ALambda params bodies)
+        else throwError eLamDupe)    >>
+    commit eLamPClose closecurly     >>
+    commit eLamBody (many1 form)     >>= \bodies ->
+    return (ALambda params bodies)   <*
+    commit eLamClose closecurly  
   where
@@
 special form can be parsed.  We extend the \verb+special+ 
-parser with a third alternative, which throws an error if it can parse
-an open curly brace:
-\begin{verbatim}
-special = define   <+> 
-          lambda   <+> 
-          (ocurly >>= \o -> throwError (esName, o))
+parser to report an error if it matches an open brace but not define or lambda:
+\begin{verbatim}
+special = 
+    opencurly  *>  commit eSpecial (define <|> lambda)
 \end{verbatim}
@@
 must first munch any leading whitespace or comments, as usual, and then check 
-whether there are any tokens left.  If there are, it should report an error at 
-the location of the first remaining token:
+whether there are any characters left; if there are, it should report an error:
 \begin{verbatim}
 endCheck = 
-    many0 (whitespace <+> comment) *>
-    get >>= \xs -> case xs of
-                        (t:_) -> throwError (ewUnp, t)
-                        []    -> pure ()
+    many0 (whitespace <|> comment) *>
+    switch item
 \end{verbatim}
@@
 \begin{verbatim}
-woof = many0 form <* endCheck
+woof = many0 form <* commit eWoof endCheck
 \end{verbatim}
@@
 \begin{verbatim}
-*Main> let p = runParser woof . countLineCol
-
-*Main> p "{define fgh {lambda {f x} (f x)}}"
+*Main> runParser woof "{define fgh {lambda {f x} (f x)}}"
 Right (Just ([ADefine "fgh" (ALambda ["f","x"] 
@@
 \begin{verbatim}
-*Main> p "()"
-Left ("application: missing operator",('(',1,1))
-
-*Main> p "(a b"
-Left ("application: missing close parenthesis",('(',1,1))
-
-*Main> p "{define (a b c)}"
-Left ("define: missing symbol",('{',1,1))
-
-*Main> p "{define a}"
-Left ("define: missing form",('{',1,1))
-
-*Main> p "{define a b"
-Left ("define: missing close curly",('{',1,1))
-
-*Main> p "{lambda (a b c)}"
-Left ("lambda: missing parameter list",('{',1,1))
-
-*Main> p "{lambda {a b a} (a b c)}"
-Left ("lambda: duplicate parameter names",('{',1,9))
-
-*Main> p "{lambda {a b (c d)}"
-Left ("lambda: missing parameter list close curly",('{',1,9))
-
-*Main> p "{lambda {a b}}"
-Left ("lambda: missing body form",('{',1,1))
-
-*Main> p "{lambda {a b} (c d)"
-Left ("lambda: missing close curly",('{',1,1))
-
-*Main> p "{defin x y}"
-Left ("special form: unable to parse",('{',1,1))
-
-*Main> p "a,b"
-Left ("woof: unparsed input",(',',1,2))
-\end{verbatim}
+*Main> runParser woof "(a b"
+Left "application: missing close parenthesis"
+
+*Main> runParser woof "{define (a b c)}"
+Left "define: missing symbol"
+\end{verbatim}
+
+And so on.
+
+\section{Example 3: reporting position}
+
+... new code ...
 

### 2013-06-21 src/parse2.hs

@@ -162,7 +162,7 @@ lambda =
     distinct names = length names == length (nub names)
 
 special = 
-    opencurly  *>  (define <|> lambda <|> throwError eSpecial)
+    opencurly  *>  commit eSpecial (define <|> lambda)
 
 form = fmap ASymbol symbol <|> application <|> special
 

### 2013-06-21 article.tex
Parser article:  start position reporting section

@@
 
-... new code ...
+Knowing that a close-parenthesis is missing, somewhere deep inside a nested mess
+of forms isn't very helpful -- you also want to know where it happened.  Let's 
+add that to our parser.
+
+\subsection{Parser type}
+
+Once again, we extend the monad transformer stack with another layer -- a second
+state layer, in which the line and column position will be kept as simple Ints.
+The error type is also extended to include a position as well as a message:
+\begin{verbatim}
+type Pos = (Int, Int)
+type Err = (String, Pos)
+
+type Parser t a = StateT [t] (StateT Pos (MaybeT (Either Err))) a 
+\end{verbatim}
+
+And \verb+runParser+ is extended to support the new stack; it assumes we always
+start parsing at line 1, column 1:
+\begin{verbatim}
+runParser :: Parser t a -> [t] -> Either Err (Maybe ((a, [t]), Pos))
+runParser p ts = runMaybeT $ runStateT (runStateT p ts) (1, 1)
+\end{verbatim}
+
+Now that we have two separate state layers, we need to be sure not to mix them 
+up (fortunately, Haskell's type system will helpfully complain if I mess up);
+the \verb+lift+ method of the \verb+MonadTrans+ type class allows us to access
+the position state:
+\begin{verbatim}
+getState :: Parser Char Pos
+getState = lift get
+
+putState :: Pos -> Parser Char ()
+putState = lift . put
+
+updateState :: (Pos -> Pos) -> Parser Char ()
+updateState f = getState >>= (putState . f)
+\end{verbatim}
+
+\subsection{}
 

### 2013-06-24 src/parse1.hs
Parsers: get rid of unnecessary differences between parser versions to make it easier to compare them and focus on the important differences

@@ -77,13 +77,15 @@ example = "{define \n\
 whitespace = many1 $ satisfy (flip elem " \n\t\r\f")
 comment = pure (:) <*> literal ';' <*> many0 (not1 $ literal '\n')
 
-munch p = many0 (whitespace <|> comment) *> p
+junk = many0 (whitespace <|> comment) 
 
-opencurly  = munch $ literal '{'
-closecurly = munch $ literal '}'
-openparen  = munch $ literal '('
-closeparen = munch $ literal ')'
-symbol = munch $ many1 char
+tok p = p <* junk
+
+opencurly  = tok $ literal '{'
+closecurly = tok $ literal '}'
+openparen  = tok $ literal '('
+closeparen = tok $ literal ')'
+symbol = tok $ many1 char
   where char = satisfy (flip elem (['a' .. 'z'] ++ ['A' .. 'Z']))
 
 application =
@@ -94,7 +96,6 @@ application =
     closeparen
 
 define =
-    opencurly                    *>
     check (== "define") symbol   *>
     pure ADefine                <*>
     symbol                      <*>
@@ -102,7 +103,6 @@ define =
     closecurly
 
 lambda = 
-    opencurly                       *>
     check (== "lambda") symbol      *>
     opencurly                       *>
     pure ALambda                   <*>
@@ -113,11 +113,13 @@ lambda =
   where
     distinct names = length names == length (nub names)
 
-special = define <|> lambda
+special = opencurly *> (define <|> lambda)
 
 form = fmap ASymbol symbol <|> application <|> special
 
-woof = many1 form <* munch end
+endCheck = switch item
+
+woof = junk *> many0 form <* endCheck
 
 test = runParser woof example == r
   where r = Just ([ADefine "f" (ALambda ["x","y"] [AApp (ASymbol "plus") [ASymbol "x",ASymbol "y"]]),

### 2013-06-24 src/parse2.hs

@@ -110,13 +110,15 @@ type Error = String
 whitespace = many1 $ satisfy (flip elem " \n\t\r\f")
 comment = pure (:) <*> literal ';' <*> many0 (not1 $ literal '\n')
 
-munch p = many0 (whitespace <|> comment) *> p
+junk = many0 (whitespace <|> comment) 
 
-opencurly  = munch $ literal '{'
-closecurly = munch $ literal '}'
-openparen  = munch $ literal '('
-closeparen = munch $ literal ')'
-symbol = munch $ many1 char
+tok p = p <* junk
+
+opencurly  = tok $ literal '{'
+closecurly = tok $ literal '}'
+openparen  = tok $ literal '('
+closeparen = tok $ literal ')'
+symbol = tok $ many1 char
   where char = satisfy (flip elem (['a' .. 'z'] ++ ['A' .. 'Z']))
 
 eAppOper    =  "application: missing operator"
@@ -166,9 +168,7 @@ special =
 
 form = fmap ASymbol symbol <|> application <|> special
 
-endCheck = 
-    many0 (whitespace <|> comment) *>
-    switch item
+endCheck = switch item
 
 woof :: Parser String Char [AST]
-woof = many0 form <* commit eWoof endCheck
+woof = junk *> many0 form <* commit eWoof endCheck

### 2013-06-24 article.tex
Parsers:
 - review section showing how transformers helped the code

 - merging code changes into article

 - parser version 3 explanation

@@
 With the additional rule that whitespace and comments may appear in any 
-amount before any token.  Tokens are:
+amount before or after tokens.  Tokens are:
 \begin{itemize}
@@
 But what about the spec that said that comments and whitespace can occur 
-before any token -- will our parser be able to handle that?
-\begin{verbatim}
-*Main> map (runParser  opencurly) ["}abc", "(abc", "abc"]
-Nothing
-\end{verbatim}
-
-Of course not -- we haven't told it how to skip comments and whitespace yet!
-Let's do that now, first by defining the whitespace and comment patterns, then
-by creating a \verb+munch+ combinator that takes in a parser, 
-throws away all leading junk, and finally runs the parser.  Also, we rename
+anywhere before or after tokens -- how should that be handled?
+
+First we need to define the whitespace and comment patterns -- and also rename
 the \verb+some+ and \verb+many+ functions from Control.Alternative to more
@@
 comment = pure (:) <*> literal ';' <*> many0 (not1 $ literal '\n')
-
-munch p = many0 (whitespace <+> comment) *> p
+\end{verbatim}
+
+So what we want to throw away are comments and whitespace; and let's throw them 
+away after every token, using a combinator which first parses a token, and then
+throws away junk:
+\begin{verbatim}
+junk = many0 (whitespace <|> comment)
+
+tok p = p <* junk
 \end{verbatim}
@@
 
-And \verb+runParser+ is extended to support the new stack; it assumes we always
+\verb+runParser+ is extended to support the new stack; it assumes we always
 start parsing at line 1, column 1:
@@
 \end{verbatim}
+
+We also need to update \verb+item+, such that when it consumes a character, 
+it updates the line/column position.  We'll do this by splitting it into two
+parts, the part that consumes a single character, and the part that updates the
+position:
+\begin{verbatim}
+basic_item :: (MonadState [t] m, Alternative m) => m t
+basic_item =
+    get >>= \xs -> case xs of
+                        (t:ts) -> put ts *> pure t;
+                        []     -> empty;
+
+item :: Parser Char Char
+item = basic_item >>= \x -> updateState (f x) >> pure x
+  where f '\n' (ln, c) = (ln + 1, 1)
+        f _    (ln, c) = (ln, c + 1)
+\end{verbatim}
+
+For simplicity, only '\n' characters will count as newlines; every other character,
+including tabs, will count as a single column on the same line.
 
@@
 the actual monad stack we use.
+
+\section{Monad transformer motivation}
+Let's take a quick look at what we've done from a different perspective that
+may help motivate why monad transformers were used.
+
+In the first parser, we could parse valid input, but invalid input left us with
+absolutely no idea what the problem was or where:
+\begin{verbatim}
+*Main> runParser woof " {define const {lambda {x x} x}} "
+Nothing
+\end{verbatim}
+
+The second parser reported what the problem was, but not where:
+\begin{verbatim}
+*Main> runParser woof " {define const {lambda {x x} x}} "
+Left "lambda: duplicate parameter names"
+\end{verbatim}
+
+And the third parser was able to report both what and where the problem occurred:
+\begin{verbatim}
+*Main> runParser woof " {define const {lambda {x x} x}} "
+Left ("lambda: duplicate parameter names",(1,25))
+\end{verbatim}
+
+Thus, a key to the usefulness of the last parser is that it supports monadic
+effects -- backtracking, error reporting, and state.  The parsers which supported 
+fewer effects were not as useful.
+
+This emphasizes some major advantages of using monad transformers:
+\begin{itemize}
+ \item it's easy to create stacks supporting multiple effects
+ \item type class instances are taken care of by the library, so stacks already
+       support many useful combinators
+ \item it's easy to change the composition and order of stacks
+\end{itemize}
+
+\Matt{hmm, the last point can be a disadvantage as well}
+
+We benefitted from these advantages when many of the parsers didn't have to change 
+at all from one version to the next (although some of their types did change).
 

### 2013-06-24 src/parse3.hs
Parser:  position reporting with more elegant combinator

@@ -152,52 +152,45 @@ eSpecial    =  "special form: unable to parse"
 eWoof       =  "woof: unparsed input"
 -- other possibilities:  non-symbol in parameter list
 
+cut :: String -> Parser Char a -> Parser Char a
+cut message parser = 
+    getState >>= \p ->
+    commit (message, p) parser
+
+
 application =
-    openparen                           >>
-    getState                            >>= \p1 -> 
-    commit (eAppOper, p1) form          >>= \op ->
-    many0 form                          >>= \args ->
-    getState                            >>= \p2 ->
-    commit (eAppClose, p2) closeparen   >>
+    openparen                     >>
+    cut eAppOper form             >>= \op ->
+    many0 form                    >>= \args ->
+    cut eAppClose closeparen      >>
     return (AApp op args)
     
 define =
-    check (== "define") symbol          >>
-    getState                            >>= \p1 ->
-    commit (eDefSym, p1) symbol         >>= \s  ->
-    getState                            >>= \p2 ->
-    commit (eDefForm, p2) form          >>= \f  ->
-    getState                            >>= \p3 ->
-    commit (eDefClose, p3) closecurly   >>
+    check (== "define") symbol    >>
+    cut eDefSym symbol            >>= \s ->
+    cut eDefForm form             >>= \f ->
+    cut eDefClose closecurly      >>
     return (ADefine s f)
     
 lambda =
-    check (== "lambda") symbol             >>
-    getState                               >>= \p1 ->
-    commit (eLamParam, p1) opencurly       >>
-    getState                               >>= \p2 ->
-    many0 symbol                           >>= \params ->
+    check (== "lambda") symbol     >>
+    cut eLamParam opencurly        >>
+    many0 symbol                   >>= \params ->
     (if distinct params 
         then return ()
-        else throwError (eLamDupe, p2))    >>
-    getState                               >>= \p3 ->
-    commit (eLamPClose, p3) closecurly     >>
-    getState                               >>= \p4 ->
-    commit (eLamBody, p4) (many1 form)     >>= \bodies ->
-    getState                               >>= \p5 ->
-    commit (eLamClose, p5) closecurly      >>
+        else cut eLamDupe empty)      >>
+    cut eLamPClose closecurly         >>
+    cut eLamBody (many1 form)         >>= \bodies ->
+    cut eLamClose closecurly          >>
     return (ALambda params bodies)
   where
[…]

### 2013-06-24 src/parse1.hs
Parsers:  factor out closecurly into special form rule

@@ -99,8 +99,7 @@ define =
     check (== "define") symbol   *>
     pure ADefine                <*>
     symbol                      <*>
-    form                        <*
-    closecurly
+    form
 
 lambda = 
     check (== "lambda") symbol      *>
@@ -108,12 +107,11 @@ lambda =
     pure ALambda                   <*>
     check distinct (many0 symbol)  <*>
     (closecurly                     *>
-     many1 form                    <*
-     closecurly)
+     many1 form)
   where
     distinct names = length names == length (nub names)
 
-special = opencurly *> (define <|> lambda)
+special = opencurly *> (define <|> lambda) <* closecurly
 
 form = fmap ASymbol symbol <|> application <|> special
 

### 2013-06-24 src/parse2.hs

@@ -125,15 +125,14 @@ eAppOper    =  "application: missing operator"
 eAppClose   =  "application: missing close parenthesis"
 eDefSym     =  "define: missing symbol"
 eDefForm    =  "define: missing form"
-eDefClose   =  "define: missing close curly"
 eLamParam   =  "lambda: missing parameter list"
 eLamDupe    =  "lambda: duplicate parameter names"
 eLamPClose  =  "lambda: missing parameter list close curly"
 eLamBody    =  "lambda: missing body form"
-eLamClose   =  "lambda: missing close curly"
+eSpecClose  =  "special form: missing close curly"
 eSpecial    =  "special form: unable to parse"
 eWoof       =  "woof: unparsed input"
--- other possibilities:  non-symbol in parameter list
+
 
 application =
     openparen                     >>
@@ -146,8 +145,7 @@ define =
     check (== "define") symbol    *>
     pure ADefine                 <*>
     commit eDefSym symbol        <*>
-    commit eDefForm form         <*
-    commit eDefClose closecurly
+    commit eDefForm form
     
 lambda =
     check (== "lambda") symbol       >>
@@ -158,13 +156,12 @@ lambda =
         else throwError eLamDupe)    >>
     commit eLamPClose closecurly     >>
     commit eLamBody (many1 form)     >>= \bodies ->
-    return (ALambda params bodies)   <*
-    commit eLamClose closecurly  
+    return (ALambda params bodies)
   where
     distinct names = length names == length (nub names)
 
 special = 
-    opencurly  *>  commit eSpecial (define <|> lambda)
+    opencurly  *>  commit eSpecial (define <|> lambda)  <*  commit eSpecClose closecurly
 
 form = fmap ASymbol symbol <|> application <|> special
 

### 2013-06-24 src/parse3.hs

@@ -142,15 +142,14 @@ eAppOper    =  "application: missing operator"
 eAppClose   =  "application: missing close parenthesis"
 eDefSym     =  "define: missing symbol"
 eDefForm    =  "define: missing form"
-eDefClose   =  "define: missing close curly"
 eLamParam   =  "lambda: missing parameter list"
 eLamDupe    =  "lambda: duplicate parameter names"
 eLamPClose  =  "lambda: missing parameter list close curly"
 eLamBody    =  "lambda: missing body form"
-eLamClose   =  "lambda: missing close curly"
+eSpecClose  =  "special form: missing close curly"
 eSpecial    =  "special form: unable to parse"
 eWoof       =  "woof: unparsed input"
--- other possibilities:  non-symbol in parameter list
+
 
 cut :: String -> Parser Char a -> Parser Char a
 cut message parser = 
@@ -169,7 +168,6 @@ define =
     check (== "define") symbol    >>
     cut eDefSym symbol            >>= \s ->
     cut eDefForm form             >>= \f ->
-    cut eDefClose closecurly      >>
     return (ADefine s f)
     
 lambda =
@@ -181,13 +179,12 @@ lambda =
         else cut eLamDupe empty)      >>
     cut eLamPClose closecurly         >>
     cut eLamBody (many1 form)         >>= \bodies ->
-    cut eLamClose closecurly          >>
     return (ALambda params bodies)
   where
     distinct names = length names == length (nub names)
 
 special = 
-    opencurly  *>  cut eSpecial (define <|> lambda)
+    opencurly  *>  cut eSpecial (define <|> lambda)  <*  cut eSpecClose closecurly
 
 form = fmap ASymbol symbol <|> application <|> special
 

### 2013-06-24 article.tex
Parsers:  use refurbished code in article examples, and improve explanations

@@
 
-And some examples to make sure it's behaving correctly: \EZY{An example?! Why not a QuickCheck property? \verb|;-)|}
+It accepts open curly brackets and rejects everything else:
 \begin{verbatim}
@@
 
-Good, it correctly accepts open curly brackets and rejects everything else.
 But what about the spec that said that comments and whitespace can occur 
@@
 
-So what we want to throw away are comments and whitespace; and let's throw them 
-away after every token, using a combinator which first parses a token, and then
-throws away junk:
+We want to throw away comments and whitespace, and let's throw them away after 
+every token.  Let's do that using a combinator which first parses a token, and 
+then throws away junk:
 \begin{verbatim}
@@
 tok p = p <* junk
-\end{verbatim}
-
-Here are these three parsers in action:
-\begin{verbatim}
-*Main> runParser whitespace "\n \t   \tabc"
-Just ("\n \t   \t","abc")
-
-*Main> runParser whitespace "abc"
-Nothing
-
-*Main> runParser comment ";123  \t\nabc"
-Just (";123  \t","\nabc")
-
-*Main> runParser (munch $ literal 'a') "abc"
-Just ('a',"bc")
-
-*Main> runParser (munch $ literal 'a') ";comment!\n   abc"
-Just ('a',"bc")
-
-*Main> runParser (munch $ literal 'a') ";comment!\n   bca"
-Nothing
-
-*Main> runParser (munch $ literal 'a') "bca"
-Nothing
 \end{verbatim}
@@
 
-With \verb+munch+ in hand, we fix \verb+opencurly+ and try it out:
-\begin{verbatim}
-opencurly = munch $ literal '{'
-
-*Main> runParser opencurly ";abc\n{def"
-Just ('{', "def")
-
-*Main> runParser opencurly "{def"
+With \verb+tok+ in hand, we fix \verb+opencurly+ and try it out:
+\begin{verbatim}
+opencurly = tok $ literal '{'
+
+*Main> runParser opencurly "{;abc\ndef"
 Just ('{', "def")
@@
 Nothing
-
-*Main> runParser opencurly "}def"
-Nothing
-\end{verbatim}
-
-Excellent!  The \verb+opencurly+ parser correctly throws away whitespace 
-and comments and succeeds, consuming one character, when the first non-junk 
-character is an open curly brace, failing otherwise.
+\end{verbatim}
+
+Excellent!  The \verb+opencurly+ parser matches an open curly brace, failing
+otherwise, and discard trailing whitespace and comments.
 
@@
 \begin{verbatim}
-closecurly = munch $ literal '}'
-openparen  = munch $ literal '('
-closeparen = munch $ literal ')'
+closecurly = tok $ literal '}'
+openparen  = tok $ literal '('
+closeparen = tok $ literal ')'
 \end{verbatim}
@@
 The parser for our last token, Symbol, is more complicated because any number of
-alphabetical characters is valid.  We'll use \verb+satisfy+ and an appropriate
+alphabetical characters is valid.  We'll use \verb+satisfy+ and a
 predicate to check that a character is valid, and \verb+many1+ to 
-match any positive number of valid characters:
-\begin{verbatim}
-symbol = munch $ many1 char
+match one or more valid characters:
+\begin{verbatim}
+symbol = tok $ many1 char
   where char = satisfy (flip elem (['a' .. 'z'] ++ ['A' .. 'Z']))
@@
 
-*Main> runParser symbol ";sdfasdfsa\n  abc123"
+*Main> runParser symbol "abc;sdfasdfsa\n  123"
 Just ("abc","123")
 
-*Main> runParser symbol ";sdfasdfsa\n  123abc"
+*Main> runParser symbol "123;sdfasdfsa\n  abc"
 Nothing
@@
 
-Note that we're again using \verb+munch+ to throw away leading junk.
+Note that we again use \verb+tok+ to throw away trailing junk.
 
@@
 
-Next, we move on to \verb+define+.
-A straightforward translation from the grammar produces:
+Next, we move on to the special forms.  All special forms are delimited with
+curly braces, and lambda and define are our special forms, so we have:
+
+\begin{verbatim}
+special = opencurly *> (define <|> lambda) <* closecurly
+\end{verbatim}
+
+A straightforward translation of define from the grammar produces:
 \begin{verbatim}
 define =
-    opencurly                    *>
     check (== "define") symbol   *>
@@
     symbol                      <*>
-    form                        <*
-    closecurly
+    form
 \end{verbatim}
@@
 lambda = 
-    opencurly                       *>
     check (== "lambda") symbol      *>
@@
     (closecurly                     *>
-     many1 form                    <*
-     closecurly)
+     many1 form)
   where
@@
 \end{verbatim}
+
 If the symbols are distinct, the parameter list subparser will succeed,
@@
 \subsection{Finishing touches}
-All we have left to do are the \verb+special+, \verb+form+ (which 
-we've already used to build the previous parsers), and \verb+woof+ parsers.  
-The special parser is simple -- it parses either define or lambda:
-\begin{verbatim}
-special = define <+> lambda
-\end{verbatim}
+All we have left to do are \verb+form+ (which we've already used recursively
+to build the previous parsers) and \verb+woof+.
 
@@
 \begin{verbatim}
-end = switch item
-\end{verbatim}
-
-We also need to munch any trailing comments and whitespace -- otherwise \verb+end+
-would fail -- so the final Woof parser is:
-\begin{verbatim}
-woof = many1 form <* munch end
+endCheck = switch item
+\end{verbatim}
+
+We also need to discard leading comments and whitespace, so the final Woof 
+parser is:
+\begin{verbatim}
+woof = junk *> many0 form <* endCheck
 \end{verbatim}
@@
 \subsection{Examples}
-Let's look at a few examples of our parser in action:
+Let's look at a few examples of our parser in action.  Here are a couple of
+successful examples:
 \begin{verbatim}
@@
        ADefine "q" (AApp (ASymbol "f") [ASymbol "x"])],"")
-
+\end{verbatim}
+
+And a failing one:
+\begin{verbatim}
 *Main> runParser woof "a b c {define q (f x)},"
@@
 \end{verbatim}
-Note that the first two parses consume the entire input, while the last one 
-fails because it doesn't recognize the trailing comma.
+
+Note that while the first two parses consume the entire input, while the last 
+one fails because it doesn't recognize the trailing comma.
 
@@
   \item define: missing form:  \begin{verbatim}{define a}\end{verbatim}
-  \item define: missing close curly:  \begin{verbatim}{define a b\end{verbatim}
   \item lambda: missing parameter list:  \begin{verbatim}{lambda (a b c)}\end{verbatim}
@@
   \item lambda: missing body form:  \begin{verbatim}{lambda {a b}}\end{verbatim}
-  \item lambda: missing close curly:  \begin{verbatim}{lambda {a b} (c d)\end{verbatim}
   \item special form: unable to parse:  \begin{verbatim}{defin x y}\end{verbatim}
+  \item special form: missing close curly:  \begin{verbatim}{define x y\end{verbatim}
   \item woof: unparsed input:  \begin{verbatim}a,b\end{verbatim}
 \end{itemize}
+
 To allow error reporting in our parsers, we first need to extend our monad stack, 
@@
 will not be performed.
-\EZY{Question a reader might have: what is the difference between failure and error?  You give a little guidance later but give some intuition now!}
+\EZY{Question a reader might have: what is the difference between failure and error?  
+You give a little guidance later but give some intuition now!}
 Note that successful and failing parses are wrapped in an additional \verb+Right+
@@
 \end{verbatim}
+
 We'll take advantage of this backtracking restriction to produce accurate error
@@
 eDefForm    =  "define: missing form"
-eDefClose   =  "define: missing close curly"
 eLamParam   =  "lambda: missing parameter list"
@@
 eLamBody    =  "lambda: missing body form"
-eLamClose   =  "lambda: missing close curly"
+eSpecClose  =  "special form: missing close curly"
 eSpecial    =  "special form: unable to parse"
@@
 \begin{verbatim}
-type Error = (String, Token)
-\end{verbatim}
-
-Now let's start with \verb+application+ as our first modified, error-reporting 
-parser:
+type Error = String
+\end{verbatim}
+
+We'll first add error-reporting to \verb+application+:
 \begin{verbatim}
@@
     commit eDefSym symbol        <*>
-    commit eDefForm form         <*
-    commit eDefClose closecurly
+    commit eDefForm form
 
@@
     commit eLamBody (many1 form)     >>= \bodies ->
-    return (ALambda params bodies)   <*
-    commit eLamClose closecurly  
+    return (ALambda params bodies)
   where
@@
 special = 
-    opencurly  *>  commit eSpecial (define <|> lambda)
+    opencurly  *>  commit eSpecial (define <|> lambda)  <*  commit eSpecClose closecurly
 \end{verbatim}
@@
 When we can no longer parse any forms, instead of failing, we'd like to report 
-an error if there's any input left, and otherwise succeed.  Our ending parser
-must first munch any leading whitespace or comments, as usual, and then check 
-whether there are any characters left; if there are, it should report an error:
-\begin{verbatim}
-endCheck = 
-    many0 (whitespace <|> comment) *>
-    switch item
-\end{verbatim}
-The \verb+woof+ parser uses \verb+endCheck+ after consuming all possible forms:
-\begin{verbatim}
-woof = many0 form <* commit eWoof endCheck
+an error if there's any input left, and otherwise succeed:
+\begin{verbatim}
+woof = junk *> many0 form <* commit eWoof endCheck
 \end{verbatim}
@@
 
-Knowing that a close-parenthesis is missing, somewhere deep inside a nested mess
+Knowing that a close-parenthesis is missing somewhere deep inside a nested mess
 of forms isn't very helpful -- you also want to know where it happened.  Let's 

### 2013-06-24 monoids.tex

@@
 The \prob{Scheduling} problem is shown graphically in Figure \ref{fig:scheduling}.
-Processors are drawn as bins (labeled $b_1 ... b_4$), and tasks are drawn as green blocks (labeled $t_1 ... t_{10}$) and placed inside a bin.
-The height of the task $t_i$ represents the length of time required to process that task.
-Our goal is to minimize the \textcolor{red}{red} processing time.
+Processors are drawn as bins, and tasks are drawn as green blocks inside the bins.
+The height of each task represents the length of time required to process it.
+Our goal is to minimize the \textcolor{red}{processing time} given a fixed \textcolor{blue}{number of processors}.
 
@@
 There is another closely related problem called \prob{BinPacking} that is easily confused with \prob{Scheduling}.
-In \prob{BinPacking}, instead of fixing the number of bins and minimizing the amount in the bins, we fix the capacity of each bin and minimize the total number of bins used.
+In \prob{BinPacking}, instead of fixing the number of bins and minimizing the amount in the bins, we fix the \textcolor{blue}{bin size} and minimize the total \textcolor{red}{number of bins} used.
 Compare Figure \ref{fig:binpacking} below and Figure \ref{fig:scheduling} above to see the difference.
@@
 Therefore, I had no chance of creating an optimal grouping for my professor---the class had 100 students, and $2^{100}$ is a \textit{big} number!
-\EZY{Well, it's not at all clear that a SAT solver wouldn't be able to find some sort of interesting structure in your problem domain which lets it find the optimal grouping!
+%\%EZY{Well, it's not at all clear that a SAT solver wouldn't be able to find some sort of interesting structure in your problem domain which lets it find the optimal grouping!
 But yes, the brute force solution is infeasible.}
-As a result, I turned to approximation algorithms.
+So I turned to approximation algorithms.
 % These algorithms give us solutions that are guaranteed to be within a certain factor of optimal.
@@
 % For a survey on variants of the Scheduling problem, see Chapter 1 of \textit{Approximation Algorithms for NP Hard Problems} \cite{hochbaum96scheduling}.
-By making this small sacrifice in accuracy, we get an algorithm that runs in time $\Theta(n\log n)$ instead of $\Theta(2^n)$.
-Much better!
+By making this small sacrifice in accuracy, we get an algorithm that runs in time $\Theta(n\log n)$ instead of $\Theta(2^n)$.  Much better!  
 In the rest of this article, we'll take a detailed look at a Haskell implementation of the LPTF algorithm, and then briefly use similar techniques to solve the \prob{BinPacking} problem.
@@
 \end{spec}
-Any time I see see a function of this form, I ask myself, ``Can it be implemented using the \h{HomTrainer} type class?'' \EZY{HomTrainer is not in the base library, so it's worth mentioning where you can get it from.
+Anytime I see see a function of this form, I ask myself, ``Can it be implemented using HLearn's \h{HomTrainer} type class?'' %\EZY{HomTrainer is not in the base library, so it's worth mentioning where you can get it from.
 You did say so in the abstract (it's HLearn), but it's worth repeating here, with a cite.}
+%(This class can be found in the \h{HLearn} library \cite{hlearn}.)
 \h{HomTrainer}s are useful because the compiler automatically derives online and parallel algorithms for all instances.
-For LPTF, the answer is ``yes.''
-This section will cover what exactly that means.
-
+In this section, we'll get a big picture view of how this class will help us solve the \prob{Scheduling} problem.
 We start by looking at the format of a \h{HomTrainer} instance as shown graphically in Figure \ref{fig:HomTrainer} below.
@@
 \end{spec}
-Vectors have a similar definition, and both lists and vectors are examples of free monoids.
-\EZY{This is a funny claim to make. The conventional definition of a vector is an nat-indexed
-data type, i.e. the fixed sized lists.  So if you try to define a Monoid instance on them,
-you'll find that you can't get the types to line up, because the size changes when you append.
-You could say, ``Well, these are vectors where the size is unknown'', but that is isomorphic
-to list, so there's no point in mentioning them.}
-Free monoids are monoids that can be generated by some underlying set.
-In our monoid instance for lists above, the underlying set is the type variable \h{a}.
-For the \h{HomTrainer}, when we say that our data set must be a free monoid, all we mean is that it is a collection of some data points
-(\eg\ a list of tasks).
-
-A homomorphism from the free monoid is a function that ``preserves the free monoid's structure.'' \EZY{Well, you can have a monoid homomorphism between two monoids which are not free; the point of being a free monoid is that there \emph{always} exists a monoid homomorphism to any other monoid.}
+%Vectors have a similar definition, and both lists and vectors are examples of free monoids.
+Lists are an example of free monoids because they can be generated by any underlying type.
+%In our monoid instance for lists above, the underlying set is the type variable \h{a}.
+For the \h{HomTrainer}, when we say that our data set must be a free monoid, all we mean is that it is a collection of some data points.
+%(\eg\ a list of tasks).
+For the \prob{Scheduling} problem, it is just a list of tasks.
+
+A homomorphism from the free monoid is a function that ``preserves the free monoid's structure.'' %\EZY{Well, you can have a monoid homomorphism between two monoids which are not free; the point of being a free monoid is that there \emph{always} exists a monoid homomorphism to any other monoid.}
 More formally, if the function is called \h{train}, then it obeys the law that for all \lstinline{xs} and \lstinline{ys} of type \lstinline{[a]}:
@@
 Figure \ref{fig:LPTF-monoid} shows this in picture form with a commutative diagram.
-This means that it doesn't matter whether we take the orange path (first train schedules from our data sets, then combine the schedules with \h{mappend}) or the purple path (first concatenate our data sets, then train a schedule on the result).
+\definecolor{darkorange}{RGB}{127,63,0}
+This means that it doesn't matter whether we take the \textbf{\textcolor{orange}{orange}} path (first train schedules from our data sets, then combine the schedules with \h{mappend}) or the \textbf{\textcolor{purple}{purple}} path (first concatenate our data sets, then train a schedule on the result).
 Either way, we get the exact same answer.
@@
 
-\EZY{At the beginning of this section, you said ``Anytime I see see a function of this form, I ask myself, “Can it be implemented
-using the HomTrainer type class?” HomTrainers are useful because the compiler
-automatically derives online and parallel algorithms for all instances. For LPTF,
-the answer is “yes.” This section will cover what exactly that means.''  I have reached the end of the section. Is the question answered? Maybe you should recapitulate.}
+%\EZY{At the beginning of this section, you said ``Anytime I see see a function of this form, I ask myself, “Can it be implemented
+%using the HomTrainer type class?” HomTrainers are useful because the compiler
+%automatically derives online and parallel algorithms for all instances. For LPTF,
+%the answer is “yes.” This section will cover what exactly that means.''  I have reached the end of the section. Is the question answered? Maybe you should recapitulate.}
+
+% Mike: Hopefully this is fixed by rewording the beginning part.
 
@@
 
-\EZY{I think it would be clearer if you split up the code fragment and interspersed it with the commentary. That way, the reader doesn't have to continually flip back and forth.}
+%\EZY{I think it would be clearer if you split up the code fragment and interspersed it with the commentary. That way, the reader doesn't have to continually flip back and forth.}
+% Mike: I agree, but I can't figure out a way to do that that makes it clear what code is actually in the file and what code is just library code that I'm presenting.
 
@@
 What does that mean?
-In mathematics, a type has a norm if it has a ``size'' of some sort. \EZY{The laws the norm must obey are not too hard to state, maybe mention them briefly?}
+In mathematics, a type has a norm if it has a ``size'' of some sort. %\EZY{The laws the norm must obey are not too hard to state, maybe mention them briefly?}
+% Mike: I didn't mention them because they require talking about vector spaces, which I think would complicate the discussion needlessly.
 In HLearn, the \h{Norm} type class is defined in two parts.
@@
 % \end{spec}
-using the \h{Cassava} package.\EZY{cite?}
+using the \h{Cassava} package \cite{cassava}.
 The details of how this works aren't too important.
@@
 Scheduling has two member variables.
-The first is a \h{SortedVector}. \EZY{Is this why you said vectors form a free monoid? This is not the monoid you are looking.}
+The first is a \h{SortedVector}. %\EZY{Is this why you said vectors form a free monoid? This is not the monoid you are looking.}
 This custom data type is a wrapper around the \h{vector} package's \h{Vector} type that maintains the invariant that items are always sorted.
@@
 In fact, it takes $\Theta(n)$ time, where $n$ is the size of both input vectors put together.
-The \h{HomTrainer} type class can handle these non-constant \h{mappend} operations elegantly.
-\EZY{I do not know what you mean here. Do you mean, using standard techniques for solving
-recurrences, we can still calculate the new runtime?}
-By looking up in Table \ref{table:rt}, we can see the run times of the derived algorithms.
+The \h{HomTrainer} type class makes reasoning about these non-constant \h{mappend} operations easy.
+%\EZY{I do not know what you mean here. Do you mean, using standard techniques for solving
+%recurrences, we can still calculate the new runtime?}
+By looking up in Table \ref{table:rt}, we can find the run times of the derived algorithms.
 Notice that if the monoid operation takes time $\Theta(n)$, then our batch trainer will take time $\Theta(n\log n)$, and this is exactly what we would expect for a sorting.
@@
 \begin{table}[H]
-\caption{Given a run time for \h{mappend}, you can calculate the run time of the automatically generated functions using this table.}
+\caption{Given a run time for \h{mappend}, you can calculate the run time of the automatically generated functions using this table.
+The variable $n$ is the total number of data points being trained on or combined, and the variable $p$ is the number of processors available.}
 \label{table:rt}
@@
 $\Theta(1)$ & $\Theta(n)$ & $\Theta\left(\frac{n}{p}+\log p\right)$ & $\Theta(1)$ \\
-$\Theta(\log^a n)$, $a>0$ & $\Theta(n)$ & $\Theta\left(\frac{n}{p}+(\log^a n)(\log p)\right)$ & $\Theta(\log^a n)$ \\
+$\Theta(\log n)$ & $\Theta(n)$ & $\Theta\left(\frac{n}{p}+(\log n)(\log p)\right)$ & $\Theta(\log n)$ \\
 $\Theta(n)$ & $\Theta(n\log n)$ & $\Theta\left(\frac{n}{p}\log\frac{n}{p}+n\right)$ & $\Theta(n)$ \\

### 2013-06-24 src/parse1.hs
Parsers:  stupid mistake -- duplicate endCheck parser

@@ -115,9 +115,7 @@ special = opencurly *> (define <|> lambda) <* closecurly
 
 form = fmap ASymbol symbol <|> application <|> special
 
-endCheck = switch item
-
-woof = junk *> many0 form <* endCheck
+woof = junk *> many0 form <* end
 
 test = runParser woof example == r
   where r = Just ([ADefine "f" (ALambda ["x","y"] [AApp (ASymbol "plus") [ASymbol "x",ASymbol "y"]]),

### 2013-06-24 src/parse2.hs

@@ -165,7 +165,5 @@ special =
 
 form = fmap ASymbol symbol <|> application <|> special
 
-endCheck = switch item
-
 woof :: Parser String Char [AST]
-woof = junk *> many0 form <* commit eWoof endCheck
+woof = junk *> many0 form <* commit eWoof end

### 2013-06-24 src/parse3.hs

@@ -77,6 +77,9 @@ instance Functor m => Switch (MaybeT m) where
 not1 :: Parser Char a -> Parser Char Char
 not1 p = switch p *> item
 
+end :: Parser Char ()
+end = switch item
+
 
 class Monad m => MonadError e m | m -> e where
   throwError :: e -> m a
@@ -163,13 +166,13 @@ application =
     many0 form                    >>= \args ->
     cut eAppClose closeparen      >>
     return (AApp op args)
-    
+
 define =
-    check (== "define") symbol    >>
-    cut eDefSym symbol            >>= \s ->
-    cut eDefForm form             >>= \f ->
-    return (ADefine s f)
-    
+    check (== "define") symbol    *>
+    pure ADefine                 <*>
+    cut eDefSym symbol           <*>
+    cut eDefForm form
+
 lambda =
     check (== "lambda") symbol     >>
     cut eLamParam opencurly        >>
@@ -188,6 +191,4 @@ special =
 
 form = fmap ASymbol symbol <|> application <|> special
 
-endCheck = switch item
-
-woof = junk *> many0 form <* cut eWoof endCheck
+woof = junk *> many0 form <* cut eWoof end

### 2013-06-24 article.tex
Parsers:  finish up position reporting section and start Parsec comparison section (needs work)

@@
 
-\subsection{}
-
-\Red{It's worth noting that up until this point, we haven't talked about how transformers contribute to the design of the parser library.  Maybe give the juicy bits earlier.}
+\subsection{Position reporting}
+To actually report the position when an error is detected, we introduce a new 
+combinator, which I'll call \verb+cut+:
+\Matt{a horrible name}
+\begin{verbatim}
+cut :: String -> Parser Char a -> Parser Char a
+cut message parser = 
+    getState >>= \p ->
+    commit (message, p) parser
+\end{verbatim}
+
+It checks the current position, runs a parser, and if the parser fails, reports
+an error consiting of a message and the position.
+
+To take advantage of this combinator, we essentially replace every use of 
+\verb+commit+ with \verb+cut+:
+\begin{verbatim}
+application =
+    openparen                     >>
+    cut eAppOper form             >>= \op ->
+    many0 form                    >>= \args ->
+    cut eAppClose closeparen      >>
+    return (AApp op args)
+
+define =
+    check (== "define") symbol    *>
+    pure ADefine                 <*>
+    cut eDefSym symbol           <*>
+    cut eDefForm form
+
+lambda =
+    check (== "lambda") symbol     >>
+    cut eLamParam opencurly        >>
+    many0 symbol                   >>= \params ->
+    (if distinct params 
+        then return ()
+        else cut eLamDupe empty)      >>
+    cut eLamPClose closecurly         >>
+    cut eLamBody (many1 form)         >>= \bodies ->
+    return (ALambda params bodies)
+  where
+    distinct names = length names == length (nub names)
+
+special = 
+    opencurly  *>  cut eSpecial (define <|> lambda)  <*  cut eSpecClose closecurly
+
+form = fmap ASymbol symbol <|> application <|> special
+
+woof = junk *> many0 form <* cut eWoof end
+\end{verbatim}
+
+Now the parsers report useful errors:
+\begin{verbatim}
+*Main> runParser woof " abc 123 "
+Left ("woof: unparsed input",(1,6))
+
+*Main> runParser woof " (((abc ())) "
+Left ("application: missing operator",(1,10))
+
+*Main> runParser woof "{define const \n  {lambda {x x} \n   x}"
+Left ("lambda: duplicate parameter names",(2,15))
+\end{verbatim}
+
+\Red{It's worth noting that up until this point, 
+we haven't talked about how transformers contribute to the design of the parser library.  
+Maybe give the juicy bits earlier.}
 
@@
 at all from one version to the next (although some of their types did change).
+
+\Section{Parsec comparison}
+A quick word about Parsec \cite{leijen}, an important, battle-tested parser 
+combinator library for Haskell.
+
+One major difference between the approach I've described and Parsec's approach
+is with respect to backtracking and error-reporting.  Parsec's default is 
+predictive parsing using a single token of lookahead for efficiency.  This can
+be changed using its \verb+try+ combinator.
+
+On the other hand, my approach by default uses unlimited lookahead, which can 
+be modified using the \verb+throwError+ and \verb+commit+ combinators.
 

### 2013-06-24 article.tex
Parsers: improve Parsec comparison

@@
 \Section{Parsec comparison}
-A quick word about Parsec \cite{leijen}, an important, battle-tested parser 
-combinator library for Haskell.
-
-One major difference between the approach I've described and Parsec's approach
-is with respect to backtracking and error-reporting.  Parsec's default is 
-predictive parsing using a single token of lookahead for efficiency.  This can
-be changed using its \verb+try+ combinator.
-
-On the other hand, my approach by default uses unlimited lookahead, which can 
-be modified using the \verb+throwError+ and \verb+commit+ combinators.
+A quick word about Parsec \cite{leijen}, a popular and battle-tested parser 
+combinator library for Haskell:
+
+The approach I've described differs from Parsec's approach with respect to 
+backtracking and error-reporting.  By default, Parsec generates LL(1) parsers --
+this means it uses predictive parsing with only a single token of lookahead.  This
+is useful for both efficiency and better error reporting.  However, often a 
+single token is not enough to unambiguously determine what rule should be tried;
+for these cases, one must use its \verb+try+ combinator.  It is up to the
+programmer to determine when it's necessary to add \verb+try+ to a parser
+implementation.
+
+On the other hand, my approach by default uses unlimited lookahead; it also 
+doesn't automatically generate meaningful errors.  It is up to the programmer 
+to generate useful errors using the \verb+throwError+ and \verb+commit+ 
+combinators.
 

### 2013-07-08 article.tex
Edits
Signed-off-by: Edward Z. Yang <ezyang@mit.edu>

@@
 \section{Parser Combinators}
-
-\EZY{It's probably worth rephrasing what is stated in the abstract here,
-for those of us who are in the habit of skipping abstracts; it also helps
-build continuity.}
-\Matt{Could you give me any pointers on how to do this?  I can't figure out
-an effective way of doing it without just repeating the abstract verbatim.}
-\EZY{Hmm, I'd probably convert the abstract into the intro, and write a new,
-shorter abstract, like ``Monad transformers can be used to ...''}
 
@@
 \subsection{Parser datatype}
-\EZY{These sections are very terse.  If the reader is expected to already be familiar with
-most of the moving bits of a parsing library, perhaps this section can be boiled down some more.
-It is a very curious audience the current presentation targets: a reader who knows how to
-implement a parser library ``from scratch'', but doesn't know how to use
-monad transformers to build them.  Perhaps the components of this
-presentation which are key to understanding how the monad transformer works should be
-fleshed out more.}
-\Matt{Good point, I did not have a specific audience in mind.  
-The article requires familiarity with monad transformers, but I don't think it should 
-require familiarity with parser combinators.  I'll work on improving this section with
-that in mind.}
 
@@
 
-But this ignores the possibility of failure.  To allow for failure, the return value needs to
+We also want parsers to return a value, if they succeed.  So let's extend the return type
+again to allow for a value, which also requires a second type parameter:
+\begin{verbatim}
+newtype Parser t a = 
+    Parser {getParser :: [t] -> ([t], a)}
+\end{verbatim}
+
+However, this ignores the possibility of failure.  To allow for failure, the return value needs to
 be wrapped in a Maybe:
@@
 newtype Parser t = 
-    Parser {getParser :: [t] -> Maybe [t]}
-\end{verbatim}
-
-But we also want parsers to return a value, if they succeed.  So let's extend the return type
-again to allow for a value, which also requires a second type parameter:
-\begin{verbatim}
-newtype Parser t a = 
     Parser {getParser :: [t] -> Maybe ([t], a)}
@@
 
-We've used \verb+newtype+ instead of \verb+type+ so that we can create type class instances
-for our parsers later on.
+This gives us a simple, minimal definition of a parser.  We've used
+\verb+newtype+ instead of \verb+type+ so that we can create type class
+instances for our parsers later on.
 
@@
 For running multiple parsers in sequence, the combinators from the Monad and
-Applicative type classes are great -- which means we need instances for
-them.
+Applicative type classes are very useful; our next step is to implement them.
 
@@
 \end{verbatim}
+That concludes the implementation of a simple parser.
 In the rest of this article, we'll rework and augment these basic parser 
@@
     for Maybe, Error, and State building blocks; using these, we can describe
-    describe the semantics of our parsers
+    describe the semantics of our parsers, and
  \item type class instances -- in addition to providing the building blocks,
     the mtl provides instances for common type classes, so we only have to 
-    worry about parser-specific operations
+    worry about parser-specific operations.
 \end{itemize}

### 2013-07-08 article.tex
remove resolved comments

--- 2013-06-24 article.tex
@@
 \section{Introducing Woof:  a Simple Lisp}
-
-\EZY{An alternate structure you can use here is to introduce Woof first, before talking about monad transformers.  
-After all, the example parser doesn't require you to know what is going on in transformer-land.}
 
@@
 will not be performed.
-\EZY{Question a reader might have: what is the difference between failure and error?  
-You give a little guidance later but give some intuition now!}
 Note that successful and failing parses are wrapped in an additional \verb+Right+
@@
 definition of 'failure' is provided by the Alternative instance of \verb+m+.
-\EZY{Give a little intuition on why the operator is called commit earlier;
-you don't get around it until later.}
 The reason for its name is that it restricts backtracking and alternatives;
@@
 Note that \verb+commit+ isn't used until we can be sure that we're in the right 
-rule: \EZY{Here it is}  if, in the \verb+define+ parser, we committed to parsing the ``define" 
+rule: if, in the \verb+define+ parser, we committed to parsing the ``define" 
 keyword, we wouldn't be able to backtrack and try the \verb+lambda+ parser if 
@@
 combinator, which I'll call \verb+cut+:
-\Matt{a horrible name}
 \begin{verbatim}
@@
 
-\Matt{hmm, the last point can be a disadvantage as well}
-
 We benefitted from these advantages when many of the parsers didn't have to change 

### 2013-07-20 article.tex
Make it build
Signed-off-by: Edward Z. Yang <ezyang@mit.edu>

@@
 
-For simplicity, only '\n' characters will count as newlines; every other character,
+For simplicity, only '\\n' characters will count as newlines; every other character,
 including tabs, will count as a single column on the same line.
@@
 
-\Section{Parsec comparison}
+\section{Parsec comparison}
 A quick word about Parsec \cite{leijen}, a popular and battle-tested parser 

### 2013-07-30 monoids.tex
Fixed reference to HLearn library
added reference to functor and monad structures

@@
 In particular, we'll look at the Haskell code I used to solve the problem of grouping my students.
-In order to run the code, you'll need to download the latest \h{HLearn-approximations} library:
+In order to run the code, you'll need to download the latest \h{HLearn-approximation} library:
 
 \begin{lstlisting}
-cabal install HLearn-approximations-1.0.0
+cabal install HLearn-approximation-1.0.0
 \end{lstlisting}
@@
 
-Finally, we've looked at the monoid structure for \h{Scheduling} and \h{BinPacking}, but these types also have Abelian group and $\mathbb{Z}$-module structure as well.
+Finally, we've looked at the monoid structure for \h{Scheduling} and \h{BinPacking}, but these types also have Abelian group, $\mathbb{Z}$-module, functor, and monad structure as well.
 I'll let you explore the documentation available on the GitHub repository \cite{hlearn} 

### 2013-08-05 article.tex
parsing: another reference for restricting backtracking
Parsers:  fix egregious typo (change all `<+>`s to `<|>`s)

--- 2013-07-11 article.tex
@@
 \begin{verbatim}
-form = symbol <+> application <+> special
+form = symbol <|> application <|> special
 \end{verbatim}
@@
 \begin{verbatim}
-form = fmap ASymbol symbol <+> application <+> special
+form = fmap ASymbol symbol <|> application <|> special
 \end{verbatim}
@@
 parses the input, or the parsing stops.  It's based on the \verb+nofail+
-combinator of \cite{hutton1992higher}.
+combinator of \cite{hutton1992higher} and also on \cite{Mizushima}.
 

### 2013-08-05 article.tex
Edits and code fixups.
Signed-off-by: Edward Z. Yang <ezyang@mit.edu>

--- 2013-07-20 article.tex
@@
 accurately report the location and cause of the problem.  Thus, a parser must
-also decide what generates an error, what information is in the error, and how
+also decide what generates an error, what information is in the error and how
 error-reporting parsers compose.
@@
 To get a feel for what parser combinators are and how they're used to
-build parsers, we'll quickly create a minimal parser combinator library to see
-the basic principles of their design before adding monad transformers to
-the mix.  We'll divide the library into three main pieces of the code:
+build parsers, we'll first create a minimal parser combinator library.
+Next, we'll add monad transformers to the mix.  We'll divide the library
+into four main pieces:
 
 \begin{itemize}
- \item Parser datatype
- \item a primitive parser
- \item type class instances
- \item combinators
+ \item the Parser datatype,
+ \item a primitive parser,
+ \item type class instances, and
+ \item combinators.
 \end{itemize}
@@
 
-We also want parsers to return a value, if they succeed.  So let's extend the return type
-again to allow for a value, which also requires a second type parameter:
+We also want parsers to return a value, if they succeed.  To do this, we extend the return type
+to allow for a value, which also requires a second type parameter:
 \begin{verbatim}
@@
 \begin{verbatim}
-newtype Parser t = 
+newtype Parser t a = 
     Parser {getParser :: [t] -> Maybe ([t], a)}
@@
 
-The \verb+<*>+ operator of Applicative runs parsers sequentially, 
+The \verb+<*>+ operator of Applicative runs parsers sequentially
 and applies the result of the first to the result of the second; it
@@
 
-We can run parsers in sequence using the Applicative combinators.  First one parser
-is run, then the second, and the token position is threaded between the two.
+We can run parsers in sequence using the Applicative combinators.  First, one parser
+is run, then the second is run, and the token position is threaded between the two.
 Note that both parsers must succeed in order for the combined parser to succeed:
@@
 \end{verbatim}
-Examples:
+Here are some examples of this code in action:
 \begin{verbatim}
@@
 \end{verbatim}
-That concludes the implementation of a simple parser.
+This concludes the implementation of a simple parser.
 In the rest of this article, we'll rework and augment these basic parser 
@@
 
-This is identical to the Parser type given before, except for the order 
-of the result tuple, and so we can redefine the parser type as:
+This is identical to the Parser type given before modulo the order 
+of the result tuple, so we can redefine the parser type as:
 \begin{verbatim}
@@
 \begin{itemize}
- \item semantic building blocks -- the mtl library provides implementations
-    for Maybe, Error, and State building blocks; using these, we can describe
-    describe the semantics of our parsers, and
+ \item compositional implementation with semantic building blocks --
+    the mtl library provides implementations
+    for Maybe, Error, and State building blocks; using these, we can easily
+    describe the semantics of our parsers without writing very much code, and
  \item type class instances -- in addition to providing the building blocks,
     the mtl provides instances for common type classes, so we only have to 
-    worry about parser-specific operations.
+    implement parser-specific operations.
 \end{itemize}
@@
 
-The simple language we'll use as a motivation for building parsers is Woof, a 
-simple dialect of Lisp.  We'll be progressively adding features to a simple 
+The rest of this articule will progressively adding features to a simple 
 initial implementation on our way to creating an error-reporting parser.
-
-The language definition in pseudo-BNF \cite{bnf} is:
+The simple language we'll use as a motivation for this task is Woof, a 
+simple dialect of Lisp.
+
+The language definition of Woof in pseudo-BNF~\cite{bnf} is:
 
@@
 
-With the additional rule that whitespace and comments may appear in any 
-amount before or after tokens.  Tokens are:
+There is an additional rule that whitespace and comments may appear in any 
+amount before or after tokens.  Our tokens are:
 \begin{itemize}
@@
 \subsection{Preliminaries}
-Our first Woof parser will be responsible for determining whether input
+Our first Woof parser will only be responsible for determining whether an input
 text conforms to the language definition and building an 
@@
 \begin{verbatim}
-runParser :: Parser [t] a -> [t] -> Maybe (a, [t])
+runParser :: Parser t a -> [t] -> Maybe (a, [t])
 runParser = runStateT
@@
 \begin{verbatim}
+{-# LANGUAGE NoMonomorphismRestriction #-}
+
 literal :: (Eq t, MonadState [t] m, Alternative m) => t -> m t
@@
 
-But what about the spec that said that comments and whitespace can occur 
-anywhere before or after tokens -- how should that be handled?
-
-First we need to define the whitespace and comment patterns -- and also rename
+Recall, however, that our specification said that comments and whitespace can occur 
+anywhere before or after tokens -- how should this be handled?
+
+First, we need to define the whitespace and comment patterns -- we'll also rename
 the \verb+some+ and \verb+many+ functions from Control.Alternative to more
@@
 
-We want to throw away comments and whitespace, and let's throw them away after 
-every token.  Let's do that using a combinator which first parses a token, and 
-then throws away junk:
+We want to throw away comments and whitespace, and this process
+must occur after every token.  We can do this by using
+a combinator which first parses a token and then throws away junk:
 \begin{verbatim}
@@
 
-With \verb+tok+ in hand, we fix \verb+opencurly+ and try it out:
+With \verb+tok+ in hand, we can fix \verb+opencurly+:
 \begin{verbatim}
@@
 Excellent!  The \verb+opencurly+ parser matches an open curly brace, failing
-otherwise, and discard trailing whitespace and comments.
+otherwise, and discards trailing whitespace and comments.
 
@@
 
-The parser for our last token, Symbol, is more complicated because any number of
+The parser for our last token, Symbol, is more complicated, as any number of
 alphabetical characters is valid.  We'll use \verb+satisfy+ and a
@@
 
-The first combining form is function application.  The rule says that it's delimited
+The first combining form is function application.  The rule says that it is delimited
 by matching parentheses, in between which must appear one form as the operator, followed
@@
 produce an informative error, indicating what and where the problem was.
-This allows a user to quickly find and correct problems.
 
@@
 While our previous parsers had two different results -- Nothing and Just --
-our new parsers have three possible results:  success, failure, and error. 
+our new parsers have three possible results:  success (Right Just), failure (Right Nothing), and error (Left).
+%% EZY
 A parse failure means that a parser wasn't able to match the input, but
@@
 has been found and that parsing should immediately stop; the parser will
-not be able to recover, even if there are alternatives since backtracking
+not be able to recover even if there are alternatives, since backtracking
 will not be performed.

### 2013-08-07 Editorial.tex
Final copy
Signed-off-by: Edward Z. Yang <ezyang@mit.edu>

\documentclass{tmr}

\title{Editorial}

\author{Edward Z. Yang\email{ezyang@cs.stanford.edu}}

\begin{document}

Summer brings with it the release of three articles talking
about three different (and each arguably essential) ideas from
the Haskell landscape.  The first is an in-depth tutorial about
generalized algebraic data types, covering all of the classic
use-cases of this useful type system extension.  The second is
a meditation on the design pattern of monad transformers, and how
they can help the develop the design of a parser library.  The
final shows how even a simple API like that of a monoid can
help code achieve dramatic speed-ups.  Happy reading!

\end{document}

### 2013-08-07 article.tex
Final copy
Signed-off-by: Edward Z. Yang <ezyang@mit.edu>

@@
 
-The language definition of Woof in pseudo-BNF~\cite{bnf} is:
+The language definition of Woof in pseudo-BNF is:
 
@@
 not fit our needs -- it requires an \verb+Error+ constraint that we do not want.
-We'll implement our own MonadError type class, based on the
-standard one, and instances of it.  We'll need an instance for each member of
+We'll implement our own MonadError type class.  We'll need an instance for each member of
 our stack -- Either, StateT, and MaybeT:
 \begin{verbatim}
-class Monad m => MonadError e m | m -> e where
-  throwError :: e -> m a
-  catchError :: m a -> (e -> m a) -> m a
-
-instance MonadError e (Either e) where
+{-# LANGUAGE FlexibleContexts, MultiParamTypeClasses,
+    FlexibleInstances, TypeFamilies #-}
+
+class Monad m => MonadError m where
+  type Error m :: *
+  throwError :: Error m -> m a
+  catchError :: m a -> (Error m -> m a) -> m a
+
+instance MonadError (Either e) where
+  type Error (Either e) = e
   throwError               =  Left
@@
 
-instance MonadError e m => MonadError e (StateT s m) where
+instance MonadError m => MonadError (StateT s m) where
+  type Error (StateT s m) = Error m
   throwError      =  lift . throwError
@@
 
-instance MonadError e m => MonadError e (MaybeT m) where
+instance MonadError m => MonadError (MaybeT m) where
+  type Error (MaybeT m) = Error m
   throwError      =  lift . throwError
@@
 errors; they just pass the error on through to the next level.  That's why they
-both require that the next lower level supports MonadError.  Also, this code 
-requires some additional extensions:
-\begin{itemize}
-  \item FlexibleInstances
-  \item FunctionalDependencies
-  \item UndecidableInstances 
-\end{itemize}
+both require that the next lower level supports MonadError.  This code uses
+type families to state what the error type is; explaining type families
+in more detail is out of scope for this article, but roughly, given some MonadError \verb|m|,
+you can write down the type of error that monad supports by saying \verb|Error m|.
 
@@
 but when should an error be generated and what 
-data should it include?  \verb+throwError+ is too basic for our needs, so we'll 
+data should it include?  \verb+throwError+ will unconditionally throw an
+error, but often we will only want to conditionally throw an error. Thus, we
 introduce the \verb+commit+ combinator:
 \begin{verbatim}
-commit :: (MonadError e m, Alternative m) => e -> m a -> m a
+commit :: (MonadError m, Alternative m) => Error m -> m a -> m a
 commit err p = p <|> throwError err
@@
 
-And so on.
-
 \section{Example 3: reporting position}
@@
 Now that we have two separate state layers, we need to be sure not to mix them 
-up (fortunately, Haskell's type system will helpfully complain if I mess up);
-the \verb+lift+ method of the \verb+MonadTrans+ type class allows us to access
+up (fortunately, because the types of the states are different, Haskell's type system will helpfully complain if I mess up).
+The \verb+lift+ method of the \verb+MonadTrans+ type class allows us to access
 the position state:
@@
 
-\Red{It's worth noting that up until this point, 
-we haven't talked about how transformers contribute to the design of the parser library.  
-Maybe give the juicy bits earlier.}
-
 \section{Stack Order}
@@
 
-We benefitted from these advantages when many of the parsers didn't have to change 
+We could see these advantages when many of the parsers didn't have to change 
 at all from one version to the next (although some of their types did change).
@@
 transformer.  Or we check to make sure that all variables are in scope when
-they're used -- using a Reader transformer.  We could even capture whitespace
+they're used using a Reader transformer.  We could even capture whitespace
 and comment tokens, instead of just throwing them away, in case they contained

### 2013-08-07 src/parse3.hs

@@ -1,8 +1,7 @@
 {-# LANGUAGE   NoMonomorphismRestriction
              , FlexibleContexts       
              , FlexibleInstances    
-             , FunctionalDependencies
-             , UndecidableInstances      #-}
+             , TypeFamilies      #-}
 
 import Control.Monad.State       (MonadState (..), StateT(..))
 import Control.Monad.Trans.Class (lift)
@@ -81,27 +80,31 @@ end :: Parser Char ()
 end = switch item
 
 
-class Monad m => MonadError e m | m -> e where
-  throwError :: e -> m a
-  catchError :: m a -> (e -> m a) -> m a
+class Monad m => MonadError m where
+  type Error m :: *
+  throwError :: Error m -> m a
+  catchError :: m a -> (Error m -> m a) -> m a
 
-instance MonadError e (Either e) where
+instance MonadError (Either e) where
+  type Error (Either e) = e
   throwError               =  Left
   catchError  (Right x) _  =  Right x
   catchError  (Left e)  f  =  f e
   
-instance MonadError e m => MonadError e (StateT s m) where
+instance MonadError m => MonadError (StateT s m) where
+  type Error (StateT s m) = Error m
   throwError      =  lift . throwError
   catchError m f  =  StateT g
     where
       g s = catchError (runStateT m s) 
                        (\e -> runStateT (f e) s)
 
-instance MonadError e m => MonadError e (MaybeT m) where
+instance MonadError m => MonadError (MaybeT m) where
+  type Error (MaybeT m) = Error m
   throwError      =  lift . throwError
   catchError m f  =  MaybeT $ catchError (runMaybeT m) (runMaybeT . f)
 
-commit :: (MonadError e m, Alternative m) => e -> m a -> m a
+commit :: (MonadError m, Alternative m) => Error m -> m a -> m a
 commit err p = p <|> throwError err
 
 -- 

### 2013-08-16 gadt.tex
Added a reference to the expression problem and a reference to the paper "Data types a la carte" where a solution is described.
Fixed typos and errors in the article

@@
         (Just (Left i1), Just (Left i2)) -> Just $ Left $ i1 + i2
-        _ -> error "AddInt takes two integers"
+        _ -> fail "AddInt takes two integers"
 \end{Verbatim}
@@
 These invariants guarantee that tree is balanced.
-Indeed, the longest path from the root node (containing alternating red-black nodes) can only be twice as long as the shortest path (containing only red nodes).
+Indeed, the longest path from the root node (containing alternating red-black nodes) can only be twice as long as the shortest path (containing only black nodes).
 Thus basic operations (such as insertion and search) take $O(\log n)$ time in the worst case.
@@
 \begin{Verbatim}
-type Tree a = forall bh. Node a bh
+data Tree a = forall bh. Root (Node a bh)
 \end{Verbatim}
@@
 Another important aspect is that GADTs, just like all data types in Haskell, are closed: once they are declared, they can not be extended in the rest of the program.
-In contrast, type class instances are open.
-For example, we could split \verb|Expr e| type class into two separate type classes (this way we improve modularity):
-
-\EZY{You should probably cite Data Types a la Carte here}
+This issue is called the \italic{expression problem}~\cite{expression_problem}.
+Wouter Swierstra describes a technique to solve this problem in the article ``Data types \`{a} la carte''~\cite{a_la_carte}.
+
+In contrast, type class instances are open. For example, we could split \verb|Expr e| type class into two separate type classes (this way we improve modularity):
 
